### 2017-10-28 treestats_paper.tex
some draft

\documentclass{article}
\usepackage{fullpage}
\usepackage{color}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{natbib}

\bibliographystyle{plainnat}

\include{macros}

\newcommand{\branch}{\mathcal{B}} % branch length stat
\newcommand{\site}{\mathcal{S}} % site stat

\newcommand{\plr}[1]{{\color{blue} \it #1}}

\begin{document}

\begin{center}
    A general class of tree-based statistics,
    and how to compute them
\end{center}

\emph{Other title ideas:}

Efficient computation of a general class of DNA sequence statistics

Theory and computation of single-site statistics of DNA sequences

%%%%%%%%%%
\paragraph{Abstract}
Here we describe how to efficiently compute single-site statistics
from population genetic data
using efficient methods on tree sequences as implemented in msprime.
The method is general enough to compute any currently-defined statistic,
and we use this general framework to suggest a number of new statistics.
\plr{say what method is: sthg about weighting fns}
This formulation also makes it easy to see what statistic of the underlying genealogical trees
each sequence-based statistic is estimating,
and, if underlying trees are also available,
the same methods can be used to quickly compute these quantities.

%%%%%%%%%%%%%%%%%%%%%%%
\section*{Introduction}

\plr{utility of looking at lots of individual-based stats}
With whole-genome sequence
we can confidently infer quantities
between \emph{individuals} that previously had to be estimated
averaging over populations.
This gives us much higher resolution, both in space and along the genome.
With geographic sampling,
there is potentially a lot of information available
from looking at all pairwise comparisons.

\plr{computational challenges}
Single statistics are not too bad to compute
since you can get them from allele frequencies,
but larger summaries, like the matrix of pairwise divergences between $N$ individuals,
naively involve $O(N^2)$ comparisons.

\plr{Something about how} in many applications
we'd really like to know the trees
and sequence differences really just stand in for these.
Example: $F_{ST}$ is X; 
$\theta_W$ is X;
divergence is X.

\plr{Review of phylogenetics lit}
a good review of theoretical properties \citet{semple_and_steel}

%%%%%%%%%%%%%%%%%%%%%%%
\section*{Methods}

We consider only so-called \emph{single site} statistics.
These are statistics computed as a simple \emph{average} 
of a quantity calcuated across a number of sites.
For a concrete example, to compute an element of the site frequency spectrum --
the proportion of sites at which an allele is present in 10\% of the samples, for instance --
we could 
(a) assign each site a value, either $+1$ if it has an allele at 10\%, or zero otherwise,
and
(b) take the average of these values across all sites.
As another example,
to compute the pairwise divergence between two sequences,
we would similarly assign each site either $+1$ or 0 according to whether the sequences differ,
and average these values.

%%%%%%%%%%%%%%%%%%
\subsection*{Tree sequences}

Quick review of tree sequence terminology
and storage:
nodes,
edges,
sites,
mutations.

%%%%%%%%%%%%%%%%%%
\subsection*{Efficient computation}

First we describe how to define
and efficiently compute a quite general set of functions of a collection of sequences
using tools from tree sequences.
In the next section we will specialize these to a natural set of statistics.

\plr{Since the ``weighting function' can take negtive values, maybe ``weight'' is not a good term.  Instead?}

\paragraph{Weighted site averages}
Suppose that we have $k$ groups of samples, $A_1$, \ldots, $A_k$.
% with $n_1$, \ldots, $n_k$ individuals in each, respectively.
The \emph{weight} we compute for a given site is a sum of weights computed for each allele at that site,
and the weights are found as a function of the vector of numbers of individuals in each group inheriting that allele.
Concretely, we specify a \emph{weighting function} $w(x_1, \ldots, x_k)$,
and we say that the \emph{average site weight} of a set of $L$ sites is
\begin{align} \label{eqn:average_site_weight}
    \site_w := \frac{1}{L} \sum_{i=1}^L \sum_{s \in I_i} w(x_1(s), \ldots, x_k(s)) ,
\end{align}
where $I_i$ is the set of alleles seen at that site and
$x_j(s)$ is the number of individuals in $A_j$ that carry allele $s$.

Before we describe how to compute this,
we give the analogous topological quantity,
that is equal to the expected weighted site average
under a continuum-sites model.

\paragraph{Weighted branch length averages}
The \emph{average branch length weight}
is computed almost just as above:
\begin{align} \label{eqn:average_branch_weight}
    \branch_w :=  \frac{1}{L} \sum_{i=1}^L \sum_{e \in T_i} \ell(e) w(x_1(e), \ldots, x_k(e)) ,
\end{align}
except that now $T_i$ is the genealogical tree at site $i$,
the sum is over edges $e$ in the tree, $\ell(e)$ is the length of the edge $e$,
and $x_j(e)$ is the number of individuals in $A_j$ that fall below $e$.

\emph{Note:} somewhat more generally,
we could assign a set of weights to the individuals in each group,
and let $x_j$ denote the total weight of the individuals in $A_j$ 
inheriting from that allele or edge.
For simplicity we do not pursue this here.
\plr{this is a good use of the word ``weight'' (although these could also be negative)}

\paragraph{Computation}
The \emph{tree differences} encoded by a tree sequence allow us to efficiently update the quantities $x$ used above.
As it is more straightforward, we first describe the algorithm
to compute the branch length version.

\plr{Rework these into three algorithms: one to maintain the node weights $x$ 
    (basically already described as `tracked nodes' in msprime)
    and the other two that use this.}

\paragraph{Total branch length weight algorithm}
In a tree sequence, every branch in any tree is associated with its terminal \emph{node}.
We will keep updated an array $x$ so that for any node $e$,
the value $x[e][j]$ is the number of samples in $A_j$ below node $e$ in the current tree.
This will be zero for any node \emph{not} appearing in the current tree.
We also maintain $T$, the total weight of the current tree.
\begin{itemize}
    \item \emph{Initialize} all elements of $x$ to zero
        except that $x[a][j] = 1$ for each $a \in A_j$ and $1 \le j \le k$.
        Also set $S=0$ and $T=0$.
    \item Pop the next \emph{tree difference} off the stack: an interval with endpoints $\ell < r$,
        a set of edges $\mathcal{R}$ to remove, and a set of edges $\mathcal{A}$ to add.
    \item \emph{Remove} each edge (parent, child) in $\mathcal{R}$ from the tree,
        let $u$ = parent, and while $u != -1$,
        subtract $x[child]$ from $x[u]$ and set $u$ to be parent($u$).
    \item \emph{Add} each edge (parent, child) in $\mathcal{R}$ to the tree,
        let $u$ = parent, and while $u != -1$,
        add $x[child]$ to $x[u]$ and set $u$ to be parent($u$).
    \item For each node that has changed, let $w_\text{old}$ and $w_\text{new}$
        be the weights before and after the change in $x$,
        and add $w_\text{new} - w_text{old}$ to $T$.
    \item Add $T \times (r-\ell)$ to the running total, $S$.
    \item Return to (2).
\end{itemize}
\plr{should modify code to maintain a vector of $w$ values}
The average weighted branch length is obtained by dividing by the total number of sites.

The algorithm for sites is similar, also maintaining the array $x$,
and making use of the association of each mutation with a particular node in the tree.

\paragraph{Total site weight algorithm}
This works as the \emph{total branch length weight algorithm} except that steps (5--6) are replaced with:
\begin{itemize}
    \item[5'] \emph{Find allele weights} for each mutated site in the tree:
        \begin{itemize}
            \item Find mutation weights: 
                    set $u[-1]$ to be the value of $x$ at the root,
                    and then for each mutated node $e$,
                \begin{itemize}
                    \item Set $u[e] = x[e]$ 
                    \item Find $p(e)$, the parent mutation of $e$
                        and update $u[p(e)] -= x[e]$.
                \end{itemize}
            \item Find allele weights: for each unique allele $a$ seen,
                set $v[a]$ to the sum of $u[e]$ across all $e$ with allele $a$.
                Then add $w(v)$ to $T$.
        \end{itemize}
\end{itemize}
\plr{check that algorithm deals correctly with more than one mut per node}

%%%%%%%%%%%%%%%%%%
\subsection*{General properties}

Here we describe some properties that we would like our statistics to have.

\paragraph{They do not depend on choice of reference allele.}
This is achieved by applying the \emph{same} weighting function to every allele.
Ancestral information can be incorporated by including an outgroup or ancestral sequence
as an additional group of size one.
Equivalently, we always work with \emph{unrooted} trees.

\paragraph{They do not depend on sample size.}
We'd like our statistics to be \emph{consistent estimators}
of a single quantity, and should therefore not depend on sample size in expectation.
\plr{insert this above}
This is achieved by always computing probabilies of sampling \emph{without replacement},
i.e., replacing each occurence of $p^k$ with $(x)_k/(n)_k$.
This lets us compare values of a given statistic
that are computed on different data sets with different sample sizes.

\paragraph{Equivalence of ``site'' and ``branch length'' statistics}
The definitions above are made so that for any weighting function,
$\site_w$ is a moment estimator of $\branch_w$, i.e.,
that $\E[\site_w] = \branch_w$ for \plr{somehow say continuum of sites here}.

%%%%%%%%%%%%%%%%%%%%%%%
\section*{Results}

%%%%%%%%%%%%%%%%%%%%%%%
\section*{Discussion}

%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Acknowledgement}

\bibliography{references}

\end{document}

### 2017-10-29 treestats_paper.tex
simpler alg

@@
 This works as the \emph{total branch length weight algorithm} except that steps (5--6) are replaced with:
+\plr{gah this is terrible}
 \begin{itemize}
     \item[5'] \emph{Find allele weights} for each mutated site in the tree:
-        \begin{itemize}
-            \item Find mutation weights: 
-                    set $u[-1]$ to be the value of $x$ at the root,
-                    and then for each mutated node $e$,
-                \begin{itemize}
-                    \item Set $u[e] = x[e]$ 
-                    \item Find $p(e)$, the parent mutation of $e$
-                        and update $u[p(e)] -= x[e]$.
-                \end{itemize}
-            \item Find allele weights: for each unique allele $a$ seen,
-                set $v[a]$ to the sum of $u[e]$ across all $e$ with allele $a$.
-                Then add $w(v)$ to $T$.
-        \end{itemize}
+        if $a$ is the ancestral allele,
+        set $u[a]$ to be the value of $x$ at the root; and then,
+        for each mutation at node $e$ with derived allele $b$ 
+        and parent mutation occurring at node $p$ with derived allele $c$,
+        add $x[e]$ to $u[b]$, and subtract $x[p]$ from $u[c]$.
+    \item[6'] For each unique allele $a$ seen,
+        add $w(v[a])$ to $T$.
 \end{itemize}

### 2018-11-08 treestats_paper.tex
notes from CSHL

@@
 
-\newcommand{\branch}{\mathcal{B}} % branch length stat
+\newcommand{\tskit}{{\texttt{tskit}}}
+\newcommand{\branch}{\mathcal{B}} % branch stat
 \newcommand{\site}{\mathcal{S}} % site stat
+\newcommand{\node}{\mathcal{N}} % node stat
+
+\newcommand{\nw}{{w}} % node weights
+\newcommand{\iw}{{\tilde w}} % initial weights
+\newcommand{\aw}{{\bar w}} % allele weights
+\newcommand{\Tw}{{W}} % allele weights
 
@@
 
+%%% OUTLINE
+% 1. Framework and algorithms for computing general stats
+% 
+%     * Framework:
+% 
+%         - single-site stats are functions of the genotype vector
+%         - in infinite sites, the genotype vector of a mutation is determined by the nodes below it
+%         - ex: divergence as a sum over trees and branches
+%         - ex: ancestry proportions
+%         - general definition, using summary function of weights propagated up the tree
+%             * recursion relating weights on nodes of two forests that differ by a branch
+% 
+%     * Review of tree sequences (quick)
+% 
+%     * Algorithms:
+% 
+%         - weight propagation across tree differences
+%         - count by node length ("Node Statistics")
+%         - count by branch area ("Branch Statistics")
+%         - count by site ("Site Statistics")
+%         - note on outputs (by node, by site, in windows)
+% 
+% 2. Examples: in sections, one of each type of output
+% 
+%     - ancestry proportions
+%     - list of standard summary statistics: divergence, Fst, f4, covariance
+%         * performance for one of these
+%     - PCA: Krylov method
+%         * performance
+% 
+% 3. Correspondence to tree statistics
+% 
+%     - math: in infinite sites, neutral model, E[site] = branch
+%     - plot from simulation showing branch stats are less noisy versions of the site stats
+%     - show time profile of PC1 as computed from rare variants and common variants
+
 %%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\plr{utility of looking at lots of individual-based stats}
-With whole-genome sequence
-we can confidently infer quantities
-between \emph{individuals} that previously had to be estimated
+\plr{need something different here}
+Today's vast quantity of whole-genome sequence
+makes it possible to confidently measure quantities
+between \emph{individuals} that previously had to be roughly estimated
 averaging over populations.
-This gives us much higher resolution, both in space and along the genome.
-With geographic sampling,
-there is potentially a lot of information available
-from looking at all pairwise comparisons.
+For instance, XXX estimated $F_{ST}$ between XX from allozyme sequence,
+but now we know that $F_{ST}$ can be viewed as $1 - t_{within}/t_{total}$ \citep{slatkin_fst}
+which can therefore be estimated precisely by $1 - \pi_{\within}/\pi_{total}$.
 
-\plr{computational challenges}
-Single statistics are not too bad to compute
-since you can get them from allele frequencies,
-but larger summaries, like the matrix of pairwise divergences between $N$ individuals,
-naively involve $O(N^2)$ comparisons.
+% {computational challenges}
+Computation is beginning to be a major problem:
+whole-genome sequence on the scale of the UK Biobank,
+for instance, might have 500K individuals genotyped at 10M loci (??),
+resulting in a genotype matrix of $5 \times 10^{12}$ entries.
+Computing a single summary from this matrix is still relatively quick,
+once loaded into memory,
+but for many applications we are now interested in computing a large number of statistics
+(for instance, consider the roughly $10^{11}$ different possible pairwise divergences).
+Indeed, many sorts of modern inference machinery, such as deep learning,
+depend on being able to compute enough statistics that this becomes a serious bottleneck in practice.
 
-\plr{Something about how} in many applications
-we'd really like to know the trees
-and sequence differences really just stand in for these.
-Example: $F_{ST}$ is X; 
-$\theta_W$ is X;
-divergence is X.
+The \emph{succinct tree sequence} is a data structure that efficiently encodes
+the genealogical trees that describe how a set of individuals are related to each other
+at each point along their genome.
+The data structure was introduced for coalescent simulation by \citet{kelleher2016efficient},
+and refined for forwards simulation by \citet{kelleher2018efficient}.
+These genealogical trees change as one looks across the genome
+because of the results of recombination in ancestors.
+However, nearby genealogical trees tend to share a lot of common structure,
+and single genealogical relationships (e.g., individual $x$ inherits from individual $y$)
+are often shared across relatively long distances,
+which manifest as shared \emph{edges} across many adjacent trees in the tree sequence.
+\citet{kelleher2016efficient}
+showed how this this redundancy can be used
+to both store genome sequence and the associated genealogical relationships extremely compactly,
+as well as to compute statistics extremely efficiently.
+This paper generalizes the latter strategy
+to a much broader class of statistics.
 
-\plr{Review of phylogenetics lit}
-a good review of theoretical properties \citet{semple_and_steel}
+\plr{Review of tree shape stuff in phylogenetics}
+a good review of theoretical properties \citet{semple_and_steel}.
+
+% summary
+In this paper, we present a general framework for defining and efficiently computing
+``single-site'' genetic statistics,
+i.e., statistics of aligned genome sequence that can be expressed as averages over values computed
+separately for each site.
+Patterns of genetic polymorphism in a population 
+are determined by the population's history of Mendelian inheritance,
+as summarized by the underlying genealogical trees,
+and so it is natural (and helpful) to express these statistics
+in terms of the structure of the trees themselves.
+This perspective turns out to both suggest efficient algorithms 
+for computing these statistics on tree sequences,
+and allows us to treat single-site genetic statistics in a more general framework
+of summaries of tree shape and inheritance.
+The methods are implemented in the python package \tskit.
 
 %%%%%%%%%%%%%%%%%%%%%%%
-\section*{Methods}
+\section*{Framework and algorithms}
 
-We consider only so-called \emph{single site} statistics.
-These are statistics computed as a simple \emph{average} 
-of a quantity calcuated across a number of sites.
-For a concrete example, to compute an element of the site frequency spectrum --
-the proportion of sites at which an allele is present in 10\% of the samples, for instance --
-we could 
-(a) assign each site a value, either $+1$ if it has an allele at 10\%, or zero otherwise,
-and
-(b) take the average of these values across all sites.
-As another example,
-to compute the pairwise divergence between two sequences,
-we would similarly assign each site either $+1$ or 0 according to whether the sequences differ,
-and average these values.
+In this paper, 
+we consider only so-called \emph{single site} statistics
+(which we will define concretely shortly).
+Extensions to statistics involving the pairwise joint distribution of genotypes across sites,
+such as linkage disequlibrium,
+are straightforward, and are planned for future work.
+Haplotype-based statistics may require a different class of algorithms.
+
+%%%%%%%%%%%%%%%%%%
+\subsection*{Theoretical framework}
+
+Suppose we have genotyped a collection of $N$ (homologous) chromosomes
+at a sequence of $L$ loci along the genome.
+Suppose first that we wish to compute an additive function of the allele frequency spectrum.
+In other words, suppose that the frequency of allele $a$ at a given site is $p_i(a)$,
+and that the allele frequency spectrum 
+-- i.e., the proportion of alleles whose frequency is $p$ in our sample -- is $f(p)$,
+and for some function $h()$ we want to compute $\sum_p f(p) h(p)$.
+Since this is a simple average over sites,
+we can compute this, summing over the $L$ sites, as 
+$\sum_p f(p) h(p) = (1/L) \sum_{i=1}^L \sum_a h(p_i(a))$.
+This definition applies to multiallelic sites by assuming that for each allele
+we can only see how many samples carry that allele,
+not how the remaining samples are partitioned between one or more other alleles.
+(Note also that this does not differentiate ancestral and derived alleles,
+and so is a function of the \emph{folded} allele frequency spectrum.)
+If we know where each mutation has occurred on the genealogy,
+we can easily find the frequency of the corresponding allele by counting how many samples
+occur below that mutation in the tree (and removing those that subsequently mutate).
+
+There are $2^n$ possible genotype patterns for $n$ haploid samples genotyped at a biallelic locus,
+and so a general statistic of biallelic genotypes could be represented by a vector of length $2^n$.
+A more convenient way to parameterize many functions of genotype patterns
+uses three ingredients:
+\begin{itemize}
+
+    \item[(a)] A list of $K$ vectors of \emph{initial node weights}, $(\iw_1, \ldots, \iw_k)$, 
+        where $\iw_i(j)$ is the ``initial'' value of the $i^\text{th}$ weight assigned to node $j$.
+
+    \item[(b)] The \emph{weight} $\nw$ of a node in a tree
+        is a $k$-vector of weights is $\nw = (\nw_1, \ldots, \nw_k)$,
+        and is equal to the sum of initial weights of all nodes below it in the tree
+        (including itself).
+
+    \item[(c)] A \emph{summary function} $F : \R^k \to \R$ so that 
+        $F(\nw)$ is the ``summary'' statistic we compute
+
+\end{itemize}
+Roughly speaking, we use these to compute a statistic
+by first propagating the initial weights up the trees,
+and then evaluate the summary function on the resulting vectors of weights
+for each allele.
+
+% Sum weights for each allele.
+Concretely, to compute the statistic at a given site $j$,
+for each allele $a$ seen at this site,
+let $\aw_j(a)$, the weight associated with this allele,
+be the sum of the initial weights over all nodes who inherit allele $a$.
+If the site has only one mutation, which produced allele $a$ 
+at node $i$, then $\aw_j(a) = \nw(i)$.
+Then, the statistic computed at site $j$ is
+\begin{align}
+    \site(F, \iw)_j
+    &=
+    \sum_a \aw_j(a) ,
+\end{align}
+and a summary statistic across $L$ sites is
+\begin{align}
+    \bar \site(F, \iw)
+    &=
+    \frac{1}{L} \sum_{j=1}^L \sum_a \aw_j(a) .
+\end{align}
+We call this the \textbf{site statistic} associated with $\iw$ and $F$.
+
+\plr{Figure: example of propagating weights and labeling nodes with weights,
+    in the case that weights are number of samples you are ancestor of.
+    Also, this is a node statistic.}
+
+\plr{Figure: example of branch and site statistics}
+
+This same framework can be used to compute other properties of interest.
+For instance, suppse that we'd like to compute
+what proportion of the genomes of the final generation were contributed by each ancestor.
+In the same framework we can do ancestry proportions.
+\plr{EXPLAIN, DEFINE NODE STATS}
+
+\plr{Introduce branch stats.}
+The \textbf{branch statistic} associated with the branch of length $t$ 
+above node $n$ on a tree of length $\ell$,
+with total weight $\Tw = \sum_j \iw(j)$
+and node weights $\nw$,
+is
+\begin{align}
+    \branch(F, \iw)_n
+    &=
+    \ell t ( F(\Tw - \nw(n)) + F(\nw(n)) ) .
+\end{align}
+This is the summary value that would be added to a site statistic 
+if a single mutation occurred on this branch:
+we must apply the summary function 
+to both the weight below $n$ as well as the weight \emph{not} below $n$
+to account for both derived and ancestral alleles.
+
+\paragraph{Summary}
+We have defined three types of statistic using the same weighting scheme:
+\begin{enumerate}
+    \item \textbf{Site statistics} 
+        -- summarizes weights below each allele at each polymorphic site.
+    \item \textbf{Node statistics} 
+        -- summarizes total weight below each node.
+    \item \textbf{Branch statistics}
+        -- summarizes total weight below each branch, multiplied by branch length.
+\end{enumerate}
+
+\plr{TO-DO:} 
+\begin{itemize}
+
+    \item How will polarization work? 
+        Right now, Site and Branch stats are polarized but Node statistics are not.
+        (To make Node statistics unpolarized, we need to also add $F(\Tw - w)$ 
+        to the weight of each node.)
+        Perhaps this is OK -- Site and Branch stats should correspond to each other,
+        but Node statistics are different, so could have different behavior.
+        BUT, people will need to compute polarized Site and Branch stats;
+        we need to make it easy for them somewhere.
+
+    \item What about normalization?
+        In the initial implementation, Node statistics were divided by the length
+        for which the node was an ancestor to any sample, rather than just total sequence length
+        To make these more analogous to the other stats, we think we want to remove this.
+        The denominator for the original case is easily computable, by using 
+        $F(x) = 1$ if $x>0$ and $F(x)=0$ otherwise, with appropriate (positive) weights.
+
+\end{itemize}
 
@@
 mutations.
+
+%%%%%%%%%%%%%%%%%%
+\subsection*{Algorithms for fast computation}
+
+%%%%%%%%%%%%%%%%%%
+\section*{Case studies}
+% AKA "examples"
+
+%%%%%%%%%%%%%%%%%%
+\subsection*{Ancestry proportions}
+
+%%%%%%%%%%%%%%%%%%
+\subsection*{Population genetics statistics}
+
+%%%%%%%%%%%%%%%%%%
+\subsection*{Genomic PCA}
+
+%%%%%%%%%%%%%%%%%%
+\section*{Duality of site and branch statistics}
+
+Under a neutral, infinite-sites model of mutation with constant mutation rate across time,
+the expected number of mutations per branch is proportional to its length.
+This implies an isormorphism between ``site'' and ``branch'' statistics defined above.
+
+%%%%%%%%%%%%%%%%%%
+\subsection*{The duality}
+
+%%%%%%%%%%%%%%%%%%
+\subsection*{Separating mutational from genealogical noise}
+
+%%%%%%%%%%%%%%%%%%
+\subsection*{Visualizing the time profile of PCA}
+
+%%%%%%%%%%%%%%%%%%%%%%%
+\section*{Discussion}
+
+\paragraph{Diploid statistics}
+The statistics we develop here are in essence \emph{haploid} statistics.
+These tools can be used to compute a statistic about polyploid individuals,
+but they do not scale to large numbers of individuals.
+For instance, to compute the heterozygosity of a set of samples
+-- this is $(1/N) \sum_{i=1}^N 2 p_i (1-p_i)$, where $p_i$ is the derived allele frequency
+in the $i^\text{th}$ individual --
+one would need to have an entry in the weight vector for \emph{each individual}.
+The genotype of a diploid at a site with a single mutation that occurred on the branch above node $n$
+is determined by where that node falls relative to the diploid's two haplotypes in the tree:
+if $n$ is on the path from either haplotype to their MRCA, the individual is heterozygous;
+if $n$ is on the path from the MRCA to the root, the individual is homozygous derived;
+and the individual is homozygous ancestral otherwise.
+This suggests the following generalization of our current scheme for diploids:
+for a list $(I_1, \ldots, I_N)$ of pairs of nodes,
+and two vectors $w^{(1)}$ and $w^{(2)}$ of weights of length $N$,
+add weight $w^{(1)}_i$ to each node on the path between the two nodes of $I_i$,
+and weight $w^{(2)}_i$ to each node on the path from the MRCA of $I_i$ to the root.
+With $k$ such weighting schemes, we would then need the summary function $F$ 
+to take $k \times 2$ values as input.
+Efficiently updating these ``diploid'' weights on the nodes should be possible,
+but is not a straightforward extension of our current method.
+
+%%%%%%%%%%%%%%%%%%%%%%%
+\subsection*{Acknowledgements}
+
+\bibliography{references}
+
+\newpage
+\appendix
+
+{\Large \plr{OLD STUFF}}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Results}
-
-%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Discussion}
-
-%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Acknowledgement}
-
-\bibliography{references}
-
 \end{document}

### 2018-11-09 treestats_paper.tex
covariance notes

@@
 but now we know that $F_{ST}$ can be viewed as $1 - t_{within}/t_{total}$ \citep{slatkin_fst}
-which can therefore be estimated precisely by $1 - \pi_{\within}/\pi_{total}$.
+which can therefore be estimated precisely by $1 - \pi_{within}/\pi_{total}$.
 
@@
 
+Let $G$ denote the genotype matrix, so that $G_{ja} \in \{0,1\}$ 
+is the genotype of the $a^\text{th}$ individual
+at the $j^\text{th}$ site, with `1` denoting the derived state
+(we assume here biallelic markers).
+Also let $P_j = (1/n) \sum_a G{ja}$ denote the vector of allele frequencies.
+Then the \emph{genetic covariance} between samples $a$ and $b$ is
+\begin{align*}
+    C_{ab} 
+        &= \frac{1}{L} \sum_j (G_{ja} - P_j) (G_{jb} - P_j) . %  \\
+        % &= \frac{1}{L} \left( (G - P \bone^T)^T (G - P \bone^T) \right)_{ab} .
+\end{align*}
+This can be computed by letting
+$\iw_a = \delta_a$ and $\iw_b = \delta_b$ 
+(these weights mark nodes above $a$ and $b$ respectively)
+and $\iw_t = \bone / n$
+(this weight gives the proportion of samples below each node).
+Then, with 
+$F(\nw_a, \nw_b, \nw_t) = (\nw_a - \nw_t) (\nw_b - \nw_t)$,
+we can compute the covariance as
+\begin{align*}
+    C_{ab} = \frac{1}{2} \bar \site(F, (\iw_a, \iw_b, \iw_t)) .
+\end{align*}
+This follows, including the factor of two, 
+because the summary function applied to the ancestral allele
+at a site with derived allele frequencies $p_a$, $p_b$, and $p_t$
+in sample $a$, sample $b$, and total, respectively, is
+$F(1 - p_a, 1 - p_b, 1 - p_t) = F(p_a, p_b, p_t)$.
+
+Computing the full covariance matrix for a large number of samples may be infeasible.
+However, we can efficiently find $y^T C z$ for arbitrary vectors $y$ and $z$.
+This works because if we set the initial weights to $y$,
+then each allele's weight is equal to the sum of the entries of $y$
+corresponding to samples carrying that allele.
+Then, if we define 
+$F_{yz}(u, v, t) = (u - (\sum_a y_a) t) (v - (\sum_b z_b) t)$,
+then
+\begin{align} \label{eqn:yCz}
+    \frac{1}{2} \bar \site(F_{yz}, (y, z, \iw_t))
+        &= \frac{1}{L} \sum_j (\sum_a y_a G_{ja} - (\sum_a y_a) P_j) 
+                    (\sum_b z_b G_{jb} - (\sum_b z_b) P_j) \\
+        &= y^T C z .
+\end{align}
+
+\plr{Hmph: to actually compute $Cz$ we need one weight vector for each sample, again.
+    TODO: look up if you can do Krylov just by computing $y^t C z$ for a reasonable number of vectors.  
+    We could compute $Cz$ quickly if we could propagate down the tree:
+    we'd have $(Cz)_a$ equal to the sum of the weights at all nodes \emph{above} $a$. }
+
+Where does the signal from an eigenvector come from?
+Let $y$ be an eigenvector of $C$ with eigenvalue $\lambda$, normalized so that $\|y\|^2 = 1$.
+Since $Cy = \lambda y$,
+we can write $1 = \sum_a y_a^2 = (1/\lambda) y^T C y$.
+However, equation \eqref{eqn:yCz} gives $y^T C y$ as a sum over \emph{sites},
+and thus distributes the variance in $G$ associated with $y$
+across the tree sequence.
+(This is true because of the singular value decomposition of $G$.)
+
 %%%%%%%%%%%%%%%%%%

### 2019-01-28 treestats_paper.tex
Tidy up refs and TS intro.

@@
 Here we describe how to efficiently compute single-site statistics
-from population genetic data
-using efficient methods on tree sequences as implemented in msprime.
+from population genetic data using the succinct tree sequence
+encoding for correlated genealogies.
 The method is general enough to compute any currently-defined statistic,
@@
 For instance, XXX estimated $F_{ST}$ between XX from allozyme sequence,
+% JK: FIXME what's the slatkin ref here?
 but now we know that $F_{ST}$ can be viewed as $1 - t_{within}/t_{total}$ \citep{slatkin_fst}
@@
 
-The \emph{succinct tree sequence} is a data structure that efficiently encodes
-the genealogical trees that describe how a set of individuals are related to each other
-at each point along their genome.
-The data structure was introduced for coalescent simulation by \citet{kelleher2016efficient},
-and refined for forwards simulation by \citet{kelleher2018efficient}.
-These genealogical trees change as one looks across the genome
+The \emph{succinct tree sequence} (or tree sequence, for brevity) is a data
+structure that efficiently encodes the genealogical trees describing how a set of
+individuals are related to each other at each point along their genome. The
+data structure was introduced in the context of coalescent
+simulation~\citep{kelleher2016efficient}, and lead to scalability increases of
+several orders of magnitude over existing methods. It was subsequently extended
+and refined for forwards-time simulations~\citep{kelleher2018efficient,haller2018tree}
+where similar efficiency gains were made. Recent
+work~\citep{kelleher2018inferring} has shown that tree sequence algorithms can
+also be used to massively increase the scalability of methods for inferring
+genome-wide genealogies, and shown that it is feasible to infer trees for
+millions of whole genome samples.
+
+The key to the remarkable scalability and efficiency of tree sequence
+algorithms is the way in which the shared structure in adjacent trees along
+the genome is encoded. As one looks across the genome the genealogies change
 because of the results of recombination in ancestors.
-However, nearby genealogical trees tend to share a lot of common structure,
+However, nearby trees tend to share a lot of common structure,
 and single genealogical relationships (e.g., individual $x$ inherits from individual $y$)
@@
 which manifest as shared \emph{edges} across many adjacent trees in the tree sequence.
-\citet{kelleher2016efficient}
-showed how this this redundancy can be used
+This redundancy can be used
 to both store genome sequence and the associated genealogical relationships extremely compactly,
-as well as to compute statistics extremely efficiently.
-This paper generalizes the latter strategy
-to a much broader class of statistics.
+as well as to compute statistics highly efficiently~\citet{kelleher2016efficient}.
+This paper generalizes the latter strategy to a much broader class of statistics.
 
 \plr{Review of tree shape stuff in phylogenetics}
-a good review of theoretical properties \citet{semple_and_steel}.
+a good review of theoretical properties \citet{semple2003phylogenetics}.
 

### 2019-01-28 treestats_paper.tex

@@
 
-\newcommand{\plr}[1]{{\color{blue} \it #1}}
+\newcommand{\plr}[1]{{\color{blue}\textbf{plr:} \it #1}}
+\newcommand{\jk}[1]{{\color{red}\textbf{jk:} \it #1}}
 
@@
 to both store genome sequence and the associated genealogical relationships extremely compactly,
-as well as to compute statistics highly efficiently~\citet{kelleher2016efficient}.
+as well as to compute statistics highly efficiently~\citep{kelleher2016efficient}.
 This paper generalizes the latter strategy to a much broader class of statistics.
@@
 of summaries of tree shape and inheritance.
-The methods are implemented in the python package \tskit.
+The methods are implemented in the \tskit\ library.
 
@@
 at a sequence of $L$ loci along the genome.
+\jk{Maybe use $n$ here? We use it below}
 Suppose first that we wish to compute an additive function of the allele frequency spectrum.
@@
 not how the remaining samples are partitioned between one or more other alleles.
+\jk{Not clear to me here if we're talking about frequency $0 \leq p \leq 1$ or
+`prevalance`, $0 \leq p \leq N$.}
 (Note also that this does not differentiate ancestral and derived alleles,
@@
 
+\jk{Big jump here. I don't see the connection between the previous paragraph
+and the bitstring representation of haplotype statistics}
 There are $2^n$ possible genotype patterns for $n$ haploid samples genotyped at a biallelic locus,
@@
 
-    \item[(a)] A list of $K$ vectors of \emph{initial node weights}, $(\iw_1, \ldots, \iw_k)$,
+    \item[(a)] A list of $k$ vectors of \emph{initial node weights}, $(\iw_1, \ldots, \iw_k)$,
         where $\iw_i(j)$ is the ``initial'' value of the $i^\text{th}$ weight assigned to node $j$.
+        \jk{Maybe referring to nodes with variables $u, v, w$ and keeping $i,
+        j, k$ for general indexes would help here separate the concepts here?}
 
@@
     \item[(c)] A \emph{summary function} $F : \R^k \to \R$ so that
-        $F(\nw)$ is the ``summary'' statistic we compute
+        $F(\nw)$ is the ``summary'' statistic we compute \jk{I'd find it more
+        intuitive to use $f$ as the notation for a summary function. We can use
+        something else to define the site frequency spectrum above, right? Big
+        $F$ says ``set" to me.}
 
 \end{itemize}
+
+\jk{This formulation of a $k$-vector associated with each node is very
+reminiscent of classical phlogenetics algorithms. Eg the Sankoff algorithm
+for counting the number of mutations needed to explain a set of genotypes
+under parsimony and with arbitrary weighting matrices. See
+\citep[pg.13]{felsenstein2004inferring}. We should emphasise this connection.
+Perhaps a diagram would help here also, showing a tree and a $k$-vector on each
+node?}
+
 Roughly speaking, we use these to compute a statistic

### 2019-05-27 treestats_paper.tex
examples

@@
 
-\newcommand{\nw}{{w}} % node weights
-\newcommand{\iw}{{\tilde w}} % initial weights
+\newcommand{\nw}{x} % node weights
+\newcommand{\iw}{w} % initial weights
 \newcommand{\aw}{{\bar w}} % allele weights
@@
 
-%%%%%%%%%%%%%%%%%%
-\subsection*{Theoretical framework}
-
 Suppose we have genotyped a collection of $N$ (homologous) chromosomes
@@
 occur below that mutation in the tree (and removing those that subsequently mutate).
+What we do next is a generalization of this idea.
 
-\jk{Big jump here. I don't see the connection between the previous paragraph
-and the bitstring representation of haplotype statistics}
-There are $2^n$ possible genotype patterns for $n$ haploid samples genotyped at a biallelic locus,
-and so a general statistic of biallelic genotypes could be represented by a vector of length $2^n$.
 A more convenient way to parameterize many functions of genotype patterns
 uses three ingredients:
+\plr{MODIFY and SHORTEN THIS}
 \begin{itemize}
@@
 
-    \item[(c)] A \emph{summary function} $F : \R^k \to \R$ so that
-        $F(\nw)$ is the ``summary'' statistic we compute \jk{I'd find it more
-        intuitive to use $f$ as the notation for a summary function. We can use
-        something else to define the site frequency spectrum above, right? Big
-        $F$ says ``set" to me.}
+    \item[(c)] A \emph{summary function} $f : \R^k \to \R$ so that
+        $f(\nw)$ is the ``summary'' statistic we compute \jk
 
@@
 
-% Sum weights for each allele.
-Concretely, to compute the statistic at a given site $j$,
+%%%%%%%%%%%%%%%%%%
+\subsection*{Subtree weights}
+
+We could calculate for a given tree how many of 
+a certain set of ``target'' samples are below each of its edges
+by (a) assigning each of those samples weight 1, and other nodes weight zero, then
+(b) finding the total weight of all nodes in the subtree below each edge.
+This total weight is equal to the number of target samples below that edge,
+and it can be calculated efficiently by propagating weights up the tree.
+It turns out to be useful to allow more general weights.
+
+\begin{definition}[Subtree weights]
+    Given a list of \emph{node weights} $w$, and a tree $T$,
+    define the \emph{subtree weight} $x_u$ for each node $u$ to be the sum of all node weights
+    of every node below $u$ in the tree (including $u$ itself):
+    \begin{align*}
+        \nw_u = \sum_{v \,:\, v \le_T u} \iw_v ,
+    \end{align*}
+    where $v \le_T u$ if $u$ is on the path from $v$ back to the root of $T$.
+\end{definition}
+
+We can easily update the vector $x$ as we move along the genome
+XXX as follows or described later.
+
+Next, we will describe several ways to summarize these subtree weights.
+
+%%%%%%%%%%%%%%%%%%
+\subsection*{Site statistics}
+
+The first method of summarizing uses subtree weights
+to compute a summary for each variant site in the genome.
+To compute the statistic at site $j$,
 for each allele $a$ seen at this site,
@@
 be the sum of the initial weights over all nodes who inherit allele $a$.
+Note that if there has been only one mutation at the site,
+then $\aw_j(a)$ is equal to the subtree weight of the node that the mutation producing $a$
+appeared above, 
+while more generally $\aw$ is not necessarily a subtree weight 
+but is easily computable from them.
 If the site has only one mutation, which produced allele $a$
@@
 \begin{align}
-    \site(F, \iw)_j
+    \site(f, \iw)_j
     &=
-    \sum_a \aw_j(a) ,
+    \sum_{a \in D_j} \aw_j(a) ,
 \end{align}
-and a summary statistic across $L$ sites is
+where the sum is over the \emph{derived} alleles $D_j$ at site $j$.
+A summary statistic across $L$ sites is
 \begin{align}
-    \bar \site(F, \iw)
+    \bar \site(f, \iw)
     &=
@@
 \end{align}
-We call this the \textbf{site statistic} associated with $\iw$ and $F$.
+We call this the \textbf{site statistic} associated with $\iw$ and $f$.
+
+\paragraph{Why the derived alleles?}
+We have defined the statistic to sum over only derived alleles,
+i.e., alleles not shared with the root of the tree.
+At a biallelic site, this is equivalent to knowing the derived allele frequency
+among any group of samples,
+and such a statistic will be unchanged under recoding
+if the summary function is unchanged after exchanging $p$ for $1-p$.
+(XXX what is $p$)
+
+\begin{example}[Nucleotide diversity]
+    Suppose we want to compute the mean density of nucleotide differences
+    in a group $S$ of $n$ sequences.
+    To do this, let $\iw_u = 1$ for $u \in S$ and $\iw_v = 0$ otherwise,
+    so that $\nw_u$ gives the number of nodes in $S$ below $u$,
+    and define
+    \begin{align*}
+        f(x) = \frac{2 x (n - x)}{n (n-1)} .
+    \end{align*}
+    Then $\bar \site(f, \iw)$ is mean nucleotide diversity in the region.
+\end{example}
+
+\plr{note LATER about what this does with triallelic sites}
+
+\begin{example}[Nucleotide divergence]
+    Now Suppose we want to compute the mean density of nucleotide differences
+    \emph{between} two nonoverlapping groups of samples, $S_1$ and $S_2$,
+    with $n_1$ and $n_2$ samples, respectively.
+    As before, let $\iw_{ju} = 1$ for $u \in S_j$ and $\iw_{jv} = 0$ otherwise,
+    so that $\nw_{ju}$ gives the number of nodes in $S_j$ below $u$.
+    Then we define
+    \begin{align*}
+        f(x_1, x_2) = \frac{x_1 (n_2 - x_2) + (n_1 - x_1) x_2}{n_1 n_2} .
+    \end{align*}
+    Then $\bar \site(f, \iw)$ is mean nucleotide divergence between the two in the region.
+\end{example}
+
+\begin{example}[Phenotypic correlations]
+    Suppose that each sample has a phenotypic value, denoted $p_u$ for $u \in S$,
+    and we want to compute for each site the correlation of phenotype and genotype.
+    Suppose we have normalized $p$ so that $\sum_u p_u = 0$ and $\sum_u p_u^2 = 1$.
+    Then, if $g$ is a vector of binary genotypes (so $g_u = 1$ if $u$ carries the derived allele),
+    then the covariance of $p$ with $g$ is just $\sum_u p_u g_u = \sum_{u \st g_u = 1} p_u$,
+    i.e., the sum of the phenotypes of all samples carrying the derived allele.
+    This means that if we define $\iw_{1u} = p_u$, and $\iw_{2u} = 1$ for $u \in S$ and zero otherwise,
+    and $f(x_1, x_2) = x_1 / x_2$,
+    then $\site(\iw, f)_j$ is the correlation between $p$ and the allele at site $j$.
+\end{example}
+
+\begin{example}[Patterson's $f_4$]
+    Given four disjoint groups of samples, $S_1$, $S_2$, $S_3$, and $S_4$,
+    Patterson's $f_4(S_1, S_2; S_3, S_4)$ statistic for an allele with frequency $p_i$ in group $S_i$
+    is $(p_1 - p_2)(p_3 - p_4)$ (and then this is typically averaged over sites).
+    Therefore, if as before we let $\iw_{ju} = 1$ if $u \in S_j$,
+    and $f(x_1, x_2, x_3, x_4) = (x_1 n_2 - x_2 n_1) (x_3 n_4 - x_4 n_3) / n_1 n_2 n_3 n_4$,
+    then $\bar \site(\iw, f)$ gives Patterson's $f_4$ across the region.
+\end{example}
 
@@
 
+%%%%%%%%%%%%%%%%%%
+\subsection*{Branch statistics}
+
+\plr{Introduce branch stats.}
+
+The \textbf{branch statistic} associated with the branch of length $t$
+above node $u$ on a tree of length $\ell$,
+with subtree weights $x$,
+is
+\begin{align}
+    \branch(f, \iw)_u
+    &=
+    \ell t  f(\nw(u))  .
+\end{align}
+This is the summary value that would be added to a site statistic
+if a single mutation occurred on this branch,
+multiplied by the expected number of mutations 
+(up to an overall constant related to mutation rate).
+
+\begin{example}[Mean TMRCA]
+    If we take $\iw$ and $f$ exactly as in the ``Nucleotide diversity'' example above,
+    then $f(u)$ gives the probability that the branch above $u$
+    lies on the path from two randomly chosen samples from $S$
+    on the path up to their most recent common ancestor (MRCA).
+    Therefore, $\bar \branch(f, \iw)$ for these choices
+    gives the mean total distance in the tree between two samples from $S$,
+    averaged across the sequence.
+    (This is twice the TMRCA if the samples are all from the same time.)
+\end{example}
+
+\begin{example}[Phenotypic correlation with pedigree]
+    If we take $\iw$ and $f$ as in the ``Phenotypic correlations'' example above,
+    what does $\branch(f, \iw)$ tell us?
+    The statistic tells us the \emph{expected} correlation between phenotype and any mutations
+    appearing on the tree, which is a summary of who much local relatedness
+    aligns with similarity in phenotype.
+\end{example}
+
+\begin{example}[Patterson's $f_4$]
+    Suppose that the four subsets each consist of only a single sample.
+    The summary function $f(x_1, x_2, x_3, x_4)$ for the $f_4$ statistic
+    then assigns weight 1 to any branch that separates $x_1$ and $x_3$ from $x_2$ and $x_4$,
+    and weight -1 to any branch that separates $x_1$ and $x_4$ from $x_2$ and $x_3$.
+    The statistic $\branch(f, \iw)$ therefore 
+    gives the difference in averages of these two types of branch,
+    averaged over choices of samples from the sample sets and averaged across the genome.
+\end{example}
+
+%%%%%%%%%%%%%%%%%%
+\subsection*{Node statistics}
+
+\textbf{Suppose we output per node.}
+
 This same framework can be used to compute other properties of interest.
@@
 
-\plr{Introduce branch stats.}
-The \textbf{branch statistic} associated with the branch of length $t$
-above node $n$ on a tree of length $\ell$,
-with total weight $\Tw = \sum_j \iw(j)$
-and node weights $\nw$,
-is
-\begin{align}
-    \branch(F, \iw)_n
-    &=
-    \ell t ( F(\Tw - \nw(n)) + F(\nw(n)) ) .
-\end{align}
-This is the summary value that would be added to a site statistic
-if a single mutation occurred on this branch:
-we must apply the summary function
-to both the weight below $n$ as well as the weight \emph{not} below $n$
-to account for both derived and ancestral alleles.
+\begin{example}[Ancestry proportions]
+    If $\iw$ are the indicator weights of the set $S$ of $n$ samples,
+    and then $\nw_u / n$ is the proportion of the samples in $S$ that lie below $u$.
+    Therefore, if $f(x) = x / n$,
+    then $\node(f, \iw)_u$ is the proportion of the genomes of $S$
+    that are inherited from (ancestor) $u$.
+\end{example}
 
-\paragraph{Summary}
+%%%%%%%%%%%%%%%%%%%%%
+\subsection*{Summary}
 We have defined three types of statistic using the same weighting scheme:
@@
 
-    \item How will polarization work?
-        Right now, Site and Branch stats are polarized but Node statistics are not.
-        (To make Node statistics unpolarized, we need to also add $F(\Tw - w)$
-        to the weight of each node.)
-        Perhaps this is OK -- Site and Branch stats should correspond to each other,
-        but Node statistics are different, so could have different behavior.
-        BUT, people will need to compute polarized Site and Branch stats;
-        we need to make it easy for them somewhere.
-
     \item What about normalization?
@@
         The denominator for the original case is easily computable, by using
-        $F(x) = 1$ if $x>0$ and $F(x)=0$ otherwise, with appropriate (positive) weights.
+        $f(x) = 1$ if $x>0$ and $f(x)=0$ otherwise, with appropriate (positive) weights.
 
@@
 Then, with
-$F(\nw_a, \nw_b, \nw_t) = (\nw_a - \nw_t) (\nw_b - \nw_t)$,
+$f(\nw_a, \nw_b, \nw_t) = (\nw_a - \nw_t) (\nw_b - \nw_t)$,
 we can compute the covariance as
 \begin{align*}
-    C_{ab} = \frac{1}{2} \bar \site(F, (\iw_a, \iw_b, \iw_t)) .
+    C_{ab} = \frac{1}{2} \bar \site(f, (\iw_a, \iw_b, \iw_t)) .
 \end{align*}
@@
 in sample $a$, sample $b$, and total, respectively, is
-$F(1 - p_a, 1 - p_b, 1 - p_t) = F(p_a, p_b, p_t)$.
+$f(1 - p_a, 1 - p_b, 1 - p_t) = f(p_a, p_b, p_t)$.
 
@@
 Then, if we define
-$F_{yz}(u, v, t) = (u - (\sum_a y_a) t) (v - (\sum_b z_b) t)$,
+$f_{yz}(u, v, t) = (u - (\sum_a y_a) t) (v - (\sum_b z_b) t)$,
 then
 \begin{align} \label{eqn:yCz}
-    \frac{1}{2} \bar \site(F_{yz}, (y, z, \iw_t))
+    \frac{1}{2} \bar \site(f_{yz}, (y, z, \iw_t))
         &= \frac{1}{L} \sum_j (\sum_a y_a G_{ja} - (\sum_a y_a) P_j)

### 2019-05-27 treestats_paper.tex
more words

@@
 
-Theory and computation of single-site statistics of DNA sequences
+Theory and computation of statistics of DNA sequences
 
@@
 of summaries of tree shape and inheritance.
-The methods are implemented in the \tskit\ library.
+The methods are implemented in the \tskit{} library.
 
@@
 
-In this paper,
-we consider only so-called \emph{single site} statistics
-(which we will define concretely shortly).
-Extensions to statistics involving the pairwise joint distribution of genotypes across sites,
-such as linkage disequlibrium,
-are straightforward, and are planned for future work.
-Haplotype-based statistics may require a different class of algorithms.
-
 Suppose we have genotyped a collection of $N$ (homologous) chromosomes
-at a sequence of $L$ loci along the genome.
+at a sequence of $L$ loci along the genome,
 \jk{Maybe use $n$ here? We use it below}
-Suppose first that we wish to compute an additive function of the allele frequency spectrum.
-In other words, suppose that the frequency of allele $a$ at a given site is $p_i(a)$,
-and that the allele frequency spectrum
+and wish to compute statistic of the allele frequency spectrum.
+In other words, if the frequency of the derived allele at a site $i$ is $p_i$,
+and the derived allele frequency spectrum
 -- i.e., the proportion of alleles whose frequency is $p$ in our sample -- is $f(p)$,
-and for some function $h()$ we want to compute $\sum_p f(p) h(p)$.
+then for some function $h()$ we want to compute $\sum_p f(p) h(p)$.
 Since this is a simple average over sites,
 we can compute this, summing over the $L$ sites, as
-$\sum_p f(p) h(p) = (1/L) \sum_{i=1}^L \sum_a h(p_i(a))$.
-This definition applies to multiallelic sites by assuming that for each allele
-we can only see how many samples carry that allele,
-not how the remaining samples are partitioned between one or more other alleles.
-\jk{Not clear to me here if we're talking about frequency $0 \leq p \leq 1$ or
-`prevalance`, $0 \leq p \leq N$.}
-(Note also that this does not differentiate ancestral and derived alleles,
-and so is a function of the \emph{folded} allele frequency spectrum.)
+$\sum_p f(p) h(p) = (1/L) \sum_{i=1}^L h(p_i)$.
+(For the moment we ignore sites with more than two alleles.)
 If we know where each mutation has occurred on the genealogy,
 we can easily find the frequency of the corresponding allele by counting how many samples
-occur below that mutation in the tree (and removing those that subsequently mutate).
-What we do next is a generalization of this idea.
-
-A more convenient way to parameterize many functions of genotype patterns
-uses three ingredients:
-\plr{MODIFY and SHORTEN THIS}
-\begin{itemize}
-
-    \item[(a)] A list of $k$ vectors of \emph{initial node weights}, $(\iw_1, \ldots, \iw_k)$,
-        where $\iw_i(j)$ is the ``initial'' value of the $i^\text{th}$ weight assigned to node $j$.
-        \jk{Maybe referring to nodes with variables $u, v, w$ and keeping $i,
-        j, k$ for general indexes would help here separate the concepts here?}
-
-    \item[(b)] The \emph{weight} $\nw$ of a node in a tree
-        is a $k$-vector of weights is $\nw = (\nw_1, \ldots, \nw_k)$,
-        and is equal to the sum of initial weights of all nodes below it in the tree
-        (including itself).
-
-    \item[(c)] A \emph{summary function} $f : \R^k \to \R$ so that
-        $f(\nw)$ is the ``summary'' statistic we compute \jk
-
-\end{itemize}
+occur below that mutation in the tree (and accounting for those that subsequently mutate).
+What we do here is a generalization of this idea.
 
@@
 
-Roughly speaking, we use these to compute a statistic
-by first propagating the initial weights up the trees,
-and then evaluate the summary function on the resulting vectors of weights
-for each allele.
+Roughly speaking, we will 
+(a) propagate ``weights'' additively up each tree,
+(b) summarize these weights on each branch with a ``summary function'',
+and (c) aggregate these summaries somehow
+(e.g., averaged in windows).
 
@@
 
-We could calculate for a given tree how many of 
-a certain set of ``target'' samples are below each of its edges
-by (a) assigning each of those samples weight 1, and other nodes weight zero, then
+The basic ingredients that we use to summarize tree shape and genotype patterns
+are the ``weights''.
+For instance, we could calculate for a given tree how many of 
+a certain set $S$ of samples are below each of its edges
+by (a) assigning each of these samples weight 1, and other nodes weight zero, then
 (b) finding the total weight of all nodes in the subtree below each edge.
@@
     \begin{align*}
-        \nw_u = \sum_{v \,:\, v \le_T u} \iw_v ,
+        \nw(u) = \sum_{v \,:\, v \le_T u} \iw(v) ,
     \end{align*}
@@
 \end{definition}
+
+We will use weights of the form described above frequently enough
+that we give them a special name --
+for a set $S$ of samples,
+the \emph{indicator weights of $S$} are the weights $\bone_S$ with
+$\bone_S(u) = 1$ if $u \in S$ and $\bone_S(u) = 0$ otherwise.
 
@@
 If the site has only one mutation, which produced allele $a$
-at node $i$, then $\aw_j(a) = \nw(i)$.
+at node $u$, then $\aw_j(a) = \nw(u)$.
 Then, the statistic computed at site $j$ is
@@
 i.e., alleles not shared with the root of the tree.
-At a biallelic site, this is equivalent to knowing the derived allele frequency
-among any group of samples,
-and such a statistic will be unchanged under recoding
-if the summary function is unchanged after exchanging $p$ for $1-p$.
-(XXX what is $p$)
+This allows statistics in principle to differentiate ancestral from derived alleles,
+information which is available in practice (albeit noisily).
+If all sites are biallelic, and the sum of all weights is $\bar \iw = \sum_u \iw(u)$,
+then any statistic defined for an $f$ with $f(x) = f(\bar \iw - x)$
+will in fact not change if the identity of ancestral/derived alleles is flipped.
 
@@
 \end{example}
-
-\plr{note LATER about what this does with triallelic sites}
 
@@
 
+\paragraph{Polyallelic sites}
+The statistics defined above map exactly onto those standard in population genetics
+when all sites are biallelic. 
+If there is a consensus in the field
+for how to use sites with more than two alleles,
+it is to reduce to biallelic data, by either discarding other sites
+or by marking all non-ancestral alleles as (the same) ``derived'' allele.
+We choose to define statistics in a way that still makes formal sense for polyallelic sites,
+so that for instance a site with ancestral allele $A$ and derived alleles $C$ and $T$
+would have site statistic $f(\aw(C)) + f(\aw(T))$.
+This definition is simple, but we make no claims that it is the best or most natural one.
+
 \plr{Figure: example of propagating weights and labeling nodes with weights,
@@
 
-\plr{Introduce branch stats.}
+Genetic variation is informative about many processes
+precisely because it tells us about the underlying patterns of genealogical relatedness.
+In other words, the genomes are useful in so far as they tell us about the trees.
+In the case where we actually have the trees, or a good proxy for them,
+it is natural to summarize them directly, rather than working indirectly with the genotypes.
+If we assume that no two mutations occur at the same genomic position --
+i.e., the \emph{infinite sites model} --
+then there is a natural correspondence between summaries of genotypes and summaries of tree shape.
+If mutations occur at a constant rate in time and along the genome,
+then the \emph{expected} number of mutations that occur somewhere along a branch of a tree
+over some segment of genome
+is equal to the mutation rate multiplied by the length of the segment and by the length of the branch.
+In other words, the ``area'' of a branch in a tree of the tree sequence,
+defined as its span (right minus left endpoint) multiplied by its length (parent time minus child time)
+is equal to the expected number of mutations that will land on it, up to the mutation rate.
+If the mutation rate is constant,
+then this gives us an isometry --
+measuring tree distances in branch lengths versus in numbers of mutations
+are equal in expectation, up to a constant.
 
-The \textbf{branch statistic} associated with the branch of length $t$
+With this in mind, we define the
+\textbf{branch statistic} associated with the branch of length $t$
 above node $u$ on a tree of length $\ell$,
@@
 (up to an overall constant related to mutation rate).
+
+A summary statistic across a region of the genome of length $L$ is
+the average across this bit of genome WRITE THIS SOMEHOW
+\begin{align}
+    \bar \branch(f, \iw)
+    &=
+    \frac{1}{L} \sum_{T \in [0,L]} 
+\end{align}
+We call this the \textbf{site statistic} associated with $\iw$ and $f$.
 
@@
 
+In this paper,
+we consider only so-called \emph{single site} statistics
+(which we will define concretely shortly).
+Extensions to statistics involving the pairwise joint distribution of genotypes across sites,
+such as linkage disequlibrium,
+are straightforward, and are planned for future work.
+Haplotype-based statistics may require a different class of algorithms.
+
 \paragraph{Diploid statistics}

### 2019-05-29 treestats_paper.tex
rewrote definitions

@@
 \usepackage{fullpage}
+
 \usepackage{color}
@@
 \newcommand{\tskit}{{\texttt{tskit}}}
-\newcommand{\branch}{\mathcal{B}} % branch stat
-\newcommand{\site}{\mathcal{S}} % site stat
-\newcommand{\node}{\mathcal{N}} % node stat
+\newcommand{\branch}{\mbox{Branch}} % branch stat
+\newcommand{\branchp}{\mbox{Branch}_+} % polarized
+\newcommand{\site}{\mbox{Site}} % site stat
+\newcommand{\sitep}{\mbox{Site}_+} % polarized
+\newcommand{\node}{\mbox{Node}} % node stat
+\newcommand{\nodep}{\mbox{Node}_+} % polarized
 
-\newcommand{\nw}{x} % node weights
-\newcommand{\iw}{w} % initial weights
-\newcommand{\aw}{{\bar w}} % allele weights
-\newcommand{\Tw}{{W}} % allele weights
+\newcommand{\iw}{w} % sample (initial) weights
+\newcommand{\nw}{x} % subtree (node) weights
+\newcommand{\aw}{{\bar x}} % allele weights
+\newcommand{\treespan}{\ell}
 
@@
 a good review of theoretical properties \citet{semple2003phylogenetics}.
+\jk{This formulation of a $k$-vector associated with each node is very
+reminiscent of classical phlogenetics algorithms. Eg the Sankoff algorithm
+for counting the number of mutations needed to explain a set of genotypes
+under parsimony and with arbitrary weighting matrices. See
+\citep[pg.13]{felsenstein2004inferring}. We should emphasise this connection.}
 
@@
 and wish to compute statistic of the allele frequency spectrum.
-In other words, if the frequency of the derived allele at a site $i$ is $p_i$,
-and the derived allele frequency spectrum
+In other words, if the frequency of allele $a$ at a site $i$ is $p_i(a)$,
+and the allele frequency spectrum
 -- i.e., the proportion of alleles whose frequency is $p$ in our sample -- is $f(p)$,
@@
 we can compute this, summing over the $L$ sites, as
-$\sum_p f(p) h(p) = (1/L) \sum_{i=1}^L h(p_i)$.
-(For the moment we ignore sites with more than two alleles.)
+$$\sum_p f(p) h(p) = (1/L) \sum_{i=1}^L \sum_a h(p_i(a)).$$
+(Note that since the ancestral allele is included, this is a function of the 
+\emph{folded} frequency spectrum, so is invariant to relabeling of alleles.)
 If we know where each mutation has occurred on the genealogy,
@@
 
-\jk{This formulation of a $k$-vector associated with each node is very
-reminiscent of classical phlogenetics algorithms. Eg the Sankoff algorithm
-for counting the number of mutations needed to explain a set of genotypes
-under parsimony and with arbitrary weighting matrices. See
-\citep[pg.13]{felsenstein2004inferring}. We should emphasise this connection.
-Perhaps a diagram would help here also, showing a tree and a $k$-vector on each
+\jk{Perhaps a diagram would help here also, showing a tree and a $k$-vector on each
 node?}
@@
 The basic ingredients that we use to summarize tree shape and genotype patterns
-are the ``weights''.
-For instance, we could calculate for a given tree how many of 
-a certain set $S$ of samples are below each of its edges
-by (a) assigning each of these samples weight 1, and other nodes weight zero, then
-(b) finding the total weight of all nodes in the subtree below each edge.
-This total weight is equal to the number of target samples below that edge,
-and it can be calculated efficiently by propagating weights up the tree.
-It turns out to be useful to allow more general weights.
+are a set of values associated with each sampled node, which we call the ``weights''.
+For instance, we could calculate how many of a certain set $A$ of samples 
+are below a particular edge in a tree
+by (a) assigning each of these samples weight 1 (and other nodes weight zero), then
+(b) finding the total weight of all nodes in the subtree below that edge.
+It turns out to be useful to allow more general weights (including negative ones).
+These subtree-associated weights can be efficiently propagated along a tree sequence
+because adjacent trees are simliar (and so common subtrees do not need to be updated).
+
+We will use weights of the form described above to count numbers of samples in each subtree
+frequently enough that we give them a special name --
+for a subset set $A$ of samples,
+the \emph{indicator weights of $A$} are the sample weights $\bone_A$ with
+$\bone_A(u) = 1$ if $u \in A$ and $\bone_A(u) = 0$ otherwise.
 
 \begin{definition}[Subtree weights]
-    Given a list of \emph{node weights} $w$, and a tree $T$,
-    define the \emph{subtree weight} $x_u$ for each node $u$ to be the sum of all node weights
+    Given a list of \emph{sample weights} $w$, and a tree $T$,
+    define the \emph{subtree weight} $x_u$ for each node $u$ to be the sum of all sample weights
     of every node below $u$ in the tree (including $u$ itself):
     \begin{align*}
-        \nw(u) = \sum_{v \,:\, v \le_T u} \iw(v) ,
+        \nw_T(u) = \sum_{v \,:\, v \le_T u} \iw(v) ,
     \end{align*}
@@
 
-We will use weights of the form described above frequently enough
-that we give them a special name --
-for a set $S$ of samples,
-the \emph{indicator weights of $S$} are the weights $\bone_S$ with
-$\bone_S(u) = 1$ if $u \in S$ and $\bone_S(u) = 0$ otherwise.
-
-We can easily update the vector $x$ as we move along the genome
+We can easily maintain a vector of subtree weights for each tree
+as we iterate through the tree sequence
 XXX as follows or described later.
 
-Next, we will describe several ways to summarize these subtree weights.
+Next, we will describe how to use these subtree weights,
+passed through a ``summary function'' in various ways,
+to summarize genetic variation and tree shapes across the tree sequence.
+\plr{note weights can be vectors}
+
+%%%%%%%%%%%%%%%%%%
+\subsection*{Node statistics}
+
+Propagating indicator weights of a subset of samples up the tree
+lets us easily cont the number of those samples that each node in the tree has inherited.
+Averaged across trees,
+this tells us what proportion of the sample's genomes were inherited from each node.
+Other choices of weights would tell us other information about the nodes.
+Motivated by this, we define the
+\textbf{node statistic} for node $u$
+associated with summary function $f()$ and sample weights $\iw$
+to be the sum of $f()$ applied to the weight of the subtree below node $u$,
+averaged across the region of the genome:
+\begin{align}
+    \node(f, \iw)_u
+    =
+    \sum_T \treespan(T) f(\nw_T(u)) ,
+\end{align}
+where the sum is over trees, and $\treespan(T)$ is the \emph{span} of the tree,
+i.e., the length of the segment of genome that the tree extends for.
+Below, when it is clear from context,
+we will drop the subscript $T$ and write $\nw_u$ for $\nw_T(u)$.
+If this is computed over only a ``window'' of the genome
+then $\treespan(T)$ is replaced by the length of the portion of that window
+that $T$ extends for.
+
+\begin{example}[Ancestry proportions]
+    If $\iw$ are the indicator weights of the set $S$ of $n$ samples,
+    and then $\iw_u / n$ is the proportion of the samples in $S$ that lie below $u$.
+    Therefore, if $f(x) = x / n$,
+    then $\node(f, \iw)_u$ is the proportion of the genomes of $S$
+    that are inherited from (ancestor) $u$.
+\end{example}
 
@@
 
-The first method of summarizing uses subtree weights
-to compute a summary for each variant site in the genome.
-To compute the statistic at site $j$,
-for each allele $a$ seen at this site,
-let $\aw_j(a)$, the weight associated with this allele,
-be the sum of the initial weights over all nodes who inherit allele $a$.
-Note that if there has been only one mutation at the site,
+The first method of summarizing uses weights
+to compute a summary for each site in the genome.
+These are often reported by averaging the values over windows.
+The general definition uses the \emph{allele weights} at each site in the genome,
+defined simply to be the total weight of all nodes who have inherited that allele.
+
+\begin{definition}[Allele weights]
+    The \emph{allele weight} for allele $a$ at site $j$ is the sum of the weights
+    of all nodes inheriting this allele:
+    \begin{align*}
+        \aw_j(a) = \sum_{v \st g_j(v) = a} \iw(v) ,
+    \end{align*}
+    where $g_j(v)$ is the allele carried by node $v$ at site $j$.
+\end{definition}
+
+If there has been only one mutation at the site,
 then $\aw_j(a)$ is equal to the subtree weight of the node that the mutation producing $a$
 appeared above, 
-while more generally $\aw$ is not necessarily a subtree weight 
+while more generally $\aw$ is not necessarily equal to a subtree weight 
 but is easily computable from them.
-If the site has only one mutation, which produced allele $a$
-at node $u$, then $\aw_j(a) = \nw(u)$.
-Then, the statistic computed at site $j$ is
+
+We then define the \textbf{site statistic} of site $j$ computed using a summary function $f()$
+and a set of sample weights $\iw$
+to be the sum of $f()$ applied to each of the allele weights:
 \begin{align}
@@
     &=
+    \sum_{a} f(\aw_j(a)) ,
+\end{align}
+where the sum is over all unique alleles at site $j$.
+
+Of course, we often want to summarize statistics across regions of the genome (``windows'').
+To do this, we overload notation somewhat and use a subscript $[i,j)$ to denote an average
+over the corresponding portion of the genome:
+\begin{align}
+    \site(f, \iw)_{[i,j)}
+    &=
+    \frac{1}{j-i} \sum_{k=i}^{j-1} \sum_a f(\aw_k(a)) .
+\end{align}
+Recall that we have always defined the summary function $f()$ to be zero at monomorphic sites,
+so that the sum will be only over variant sites
+(although the normalization is by total number of sites).
+
+\paragraph{What about polarization?}
+In the definition above we simply sum over all alleles at each site.
+However, sometimes it is useful to distinguish the \emph{ancestral} allele 
+(i.e., the allele at the root of the tree) from the remaining derived alleles.
+This allows statistics in principle to differentiate ancestral from derived alleles,
+information which is available in practice (albeit noisily).
+One way to make use of this information is to sum over only derived alleles.
+We say a site statistic is \textbf{polarized} if we do this,
+and so
+\begin{align}
+    \sitep(f, \iw)_j
+    &=
     \sum_{a \in D_j} \aw_j(a) ,
 \end{align}
-where the sum is over the \emph{derived} alleles $D_j$ at site $j$.
-A summary statistic across $L$ sites is
-\begin{align}
-    \bar \site(f, \iw)
-    &=
-    \frac{1}{L} \sum_{j=1}^L \sum_a \aw_j(a) .
-\end{align}
-We call this the \textbf{site statistic} associated with $\iw$ and $f$.
-
-\paragraph{Why the derived alleles?}
-We have defined the statistic to sum over only derived alleles,
-i.e., alleles not shared with the root of the tree.
-This allows statistics in principle to differentiate ancestral from derived alleles,
-information which is available in practice (albeit noisily).
-If all sites are biallelic, and the sum of all weights is $\bar \iw = \sum_u \iw(u)$,
-then any statistic defined for an $f$ with $f(x) = f(\bar \iw - x)$
-will in fact not change if the identity of ancestral/derived alleles is flipped.
+where $D_j$ denotes the set of all alleles at site $j$ except the allele at the root of the tree.
+Note that this may not be quite what is expected,
+for instance, if there has been a backmutation to the ancestral allele at some point in the tree,
+or if there have been mutations to distinct alleles on different parts of the tree
+such that the ancestral allele is no longer present.
+However, since these situations depend on multiple mutations occurring at a single site,
+they are relatively rare in practice.
 
@@
     in a group $S$ of $n$ sequences.
-    To do this, let $\iw_u = 1$ for $u \in S$ and $\iw_v = 0$ otherwise,
+    To do this, 
+    % let $\iw_u = 1$ for $u \in S$ and $\iw_v = 0$ otherwise,
+    let $\iw = \bone_S$,
     so that $\nw_u$ gives the number of nodes in $S$ below $u$,
@@
     \begin{align*}
-        f(x) = \frac{2 x (n - x)}{n (n-1)} .
+        f(x) = \frac{x (n - x)}{n (n-1)} .
     \end{align*}
@@
     with $n_1$ and $n_2$ samples, respectively.
-    As before, let $\iw_{ju} = 1$ for $u \in S_j$ and $\iw_{jv} = 0$ otherwise,
+    As before, 
+    let $\iw_j = \bone_{S_j}$,
+    % let $\iw_{ju} = 1$ for $u \in S_j$ and $\iw_{jv} = 0$ otherwise,
     so that $\nw_{ju}$ gives the number of nodes in $S_j$ below $u$.
@@
     \begin{align*}
-        f(x_1, x_2) = \frac{x_1 (n_2 - x_2) + (n_1 - x_1) x_2}{n_1 n_2} .
+        f(x_1, x_2) = \frac{x_1 (n_2 - x_2)}{n_1 n_2} .
     \end{align*}
@@
 \begin{example}[Phenotypic correlations]
-    Suppose that each sample has a phenotypic value, denoted $p_u$ for $u \in S$,
-    and we want to compute for each site the correlation of phenotype and genotype.
-    Suppose we have normalized $p$ so that $\sum_u p_u = 0$ and $\sum_u p_u^2 = 1$.
-    Then, if $g$ is a vector of binary genotypes (so $g_u = 1$ if $u$ carries the derived allele),
-    then the covariance of $p$ with $g$ is just $\sum_u p_u g_u = \sum_{u \st g_u = 1} p_u$,
+    Suppose that for each sample $u$ we have a numeric phenotype, denoted $z_u$,
+    and we want to compute the correlation between this phenotype
+    and the genotype at each site in the tree sequence.
+    For convenience, suppose $z$ is normalized so that $\sum_u z_u = 0$ and $\sum_u z_u^2 = 1$.
+    Then, if $g_j$ is a vector of binary genotypes (so $g_j(u) = 1$ if $u$ carries the derived allele),
+    then the covariance of $z$ with $g_j$ is just $\sum_u z_u g_j(u) = \sum_{u \st g_j(u) = 1} z_u$,
     i.e., the sum of the phenotypes of all samples carrying the derived allele.
-    This means that if we define $\iw_{1u} = p_u$, and $\iw_{2u} = 1$ for $u \in S$ and zero otherwise,
-    and $f(x_1, x_2) = x_1 / x_2$,
-    then $\site(\iw, f)_j$ is the correlation between $p$ and the allele at site $j$.
+    Since the phenotypes sum to zero, this is also equal to 
+    $- \sum_{u \st g_j(u) = 0} z_u$.
+    Therefore, if we write $p_j = \sum_u g_j(u)$ for the derived allele frequency,
+    we can write the squared correlation as the sum across alleles as
+    \begin{align*}
+        r_j^2 =
+        \frac{\left( \frac{1}{n} \sum_{u \st g_j(u) = 0} z_u\right)^2}{2p(1-p)} 
+        + \frac{\left( \frac{1}{n} \sum_{u \st g_j(u) = 1} z_u\right)^2}{2p(1-p)}  .
+    \end{align*}
+    We can compute this as a site statistic by defining $\iw_{1u} = z_u$, and $\iw_{2u} = 1$,
+    and $f(x_1, x_2) = x_1^2 / 2 x_2 (n - x_2)$:
+    then $\site(\iw, f)_j = r_j^2$ is the squared correlation between $z$ and the allele at site $j$.
 \end{example}
@@
 \begin{example}[Patterson's $f_4$]
-    Given four disjoint groups of samples, $S_1$, $S_2$, $S_3$, and $S_4$,
-    Patterson's $f_4(S_1, S_2; S_3, S_4)$ statistic for an allele with frequency $p_i$ in group $S_i$
+    Given four disjoint groups of samples, $A_1$, $A_2$, $A_3$, and $A_4$,
+    Patterson's $f_4(A_1, A_2; A_3, A_4)$ statistic for an allele with frequency $p_i$ in group $A_i$
     is $(p_1 - p_2)(p_3 - p_4)$ (and then this is typically averaged over sites).
-    Therefore, if as before we let $\iw_{ju} = 1$ if $u \in S_j$,
-    and $f(x_1, x_2, x_3, x_4) = (x_1 n_2 - x_2 n_1) (x_3 n_4 - x_4 n_3) / n_1 n_2 n_3 n_4$,
-    then $\bar \site(\iw, f)$ gives Patterson's $f_4$ across the region.
+    To rewrite this as a sum over alleles, note that 
+    $p_1 - p_2 = p_1 (1 - p_2) - (1 - p_1) p_2$,
+    and so the statistic counts with positive weight
+    alleles that split $A_1$ and $A_3$ from $A_2$ and $A_4$,
+    and negative weight ones that split $A_1$ and $A_4$ from $A_2$ and $A_3$.
+    Therefore, if as before we 
+    let $\iw_j = \bone_{A_j}$
+    % let $\iw_{ju} = 1$ if $u \in A_j$
+    (so that $\iw_{ju}$ tells us the number of samples in $A_j$ descended from $u$),
+    and
+    \begin{align*}
+        f(x_1, x_2, x_3, x_4)
+        =
+        \frac{x_1}{n_1}
+        \frac{x_3}{n_3}
+        \left(1 - \frac{x_2}{n_2}\right)
+        \left(1 - \frac{x_4}{n_4}\right)
+        -
+        \left(1 - \frac{x_1}{n_1}\right)
+        \left(1 - \frac{x_4}{n_4}\right)
+        \frac{x_2}{n_2}
+        \frac{x_3}{n_3}
+    \end{align*}
+    then $\site(\iw, f)_j$ is equal to Patterson's $f_4(A_1, A_2; A_3, A_4)$
+    at site $j$.
 \end{example}
@@
 would have site statistic $f(\aw(C)) + f(\aw(T))$.
-This definition is simple, but we make no claims that it is the best or most natural one.
+This is a natural defintion in that it agrees with what you'd get
+from looking at haplotype differences.
+For instance, the definition of nucleotide divergence between sets $A_1$ and $A_2$ above
+gives the probability that a randomly chosen sample from each set differ.
+This definition makes sense even at polyallelic sites,
+and (as we discuss below) has a natural correspondence to statistics related to branch lengths.
 
@@
 
-With this in mind, we define the
-\textbf{branch statistic} associated with the branch of length $t$
-above node $u$ on a tree of length $\ell$,
-with subtree weights $x$,
-is
+This makes it natural to define 
+a statistic of tree shape by summing these expected contributions across its branches.
+Concretely, the \textbf{branch statistic} of a tree $T$
+over a region of genome of length $\ell$
+is defined to be
 \begin{align}
-    \branch(f, \iw)_u
+    \branch(f, \iw)_T
     &=
-    \ell t f(\nw(u))  .
+    \ell \sum_{u \in T} t_u f(\nw(u))  ,
 \end{align}
+where $t_u$ is the length of the branch above node $u$ in the tree $T$
+and $\nw(u)$ is the total weight of the subtree below $u$ (as defined above).
 This is the summary value that would be added to a site statistic
@@
 (up to an overall constant related to mutation rate).
+In other words,
+this gives the expected contribution of this branch to $\site(f, \iw)$,
+up to a multiplicative factor of mutation rate.
 
@@
     The statistic tells us the \emph{expected} correlation between phenotype and any mutations
-    appearing on the tree, which is a summary of who much local relatedness
+    appearing on the tree, which is a summary of how much local relatedness
     aligns with similarity in phenotype.
@@
     averaged over choices of samples from the sample sets and averaged across the genome.
-\end{example}
-
-%%%%%%%%%%%%%%%%%%
-\subsection*{Node statistics}
-
-\textbf{Suppose we output per node.}
-
-This same framework can be used to compute other properties of interest.
-For instance, suppse that we'd like to compute
-what proportion of the genomes of the final generation were contributed by each ancestor.
-In the same framework we can do ancestry proportions.
-\plr{EXPLAIN, DEFINE NODE STATS}
-
-\begin{example}[Ancestry proportions]
-    If $\iw$ are the indicator weights of the set $S$ of $n$ samples,
-    and then $\nw_u / n$ is the proportion of the samples in $S$ that lie below $u$.
-    Therefore, if $f(x) = x / n$,
-    then $\node(f, \iw)_u$ is the proportion of the genomes of $S$
-    that are inherited from (ancestor) $u$.
 \end{example}
@@
 
-\plr{Rework these into three algorithms: one to maintain the node weights $x$
+\plr{Rework these into three algorithms: one to maintain the sample weights $\iw$
     (basically already described as `tracked nodes' in msprime)

### 2019-05-30 treestats_paper.tex
description of duality

@@
 
+\newcommand{\treeseq}{\mathbb{T}} % tree sequence
 \newcommand{\iw}{w} % sample (initial) weights
+\newcommand{\tiw}{w_\text{total}} % total sample (initial) weights
 \newcommand{\nw}{x} % subtree (node) weights
 \newcommand{\aw}{{\bar x}} % allele weights
-\newcommand{\treespan}{\ell}
 
@@
 \begin{center}
-    A general class of tree-based statistics,
-    and how to compute them
+    Efficiently summarising relationships in large samples:
+    a general duality between statistics of genealogies and genomes
 \end{center}
 
-\emph{Other title ideas:}
-
-Efficient computation of a general class of DNA sequence statistics
-
-Theory and computation of statistics of DNA sequences
+\emph{Running head:} Genealogies and genomes
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%
-\section*{Framework and algorithms}
+\section*{Framework and statistics}
 
 Suppose we have genotyped a collection of $N$ (homologous) chromosomes
-at a sequence of $L$ loci along the genome,
-\jk{Maybe use $n$ here? We use it below}
+at a sequence of $G$ loci along the genome,
 and wish to compute statistic of the allele frequency spectrum.
@@
 we can compute this, summing over the $L$ sites, as
-$$\sum_p f(p) h(p) = (1/L) \sum_{i=1}^L \sum_a h(p_i(a)).$$
+$$\sum_p f(p) h(p) = (1/G) \sum_{i=1}^G \sum_a h(p_i(a)).$$
 (Note that since the ancestral allele is included, this is a function of the 
@@
 (b) summarize these weights on each branch with a ``summary function'',
-and (c) aggregate these summaries somehow
+and (c) aggregate these summaries to produce statistics
 (e.g., averaged in windows).
+
+%%%%%%%%%%%%%%%%%%
+\subsection*{Tree sequences}
+
+A tree sequence describes how a set of $n$ sampled chromosomes
+are related to each other along a (linear) genome \citep{kelleher2016efficient}.
+These trees are highly correlated,
+a fact which allows them to be stored and processed very efficiently,
+as described in \citet{kelleher2018efficient}.
+This is encoded as a sequence of trees, $\treeseq = (T_1, T_2, \ldots, T_{n(T)})$
+and a sequence of breakpoints $0 = a_0 < a_1 < \cdots < a_{n(T)} = G$,
+where $G$ is the length of the genome
+and $T_k$ describes the genealogical relationships of the samples
+over the segment of genome betwen $a_{k-1}$ (inclusive) and $a_k$ (exclusive).
+We say that tree $T_k$ \emph{covers} the (half-open) segment $[a_{k-1}, a_k)$,
+and call the length of this segment its \emph{span}, denoted $L_k = a_k - a_{k-1}$.
+We refer to the branches in each tree using the node at the bottom,
+so for instance the branch describing a relationship where $u$ is the child of $v$
+is associated with $u$.
+The length of this branch -- i.e., the difference in birth times of $v$ and $u$ --
+we denote $t_T(u)$ (dropping $T$ if it is clear from context).
+Note that although the same parent-child relationship may exist across many adjacent trees
+(this is called an ``edge''),
+rearrangements of genealogical relationships
+can cause the precise set of samples that lie below that edge to differ in different trees.
 
@@
     Given a list of \emph{sample weights} $w$, and a tree $T$,
-    define the \emph{subtree weight} $x_u$ for each node $u$ to be the sum of all sample weights
+    define the \emph{subtree weight} $\nw_T(u)$ for each node $u$ to be the sum of all sample weights
     of every node below $u$ in the tree (including $u$ itself):
@@
 
-We can easily maintain a vector of subtree weights for each tree
-as we iterate through the tree sequence
-XXX as follows or described later.
+We allow vector-valued weights,
+i.e., $\iw(v)$ may be a vector $(\iw_1(v), \ldots, \iw_m(v))$,
+so that when summarising we have access to more than one aspect of each subtree.
+Below, in section XXX, we describe how to efficiently
+maintain a list of the weights of the subtrees below each node
+as we iterate through the tree sequence.
 
@@
 to summarize genetic variation and tree shapes across the tree sequence.
-\plr{note weights can be vectors}
 
@@
 Propagating indicator weights of a subset of samples up the tree
-lets us easily cont the number of those samples that each node in the tree has inherited.
+lets us easily count the number of those samples that each node in the tree has inherited.
 Averaged across trees,
@@
 to be the sum of $f()$ applied to the weight of the subtree below node $u$,
-averaged across the region of the genome:
+averaged across the genome:
 \begin{align}
@@
     =
-    \sum_T \treespan(T) f(\nw_T(u)) ,
+    \frac{1}{G} \sum_k L_k f(\nw_{T_k}(u)) ,
 \end{align}
-where the sum is over trees, and $\treespan(T)$ is the \emph{span} of the tree,
-i.e., the length of the segment of genome that the tree extends for.
-Below, when it is clear from context,
-we will drop the subscript $T$ and write $\nw_u$ for $\nw_T(u)$.
-If this is computed over only a ``window'' of the genome
-then $\treespan(T)$ is replaced by the length of the portion of that window
-that $T$ extends for.
+where the sum is over trees, and $L_k$ is the \emph{span} of the $k^\text{th}$ tree.
+% i.e., the length of the segment of genome that the tree extends for.
+% If this is computed over only a ``window'' of the genome
+% then $L_k$ is replaced by the length of the portion of that window
+% that $T$ extends for.
 
-\begin{example}[Ancestry proportions]
+\begin{example}[Ancestry proportions] \label{ex:ancestry_props}
     If $\iw$ are the indicator weights of the set $S$ of $n$ samples,
@@
     &=
-    \frac{1}{j-i} \sum_{k=i}^{j-1} \sum_a f(\aw_k(a)) .
+    \frac{1}{j-i} \sum_{k=i}^{j-1} \site(f, \iw)_k
 \end{align}
 Recall that we have always defined the summary function $f()$ to be zero at monomorphic sites,
-so that the sum will be only over variant sites
-(although the normalization is by total number of sites).
+so that the sum will be only over polymorphic sites
+(although the normalization is always by total number of sites, to allow comparison between regions).
 
@@
 and so
-\begin{align}
+\begin{align} \label{eqn:site_unpolarized}
     \sitep(f, \iw)_j
     &=
-    \sum_{a \in D_j} \aw_j(a) ,
+    \sum_{a \in D_j} f(\aw_j(a)) ,
 \end{align}
@@
 
-\begin{example}[Nucleotide diversity]
+\begin{example}[Nucleotide diversity] \label{ex:site_diversity}
     Suppose we want to compute the mean density of nucleotide differences
@@
     let $\iw = \bone_S$,
-    so that $\nw_u$ gives the number of nodes in $S$ below $u$,
+    so that $\nw(u)$ gives the number of nodes in $S$ below $u$,
     and define
@@
     \end{align*}
-    Then $\bar \site(f, \iw)$ is mean nucleotide diversity in the region.
+    Then $\site(f, \iw)_{[a,b)}$ is mean nucleotide diversity in the region between $a$ and $b$.
 \end{example}
 
-\begin{example}[Nucleotide divergence]
-    Now Suppose we want to compute the mean density of nucleotide differences
+\begin{example}[Nucleotide divergence] \label{ex:site_divergence}
+    Now suppose we want to compute the mean density of nucleotide differences
     \emph{between} two nonoverlapping groups of samples, $S_1$ and $S_2$,
@@
     let $\iw_j = \bone_{S_j}$,
-    % let $\iw_{ju} = 1$ for $u \in S_j$ and $\iw_{jv} = 0$ otherwise,
-    so that $\nw_{ju}$ gives the number of nodes in $S_j$ below $u$.
+    so that $\nw_{j}(u)$ gives the number of nodes in $S_j$ below $u$.
     Then we define
@@
     \end{align*}
-    Then $\bar \site(f, \iw)$ is mean nucleotide divergence between the two in the region.
+    Then $\site(f, \iw)_{[a,b)}$ is mean nucleotide divergence between the two groups
+    in the region between $a$ and $b$.
 \end{example}
 
-\begin{example}[Phenotypic correlations]
-    Suppose that for each sample $u$ we have a numeric phenotype, denoted $z_u$,
+\begin{example}[Phenotypic correlations] \label{ex:site_correlations}
+    Suppose that for each sample $u$ we have a numeric phenotype, denoted $z(u)$,
     and we want to compute the correlation between this phenotype
     and the genotype at each site in the tree sequence.
-    For convenience, suppose $z$ is normalized so that $\sum_u z_u = 0$ and $\sum_u z_u^2 = 1$.
+    For convenience, suppose $z$ is normalized so that $\sum_u z(u) = 0$ and $\sum_u z(u)^2 = 1$.
     Then, if $g_j$ is a vector of binary genotypes (so $g_j(u) = 1$ if $u$ carries the derived allele),
-    then the covariance of $z$ with $g_j$ is just $\sum_u z_u g_j(u) = \sum_{u \st g_j(u) = 1} z_u$,
+    then the covariance of $z$ with $g_j$ is just $\sum_u z(u) g_j(u) = \sum_{u \st g_j(u) = 1} z(u)$,
     i.e., the sum of the phenotypes of all samples carrying the derived allele.
     Since the phenotypes sum to zero, this is also equal to 
-    $- \sum_{u \st g_j(u) = 0} z_u$.
+    $- \sum_{u \st g_j(u) = 0} z(u)$.
     Therefore, if we write $p_j = \sum_u g_j(u)$ for the derived allele frequency,
@@
         r_j^2 =
-        \frac{\left( \frac{1}{n} \sum_{u \st g_j(u) = 0} z_u\right)^2}{2p(1-p)} 
-        + \frac{\left( \frac{1}{n} \sum_{u \st g_j(u) = 1} z_u\right)^2}{2p(1-p)}  .
+        \frac{\left( \frac{1}{n} \sum_{u \st g_j(u) = 0} z(u)\right)^2}{2p(1-p)} 
+        + \frac{\left( \frac{1}{n} \sum_{u \st g_j(u) = 1} z(u)\right)^2}{2p(1-p)}  .
     \end{align*}
-    We can compute this as a site statistic by defining $\iw_{1u} = z_u$, and $\iw_{2u} = 1$,
+    We can compute this as a site statistic by defining $\iw_{1}(u) = z(u)$, and $\iw_{2}(u) = 1$,
     and $f(x_1, x_2) = x_1^2 / 2 x_2 (n - x_2)$:
@@
 
-\begin{example}[Patterson's $f_4$]
-    Given four disjoint groups of samples, $A_1$, $A_2$, $A_3$, and $A_4$,
-    Patterson's $f_4(A_1, A_2; A_3, A_4)$ statistic for an allele with frequency $p_i$ in group $A_i$
+\begin{example}[Patterson's $f_4$] \label{ex:site_f4}
+    Given four disjoint groups of samples, $S_1$, $S_2$, $S_3$, and $S_4$,
+    Patterson's $f_4(S_1, S_2; S_3, S_4)$ statistic for an allele with frequency $p_i$ in group $S_i$
     is $(p_1 - p_2)(p_3 - p_4)$ (and then this is typically averaged over sites).
@@
     and so the statistic counts with positive weight
-    alleles that split $A_1$ and $A_3$ from $A_2$ and $A_4$,
-    and negative weight ones that split $A_1$ and $A_4$ from $A_2$ and $A_3$.
+    alleles that split $S_1$ and $S_3$ from $S_2$ and $S_4$,
+    and negative weight ones that split $S_1$ and $S_4$ from $S_2$ and $S_3$.
     Therefore, if as before we 
-    let $\iw_j = \bone_{A_j}$
-    % let $\iw_{ju} = 1$ if $u \in A_j$
-    (so that $\iw_{ju}$ tells us the number of samples in $A_j$ descended from $u$),
-    and
+    let $\iw_j = \bone_{S_j}$
+    % let $\iw_{ju} = 1$ if $u \in S_j$
+    (so that $\iw_{ju}$ tells us the number of samples in $S_j$ descended from $u$),
+    and write $n_i$ for the number of samples in $S_i$,
     \begin{align*}
@@
     \end{align*}
-    then $\site(\iw, f)_j$ is equal to Patterson's $f_4(A_1, A_2; A_3, A_4)$
-    at site $j$.
+    then $\site(\iw, f)_{[a,b)}$ is equal to Patterson's $f_4(S_1, S_2; S_3, S_4)$ statistic
+    across the region of the genome between $a$ and $b$.
 \end{example}
@@
 from looking at haplotype differences.
-For instance, the definition of nucleotide divergence between sets $A_1$ and $A_2$ above
+For instance, the definition of nucleotide divergence in Example \ref{ex:site_divergence}
 gives the probability that a randomly chosen sample from each set differ.
@@
 Concretely, the \textbf{branch statistic} of a tree $T$
-over a region of genome of length $\ell$
+obtained from summary function $f()$
+and sample weights $\iw$ with total weight $\tiw = \sum_u \iw(u)$
 is defined to be
@@
     &=
-    \ell \sum_{u \in T} t_u f(\nw(u))  ,
+    \sum_{k=1}^{n(T)} L_k \left\{ f(\nw_{T_k}(u)) + f(\tiw - \nw_{T_k}(u)) \right\}  ,
 \end{align}
@@
 and $\nw(u)$ is the total weight of the subtree below $u$ (as defined above).
-This is the summary value that would be added to a site statistic
-if a single mutation occurred on this branch,
-multiplied by the expected number of mutations 
-(up to an overall constant related to mutation rate).
-In other words,
-this gives the expected contribution of this branch to $\site(f, \iw)$,
-up to a multiplicative factor of mutation rate.
+The term $\tiw - \nw_{T_k}(u)$ gives the total weight \emph{not} in the subtree of $T_k$ below $u$,
+because the sum of all weights in the tree is always constant,
+and so so $\sum_u \nw_T(u) = \tiw$.
+The value $f(\nw(u))$ is the summary value that would be added to a site statistic
+if a single mutation occurred on the branch above $u$,
+and $f(\tiw - \nw_{T_k}(u))$ is the value that would be added due to its complementary allele,
+so $\branch(f, \iw)_T$ gives the expected contribution of this branch to $\site(f, \iw)$,
+per unit of sequence length that the tree $T$ covers.
 
-A summary statistic across a region of the genome of length $L$ is
-the average across this bit of genome WRITE THIS SOMEHOW
+In practice it is probably most useful to average branch statistics
+over a region of the genome,
+which we do by averaging the tree statistics over the region
+with weights proportional to the trees' spans.
+We write this with similar notation:
 \begin{align}
-    \bar \branch(f, \iw)
+    \branch(f, \iw)_{[a,b)}
     &=
-    \frac{1}{L} \sum_{T \in [0,L]} 
+    \frac{1}{b-a} \sum_T \ell_T(a,b) \branch(f, \iw)_T ,
 \end{align}
-We call this the \textbf{site statistic} associated with $\iw$ and $f$.
+where $\ell_T(a,b)$ is the length of the region in $[a,b)$ that the tree $T$ extends for
+(i.e., if $T$ covers the half-open interval $[c,d)$,
+then $\ell_T(a,b) = \max(0.0, \min(b,d) - \max(a,c))$).
 
-\begin{example}[Mean TMRCA]
+The ``unpolarized'' version of this statistic, corresponding to equation \ref{eqn:site_unpolarized},
+omits the second term:
+\begin{align} \label{eqn:branch_unpolarized}
+    \branch_+(f, \iw)_T 
+    &=
+    \sum_{k=1}^{n(T)} L_k f(\nw_{T_k}(u)) .
+\end{align}
+However, the correspondence between the unpolarized statistics is less tight
+than for the polarized ones.
+
+\begin{example}[Mean TMRCA] \label{ex:branch_diversity}
     If we take $\iw$ and $f$ exactly as in the ``Nucleotide diversity'' example above,
@@
 
-\begin{example}[Phenotypic correlation with pedigree]
+\begin{example}[Phenotypic correlation with pedigree] \label{ex:branch_correlation}
     If we take $\iw$ and $f$ as in the ``Phenotypic correlations'' example above,
@@
 
-\begin{example}[Patterson's $f_4$]
+\begin{example}[Patterson's $f_4$] \label{ex:branch_f4}
     Suppose that the four subsets each consist of only a single sample.
@@
 \end{example}
+
+%%%%%%%%%%%%%%%%%%
+\section*{Duality of site and branch statistics}
+
+Under a neutral, infinite-sites model of mutation with constant mutation rate across time,
+the expected number of mutations per branch is proportional to its length.
+This implies an isormorphism between ``site'' and ``branch'' statistics defined above,
+which is discussed in more detail in \citet{ralph2019empirical}.
+For instance, the site statistic of Example \ref{ex:site_diversity} (genetic diversity)
+and the branch statistic of Example \ref{ex:branch_diversity} (mean TMRCA)
+use the same summary function $f(x) = x(n-x)/n(n-1)$.
+These are closely related because under an infinite-sites model of mutation,
+two sequences differ at a site only if there has been a mutation somewhere on the branches going back
+to their most recent common ancestor, 
+and so if mutations occur with constant rate, 
+the expected value of genetic diversity, 
+averaging over mutational noise give the tree sequence,
+is equal to the mutation rate multiplied by the average distance between the two in the trees.
+
+This relationship is true more generally.
+In fact, for any region of the genome between $a$ and $b$,
+\begin{align}
+    \branch(f, \iw)_{[a,b)}
+    =
+    \E\left[ \site(f, \iw)_{[a,b)} \right] ,
+\end{align}
+where the expected value averages over infinite-sites mutations with rate 1
+(and so lengths are measured in units of expected numbers of mutations).
+Let's unpack this statement a bit more: what exactly is the mutational model?
+First, we are taking the mutation rate to be 1, i.e.,
+the expected number of mutations that occur on a region of the genome of length $\ell$
+over $t$ units of time is equal to $\ell t$.
+This just amounts to a change of units --
+for instance, if genome length is measured in nucleotides,
+and the probability of mutation is $\mu$ per nucleotide and per generation,
+then we are measuring times in units of $1/\mu$ generations.
+Second, we are assuming that the probability of per-site mutation is low enough
+that no two mutations occur at the same site
+-- the fact that they do, occasionally, means that this is an approximation.
+Third, we are assuming that mutation rates are constant through time and across the genome.
+Of course, the statement remains true if we can measure distance along the genome and time 
+in a way that mutation rates are constant, but how these vary is generally unknown.
+
+Note that the expected product of two site statistics,
+$\E[\site(f, \iw) \site(g, \iw')]$
+is not equal to $\branch(f, \iw) \branch(g, \iw')$,
+because they are not independent.
+However, it is always possible to define a branch statistic that 
+gives the expected value of the product.
+How to do this is described in \citet{ralph2019empirical}.
+
+This correspondance is useful for several things:
+(1) removing noise of mutations
+(2) understanding what statistics tell us about genealogies
+
+%%%%%%%%%%%%%%%%%%
+\subsection*{Separating mutational from genealogical noise}
+
+\plr{figure showing noise of some site statistic decreasing and approaching the branch statistic
+    as the mutation rate increases}
+
+%%%%%%%%%%%%%%%%%%
+\section*{Efficient algorithms and implementation}
+
+%%%%%%%%%%%%%%%%%%%%%
+\subsection*{Algorithms}
+write-up of algorithm for maintaining branch statistic across trees
+
+%%%%%%%%%%%%%%%%%%%%%
+\subsection*{Speed}
+
+compare speed to scikit-allel
+
+%%%%%%%%%%%%%%%%%%%%%%%
+\section{misc stuff}
 
@@
 (This is true because of the singular value decomposition of $G$.)
-
-%%%%%%%%%%%%%%%%%%
-\section*{Duality of site and branch statistics}
-
-Under a neutral, infinite-sites model of mutation with constant mutation rate across time,
-the expected number of mutations per branch is proportional to its length.
-This implies an isormorphism between ``site'' and ``branch'' statistics defined above.
-
-%%%%%%%%%%%%%%%%%%
-\subsection*{The duality}
-
-%%%%%%%%%%%%%%%%%%
-\subsection*{Separating mutational from genealogical noise}
 

### 2019-05-30 treestats_paper.tex
tweak to L

@@
 
-Suppose we have genotyped a collection of $N$ (homologous) chromosomes
-at a sequence of $G$ loci along the genome,
-and wish to compute statistic of the allele frequency spectrum.
+Suppose we have sequenced a linear genome of $L$ bases from $n$ (homologous) sampled chromosomes,
+and wish to compute a statistic of the allele frequency spectrum.
 In other words, if the frequency of allele $a$ at a site $i$ is $p_i(a)$,
 and the allele frequency spectrum
--- i.e., the proportion of alleles whose frequency is $p$ in our sample -- is $f(p)$,
-then for some function $h()$ we want to compute $\sum_p f(p) h(p)$.
-Since this is a simple average over sites,
-we can compute this, summing over the $L$ sites, as
-$$\sum_p f(p) h(p) = (1/G) \sum_{i=1}^G \sum_a h(p_i(a)).$$
+-- i.e., the proportion of alleles whose frequency is $p$ in our sample -- is $h(p)$,
+then for some function $f()$, we want to compute $\sum_p h(p) f(p)$.
+Since this is a simple average over sites, and monomorphic sites do not contribute,
+we can compute this, summing over the $L$ loci, as
+$$\sum_p h(p) f(p) = (1/L) \sum_{i=1}^L \sum_a f(p_i(a)).$$
 (Note that since the ancestral allele is included, this is a function of the 

### 2019-05-30 treestats_paper.tex
Remove trailing whitespace.
Minor edits.

@@
 a good review of theoretical properties \citet{semple2003phylogenetics}.
-\jk{This formulation of a $k$-vector associated with each node is very
-reminiscent of classical phlogenetics algorithms. Eg the Sankoff algorithm
-for counting the number of mutations needed to explain a set of genotypes
-under parsimony and with arbitrary weighting matrices. See
-\citep[pg.13]{felsenstein2004inferring}. We should emphasise this connection.}
 
@@
 we can compute this, summing over the $L$ loci, as
-$$\sum_p h(p) f(p) = (1/L) \sum_{i=1}^L \sum_a f(p_i(a)).$$
+$$\sum_p h(p) f(p) = \frac{1}{L} \sum_{i=1}^L \sum_a f(p_i(a)).$$
 (Note that since the ancestral allele is included, this is a function of the
@@
 What we do here is a generalization of this idea.
-
-\jk{Perhaps a diagram would help here also, showing a tree and a $k$-vector on each
-node?}
 
@@
 A tree sequence describes how a set of $n$ sampled chromosomes
-are related to each other along a (linear) genome \citep{kelleher2016efficient}.
+are related to each other along a (linear) genome of length $L$ \citep{kelleher2016efficient}.
 These trees are highly correlated,
@@
 as described in \citet{kelleher2018efficient}.
-This is encoded as a sequence of trees, $\treeseq = (T_1, T_2, \ldots, T_{n(T)})$
-and a sequence of breakpoints $0 = a_0 < a_1 < \cdots < a_{n(T)} = G$,
-where $G$ is the length of the genome
-and $T_k$ describes the genealogical relationships of the samples
+This is encoded as a sequence of trees, $\treeseq = (T_1, T_2, \ldots,
+T_{n(\treeseq)})$
+and a sequence of breakpoints $0 = a_0 < a_1 < \cdots < a_{n(T)} = L$,
+where $T_k$ describes the genealogical relationships of the samples
 over the segment of genome betwen $a_{k-1}$ (inclusive) and $a_k$ (exclusive).
@@
 and call the length of this segment its \emph{span}, denoted $L_k = a_k - a_{k-1}$.
-We refer to the branches in each tree using the node at the bottom,
+We refer to the branches in each tree using the most recent node,
 so for instance the branch describing a relationship where $u$ is the child of $v$
@@
 rearrangements of genealogical relationships
-can cause the precise set of samples that lie below that edge to differ in different trees.
+can cause the precise set of samples that lie below that edge to differ across
+trees.
 
 %%%%%%%%%%%%%%%%%%
-\subsection*{Subtree weights}
+\subsection*{Defining a statistic}
+
+\jk{I think we're missing an overall overview of how to specify a statistic
+and basic definitions of the framework itself. Before getting into the
+specifics below, it would be good to give an more abstract idea of
+what's happening. Perhaps with reference to a diagram?}
 
@@
 \begin{definition}[Subtree weights]
-    Given a list of \emph{sample weights} $w$, and a tree $T$,
+    Given a list of \emph{sample weights} $w$ and a tree $T$,
     define the \emph{subtree weight} $\nw_T(u)$ for each node $u$ to be the sum of all sample weights
-    of every node below $u$ in the tree (including $u$ itself):
+    of every node descending from (and including) $u$ in the tree:
     \begin{align*}
@@
     \end{align*}
-    where $v \le_T u$ if $u$ is on the path from $v$ back to the root of $T$.
+    where $v \le_T u$ if $u$ is on the path from $v$ to root in $T$.
 \end{definition}
@@
 so that when summarising we have access to more than one aspect of each subtree.
-Below, in section XXX, we describe how to efficiently
-maintain a list of the weights of the subtrees below each node
-as we iterate through the tree sequence.
+% Below, in section XXX, we describe how to efficiently
+% maintain a list of the weights of the subtrees below each node
+% as we iterate through the tree sequence.
 
@@
 
+\begin{definition}[Summary function]
+    Given a weight vector $(w_1, \dots, w_k)$, the summary function
+    $f: \mathbb{R}^k \rightarrow \mathbb{R}^m$ computes a vector
+    $(\sigma_1, \dots, \sigma_m)$ summarising these weights.
+\end{definition}
+
+\jk{This definition is pretty weak, but wanted something in there.}
+
 %%%%%%%%%%%%%%%%%%
@@
 Propagating indicator weights of a subset of samples up the tree
-lets us easily count the number of those samples that each node in the tree has inherited.
-Averaged across trees,
+lets us easily count the number of those samples that inherit from each node in
+the tree. Averaged across trees,
 this tells us what proportion of the sample's genomes were inherited from each node.
@@
     =
-    \frac{1}{G} \sum_k L_k f(\nw_{T_k}(u)) ,
+    \frac{1}{L} \sum_{k=1}^{n(\treeseq)} L_k f(\nw_{T_k}(u)).
 \end{align}
-where the sum is over trees, and $L_k$ is the \emph{span} of the $k^\text{th}$ tree.
 % i.e., the length of the segment of genome that the tree extends for.
@@
 Note that this may not be quite what is expected,
-for instance, if there has been a backmutation to the ancestral allele at some point in the tree,
+for instance, if there has been a back mutation to the ancestral allele at some point in the tree,
 or if there have been mutations to distinct alleles on different parts of the tree
@@
     &=
-    \sum_{k=1}^{n(T)} L_k \left\{ f(\nw_{T_k}(u)) + f(\tiw - \nw_{T_k}(u)) \right\}  ,
+    \sum_{k=1}^{n(\treeseq)} L_k \left\{ f(\nw_{T_k}(u)) + f(\tiw - \nw_{T_k}(u)) \right\}  ,
 \end{align}
@@
 because the sum of all weights in the tree is always constant,
-and so so $\sum_u \nw_T(u) = \tiw$.
+and so $\sum_u \nw_T(u) = \tiw$.
 The value $f(\nw(u))$ is the summary value that would be added to a site statistic
@@
 (i.e., if $T$ covers the half-open interval $[c,d)$,
-then $\ell_T(a,b) = \max(0.0, \min(b,d) - \max(a,c))$).
+then $\ell_T(a,b) = \max(0, \min(b,d) - \max(a,c))$).
 

### 2019-06-28 treestats_paper.tex
cor refs

@@
     then the covariance of $z$ with $g_j$ is just $\sum_u z(u) g_j(u) = \sum_{u \st g_j(u) = 1} z(u)$,
-    i.e., the sum of the phenotypes of all samples carrying the derived allele.
+    i.e., the sum of the phenotypes of all samples carrying the derived allele,
+    divided $\sqrt{n - 1}$, where $n$ is the number of samples.
     Since the phenotypes sum to zero, this is also equal to 
@@
         r_j^2 =
-        \frac{\left( \frac{1}{n} \sum_{u \st g_j(u) = 0} z(u)\right)^2}{2p(1-p)}
-        + \frac{\left( \frac{1}{n} \sum_{u \st g_j(u) = 1} z(u)\right)^2}{2p(1-p)}  .
+        \frac{\left( \sum_{u \st g_j(u) = 0} z(u)\right)^2}{2p(1-p)n(n-1)} 
+        + \frac{\left( \sum_{u \st g_j(u) = 1} z(u)\right)^2}{2p(1-p)n(n-1)}  .
     \end{align*}
-    We can compute this as a site statistic by defining $\iw_{1}(u) = z(u)$, and $\iw_{2}(u) = 1$,
-    and $f(x_1, x_2) = x_1^2 / 2 x_2 (n - x_2)$:
+    We can compute this as a site statistic by defining $\iw_{1}(u) = z(u)$, and $\iw_{2}(u) = 1/n$,
+    and $f(x_1, x_2) = x_1^2 / (2 x_2 (1 - x_2) n (n-1))$:
     then $\site(\iw, f)_j = r_j^2$ is the squared correlation between $z$ and the allele at site $j$.
 \end{example}
+
+\textbf{Note:} computing correlations with branches instead of SNPs is like stuff done in \citet{zollner2005coalescent} and \citet{minichiello2006mapping}, maybe.
 
@@
 \subsection*{Acknowledgements}
+Thanks to Graham Coop for useful suggestions.
 

### 2019-07-01 treestats_paper.tex
math for regression

@@
 
+%%%%%%%%%%%
+\appendix
+
+\section{Linear regression}
+
+Let $h$ be a trait, $Z$ be a matrix of covariates, and $g$ be a vector denoting inheritance
+(so, with $n$ samples, $h$ and $g$ are both $n$-vectors and $Z$ is an $n \times k$ matrix).
+We would like to find the coefficient of $g$ in the linear regression of $h$ against $g$ and $Z$,
+without doing full multivariate regression for every new $g$,
+using the fact that $Z$ is always the same.
+Suppose that $Z^T Z = I$ and that the vector of all ones is in the span of the columns of $Z$,
+although in the implementation we post-process $Z$ to make this the case.
+Then, let $a$ be the number and $b$ be the $k$-vector satisfying
+\begin{align*}
+    a, b = \text{argmin}\left\{ \sum_i \left( w_i - a g_i - \sum_j Z_{ij} b_j \right)^2 \right\}
+\end{align*}
+Writing this in block matrix notation, $a$ and $b$ minimise
+\begin{align*}
+    \left\| 
+        \left[ \begin{array}{@{}c@{}} h \end{array}\right]
+            - 
+        \left[ \begin{array}{@{}c|c@{}} g & Z \end{array} \right]
+            \left[ \begin{array}{@{}c@{}} a \\ \hline b \end{array} \right]
+    \right\|^2 .
+\end{align*}
+
+Letting $B = [g | Z]$, the solution to this is
+\begin{align*}
+    \left[\begin{array}{@{}c@{}}a\\\hline b\end{array}\right] 
+        = (B^T B)^{-1} B^T h ,
+\end{align*}
+as long as $B^T B$ is invertible (which we assume to be the case).
+Let $p = g^T g$ be the allele frequency, 
+let $u = g^T Z$ be the vector giving sums of the covariates of all samples carrying the allele,
+and
+\begin{align*}
+    \alpha
+    &=
+        (g^T g - g^T Z (Z^T Z)^{-1} Z^T g)^{-1} \\
+    &=
+        \frac{1}{
+            p - \sum_j u_j^2
+        } .
+\end{align*}
+By the inversion formula for a block two-by-two matrix,
+since we have assumed that that $Z^T Z = I$,
+\begin{align*}
+    (B^T B)^{-1}
+    &=
+    \left[
+        \begin{array}{@{}cc@{}}
+            p & u \\
+            u^T & I
+        \end{array}
+    \right]^{-1} \\
+    &=
+    \left[
+        \begin{array}{@{}cc@{}}
+            \alpha
+            &
+            - \alpha u
+            \\
+            - \alpha u^T
+            &
+            p \left(p I - u u^T \right)^{-1}
+        \end{array}
+    \right] .
+\end{align*}
+Now, the regression coefficient we seek is,
+with $h_g = g^T h$,
+\begin{align*}
+    a
+    &=
+    \alpha \left(
+        h_g - u Z^T h
+    \right) .
+\end{align*}
+
+To compute this in the framework above,
+first add a column of 1s to the covariates $Z$,
+then decorrelate the resulting matrix, so that now $Z^T Z = I$.
+Then, put this normalised version of $Z$ 
+into the first $k$ columns of the weight matrix (so that $w_j(u) = Z_{uj}$),
+set the $(k+1)^\text{st}$ column to the trait (so that $w_{k+1}(u) = w_u$),
+and the final column to all $1/n$ (so $w_{k+2}(u) = 1/n$).
+Also let $Z^T h = v$ be precomputed.
+Then the sum of the traits of samples with the focal genotype is $h_g = x_{k+1}$,
+and the allele frequency is $p = x_{k+2}$,
+so that
+\begin{align*}
+    a
+    &=
+    \frac{
+        h_g - \sum_{j=1}^k x_j v_j
+    }{
+        p - \sum_{j=1}^k x_j^2
+    } .
+\end{align*}
+
+In practice we square this and divide by two,
+so that for biallelic loci the two alleles contribute an equal amount.
+For loci with more than two alleles,
+it would be more satisfying to return the proportion of variance in the trait
+that is explained by \emph{all} of the alleles;
+however, this would be more involved
+(it would entail inversion of a $3 \times 3$ matrix for each locus).
+
+% # NUMERICAL CHECK
+% n <- 20
+% k <- 5
+% h <- rnorm(n)
+% g <- rbinom(n, size=1, prob=0.5)
+% oZ <- cbind(matrix(rnorm(n*k), nrow=n), rep(1,n))
+% colnames(oZ) <- c(paste0("oZ", 1:k), "1")
+% Z <- oZ %*% solve(chol(crossprod(oZ)))
+% colnames(Z) <- paste0("Z", 1:(k+1))
+% lm_a <- coef(lm(h ~ g + oZ))[2]
+% hg <- sum(g*h)
+% hZ <- crossprod(Z, h)
+% u <- crossprod(Z, g)
+% p <- sum(g)
+% alpha <- 1/(p - sum(u^2))
+% a <- alpha * (hg - crossprod(u, hZ))
+% 
+% # check formula for regression coefs
+% # and that coefficient of g doesn't change on linear transform of Z
+% B <- cbind(g, Z)
+% H <- cbind(solve(crossprod(B), crossprod(B, h)),
+%            coef(lm(h ~ g + Z + 0)),
+%            coef(lm(h ~ g + oZ + 0)))
+% stopifnot(all(abs(diff(H[1,])) < 2e-15))
+% # check for alpha, an other bits
+% stopifnot(all(abs(crossprod(B, h) - c(hg, hZ)) < 2e-15))
+% stopifnot(all(abs(- alpha * u - solve(crossprod(B))[1,-1]) < 2e-15))
+% stopifnot(abs(solve(crossprod(B))[1,1] - alpha) < 2e-15)
+% stopifnot(all(abs(p*solve(p*diag(k+1) - tcrossprod(u)) - solve(crossprod(B))[-1,-1]) < 2e-15))
+% # and the final answer
+% stopifnot(abs(a - lm_a) < 2e-15)
+
 \end{document}

### 2019-07-03 treestats_paper.tex
math corrections

@@
 as long as $B^T B$ is invertible (which we assume to be the case).
-Let $p = g^T g$ be the allele frequency, 
+Let $m = g^T g$ be the number of alleles in the sample coded 1,
 let $u = g^T Z$ be the vector giving sums of the covariates of all samples carrying the allele,
@@
     &=
-        (g^T g - g^T Z (Z^T Z)^{-1} Z^T g)^{-1} \\
-    &=
-        \frac{1}{
-            p - \sum_j u_j^2
-        } .
-\end{align*}
-By the inversion formula for a block two-by-two matrix,
+        (g^T g - g^T Z (Z^T Z)^{-1} Z^T g) \\
+    &=
+        m - \sum_j u_j^2 .
+\end{align*}
+$\alpha$ is the norm of the component of $g$ not in the subspace spanned by the columns of $Z$,
+so if $\alpha = 0$ then we want to return $a=0$.
+
+Otherwise, by the inversion formula for a block two-by-two matrix,
 since we have assumed that that $Z^T Z = I$,
@@
         \begin{array}{@{}cc@{}}
-            p & u \\
+            m & u \\
             u^T & I
@@
         \begin{array}{@{}cc@{}}
-            \alpha
+            1/\alpha
             &
-            - \alpha u
+            - u / \alpha
             \\
-            - \alpha u^T
+            - u^T / \alpha
             &
-            p \left(p I - u u^T \right)^{-1}
+            m \left(m I - u u^T \right)^{-1}
         \end{array}
@@
     &=
-    \alpha \left(
+    \frac{1}{\alpha} \left(
         h_g - u Z^T h
@@
 set the $(k+1)^\text{st}$ column to the trait (so that $w_{k+1}(u) = w_u$),
-and the final column to all $1/n$ (so $w_{k+2}(u) = 1/n$).
+and the final column to all $1$s (so $w_{k+2}(u) = 1$).
 Also let $Z^T h = v$ be precomputed.
 Then the sum of the traits of samples with the focal genotype is $h_g = x_{k+1}$,
-and the allele frequency is $p = x_{k+2}$,
+and the allele count is $m = x_{k+2}$,
 so that
@@
     }{
-        p - \sum_{j=1}^k x_j^2
-    } .
+        m - \sum_{j=1}^k x_j^2 } .
 \end{align*}
@@
 % p <- sum(g)
-% alpha <- 1/(p - sum(u^2))
-% a <- alpha * (hg - crossprod(u, hZ))
+% alpha <- (p - sum(u^2))
+% a <- (1/alpha) * (hg - crossprod(u, hZ))
 % 
@@
 % stopifnot(all(abs(crossprod(B, h) - c(hg, hZ)) < 2e-15))
-% stopifnot(all(abs(- alpha * u - solve(crossprod(B))[1,-1]) < 2e-15))
-% stopifnot(abs(solve(crossprod(B))[1,1] - alpha) < 2e-15)
+% stopifnot(all(abs(- (1/alpha) * u - solve(crossprod(B))[1,-1]) < 2e-15))
+% stopifnot(abs(solve(crossprod(B))[1,1] - (1/alpha)) < 2e-15)
 % stopifnot(all(abs(p*solve(p*diag(k+1) - tcrossprod(u)) - solve(crossprod(B))[-1,-1]) < 2e-15))

### 2019-07-13 figures/plot_sweep.py
plot diversity around a sweep

@@ -1,0 +1,71 @@
+import pyslim, tskit, msprime
+import numpy as np
+import os, sys
+
+import matplotlib
+matplotlib.use('Agg')
+import matplotlib.pyplot as plt
+
+usage = """
+Usage:
+    {} treefile Ne
+""".format(sys.argv[0])
+
+if len(sys.argv) != 3:
+    raise ValueError(usage)
+
+treefile = sys.argv[1]
+Ne = int(sys.argv[2])
+
+seed = 23
+mut_rate = 1e-9
+
+outbase = ".".join(treefile.split(".")[:-1]) + ".{}".format(mut_rate)
+newtrees = outbase + ".recap.trees"
+
+try:
+    ts = pyslim.load(newtrees)
+except:
+    ots = pyslim.load(treefile)
+    rts = ots.recapitate(1e-8, Ne=Ne)
+    mts = pyslim.SlimTreeSequence(msprime.mutate(rts, rate=mut_rate, random_seed=seed, keep=True))
+    ts = mts.simplify()
+    ts.dump(newtrees)
+
+# get swept mutations and selection coefficients
+swept_id = np.where(np.diff(ts.tables.mutations.metadata_offset) > 0)[0]
+swept = [ts.mutation(m) for m in swept_id]
+s = np.array([mut.metadata[0].selection_coeff for mut in swept])
+spos = np.array([ts.site(mut.site).position for mut in swept])
+
+windows = np.arange(0, ts.sequence_length, 5e5)
+windows[-1] = ts.sequence_length
+x = windows[:-1] + np.diff(windows)/2
+site_div = ts.diversity([ts.samples()], windows=windows, mode='site')
+branch_div = mut_rate * ts.diversity([ts.samples()], windows=windows, mode='branch')
+
+fig = plt.figure(figsize=(6, 4))
+ax = fig.add_subplot(111)
+maxy = 1.1 * max(np.max(np.abs(site_div)), np.max(np.abs(branch_div)))
+ax.set_ylim(0.0, maxy)
+sl, = ax.plot(x, site_div,
+             alpha=1.0,
+             color='black',
+             label="site")
+bl, = ax.plot(x, branch_div,
+             alpha=1.0,
+             color='red',
+             label="branch")
+sl = ax.axvline(spos[0], linestyle=":", label="sweep")
+for k, (pos, sval) in enumerate(zip(spos, s)):
+    if k > 0:
[…]

### 2019-07-14 figures/plot_sweep.py
a good fig

@@ -17,10 +17,10 @@ if len(sys.argv) != 3:
 treefile = sys.argv[1]
 Ne = int(sys.argv[2])
 
-seed = 23
 mut_rate = 1e-9
+nreps = 20
 
-outbase = ".".join(treefile.split(".")[:-1]) + ".{}".format(mut_rate)
+outbase = ".".join(treefile.split(".")[:-1])
 newtrees = outbase + ".recap.trees"
 
 try:
@@ -28,44 +28,203 @@ try:
 except:
     ots = pyslim.load(treefile)
     rts = ots.recapitate(1e-8, Ne=Ne)
-    mts = pyslim.SlimTreeSequence(msprime.mutate(rts, rate=mut_rate, random_seed=seed, keep=True))
-    ts = mts.simplify()
+    ts = rts.simplify()
     ts.dump(newtrees)
 
 # get swept mutations and selection coefficients
-swept_id = np.where(np.diff(ts.tables.mutations.metadata_offset) > 0)[0]
-swept = [ts.mutation(m) for m in swept_id]
-s = np.array([mut.metadata[0].selection_coeff for mut in swept])
-spos = np.array([ts.site(mut.site).position for mut in swept])
+s = np.array([mut.metadata[0].selection_coeff for mut in ts.mutations()])
+spos = np.array([ts.site(mut.site).position for mut in ts.mutations()])
 
 windows = np.arange(0, ts.sequence_length, 5e5)
 windows[-1] = ts.sequence_length
 x = windows[:-1] + np.diff(windows)/2
-site_div = ts.diversity([ts.samples()], windows=windows, mode='site')
-branch_div = mut_rate * ts.diversity([ts.samples()], windows=windows, mode='branch')
 
-fig = plt.figure(figsize=(6, 4))
-ax = fig.add_subplot(111)
-maxy = 1.1 * max(np.max(np.abs(site_div)), np.max(np.abs(branch_div)))
+########################
+# Plot lines for 
+
+all_branch_div = ts.diversity([ts.samples()], windows=windows, mode='branch')
+
+# independent versions of the same rate, low
+
+low_mut_rate = mut_rate
+all_site_div = np.zeros((len(windows) - 1, nreps))
+for k in range(nreps):
+    mts = pyslim.SlimTreeSequence(msprime.mutate(ts, rate=low_mut_rate, keep=True))
[…]

### 2019-07-14 treestats_paper.tex
bigger pop size

@@
 \usepackage{fullpage}
+\usepackage{graphicx}
 
@@
 
+\paragraph{Example: a selective sweep}
+In this view,
+site statistics are noisy approximations to the corresponding branch statistic
+-- but, how noisy?
+How big is the contribution of mutation to the overall sampling variance of a statistic?
+Figure~\ref{fig:sweep_duality} shows this in a particular situation:
+diversity along the chromosome following a few selective sweeps (two full sweeps and three partial).
+The top two plots compare ``branch'' diversity -- i.e., as computed only with tree shape --
+to ``site'' diversity computed from sequence generated by 20 independent assignments of mutations to the same tree sequence,
+with mutation rates $10^{-9}$ and $10^{-8}$, respectively.
+We see that as the mutation rate increases, the signal of decrease in diversity around swept loci becomes more clear,
+and approaches ``branch'' diversity.
+These were computed using the entire population of 1,000 individuals;
+how does sampling variance contribute?
+Not much, it turns out -- the bottom plot shows both site and branch diversity
+computed from 20 nonoverlapping groups of 100 samples.
+Neither site or branch diversity vary much between these samples,
+implying that the subsample gives us a good estimate of the whole-population values of each.
+However, as we see in the top figure,
+whole-population site diversity is itself only a quite noisy estimator of branch diversity.
+
+\begin{figure}
+    \includegraphics{{figures/swept.1000.999.1e-09.diversity}.pdf}
+    \caption{
+        Mean genetic diversity and time to most recent common ancestor
+        in 500Kb windows along a 50Mb genome (0.5 Morgans) following several selective sweeps.
+        In each case, ``site'' is mean genetic diversity (Tajima's $\pi$) divided by mutation rate,
+        and ``branch'' is the corresponding branch statistic described above.
+        The tree sequence was produced by simulating mutations under positive selection with mutation rate $10^{-12}$ 
+        in a population of size 1000 for 400 generations using SLiM,
+        followed by recapitation with $N_e=1000$.
+        The selected alleles at the marked sites have selection coefficients between 0.08 and 0.25,
+        and are at frequencies 96.8\%, 100\%, 16.25\%, 100\%, and 82.6\% in the final generation, respectively.
+        All curves use the same tree sequence, including selected mutations,
+        and with additional neutral mutations added.
+        \textbf{(top)}
+        Diversity within the entire population,
+        computed using for mode ``site'' from 20 independent assignments of mutations to the same tree sequence with mutation rate $\mu = 10^{-9}$.
+        \textbf{(middle)}
+        As in the top panel, but with mutation rate $\mu=10^{-8}$,
+        showing that as mutation rate increases, mode ``site'' (divided by $\mu$) converges to mode ``branch''.
+        \textbf{(bottom)}
+        ``Site'' and ``branch'' diversity within 20 disjoint samples of size 100 each,
+        computed on a single tree sequence with mutation rate $\mu = 10^{-9}$.
+        \label{fig:sweep_duality}
+    }
+\end{figure}
+
 %%%%%%%%%%%%%%%%%%

### 2019-07-14 figures/plot_sweep.py

@@ -34,6 +34,9 @@ except:
 # get swept mutations and selection coefficients
 s = np.array([mut.metadata[0].selection_coeff for mut in ts.mutations()])
 spos = np.array([ts.site(mut.site).position for mut in ts.mutations()])
+freqs = np.array([ts.at(ts.site(mut.site).position).num_samples(mut.node) for mut in ts.mutations()])
+spos = spos[freqs > 0.5 * ts.num_samples]
+s = s[freqs > 0.5 * ts.num_samples]
 
 windows = np.arange(0, ts.sequence_length, 5e5)
 windows[-1] = ts.sequence_length
@@ -70,9 +73,7 @@ sample_sets = []
 for k in range(nreps):
     sample_sets.append(sord[k * sample_size + np.arange(sample_size)])
 
-subts = ts.simplify([x for y in sample_sets for x in y])
-
-mts = pyslim.SlimTreeSequence(msprime.mutate(subts, rate=mut_rate, keep=True))
+mts = pyslim.SlimTreeSequence(msprime.mutate(ts, rate=mut_rate, keep=True))
 sub_branch_div = ts.diversity(sample_sets, windows=windows, mode='branch')
 sub_site_div = (1/mut_rate) * mts.diversity(sample_sets, windows=windows, mode='site')
 
@@ -105,7 +106,7 @@ for k, (pos, sval) in enumerate(zip(spos, s)):
     # at = ax.text(pos, 0.95*maxy, "s={:.2}".format(sval))
 
 ax.set_xlabel("chromosome position (bp)")
-ax.set_ylabel("mean diversity (π)")
+ax.set_ylabel("mean TMRCA (π/μ)")
 leg = ax.legend(
          loc = "upper left",
          bbox_to_anchor = (1.01, 1.01),
@@ -137,7 +138,7 @@ for k, (pos, sval) in enumerate(zip(spos, s)):
     # at = ax.text(pos, 0.95*maxy, "s={:.2}".format(sval))
 
 ax.set_xlabel("chromosome position (bp)")
-ax.set_ylabel("mean diversity (π)")
+ax.set_ylabel("mean TMRCA (π/μ)")
 leg = ax.legend(
          loc = "upper left",
          bbox_to_anchor = (1.01, 1.01),
@@ -173,7 +174,7 @@ for k, (pos, sval) in enumerate(zip(spos, s)):
[…]

### 2019-07-14 figures/plot_sweep.py
removed pyslims to speed this up

@@ -8,16 +8,16 @@ import matplotlib.pyplot as plt
 
 usage = """
 Usage:
-    {} treefile Ne
+    {} treefile Ne mut_rate
 """.format(sys.argv[0])
 
-if len(sys.argv) != 3:
+if len(sys.argv) != 4:
     raise ValueError(usage)
 
 treefile = sys.argv[1]
 Ne = int(sys.argv[2])
+mut_rate = float(sys.argv[3])
 
-mut_rate = 1e-9
 nreps = 20
 
 outbase = ".".join(treefile.split(".")[:-1])
@@ -52,7 +52,7 @@ all_branch_div = ts.diversity([ts.samples()], windows=windows, mode='branch')
 low_mut_rate = mut_rate
 all_site_div = np.zeros((len(windows) - 1, nreps))
 for k in range(nreps):
-    mts = pyslim.SlimTreeSequence(msprime.mutate(ts, rate=low_mut_rate, keep=True))
+    mts = msprime.mutate(ts, rate=low_mut_rate, keep=True)
     all_site_div[:, k] = (1/low_mut_rate) * mts.diversity([ts.samples()], windows=windows, mode='site')[:, 0]
 
 # and independent versions of the same rate, high
@@ -60,7 +60,7 @@ for k in range(nreps):
 high_mut_rate = mut_rate * 10
 all_site_div_high = np.zeros((len(windows) - 1, nreps))
 for k in range(nreps):
-    mts = pyslim.SlimTreeSequence(msprime.mutate(ts, rate=high_mut_rate, keep=True))
+    mts = msprime.mutate(ts, rate=high_mut_rate, keep=True)
     all_site_div_high[:, k] = (1/high_mut_rate) * mts.diversity([ts.samples()], windows=windows, mode='site')[:, 0]
 
 # and for independent subsamples
@@ -73,7 +73,7 @@ sample_sets = []
 for k in range(nreps):
     sample_sets.append(sord[k * sample_size + np.arange(sample_size)])
 
-mts = pyslim.SlimTreeSequence(msprime.mutate(ts, rate=mut_rate, keep=True))
+mts = msprime.mutate(ts, rate=mut_rate, keep=True)
 sub_branch_div = ts.diversity(sample_sets, windows=windows, mode='branch')
 sub_site_div = (1/mut_rate) * mts.diversity(sample_sets, windows=windows, mode='site')
 
@@ -199,7 +199,7 @@ if False:
     for k in range(len(mut_rates)):
[…]

### 2019-07-30 treestats_paper.tex
supp fig
edits to intro

@@
 \usepackage{graphicx}
+\usepackage[hidelinks]{hyperref}
 
@@
 
-\begin{center}
+\title{
     Efficiently summarising relationships in large samples:
-    a general duality between statistics of genealogies and genomes
-\end{center}
-
-\emph{Running head:} Genealogies and genomes
+    a general duality between statistics of genealogies and genomes}
+\author{Peter Ralph, Kevin Thornton, and Jerome Kelleher}
+\maketitle
+
+% \emph{Running head:} Genealogies and genomes
 
 %%%%%%%%%%
-\paragraph{Abstract}
-Here we describe how to efficiently compute single-site statistics
-from population genetic data using the succinct tree sequence
-encoding for correlated genealogies.
-The method is general enough to compute any currently-defined statistic,
-and we use this general framework to suggest a number of new statistics.
-\plr{say what method is: sthg about weighting fns}
+\section*{Abstract}
+
+As a genetic mutation is passed down across generations,
+it distinguishes those genomes that have inherited it from those that have not,
+thus providing a glimpse of the genealogical tree that relates the genomes to each other at that site.
+Summaries of genetic variation thus also aim to summarize the underlying genealogies.
+Here we exploit this duality to create a method that efficiently computes single-site statistics
+from population genetic data using the succinct tree sequence encoding
+of genome sequence and genealogies.
+We use the framework to provide fast implementations of many currently-defined statistics
+(whose run time scales with the logarithm of the number of samples),
+and to suggest several statistics.
+The general approach aggregates ``sample weights'' up each marginal tree,
+which are then summarised using a ``summary function'';
+different statistics result from different choices of weight or function.
+The results can be summarised in three different ways:
+by \emph{site}, which corresponds to statistics calculated from genome sequence,
+by \emph{branch}, which gives the expected value of the site statistic
+with a large number of infinite-sites mutations,
+or by \emph{node}, which summarises the contribution of each node in the tree to the statistics.
 This formulation also makes it easy to see what statistic of the underlying genealogical trees
-each sequence-based statistic is estimating,
-and, if underlying trees are also available,
-the same methods can be used to quickly compute these quantities.
+each sequence-based statistic is estimating.
+
+\plr{add bits for speedup; data}
 
@@
 
-\plr{need something different here}
+It was once a major undertaking to estimate a single summary
+of the many genetic relationships between the individuals of a sample,
+e.g., to estimate $F_{ST}$ between two populations using a dozen allozymes.
 Today's vast quantity of whole-genome sequence
-makes it possible to confidently measure quantities
-between \emph{individuals} that previously had to be roughly estimated
-averaging over populations.
-For instance, XXX estimated $F_{ST}$ between XX from allozyme sequence,
-% JK: FIXME what's the slatkin ref here?
-but now we know that $F_{ST}$ can be viewed as $1 - t_{within}/t_{total}$ \citep{slatkin_fst}
-which can therefore be estimated precisely by $1 - \pi_{within}/\pi_{total}$.
-
-% {computational challenges}
+makes it possible to confidently estimate many more properties of genealogical relationships
+in local regions of genomes \citep{fst_landscapes},
+or between individuals rather than populations \citep{individual_stats}.
 Computation is beginning to be a major problem:
 whole-genome sequence on the scale of the UK Biobank,
-for instance, might have 500K individuals genotyped at 10M loci (??),
+for instance, might have 500K individuals genotyped at 10M loci \citep{ukbb},
 resulting in a genotype matrix of $5 \times 10^{12}$ entries.
@@
 once loaded into memory,
-but for many applications we are now interested in computing a large number of statistics
-(for instance, consider the roughly $10^{11}$ different possible pairwise divergences).
+but for many applications we are now interested in computing a large number of statistics,
+or iteratively refining a single computation.
 Indeed, many sorts of modern inference machinery, such as deep learning,
@@
 
-The \emph{succinct tree sequence} (or tree sequence, for brevity) is a data
+The \emph{succinct tree sequence} (or \emph{tree sequence}, for brevity) is a data
 structure that efficiently encodes the genealogical trees describing how a set of
@@
 data structure was introduced in the context of coalescent
-simulation~\citep{kelleher2016efficient}, and lead to scalability increases of
+simulation~\citep{kelleher2016efficient}, and led to scalability increases of
 several orders of magnitude over existing methods. It was subsequently extended
@@
 The key to the remarkable scalability and efficiency of tree sequence
-algorithms is the way in which the shared structure in adjacent trees along
-the genome is encoded. As one looks across the genome the genealogies change
-because of the results of recombination in ancestors.
+algorithms is the way that shared structure in adjacent trees along the genome is encoded. 
+As one looks across the genome, genealogies change because of the results of recombination in ancestors.
 However, nearby trees tend to share a lot of common structure,
@@
 which manifest as shared \emph{edges} across many adjacent trees in the tree sequence.
-This redundancy can be used
-to both store genome sequence and the associated genealogical relationships extremely compactly,
-as well as to compute statistics highly efficiently~\citep{kelleher2016efficient}.
-This paper generalizes the latter strategy to a much broader class of statistics.
-
-\plr{Review of tree shape stuff in phylogenetics}
-a good review of theoretical properties \citet{semple2003phylogenetics}.
+\citep{kelleher2016efficient} used this redundancy
+to store genome sequence and the associated genealogical relationships extremely compactly,
+as well as to compute several population genetics statistics highly efficiently.
+This paper generalizes the strategy used by \citep{kelleher2016efficient} for statistics computation
+to a much broader class of statistics.
+
+The trees describing genealogical relationships as one moves along the genome
+-- also described by the Ancestral Recombination Graph \citep[][or ARG,]{arg} --
+contain a good deal of information about the history and ongoing dynamics of a population.
+Given sufficiently many mutations, it would be possible to infer the genealogical tree
+at each point along the genome \citep{felsenstein_book},
+and each mutation gives a small piece of information about the tree at that site
+\citep{semple2003phylogenetics}.
+However, population genetics has the task of summarizing relationships across \emph{many} trees,
+each only partially known.
+The basic relationship between tree shape and summaries of genetic variation
+is that if mutations are neutral then the expected number that occur on a given branch
+is proportional to the length of that branch,
+thus establishing a duality between genealogical and genetic summaries,
+under the infinite-sites model of mutation \citep{ralph2019empirical}.
 
 % summary
-In this paper, we present a general framework for defining and efficiently computing
-``single-site'' genetic statistics,
+In this paper, we study only ``single-site'' genetic statistics,
 i.e., statistics of aligned genome sequence that can be expressed as averages over values computed
-separately for each site.
-Patterns of genetic polymorphism in a population
-are determined by the population's history of Mendelian inheritance,
-as summarized by the underlying genealogical trees,
-and so it is natural (and helpful) to express these statistics
-in terms of the structure of the trees themselves.
-This perspective turns out to both suggest efficient algorithms
-for computing these statistics on tree sequences,
-and allows us to treat single-site genetic statistics in a more general framework
-of summaries of tree shape and inheritance.
-The methods are implemented in the \tskit{} library.
+separately for each site,
+leaving other statistics (e.g., pairwise statistics such as linkage disequilibrium
+or haplotype statistics) for later work.
+The methods we describe are implemented, tested, and validated in the \tskit{} library,
+which is available as an open-source Python package at \url{https://github.com/tskit-dev/tskit}.
 
@@
 whole-population site diversity is itself only a quite noisy estimator of branch diversity.
+The same things are shown in Supplementary Figure \ref{sfig:sweep_duality_10K}
+for a simulation with 10,000 individuals.
 
@@
 \subsection*{Acknowledgements}
-Thanks to Graham Coop for useful suggestions.
+Thanks to Graham Coop for useful suggestions
+and to Wilder Wohns for providing the 1000 Genomes tree sequence.
 
@@
 
-\newpage
+\clearpage
 \appendix
+\setcounter{table}{0}
+\renewcommand{\thetable}{S\arabic{table}}
+\setcounter{figure}{0}
+\renewcommand{\thefigure}{S\arabic{figure}}
 
@@
 
+\section*{Supplementary figures}
+
+\begin{figure}
+    \includegraphics{{figures/swept.10000.765.1e-09.diversity}.pdf}
+    \caption{
+        As in Figure~\ref{fig:sweep_duality}, but in a population of size 10,000.
+        \label{sfig:sweep_duality_10K}
+    }
+\end{figure}
+
 \end{document}

### 2019-07-31 treestats_paper.tex
figure and more edits

@@
 
-Suppose we have sequenced a linear genome of $L$ bases from $n$ (homologous) sampled chromosomes,
-and wish to compute a statistic of the allele frequency spectrum.
-In other words, if the frequency of allele $a$ at a site $i$ is $p_i(a)$,
-and the allele frequency spectrum
--- i.e., the proportion of alleles whose frequency is $p$ in our sample -- is $h(p)$,
-then for some function $f()$, we want to compute $\sum_p h(p) f(p)$.
-Since this is a simple average over sites, and monomorphic sites do not contribute,
-we can compute this, summing over the $L$ loci, as
-$$\sum_p h(p) f(p) = \frac{1}{L} \sum_{i=1}^L \sum_a f(p_i(a)).$$
+Many population genetic summary statistics can be defined using the allele frequency spectrum
+of a collection of homologous sequences.
+Concretely, write $p_i(a)$ for the frequency of allele $a$ at a site $i$,
+and $h(p)$ for the allele frequency spectrum
+-- i.e., the number of alleles whose frequency is $p$ in our sample --
+and suppose that for some function $f()$, we want to compute $\sum_p h(p) f(p)$.
+(For instance, mean genetic diversity is computed with $h(p) = p (1-p)$.)
+Since this is a simple sum over sites,
+we can compute this by summing over the $L$ loci, as
+\begin{align*}
+    \sum_p h(p) f(p) = \sum_{i=1}^L \sum_a f(p_i(a)).
+\end{align*}
 (Note that since the ancestral allele is included, this is a function of the
-\emph{folded} frequency spectrum, so is invariant to relabeling of alleles.)
+is invariant to relabeling of alleles.)
 If we know where each mutation has occurred on the genealogy,
@@
 occur below that mutation in the tree (and accounting for those that subsequently mutate).
+
 What we do here is a generalization of this idea.
-
-Roughly speaking, we will
-(a) propagate ``weights'' additively up each tree,
-(b) summarize these weights on each branch with a ``summary function'',
-and (c) aggregate these summaries to produce statistics
+Roughly speaking, to do this we will
+propagate ``weights'' additively up each tree,
+summarize these weights on each branch with a ``summary function'',
+and aggregate these summaries to produce statistics
 (e.g., averaged in windows).
+For instance, we could calculate how many of a certain set $A$ of samples
+are below a particular edge in a tree
+by (a) assigning each of these samples weight 1 (and other nodes weight zero), then
+(b) finding the total weight of all nodes in the subtree below that edge.
+It turns out to be useful to allow more general weights (including negative ones).
+Figure \ref{fig:divergence_diagram} shows a simple example of these procedures
+of propagating weights up a tree and summarising them,
+used to calculate mean sequence divergence over a short sequence spanned by a single tree.
+
+\begin{figure}
+    \centering
+    \includegraphics{figures/divergence_diagram}
+    \caption{
+        An example of computing sequence divergence between red nodes and blue nodes
+        from seven mutations at distinct sites in a sequence of length 10 using the tree.
+        The summary function, $f(a,b) = (a/3)(1-b/2)$ gives the probability that
+        an allele found in $a$ of the red nodes and $b$ of the blue nodes is present in
+        a randomly chosen red node but not in a randomly chosen blue node.
+        \textbf{(a)} Weights are propagated up the tree by adding together the weights of any children,
+        plus the weight of the node itself, if it is a sample.
+        \textbf{(b)} In this example, weights $(a,b)$ record how many red nodes ($a$)
+        and how many blue nodes ($b$) are below each node.
+        The weights at each node is the sum of the weights of all descendant sample nodes.
+        \textbf{(c)} The summary function shown is applied to both the weight below 
+        and the weight not below each mutation,
+        so for instance the mutation above the node with weights $(2,1)$
+        contributes $f(2,1) + f(1,1) = 1/3 + 1/6 = 1/2$.
+        The final value for divergence is the sum of these values divided by 10,
+        the total number of sites.
+        \label{fig:divergence_diagram}
+    }
+\end{figure}
+
+We now begin to formalize these notions with definitions.
+
+\begin{definition}[Subtree weights]
+    Given a list of \emph{sample weights} $w$ and a tree $T$,
+    define the \emph{subtree weight} $\nw_T(u)$ for each node $u$ to be the sum of all sample weights
+    of every node descending from (and including) $u$ in the tree:
+    \begin{align*}
+        \nw_T(u) = \sum_{v \,:\, v \le_T u} \iw(v) ,
+    \end{align*}
+    where $v \le_T u$ if $u$ is on the path from $v$ to root in $T$.
+\end{definition}
+
+\begin{definition}[Summary function]
+    For a set of $k$-dimensional weights with total weight $\tiw$,
+    a summary function is a real-valued function $f(w_1, \ldots, w_k)$
+    with the property that $f(0) = f(\tiw) = 0$.
+\end{definition}
+
+We allow vector-valued weights,
+i.e., $\iw(v)$ may be a vector $(\iw_1(v), \ldots, \iw_m(v))$,
+so that when summarising we have access to more than one aspect of each subtree.
+The requirement that $f(0) = f(\tiw) = 0$ is for consistency,
+so that statistics do not depend on portions of the tree not ancestral to any of the samples.
+% Below, in section XXX, we describe how to efficiently
+% maintain a list of the weights of the subtrees below each node
+% as we iterate through the tree sequence.
+In practice we allow summary functions to be vector-valued (so as to compute many statistics at once)
+but this will not feature in our examples.
+Next, we will describe how to use these subtree weights,
+passed through a ``summary function'' in various ways,
+to summarize genetic variation and tree shapes across the tree sequence.
 
@@
 
-A tree sequence describes how a set of $n$ sampled chromosomes
-are related to each other along a (linear) genome of length $L$ \citep{kelleher2016efficient}.
-These trees are highly correlated,
-a fact which allows them to be stored and processed very efficiently,
-as described in \citet{kelleher2018efficient}.
-This is encoded as a sequence of trees, $\treeseq = (T_1, T_2, \ldots,
-T_{n(\treeseq)})$
+Before defining the different classes of statistic,
+it will help to establish a little more notation.
+A tree sequence efficiently describes how a set of $n$ sampled chromosomes
+are related to each other along a (linear) genome of length $L$ \citep{kelleher2016efficient,kelleher2018efficient}.
+% These trees are highly correlated,
+% a fact which allows them to be stored and processed very efficiently,
+% as described in \citet{kelleher2018efficient}.
+From the tree sequence can be extracted a sequence of trees,
+$\treeseq = (T_1, T_2, \ldots, T_{n(\treeseq)})$
 and a sequence of breakpoints $0 = a_0 < a_1 < \cdots < a_{n(T)} = L$,
@@
 is associated with $u$.
-The length of this branch -- i.e., the difference in birth times of $v$ and $u$ --
-we denote $t_T(u)$ (dropping $T$ if it is clear from context).
+The \emph{length} of this branch is the difference in birth times of $v$ and $u$,
+and is denoted $t_T(u)$ (dropping $T$ if it is clear from context).
 Note that although the same parent-child relationship may exist across many adjacent trees
@@
 %%%%%%%%%%%%%%%%%%
-\subsection*{Defining a statistic}
-
-\jk{I think we're missing an overall overview of how to specify a statistic
-and basic definitions of the framework itself. Before getting into the
-specifics below, it would be good to give an more abstract idea of
-what's happening. Perhaps with reference to a diagram?}
-
-The basic ingredients that we use to summarize tree shape and genotype patterns
-are a set of values associated with each sampled node, which we call the ``weights''.
-For instance, we could calculate how many of a certain set $A$ of samples
-are below a particular edge in a tree
-by (a) assigning each of these samples weight 1 (and other nodes weight zero), then
-(b) finding the total weight of all nodes in the subtree below that edge.
-It turns out to be useful to allow more general weights (including negative ones).
-These subtree-associated weights can be efficiently propagated along a tree sequence
-because adjacent trees are simliar (and so common subtrees do not need to be updated).
-
-We will use weights of the form described above to count numbers of samples in each subtree
-frequently enough that we give them a special name --
-for a subset set $A$ of samples,
-the \emph{indicator weights of $A$} are the sample weights $\bone_A$ with
-$\bone_A(u) = 1$ if $u \in A$ and $\bone_A(u) = 0$ otherwise.
-
-\begin{definition}[Subtree weights]
-    Given a list of \emph{sample weights} $w$ and a tree $T$,
-    define the \emph{subtree weight} $\nw_T(u)$ for each node $u$ to be the sum of all sample weights
-    of every node descending from (and including) $u$ in the tree:
-    \begin{align*}
-        \nw_T(u) = \sum_{v \,:\, v \le_T u} \iw(v) ,
-    \end{align*}
-    where $v \le_T u$ if $u$ is on the path from $v$ to root in $T$.
-\end{definition}
-
-We allow vector-valued weights,
-i.e., $\iw(v)$ may be a vector $(\iw_1(v), \ldots, \iw_m(v))$,
-so that when summarising we have access to more than one aspect of each subtree.
-% Below, in section XXX, we describe how to efficiently
-% maintain a list of the weights of the subtrees below each node
-% as we iterate through the tree sequence.
-
-Next, we will describe how to use these subtree weights,
-passed through a ``summary function'' in various ways,
-to summarize genetic variation and tree shapes across the tree sequence.
-
-\begin{definition}[Summary function]
-    Given a weight vector $(w_1, \dots, w_k)$, the summary function
-    $f: \mathbb{R}^k \rightarrow \mathbb{R}^m$ computes a vector
-    $(\sigma_1, \dots, \sigma_m)$ summarising these weights.
-\end{definition}
-
-\jk{This definition is pretty weak, but wanted something in there.}
-
-%%%%%%%%%%%%%%%%%%
 \subsection*{Node statistics}
 
-Propagating indicator weights of a subset of samples up the tree
-lets us easily count the number of those samples that inherit from each node in
-the tree. Averaged across trees,
+Perhaps the most natural way to summarise weights on the tree
+is simply to examine the values at each node.
+This would allow us, for instance, to 
+count the number of samples that inherit from each node in the tree. 
+Averaged across trees,
 this tells us what proportion of the sample's genomes were inherited from each node.
-Other choices of weights would tell us other information about the nodes.
 Motivated by this, we define the
@@
 
+We will use weights that count number of samples below each node
+sufficiently often that we give them a special name --
+for a subset set $A$ of samples,
+the \emph{indicator weights of $A$} are the sample weights $\bone_A$ with
+$\bone_A(u) = 1$ if $u \in A$ and $\bone_A(u) = 0$ otherwise.
+
 \begin{example}[Ancestry proportions] \label{ex:ancestry_props}
@@
 
+Node statistics only make sense in the context of the tree sequence,
+but provide a useful bridge to the next type of statistic,
+which are defined directly in terms of the genotypes
+(and so do not depend on the tree sequence).
+
 %%%%%%%%%%%%%%%%%%
@@
 
-The first method of summarizing uses weights
+The next mode of summarizing uses weights
 to compute a summary for each site in the genome.
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%
-\section{misc stuff}
-
-%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Summary}
-We have defined three types of statistic using the same weighting scheme:
-\begin{enumerate}
-    \item \textbf{Site statistics}
-        -- summarizes weights below each allele at each polymorphic site.
-    \item \textbf{Node statistics}
-        -- summarizes total weight below each node.
-    \item \textbf{Branch statistics}
-        -- summarizes total weight below each branch, multiplied by branch length.
-\end{enumerate}
-
-\plr{TO-DO:}
-\begin{itemize}
-
-    \item What about normalization?
-        In the initial implementation, Node statistics were divided by the length
-        for which the node was an ancestor to any sample, rather than just total sequence length
-        To make these more analogous to the other stats, we think we want to remove this.
-        The denominator for the original case is easily computable, by using
-        $f(x) = 1$ if $x>0$ and $f(x)=0$ otherwise, with appropriate (positive) weights.
-
-\end{itemize}
-
-%%%%%%%%%%%%%%%%%%
-\subsection*{Tree sequences}
-
-Quick review of tree sequence terminology
-and storage:
-nodes,
-edges,
-sites,
-mutations.
-
-%%%%%%%%%%%%%%%%%%
-\subsection*{Algorithms for fast computation}
-
-%%%%%%%%%%%%%%%%%%
-\section*{Case studies}
-% AKA "examples"
-
-%%%%%%%%%%%%%%%%%%
-\subsection*{Ancestry proportions}
-
-%%%%%%%%%%%%%%%%%%
-\subsection*{Population genetics statistics}
-
-%%%%%%%%%%%%%%%%%%
-\subsection*{Genomic PCA}
-
-Let $G$ denote the genotype matrix, so that $G_{ja} \in \{0,1\}$
-is the genotype of the $a^\text{th}$ individual
-at the $j^\text{th}$ site, with `1` denoting the derived state
-(we assume here biallelic markers).
-Also let $P_j = (1/n) \sum_a G{ja}$ denote the vector of allele frequencies.
-Then the \emph{genetic covariance} between samples $a$ and $b$ is
-\begin{align*}
-    C_{ab}
-        &= \frac{1}{L} \sum_j (G_{ja} - P_j) (G_{jb} - P_j) . %  \\
-        % &= \frac{1}{L} \left( (G - P \bone^T)^T (G - P \bone^T) \right)_{ab} .
-\end{align*}
-This can be computed by letting
-$\iw_a = \delta_a$ and $\iw_b = \delta_b$
-(these weights mark nodes above $a$ and $b$ respectively)
-and $\iw_t = \bone / n$
-(this weight gives the proportion of samples below each node).
-Then, with
-$f(\nw_a, \nw_b, \nw_t) = (\nw_a - \nw_t) (\nw_b - \nw_t)$,
-we can compute the covariance as
-\begin{align*}
-    C_{ab} = \frac{1}{2} \bar \site(f, (\iw_a, \iw_b, \iw_t)) .
-\end{align*}
-This follows, including the factor of two,
-because the summary function applied to the ancestral allele
-at a site with derived allele frequencies $p_a$, $p_b$, and $p_t$
-in sample $a$, sample $b$, and total, respectively, is
-$f(1 - p_a, 1 - p_b, 1 - p_t) = f(p_a, p_b, p_t)$.
-
-Computing the full covariance matrix for a large number of samples may be infeasible.
-However, we can efficiently find $y^T C z$ for arbitrary vectors $y$ and $z$.
-This works because if we set the initial weights to $y$,
-then each allele's weight is equal to the sum of the entries of $y$
-corresponding to samples carrying that allele.
-Then, if we define
-$f_{yz}(u, v, t) = (u - (\sum_a y_a) t) (v - (\sum_b z_b) t)$,
-then
-\begin{align} \label{eqn:yCz}
-    \frac{1}{2} \bar \site(f_{yz}, (y, z, \iw_t))
-        &= \frac{1}{L} \sum_j (\sum_a y_a G_{ja} - (\sum_a y_a) P_j)
-                    (\sum_b z_b G_{jb} - (\sum_b z_b) P_j) \\
-        &= y^T C z .
-\end{align}
-
-\plr{Hmph: to actually compute $Cz$ we need one weight vector for each sample, again.
-    TODO: look up if you can do Krylov just by computing $y^t C z$ for a reasonable number of vectors.
-    We could compute $Cz$ quickly if we could propagate down the tree:
-    we'd have $(Cz)_a$ equal to the sum of the weights at all nodes \emph{above} $a$. }
-
-Where does the signal from an eigenvector come from?
-Let $y$ be an eigenvector of $C$ with eigenvalue $\lambda$, normalized so that $\|y\|^2 = 1$.
-Since $Cy = \lambda y$,
-we can write $1 = \sum_a y_a^2 = (1/\lambda) y^T C y$.
-However, equation \eqref{eqn:yCz} gives $y^T C y$ as a sum over \emph{sites},
-and thus distributes the variance in $G$ associated with $y$
-across the tree sequence.
-(This is true because of the singular value decomposition of $G$.)
-
 %%%%%%%%%%%%%%%%%%
@@
 \renewcommand{\thefigure}{S\arabic{figure}}
-
-{\Large \plr{OLD STUFF}}
-
-%%%%%%%%%%%%%%%%%%
-\subsection*{Efficient computation}
-
-First we describe how to define
-and efficiently compute a quite general set of functions of a collection of sequences
-using tools from tree sequences.
-In the next section we will specialize these to a natural set of statistics.
-
-\plr{Since the ``weighting function' can take negtive values, maybe ``weight'' is not a good term.  Instead?}
-
-\paragraph{Weighted site averages}
-Suppose that we have $k$ groups of samples, $A_1$, \ldots, $A_k$.
-% with $n_1$, \ldots, $n_k$ individuals in each, respectively.
-The \emph{weight} we compute for a given site is a sum of weights computed for each allele at that site,
-and the weights are found as a function of the vector of numbers of individuals in each group inheriting that allele.
-Concretely, we specify a \emph{weighting function} $w(x_1, \ldots, x_k)$,
-and we say that the \emph{average site weight} of a set of $L$ sites is
-\begin{align} \label{eqn:average_site_weight}
-    \site_w := \frac{1}{L} \sum_{i=1}^L \sum_{s \in I_i} w(x_1(s), \ldots, x_k(s)) ,
-\end{align}
-where $I_i$ is the set of alleles seen at that site and
-$x_j(s)$ is the number of individuals in $A_j$ that carry allele $s$.
-
-Before we describe how to compute this,
-we give the analogous topological quantity,
-that is equal to the expected weighted site average
-under a continuum-sites model.
-
-\paragraph{Weighted branch length averages}
-The \emph{average branch length weight}
-is computed almost just as above:
-\begin{align} \label{eqn:average_branch_weight}
-    \branch_w :=  \frac{1}{L} \sum_{i=1}^L \sum_{e \in T_i} \ell(e) w(x_1(e), \ldots, x_k(e)) ,
-\end{align}
-except that now $T_i$ is the genealogical tree at site $i$,
-the sum is over edges $e$ in the tree, $\ell(e)$ is the length of the edge $e$,
-and $x_j(e)$ is the number of individuals in $A_j$ that fall below $e$.
-
-\emph{Note:} somewhat more generally,
-we could assign a set of weights to the individuals in each group,
-and let $x_j$ denote the total weight of the individuals in $A_j$
-inheriting from that allele or edge.
-For simplicity we do not pursue this here.
-\plr{this is a good use of the word ``weight'' (although these could also be negative)}
-
-\paragraph{Computation}
-The \emph{tree differences} encoded by a tree sequence allow us to efficiently update the quantities $x$ used above.
-As it is more straightforward, we first describe the algorithm
-to compute the branch length version.
-
-\plr{Rework these into three algorithms: one to maintain the sample weights $\iw$
-    (basically already described as `tracked nodes' in msprime)
-    and the other two that use this.}
-
-\paragraph{Total branch length weight algorithm}
-In a tree sequence, every branch in any tree is associated with its terminal \emph{node}.
-We will keep updated an array $x$ so that for any node $e$,
-the value $x[e][j]$ is the number of samples in $A_j$ below node $e$ in the current tree.
-This will be zero for any node \emph{not} appearing in the current tree.
-We also maintain $T$, the total weight of the current tree.
-\begin{itemize}
-    \item \emph{Initialize} all elements of $x$ to zero
-        except that $x[a][j] = 1$ for each $a \in A_j$ and $1 \le j \le k$.
-        Also set $S=0$ and $T=0$.
-    \item Pop the next \emph{tree difference} off the stack: an interval with endpoints $\ell < r$,
-        a set of edges $\mathcal{R}$ to remove, and a set of edges $\mathcal{A}$ to add.
-    \item \emph{Remove} each edge (parent, child) in $\mathcal{R}$ from the tree,
-        let $u$ = parent, and while $u != -1$,
-        subtract $x[child]$ from $x[u]$ and set $u$ to be parent($u$).
-    \item \emph{Add} each edge (parent, child) in $\mathcal{R}$ to the tree,
-        let $u$ = parent, and while $u != -1$,
-        add $x[child]$ to $x[u]$ and set $u$ to be parent($u$).
-    \item For each node that has changed, let $w_\text{old}$ and $w_\text{new}$
-        be the weights before and after the change in $x$,
-        and add $w_\text{new} - w_text{old}$ to $T$.
-    \item Add $T \times (r-\ell)$ to the running total, $S$.
-    \item Return to (2).
-\end{itemize}
-\plr{should modify code to maintain a vector of $w$ values}
-The average weighted branch length is obtained by dividing by the total number of sites.
-
-The algorithm for sites is similar, also maintaining the array $x$,
-and making use of the association of each mutation with a particular node in the tree.
-
-\paragraph{Total site weight algorithm}
-This works as the \emph{total branch length weight algorithm} except that steps (5--6) are replaced with:
-\plr{gah this is terrible}
-\begin{itemize}
-    \item[5'] \emph{Find allele weights} for each mutated site in the tree:
-        if $a$ is the ancestral allele,
-        set $u[a]$ to be the value of $x$ at the root; and then,
-        for each mutation at node $e$ with derived allele $b$
-        and parent mutation occurring at node $p$ with derived allele $c$,
-        add $x[e]$ to $u[b]$, and subtract $x[p]$ from $u[c]$.
-    \item[6'] For each unique allele $a$ seen,
-        add $w(v[a])$ to $T$.
-\end{itemize}
-\plr{check that algorithm deals correctly with more than one mut per node}
-
-%%%%%%%%%%%%%%%%%%
-\subsection*{General properties}
-
-Here we describe some properties that we would like our statistics to have.
-
-\paragraph{They do not depend on choice of reference allele.}
-This is achieved by applying the \emph{same} weighting function to every allele.
-Ancestral information can be incorporated by including an outgroup or ancestral sequence
-as an additional group of size one.
-Equivalently, we always work with \emph{unrooted} trees.
-
-\paragraph{They do not depend on sample size.}
-We'd like our statistics to be \emph{consistent estimators}
-of a single quantity, and should therefore not depend on sample size in expectation.
-\plr{insert this above}
-This is achieved by always computing probabilies of sampling \emph{without replacement},
-i.e., replacing each occurence of $p^k$ with $(x)_k/(n)_k$.
-This lets us compare values of a given statistic
-that are computed on different data sets with different sample sizes.
-
-\paragraph{Equivalence of ``site'' and ``branch length'' statistics}
-The definitions above are made so that for any weighting function,
-$\site_w$ is a moment estimator of $\branch_w$, i.e.,
-that $\E[\site_w] = \branch_w$ for \plr{somehow say continuum of sites here}.
 
@@
 \end{document}
+
+%%%%%%%%%%%%%%%%%%
+\subsection*{Genomic PCA}
+
+Let $G$ denote the genotype matrix, so that $G_{ja} \in \{0,1\}$
+is the genotype of the $a^\text{th}$ individual
+at the $j^\text{th}$ site, with `1` denoting the derived state
+(we assume here biallelic markers).
+Also let $P_j = (1/n) \sum_a G{ja}$ denote the vector of allele frequencies.
+Then the \emph{genetic covariance} between samples $a$ and $b$ is
+\begin{align*}
+    C_{ab}
+        &= \frac{1}{L} \sum_j (G_{ja} - P_j) (G_{jb} - P_j) . %  \\
+        % &= \frac{1}{L} \left( (G - P \bone^T)^T (G - P \bone^T) \right)_{ab} .
+\end{align*}
+This can be computed by letting
+$\iw_a = \delta_a$ and $\iw_b = \delta_b$
+(these weights mark nodes above $a$ and $b$ respectively)
+and $\iw_t = \bone / n$
+(this weight gives the proportion of samples below each node).
+Then, with
+$f(\nw_a, \nw_b, \nw_t) = (\nw_a - \nw_t) (\nw_b - \nw_t)$,
+we can compute the covariance as
+\begin{align*}
+    C_{ab} = \frac{1}{2} \bar \site(f, (\iw_a, \iw_b, \iw_t)) .
+\end{align*}
+This follows, including the factor of two,
+because the summary function applied to the ancestral allele
+at a site with derived allele frequencies $p_a$, $p_b$, and $p_t$
+in sample $a$, sample $b$, and total, respectively, is
+$f(1 - p_a, 1 - p_b, 1 - p_t) = f(p_a, p_b, p_t)$.
+
+Computing the full covariance matrix for a large number of samples may be infeasible.
+However, we can efficiently find $y^T C z$ for arbitrary vectors $y$ and $z$.
+This works because if we set the initial weights to $y$,
+then each allele's weight is equal to the sum of the entries of $y$
+corresponding to samples carrying that allele.
+Then, if we define
+$f_{yz}(u, v, t) = (u - (\sum_a y_a) t) (v - (\sum_b z_b) t)$,
+then
+\begin{align} \label{eqn:yCz}
+    \frac{1}{2} \bar \site(f_{yz}, (y, z, \iw_t))
+        &= \frac{1}{L} \sum_j (\sum_a y_a G_{ja} - (\sum_a y_a) P_j)
+                    (\sum_b z_b G_{jb} - (\sum_b z_b) P_j) \\
+        &= y^T C z .
+\end{align}
+
+\plr{Hmph: to actually compute $Cz$ we need one weight vector for each sample, again.
+    TODO: look up if you can do Krylov just by computing $y^t C z$ for a reasonable number of vectors.
+    We could compute $Cz$ quickly if we could propagate down the tree:
+    we'd have $(Cz)_a$ equal to the sum of the weights at all nodes \emph{above} $a$. }
+
+Where does the signal from an eigenvector come from?
+Let $y$ be an eigenvector of $C$ with eigenvalue $\lambda$, normalized so that $\|y\|^2 = 1$.
+Since $Cy = \lambda y$,
+we can write $1 = \sum_a y_a^2 = (1/\lambda) y^T C y$.
+However, equation \eqref{eqn:yCz} gives $y^T C y$ as a sum over \emph{sites},
+and thus distributes the variance in $G$ associated with $y$
+across the tree sequence.
+(This is true because of the singular value decomposition of $G$.)

### 2019-08-01 treestats_paper.tex
edits

@@
 \newcommand{\branch}{\mbox{Branch}} % branch stat
-\newcommand{\branchp}{\mbox{Branch}_+} % polarized
+\newcommand{\branchp}{\mbox{Branch}_+} % polarised
 \newcommand{\site}{\mbox{Site}} % site stat
-\newcommand{\sitep}{\mbox{Site}_+} % polarized
+\newcommand{\sitep}{\mbox{Site}_+} % polarised
 \newcommand{\node}{\mbox{Node}} % node stat
-\newcommand{\nodep}{\mbox{Node}_+} % polarized
+\newcommand{\nodep}{\mbox{Node}_+} % polarised
 
@@
 makes it possible to confidently estimate many more properties of genealogical relationships
-in local regions of genomes \citep{fst_landscapes},
+in local regions of genomes \citep{stankowski2019widespread,fst_landscapes},
 or between individuals rather than populations \citep{individual_stats}.
@@
 
+\plr{previous work stuff here}
+Computing correlations with branches instead of SNPs is like stuff done in \citet{zollner2005coalescent} and \citet{minichiello2006mapping}, maybe.
+Cite dynamic graph algorithm something something.
+
 % summary
@@
 
-\begin{figure}
-    \centering
-    \includegraphics{figures/divergence_diagram}
-    \caption{
-        An example of computing sequence divergence between red nodes and blue nodes
-        from seven mutations at distinct sites in a sequence of length 10 using the tree.
-        The summary function, $f(a,b) = (a/3)(1-b/2)$ gives the probability that
-        an allele found in $a$ of the red nodes and $b$ of the blue nodes is present in
-        a randomly chosen red node but not in a randomly chosen blue node.
-        \textbf{(a)} Weights are propagated up the tree by adding together the weights of any children,
-        plus the weight of the node itself, if it is a sample.
-        \textbf{(b)} In this example, weights $(a,b)$ record how many red nodes ($a$)
-        and how many blue nodes ($b$) are below each node.
-        The weights at each node is the sum of the weights of all descendant sample nodes.
-        \textbf{(c)} The summary function shown is applied to both the weight below 
-        and the weight not below each mutation,
-        so for instance the mutation above the node with weights $(2,1)$
-        contributes $f(2,1) + f(1,1) = 1/3 + 1/6 = 1/2$.
-        The final value for divergence is the sum of these values divided by 10,
-        the total number of sites.
-        \label{fig:divergence_diagram}
-    }
-\end{figure}
-
 We now begin to formalize these notions with definitions.
 
-\begin{definition}[Subtree weights]
-    Given a list of \emph{sample weights} $w$ and a tree $T$,
-    define the \emph{subtree weight} $\nw_T(u)$ for each node $u$ to be the sum of all sample weights
-    of every node descending from (and including) $u$ in the tree:
+\begin{definition}[Sample and subtree weights]
+    The \emph{samples} of a tree sequence are the focal nodes, 
+    and appear in the trees at every point along the genome.
+    A list of \emph{sample weights} $\iw$ assigns a numeric value $\iw(v)$
+    to every sample node.
+    Given these and a tree $T$,
+    the \emph{subtree weight} $\nw_T(u)$ for a node $u$ is the sum of all sample weights
+    of every sample node that is descended $u$ in the tree (including $u$, if it is a sample):
     \begin{align*}
@@
 to summarize genetic variation and tree shapes across the tree sequence.
+
+\begin{figure}
+    \centering
+    \includegraphics{figures/divergence_diagram}
+    \caption{
+        An example of computing sequence divergence between red nodes and blue nodes
+        from seven mutations at distinct sites in a sequence of length 10 using the tree.
+        The summary function, $f(a,b) = (a/3)(1-b/2)$ gives the probability that
+        an allele found in $a$ of the red nodes and $b$ of the blue nodes is present in
+        a randomly chosen red node but not in a randomly chosen blue node.
+        \textbf{(a)} Weights are assigned to each subtree by adding together the weights of any children,
+        plus the weight of the node itself, if it is a sample.
+        \textbf{(b)} In this example, subtree weights $(a,b)$ record how many red nodes ($a$)
+        and how many blue nodes ($b$) lie in that subtree.
+        \textbf{(c)} The summary function shown is applied to both the weight below 
+        and the weight not below each mutation,
+        so for instance the mutation above the node with weights $(2,1)$
+        contributes $f(2,1) + f(1,1) = 1/3 + 1/6 = 1/2$.
+        The final value for divergence is the sum of these values divided by 10,
+        the total number of sites.
+        \label{fig:divergence_diagram}
+    }
+\end{figure}
 
@@
 associated with summary function $f()$ and sample weights $\iw$
-to be the sum of $f()$ applied to the weight of the subtree below node $u$,
+to be the sum of $f()$ applied to the weight of the subtree below node $u$
+and to the remaining weight,
 averaged across the genome:
@@
     =
-    \frac{1}{L} \sum_{k=1}^{n(\treeseq)} L_k f(\nw_{T_k}(u)).
+    \frac{1}{L} \sum_{k=1}^{n(\treeseq)} L_k \left( f(\nw_{T_k}(u)) + f(\tiw - \nw_{T_k}(u)) \right).
 \end{align}
+The weight $\nw_{T_k}(u)$ is the subtree weight below $u$ in the tree $T_k$,
+and $\tiw - \nw_{T_k}(u)$ is the sum of all weights \emph{not} in the subtree below $u$ in $T_k$.
 % i.e., the length of the segment of genome that the tree extends for.
@@
 
+\paragraph{Polarisation}
+In this definition we have associated with each node the summary of both the weight below the node
+\emph{and} the weight not below the node.
+Would it not be more natural to only include the first term, $f(\nw_{T_k}(u))$?
+The reason for this definition is that it does not depend on the location of the root of the tree,
+which in practice may be unknown.
+Statistics of genotype sequence defined below in an analogous way
+will not depend on knowledge of which allele is ancestral, i.e., on \emph{polarisation} of alleles.
+For this reason, the node statistic above is the \emph{unpolarised} version,
+while a \emph{polarised} node statistic is defined without the second term:
+\begin{align}
+    \text{(polarised)} \qquad
+    \nodep(f, \iw)_u
+    =
+    \frac{1}{L} \sum_{k=1}^{n(\treeseq)} L_k f(\nw_{T_k}(u)) .
+\end{align}
+
 We will use weights that count number of samples below each node
 sufficiently often that we give them a special name --
-for a subset set $A$ of samples,
-the \emph{indicator weights of $A$} are the sample weights $\bone_A$ with
-$\bone_A(u) = 1$ if $u \in A$ and $\bone_A(u) = 0$ otherwise.
+for a subset set $S$ of samples,
+the \emph{indicator weights of $S$} are the sample weights $\bone_S$ with
+$\bone_S(u) = 1$ if $u \in S$ and $\bone_S(u) = 0$ otherwise.
 
@@
     Therefore, if $f(x) = x / n$,
-    then $\node(f, \iw)_u$ is the proportion of the genomes of $S$
+    then $\nodep(f, \iw)_u$ is the proportion of the genomes of $S$
     that are inherited from (ancestor) $u$.
@@
 but provide a useful bridge to the next type of statistic,
-which are defined directly in terms of the genotypes
-(and so do not depend on the tree sequence).
+which are defined directly in terms of the genotypes.
 
@@
 
-The next mode of summarizing uses weights
-to compute a summary for each site in the genome.
-These are often reported by averaging the values over windows.
-The general definition uses the \emph{allele weights} at each site in the genome,
-defined simply to be the total weight of all nodes who have inherited that allele.
+Now we arrive at our primary goal, which is to compute statistics from \emph{genomes}
+using this framework.
+To do this, we assume that the genetic variation data is embedded in a tree sequence,
+but the trees are used only for efficiency --
+the results will not depend on the trees in any way
+(with the only exception being if the ancestral allele is assumed to be known, as discussed below).
+
+The summaries we defined above for each node in each tree
+extend nicely to genetic variants 
+-- just as a node weight contains information about which samples are below the node and which are not,
+so we can summarise patterns of genetic variation by summing up weights of all samples
+that carry a given allele.
+So, we define \emph{allele weights} to be
+the total weight of all sample nodes who have inherited that allele.
 
@@
     The \emph{allele weight} for allele $a$ at site $j$ is the sum of the weights
-    of all nodes inheriting this allele:
+    of all sample nodes inheriting this allele:
     \begin{align*}
@@
     \end{align*}
-    where $g_j(v)$ is the allele carried by node $v$ at site $j$.
+    where the sum is over all sample nodes $v$ for which
+    $g_j(v)$, the allele carried by node $v$ at site $j$, is equal to $a$.
 \end{definition}
@@
 If there has been only one mutation at the site,
-then $\aw_j(a)$ is equal to the subtree weight of the node that the mutation producing $a$
-appeared above,
-while more generally $\aw$ is not necessarily equal to a subtree weight
+then $\aw_j(a)$ is equal to the weight of the subtree below the mutation that produced $a$.
+More generally, $\aw$ is not necessarily equal to a subtree weight
 but is easily computable from them.
 
-We then define the \textbf{site statistic} of site $j$ computed using a summary function $f()$
+Using allele weights,
+we define the \textbf{site statistic} of site $j$ computed using a summary function $f()$
 and a set of sample weights $\iw$
@@
 \end{align}
-Recall that we have always defined the summary function $f()$ to be zero at monomorphic sites,
-so that the sum will be only over polymorphic sites
-(although the normalization is always by total number of sites, to allow comparison between regions).
-
-\paragraph{What about polarization?}
+Since we have required $f(0) = f(\tiw) = 0$,
+the sum will only be affected by polymorphic sites,
+although the normalization is always by total number of sites, to allow comparison between regions.
+
+\paragraph{What about polarisation?}
 In the definition above we simply sum over all alleles at each site.
@@
 One way to make use of this information is to sum over only derived alleles.
-We say a site statistic is \textbf{polarized} if we do this,
-and so
-\begin{align} \label{eqn:site_unpolarized}
+We say a site statistic is \textbf{polarised} if we do this,
+and so define
+\begin{align} \label{eqn:site_polarised}
+    \text{(polarised)} \qquad
     \sitep(f, \iw)_j
@@
 
+Now we have defined a general, abstract, framework.
+Is it good for anything?
+Next we give some examples.
+
 \begin{example}[Nucleotide diversity] \label{ex:site_diversity}
@@
 
+\begin{example}[Segregating sites] \label{ex:segregating_sites}
+    Again, let $\iw = \bone_{S}$ for a group of $n$ sequences,
+    and now define
+    \begin{align*}
+        f(x) = \begin{cases}
+            1 - \frac{x}{n} \qquad &\text{if } x > 0 \\
+            0 \qquad &\text{otherwise} .
+        \end{cases}
+    \end{align*}
+    In computing $\site(f, \iw)_{[a,b)}$, the function $f()$ is given the number of samples in $S$
+    that inherit each allele.
+    A site with $k$ distinct alleles that segregate in $S$ with proportions $p_1 + \cdots + p_k = 1$
+    contributes $f(p_1) + \cdots + f(p_k) = (1 - p_1) + \cdots + (1 - p_k) = k - 1$ to the statistic.
+    Therefore, $\site(f, \iw)_{[a,b)}$, is the minimum number of derived mutations per unit length,
+    which is the density of segregating sites if all sites are biallelic.
+\end{example}
+
 \begin{example}[Phenotypic correlations] \label{ex:site_correlations}
@@
     and the genotype at each site in the tree sequence.
-    For convenience, suppose $z$ is normalized so that $\sum_u z(u) = 0$ and $\sum_u z(u)^2 = 1$.
+    For convenience, suppose $z$ is normalized 
+    % so that $\sum_u z(u) = 0$ and $\sum_u z(u)^2/(n-1) = 1$.
+    to have mean zero and variance 1.
     Then, if $g_j$ is a vector of binary genotypes (so $g_j(u) = 1$ if $u$ carries the derived allele),
@@
     i.e., the sum of the phenotypes of all samples carrying the derived allele,
-    divided $\sqrt{n - 1}$, where $n$ is the number of samples.
+    divided by $(n - 1)$, where $n$ is the number of samples.
     Since the phenotypes sum to zero, this is also equal to 
     $- \sum_{u \st g_j(u) = 0} z(u)$.
-    Therefore, if we write $p_j = \sum_u g_j(u)$ for the derived allele frequency,
-    we can write the squared correlation as the sum across alleles as
+    If $p_j = \sum_u g_j(u)$ is the derived allele frequency at site $j$,
+    then the variance of $g_j$ is $n p_j (1-p_j) / (n-1)$,
+    and so the squared correlation can be calculated as a sum across the two alleles:
     \begin{align*}
@@
 
-\textbf{Note:} computing correlations with branches instead of SNPs is like stuff done in \citet{zollner2005coalescent} and \citet{minichiello2006mapping}, maybe.
+%% Check that math:
+% n <- 10; z <- rnorm(10); z <- (z - mean(z))/sd(z); g <- rbinom(n, size=1, prob=1/2); p <- mean(g)
+% c(cov(z,g), sum(z*g)/(n-1))
+% c(cor(g,z), sum(g*z)/sqrt(n*(n-1)*p*(1-p)))
 
@@
     (so that $\iw_{ju}$ tells us the number of samples in $S_j$ descended from $u$),
-    and write $n_i$ for the number of samples in $S_i$,
+    and write $n_i$ for the number of samples in $S_i$, and
     \begin{align*}
@@
         \frac{x_2}{n_2}
-        \frac{x_3}{n_3}
+        \frac{x_3}{n_3} ,
     \end{align*}
     then $\site(\iw, f)_{[a,b)}$ is equal to Patterson's $f_4(S_1, S_2; S_3, S_4)$ statistic
-    across the region of the genome between $a$ and $b$.
+    for the region of the genome between $a$ and $b$.
 \end{example}
@@
 when all sites are biallelic.
-If there is a consensus in the field
-for how to use sites with more than two alleles,
-it is to reduce to biallelic data, by either discarding other sites
+There is not a consensus in the field
+on how to use sites with more than two alleles, but the most common practice is
+to reduce to biallelic data, by either discarding other sites
 or by marking all non-ancestral alleles as (the same) ``derived'' allele.
@@
 so that for instance a site with ancestral allele $A$ and derived alleles $C$ and $T$
-would have site statistic $f(\aw(C)) + f(\aw(T))$.
-This is a natural defintion in that it agrees with what you'd get
-from looking at haplotype differences.
+would have site statistic $f(\aw(A)) + f(\aw(C)) + f(\aw(T))$.
+This is a natural defintion in that it agrees with what is obtained by
+looking at pairwise haplotype differences.
 For instance, the definition of nucleotide divergence in Example \ref{ex:site_divergence}
@@
 and (as we discuss below) has a natural correspondence to statistics related to branch lengths.
-
-\plr{Figure: example of propagating weights and labeling nodes with weights,
-    in the case that weights are number of samples you are ancestor of.
-    Also, this is a node statistic.}
-
-\plr{Figure: example of branch and site statistics}
 
@@
 defined as its span (right minus left endpoint) multiplied by its length (parent time minus child time)
-is equal to the expected number of mutations that will land on it, up to the mutation rate.
+and scaled by the mutation rate,
+is equal to the expected number of mutations that will land on it.
 If the mutation rate is constant,
@@
 a statistic of tree shape by summing these expected contributions across its branches.
-Concretely, the \textbf{branch statistic} of a tree $T$
+We define the \textbf{branch statistic} of a single tree $T$ with span $L$
 obtained from summary function $f()$
 and sample weights $\iw$ with total weight $\tiw = \sum_u \iw(u)$
-is defined to be
+to be the sum of the area of each branch in each tree
+multiplied by the summary function applied to its subtree weight:
 \begin{align}
@@
     &=
-    \sum_{k=1}^{n(\treeseq)} L_k \left\{ f(\nw_{T_k}(u)) + f(\tiw - \nw_{T_k}(u)) \right\}  ,
+    \frac{1}{L} \sum_{u \in T} t_{u} \left( f(\nw_{T}(u)) + f(\tiw - \nw_{T}(u)) \right)  ,
 \end{align}
-where $t_u$ is the length of the branch above node $u$ in the tree $T$
-and $\nw(u)$ is the total weight of the subtree below $u$ (as defined above).
-The term $\tiw - \nw_{T_k}(u)$ gives the total weight \emph{not} in the subtree of $T_k$ below $u$,
-because the sum of all weights in the tree is always constant,
-and so $\sum_u \nw_T(u) = \tiw$.
+where $t_{u}$ is the length of the branch above node $u$ in tree $T$
+and $\nw_{T}(u)$ is the total weight of the subtree of $T$ below $u$ (as defined above).
+The term $\tiw - \nw_{T}(u)$ gives the total weight \emph{not} in the subtree of $T$ below $u$,
+because the sum of all weights in the tree is always $\tiw$.
 The value $f(\nw(u))$ is the summary value that would be added to a site statistic
 if a single mutation occurred on the branch above $u$,
-and $f(\tiw - \nw_{T_k}(u))$ is the value that would be added due to its complementary allele,
+and $f(\tiw - \nw_{T}(u))$ is the value that would be added due to its complementary allele,
 so $\branch(f, \iw)_T$ gives the expected contribution of this branch to $\site(f, \iw)$,
@@
 
-In practice it is probably most useful to average branch statistics
+This defines a statistic for a single window,
+but in practice it is probably most useful to average branch statistics
 over a region of the genome,
 which we do by averaging the tree statistics over the region
-with weights proportional to the trees' spans.
+with probabilities proportional to the trees' spans.
 We write this with similar notation:
 \begin{align}
-    \branch(f, \iw)_{[a,b)}
-    &=
-    \frac{1}{b-a} \sum_T \ell_T(a,b) \branch(f, \iw)_T ,
+    \branch(f, \iw)_{[c,d)}
+    &=
+    \frac{1}{d-c} \sum_{k=1}^{n(\treeseq)} \ell_k(c,d) \branch(f, \iw)_{T_k} ,
 \end{align}
-where $\ell_T(a,b)$ is the length of the region in $[a,b)$ that the tree $T$ extends for
-(i.e., if $T$ covers the half-open interval $[c,d)$,
-then $\ell_T(a,b) = \max(0, \min(b,d) - \max(a,c))$).
-
-The ``unpolarized'' version of this statistic, corresponding to equation \ref{eqn:site_unpolarized},
-omits the second term:
-\begin{align} \label{eqn:branch_unpolarized}
-    \branch_+(f, \iw)_T
-    &=
-    \sum_{k=1}^{n(T)} L_k f(\nw_{T_k}(u)) .
+where $\ell_k(c,d)$ is the length of the region in $[c,d)$ that the tree $T_k$ extends for
+(i.e., $T$ covers the half-open interval $[a_{k-1},a_k)$,
+so $\ell_k(c,d) = \max(0, \min(d,a_k) - \max(c,a_{k-1}))$).
+
+The \emph{polarised} version of this statistic
+is defined analogously to the Node and Site versions:
+\begin{align} \label{eqn:branch_polarised}
+    \text{(polarised)} \qquad
+    \branchp(f, \iw)_T
+    &=
+    \sum_{u \in T} f(\nw_{T}(u)) ,
 \end{align}
-However, the correspondence between the unpolarized statistics is less tight
-than for the polarized ones.
+and $\branchp(f, \iw)_{[c,d)}$ is defined using $\branchp(f, \iw)_T$ as before.
+However, the correspondence between the polarised statistics is less tight
+than for the unpolarised ones.
 
@@
     on the path up to their most recent common ancestor (MRCA).
-    Therefore, $\bar \branch(f, \iw)$ for these choices
+    Therefore, $\branch(f, \iw)$ for these choices
     gives the mean total distance in the tree between two samples from $S$,
     averaged across the sequence.
-    (This is twice the TMRCA if the samples are all from the same time.)
+    This is twice the mean TMRCA if the samples are all from the same time.
 \end{example}
@@
 $\E[\site(f, \iw) \site(g, \iw')]$
-is not equal to $\branch(f, \iw) \branch(g, \iw')$,
+is not equal to the product of the two branch statistics, $\branch(f, \iw) \branch(g, \iw')$,
 because they are not independent.
@@
 
-This correspondance is useful for several things:
-(1) removing noise of mutations
-(2) understanding what statistics tell us about genealogies
-
-\paragraph{Example: a selective sweep}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection*{A selective sweep}
+
 In this view,
@@
 %%%%%%%%%%%%%%%%%%
-\subsection*{Separating mutational from genealogical noise}
-
-\plr{figure showing noise of some site statistic decreasing and approaching the branch statistic
-    as the mutation rate increases}
+\subsection*{Empirical duality between branch and site statistics?}
+
+We do not know the true genealogies underlying real data,
+but recent methods are available to estimate them.
+To evaluate the correspondence between site and branch statistics
+in real data,
+we calculated mean sequence diversity along chromosome 20 from the 1000 Genomes dataset
+in windows of 500Kb \plr{check this}
+separately in each of its 26 populations,
+calculated using the site statistic described in Example~\ref{ex:site_diversity},
+and compared this to the corresponding branch statistic (of Example~\ref{ex:branch_diversity})
+as calculated from three inferred tree sequences:
+by tsinfer \citep{tsinfer}, which does not attenpt to estimate well-calibrated node times,
+(2) by GEVA \citep{geva}, which aims to calibrate the node times reported by tsinfer,
+and (3) by Relate \citep{relate}, an independently developed method.
+These are shown in Figure~\ref{fig:chr20}.
+
+\begin{figure}
+    \includegraphics[width=0.5\textwidth]{{figures/1kg/relate_site_diversity_chr20}.png}
+    \includegraphics[width=0.5\textwidth]{{figures/1kg/tsinfer_1kg_chr20.branch_diversity.2000000}.pdf}
+    \includegraphics[width=0.5\textwidth]{{figures/1kg/tgp_geva_chr20.branch_diversity.2000000}.pdf}
+    \includegraphics[width=0.5\textwidth]{{figures/1kg/relate_branch_diversity_chr20}.png}
+    \caption{
+        \textbf{(A)} Mean sequence diversity 
+        in 500Kb \plr{?} windows along human chromosome 20 of the 1000 Genomes Project,
+        and the dual branch statistic as calculated from tree sequences inferred by
+        \textbf{(B)} tsinfer,
+        \textbf{(C)} GEVA, and 
+        \textbf{(D)} Relate.
+        \label{fig:chr20}
+    }
+\end{figure}
+
+These are different because of XXX but similar YYY.
+As discussed above, we expect site and branch statistics to agree up to noise
+if mutation rates are constant along the genome.
+However, this caveat may well be substantial -- mutation rates are likely \emph{not} constant
+\citep{mut_rate_variation},
+although the extent of this is unknown.
 

### 2019-08-01 figures/plot_1kg_diversity.py

@@ -1,0 +1,56 @@
+import tskit, tszip, json
+import numpy as np
+import matplotlib
+matplotlib.use('Agg')
+import matplotlib.pyplot as plt
+
+window_width = 5e5
+
+popstyles = {"EAS" : (0, ()),
+             "EUR" : (0, (1, 2)),
+             "AFR" : (0, (3, 1)),
+             "AMR" : (0, (3, 2, 1, 2)),
+             "SAS" : (0, (3, 1, 3, 1, 1, 1))}
+colors = plt.get_cmap("tab20").colors
+pop_names = ['CHB', 'JPT', 'CHS', 'CDX', 'KHV', 'CEU', 'TSI', 'FIN', 'GBR', 'IBS', 'YRI', 'LWK', 'GWD', 'MSL', 'ESN', 'ASW', 'ACB', 'MXL', 'PUR', 'CLM', 'PEL', 'GIH', 'PJL', 'BEB', 'STU', 'ITU']
+popcolors = { k : colors[j % len(colors)] for j, k in enumerate(pop_names) }
+
+for chrom in range(1, 2):
+    ts = tskit.load("1kg/1kg_chr{}.trees".format(chrom))
+
+    pop_metadata = [json.loads(pop.metadata) for pop in ts.populations()]
+    pn = [x['name'] for x in pop_metadata]
+    for a,b in zip(pn, pop_names):
+        assert(a == b)
+
+    superpops = [x['super_population'] for x in pop_metadata]
+    pop = [ts.node(ind.nodes[0]).population for ind in ts.individuals()]
+
+    pop_nodes = [[] for _ in range(ts.num_populations)]
+    for ind in ts.individuals():
+        pop_nodes[pop[ind.id]].extend(ind.nodes)
+
+    num_windows = int(ts.sequence_length / window_width)
+    windows = np.linspace(0, ts.sequence_length, num_windows)
+    x = windows[:-1] + np.diff(windows)/2
+
+    bdiv = ts.diversity(pop_nodes, windows=windows, mode="branch")
+
+    fig = plt.figure(figsize=(12, 6))
+    ax = fig.add_subplot(111)
+    for j in range(bdiv.shape[1]):
+        dl = ax.plot(
+                x, bdiv[:, j], label=pop_names[j],
+                linestyle=popstyles[superpops[j]],
+                color=popcolors[pop_names[j]])
+
+    leg = ax.legend(
+             fontsize = "small",
+             loc = "upper left",
[…]

### 2019-08-01 treestats_paper.tex
new fig

@@
 
+An example is shown in Figure~\ref{fig:branch_site_diagram},
+showing how the $f_4$ site statistic assigns a weight to each mutation
+depending on the frequencies in each of the four sample sets below it,
+and how the corresponding branch statistic assigns the same weight to those branches.
+
+\begin{figure}
+    \centering
+    \includegraphics{figures/branch_site_diagram}
+    \caption{
+    Computing the $f_4(A,B;C,D)$ statistic on a tree.
+    \textbf{(a)} Each mutation is given weight equal to $(p_A - p_B)(p_C - p_D)$,
+    where $p_X$ is the proportion of the samples inheriting that mutation in set $X$
+    (shown in labels at the bottom).
+    Note that each of the mutations shown occurs at a distinct site, so there is no backmutation.
+    \textbf{(b)} Each branch is assigned weight equal to the weight that would be assigned
+    to any mutation falling in it, and then the branch $f_4$ statistic
+    is the sum over these weights, multiplied by the length of the branch
+    (and the span of the tree, not depicted).
+        \label{fig:branch_site_diagram}
+    }
+\end{figure}
+
 This defines a statistic for a single window,

### 2019-08-01 treestats_paper.tex
comparison of tsinferish things

@@
 and (3) by Relate \citep{relate}, an independently developed method.
-These are shown in Figure~\ref{fig:chr20}.
+
+The comparisons are shown in Figure~\ref{fig:chr20}.
+Of these methods, branch diversity from tsinfer shows the worst agreement with site diversity,
+which is unsurprising as tsinfer makes no attempt to provide well-calibrated node times,
+and only aims to provide consistent tree topologies.
+The two methods that infer node times differ substantially,
+with Relate showing a significantly tighter agreement than GEVA.
+Does this mean that Relate is doing a better job at inferring the tree sequence than GEVA?
+This is highly suggestive, but not entirely clear.
 
@@
 
-These are different because of XXX but similar YYY.
-As discussed above, we expect site and branch statistics to agree up to noise
-if mutation rates are constant along the genome.
-However, this caveat may well be substantial -- mutation rates are likely \emph{not} constant
-\citep{mut_rate_variation},
-although the extent of this is unknown.
+In practice, we do not expect branch and site diversity to agree exactly,
+because of local differences in mutation rate and intensities of selected mutations.
+For instance, regions in which a high proportion of potential mutations are deleterious
+are expected to have a diversity for two reasons:
+first, the deleterious mutations themselves are less likely to be found,
+and second, the indirect effect of deleterious mutations on nearby sites reduces typical tree height
+and thus diversity at even neutral, linked sites \citep{background_selection}.
+The first effect would cause site and branch statistics to differ,
+because it effectively reduces the mutation rate in the region
+and violates the assumption of independence of mutations given the trees.
+However, the second effect does not affect the correspondence between site and branch statistics,
+since it is mediated by tree shape.
+However, it is generally unknown how much mutation rate varies along the genome
+\citep[but see][]{mut_rate_varies}
+or how dense are targets of selection \citep{how_important_is_selection}.
+For this reason, it is very interesting to see how close a correspondence between site and branch
+statistics it is \emph{possible} to obtain --
+it is tempting to say that the best tree sequence would obtain
+as tight a match between site and branch statistics as possible,
+with remaining variation explained by selection and/or mutation rate variation.
 

### 2019-08-02 figures/plot_1kg_diversity.py
plot diversity as executable

@@ -1,10 +1,23 @@
+#!/usr/bin/env python3
 import tskit, tszip, json
 import numpy as np
 import matplotlib
 matplotlib.use('Agg')
 import matplotlib.pyplot as plt
+import os, sys
 
-window_width = 5e5
+usage = """
+Usage:
+    {} treefile window_width
+""".format(sys.argv[0])
+
+if len(sys.argv) != 3:
+    raise ValueError(usage)
+
+treefile = sys.argv[1]
+window_width = float(sys.argv[2])
+
+outbase = ".".join(treefile.split(".")[:-1]) + f".{int(window_width)}"
 
 popstyles = {"EAS" : (0, ()),
              "EUR" : (0, (1, 2)),
@@ -15,42 +28,41 @@ colors = plt.get_cmap("tab20").colors
 pop_names = ['CHB', 'JPT', 'CHS', 'CDX', 'KHV', 'CEU', 'TSI', 'FIN', 'GBR', 'IBS', 'YRI', 'LWK', 'GWD', 'MSL', 'ESN', 'ASW', 'ACB', 'MXL', 'PUR', 'CLM', 'PEL', 'GIH', 'PJL', 'BEB', 'STU', 'ITU']
 popcolors = { k : colors[j % len(colors)] for j, k in enumerate(pop_names) }
 
-for chrom in range(1, 2):
-    ts = tskit.load("1kg/1kg_chr{}.trees".format(chrom))
+ts = tskit.load(treefile)
 
-    pop_metadata = [json.loads(pop.metadata) for pop in ts.populations()]
-    pn = [x['name'] for x in pop_metadata]
-    for a,b in zip(pn, pop_names):
-        assert(a == b)
+pop_metadata = [json.loads(pop.metadata) for pop in ts.populations()]
+pn = [x['name'] for x in pop_metadata]
+for a,b in zip(pn, pop_names):
+    assert(a == b)
 
-    superpops = [x['super_population'] for x in pop_metadata]
-    pop = [ts.node(ind.nodes[0]).population for ind in ts.individuals()]
+superpops = [x['super_population'] for x in pop_metadata]
+pop = [ts.node(ind.nodes[0]).population for ind in ts.individuals()]
 
-    pop_nodes = [[] for _ in range(ts.num_populations)]
-    for ind in ts.individuals():
-        pop_nodes[pop[ind.id]].extend(ind.nodes)
[…]

### 2019-08-05 figures/plot_1kg_diversity.py
better plot script

@@ -8,16 +8,22 @@ import os, sys
 
 usage = """
 Usage:
-    {} treefile window_width
+    {} treefile maskfile mode window_width
+Here `maskfile` should be a fasta of 0s and 1s.
 """.format(sys.argv[0])
 
-if len(sys.argv) != 3:
+if len(sys.argv) != 5:
     raise ValueError(usage)
 
 treefile = sys.argv[1]
-window_width = float(sys.argv[2])
+maskfile = sys.argv[2]
+mode = sys.argv[3]
+window_width = float(sys.argv[4])
 
-outbase = ".".join(treefile.split(".")[:-1]) + f".{int(window_width)}"
+# min amount of data allowed to still deal with the window
+min_window_content = window_width / 3
+
+outbase = ".".join(treefile.split(".")[:-1]) + f".{mode}.{int(window_width)}"
 
 popstyles = {"EAS" : (0, ()),
              "EUR" : (0, (1, 2)),
@@ -42,11 +48,65 @@ pop_nodes = [[] for _ in range(ts.num_populations)]
 for ind in ts.individuals():
     pop_nodes[pop[ind.id]].extend(ind.nodes)
 
-num_windows = int(ts.sequence_length / window_width)
-windows = np.linspace(0, ts.sequence_length, num_windows)
-x = windows[:-1] + np.diff(windows)/2
 
-bdiv = ts.diversity(pop_nodes, windows=windows, mode="branch")
+#############
+# get the mask
+
+mask = np.genfromtxt(maskfile, delimiter=1, skip_footer=1)
+masktail = np.genfromtxt(maskfile, delimiter=1, skip_header=len(mask))
+num_added_zeros = max(0, int(ts.sequence_length) - np.prod(mask.shape) - len(masktail))
+if num_added_zeros > 0:
+    print(f"Adding {num_added_zeros} masked bases to the end of the mask.")
+
+mask = np.concatenate(
+        [mask.reshape((mask.shape[0] * mask.shape[1],)),
+         masktail,
+         np.repeat(0, num_added_zeros)])
+
+mask = mask[:int(ts.sequence_length)]
+
+starts = np.where(np.diff(mask) > 0)[0]
+ends = np.where(np.diff(mask) < 0)[0]
+lengths = ends - starts
+assert(np.all(lengths > 0))
+
+#############
+# compute the statistics in lots of little windows
+
+# define the big windows
+num_windows = int(ts.sequence_length / window_width)
[…]

### 2019-08-06 figures/plot_1kg_diversity.py
plot relate ts

@@ -32,19 +32,36 @@ popstyles = {"EAS" : (0, ()),
              "SAS" : (0, (3, 1, 3, 1, 1, 1))}
 colors = plt.get_cmap("tab20").colors
 pop_names = ['CHB', 'JPT', 'CHS', 'CDX', 'KHV', 'CEU', 'TSI', 'FIN', 'GBR', 'IBS', 'YRI', 'LWK', 'GWD', 'MSL', 'ESN', 'ASW', 'ACB', 'MXL', 'PUR', 'CLM', 'PEL', 'GIH', 'PJL', 'BEB', 'STU', 'ITU']
+superpops = ['EAS', 'EAS', 'EAS', 'EAS', 'EAS', 'EUR', 'EUR', 'EUR', 'EUR', 'EUR', 'AFR', 'AFR', 'AFR', 'AFR', 'AFR', 'AFR', 'AFR', 'AMR', 'AMR', 'AMR', 'AMR', 'SAS', 'SAS', 'SAS', 'SAS', 'SAS']
 popcolors = { k : colors[j % len(colors)] for j, k in enumerate(pop_names) }
+num_pops = len(pop_names)
 
 ts = tskit.load(treefile)
 
-pop_metadata = [json.loads(pop.metadata) for pop in ts.populations()]
-pn = [x['name'] for x in pop_metadata]
-for a,b in zip(pn, pop_names):
-    assert(a == b)
+if ts.num_populations > 0:
+    pop_metadata = [json.loads(pop.metadata) for pop in ts.populations()]
+    for a, b, c in zip(pop_metadata, pop_names, superpops):
+        assert(a['name'] == b)
+        assert(a['super_population'] == c)
 
-superpops = [x['super_population'] for x in pop_metadata]
-pop = [ts.node(ind.nodes[0]).population for ind in ts.individuals()]
+    pop = [ts.node(ind.nodes[0]).population for ind in ts.individuals()]
 
-pop_nodes = [[] for _ in range(ts.num_populations)]
+else:
+    # this should be a Relate tree sequence
+    assert(treefile.find("relate") >= 0)
+    # in which case the metadata are in this external file, in order,
+    # with one line per diploid
+    pop = []
+    with open("1kg/1000GP_Phase3_sub.sample", "r") as metafile:
+        header = metafile.readline()
+        assert(header == "ID POP GROUP SEX\n")
+        for line in metafile:
+            md = line.strip().split()
[…]

### 2019-08-06 figures/plot_1kg_diversity.py
.

@@ -1,5 +1,5 @@
 #!/usr/bin/env python3
-import tskit, tszip, json
+import tskit, json
 import numpy as np
 import matplotlib
 matplotlib.use('Agg')
@@ -84,6 +84,8 @@ mask = mask[:int(ts.sequence_length)]
 
 starts = np.where(np.diff(mask) > 0)[0]
 ends = np.where(np.diff(mask) < 0)[0]
+if len(ends) == len(starts) - 1:
+    ends = np.concatenate([ends, np.array([len(mask) - 1])])
 lengths = ends - starts
 assert(np.all(lengths > 0))
 

### 2019-08-06 treestats_paper.tex
added new plots

@@
 \begin{figure}
-    \includegraphics[width=0.5\textwidth]{{figures/1kg/relate_site_diversity_chr20}.png}
-    \includegraphics[width=0.5\textwidth]{{figures/1kg/tsinfer_1kg_chr20.branch_diversity.2000000}.pdf}
-    \includegraphics[width=0.5\textwidth]{{figures/1kg/tgp_geva_chr20.branch_diversity.2000000}.pdf}
-    \includegraphics[width=0.5\textwidth]{{figures/1kg/relate_branch_diversity_chr20}.png}
+    \includegraphics[width=0.5\textwidth]{{figures/1kg/relate_chr20.site.1000000.diversity}.pdf}
+    \includegraphics[width=0.5\textwidth]{{figures/1kg/1kg_chr20.branch.1000000.diversity}.pdf}
+    \includegraphics[width=0.5\textwidth]{{figures/1kg/tgp_geva_chr20.branch.1000000.diversity}.pdf}
+    \includegraphics[width=0.5\textwidth]{{figures/1kg/relate_chr20.branch.1000000.diversity}.pdf}
     \caption{
-        \textbf{(A)} Mean sequence diversity 
-        in 500Kb \plr{?} windows along human chromosome 20 of the 1000 Genomes Project,
+        \textbf{(top left)} Mean sequence diversity 
+        in 1Mb windows along human chromosome 20 of the 1000 Genomes Project,
         and the dual branch statistic as calculated from tree sequences inferred by
-        \textbf{(B)} tsinfer,
-        \textbf{(C)} GEVA, and 
-        \textbf{(D)} Relate.
+        \textbf{(top right)} tsinfer,
+        \textbf{(bottom left)} GEVA, and 
+        \textbf{(bottom right)} Relate.
         \label{fig:chr20}

### 2019-08-08 figures/plot_1kg_f4.py
f4 script

@@ -1,5 +1,5 @@
 #!/usr/bin/env python3
-import tskit, json
+import tskit, tszip, json
 import numpy as np
 import matplotlib
 matplotlib.use('Agg')
@@ -11,6 +11,7 @@ Usage:
     {} treefile maskfile mode window_width
 Here `maskfile` should be a fasta of 0s and 1s.
 """.format(sys.argv[0])
+
 
 if len(sys.argv) != 5:
     raise ValueError(usage)
@@ -24,19 +25,21 @@ window_width = float(sys.argv[4])
 min_window_content = window_width / 3
 
 outbase = ".".join(treefile.split(".")[:-1]) + f".{mode}.{int(window_width)}"
-divfile = f"{outbase}.diversity.tsv"
-plotfile = f"{outbase}.diversity.pdf"
+statfile = f"{outbase}.f4.tsv"
+plotfile = f"{outbase}.f4.pdf"
 
-popstyles = {"EAS" : (0, ()),
-             "EUR" : (0, (1, 2)),
-             "AFR" : (0, (3, 1)),
-             "AMR" : (0, (3, 2, 1, 2)),
-             "SAS" : (0, (3, 1, 3, 1, 1, 1))}
-colors = plt.get_cmap("tab20").colors
 pop_names = ['CHB', 'JPT', 'CHS', 'CDX', 'KHV', 'CEU', 'TSI', 'FIN', 'GBR', 'IBS', 'YRI', 'LWK', 'GWD', 'MSL', 'ESN', 'ASW', 'ACB', 'MXL', 'PUR', 'CLM', 'PEL', 'GIH', 'PJL', 'BEB', 'STU', 'ITU']
 superpops = ['EAS', 'EAS', 'EAS', 'EAS', 'EAS', 'EUR', 'EUR', 'EUR', 'EUR', 'EUR', 'AFR', 'AFR', 'AFR', 'AFR', 'AFR', 'AFR', 'AFR', 'AMR', 'AMR', 'AMR', 'AMR', 'SAS', 'SAS', 'SAS', 'SAS', 'SAS']
-popcolors = { k : colors[j % len(colors)] for j, k in enumerate(pop_names) }
 num_pops = len(pop_names)
+
+# which populations?
+f4_pops = [('PUR', 'TSI', 'GWD', 'CHB'),
+           ('ASW', 'CEU', 'MSL', 'CHB')]
+
+stat_names = [f"{a},{b};{c},{d}" for a, b, c, d in f4_pops]
[…]

### 2019-08-09 figures/bootstrap_1kg_D_f4.py
sim scripts

@@ -1,0 +1,172 @@
+#!/usr/bin/env python3
+import tskit, json
+import numpy as np
+import matplotlib
+matplotlib.use('Agg')
+import matplotlib.pyplot as plt
+import os, sys
+
+usage = """
+Usage:
+    {} treefile maskfile mode window_width
+Here `maskfile` is unused.
+""".format(sys.argv[0])
+
+
+if len(sys.argv) != 5:
+    raise ValueError(usage)
+
+treefile = sys.argv[1]
+maskfile = sys.argv[2]
+mode = sys.argv[3]
+window_width = float(sys.argv[4])
+
+outbase = ".".join(treefile.split(".")[:-1]) + f".{mode}.{int(window_width)}"
+statfile = f"{outbase}.divergences.tsv" # created by calc_1kg_divergences.py
+D_plotfile = f"{outbase}.D.jackknife.pdf"
+f4_plotfile = f"{outbase}.f4.jackknife.pdf"
+
+pop_names = ['CHB', 'JPT', 'CHS', 'CDX', 'KHV', 'CEU', 'TSI', 'FIN', 'GBR', 'IBS', 'YRI', 'LWK', 'GWD', 'MSL', 'ESN', 'ASW', 'ACB', 'MXL', 'PUR', 'CLM', 'PEL', 'GIH', 'PJL', 'BEB', 'STU', 'ITU']
+superpops = ['EAS', 'EAS', 'EAS', 'EAS', 'EAS', 'EUR', 'EUR', 'EUR', 'EUR', 'EUR', 'AFR', 'AFR', 'AFR', 'AFR', 'AFR', 'AFR', 'AFR', 'AMR', 'AMR', 'AMR', 'AMR', 'SAS', 'SAS', 'SAS', 'SAS', 'SAS']
+num_pops = len(pop_names)
+
+# which populations?
+f4_pops = [('PUR', 'TSI', 'GWD', 'CHB'),
+           ('ASW', 'CEU', 'MSL', 'CHB'),
+           ('GBR', 'IBS', 'ITU', 'JPT')]
+
+stat_names = [f"{a},{b};{c},{d}" for a, b, c, d in f4_pops]
+colors = plt.get_cmap("tab20").colors
+stat_colors = { k : colors[j] for j, k in enumerate(stat_names) }
+
+# read in already-computed divergences
+all_pairs = [(i, j) for i in range(num_pops) for j in range(i, num_pops)]
[…]

### 2019-08-09 figures/calc_1kg_divergences.py

@@ -1,0 +1,122 @@
+#!/usr/bin/env python3
+import tskit, json
+import numpy as np
+import os, sys
+
+usage = """
+Usage:
+    {} treefile maskfile mode window_width
+Here `maskfile` should be a fasta of 0s and 1s.
+""".format(sys.argv[0])
+
+
+if len(sys.argv) != 5:
+    raise ValueError(usage)
+
+treefile = sys.argv[1]
+maskfile = sys.argv[2]
+mode = sys.argv[3]
+window_width = float(sys.argv[4])
+
+# min amount of data allowed to still deal with the window
+min_window_content = window_width / 3
+
+outbase = ".".join(treefile.split(".")[:-1]) + f".{mode}.{int(window_width)}"
+statfile = f"{outbase}.divergences.tsv"
+
+pop_names = ['CHB', 'JPT', 'CHS', 'CDX', 'KHV', 'CEU', 'TSI', 'FIN', 'GBR', 'IBS', 'YRI', 'LWK', 'GWD', 'MSL', 'ESN', 'ASW', 'ACB', 'MXL', 'PUR', 'CLM', 'PEL', 'GIH', 'PJL', 'BEB', 'STU', 'ITU']
+superpops = ['EAS', 'EAS', 'EAS', 'EAS', 'EAS', 'EUR', 'EUR', 'EUR', 'EUR', 'EUR', 'AFR', 'AFR', 'AFR', 'AFR', 'AFR', 'AFR', 'AFR', 'AMR', 'AMR', 'AMR', 'AMR', 'SAS', 'SAS', 'SAS', 'SAS', 'SAS']
+num_pops = len(pop_names)
+
+ts = tskit.load(treefile)
+
+if ts.num_populations > 0:
+    pop_metadata = [json.loads(pop.metadata) for pop in ts.populations()]
+    for a, b, c in zip(pop_metadata, pop_names, superpops):
+        assert(a['name'] == b)
+        assert(a['super_population'] == c)
+
+    pop = [ts.node(ind.nodes[0]).population for ind in ts.individuals()]
+
+else:
+    # this should be a Relate tree sequence
+    assert(treefile.find("relate") >= 0)
+    # in which case the metadata are in this external file, in order,
+    # with one line per diploid
+    pop = []
+    with open("1kg/1000GP_Phase3_sub.sample", "r") as metafile:
+        header = metafile.readline()
+        assert(header == "ID POP GROUP SEX\n")
+        for line in metafile:
[…]

### 2019-08-09 figures/plot_1kg_D_and_f4.py

@@ -1,0 +1,129 @@
+#!/usr/bin/env python3
+import tskit, json
+import numpy as np
+import matplotlib
+matplotlib.use('Agg')
+import matplotlib.pyplot as plt
+import os, sys
+
+usage = """
+Usage:
+    {} treefile maskfile mode window_width
+Here `maskfile` should be a fasta of 0s and 1s.
+""".format(sys.argv[0])
+
+
+if len(sys.argv) != 5:
+    raise ValueError(usage)
+
+treefile = sys.argv[1]
+maskfile = sys.argv[2]
+mode = sys.argv[3]
+window_width = float(sys.argv[4])
+
+outbase = ".".join(treefile.split(".")[:-1]) + f".{mode}.{int(window_width)}"
+statfile = f"{outbase}.divergences.tsv" # created by calc_1kg_divergences.py
+D_plotfile = f"{outbase}.D.pdf"
+f4_plotfile = f"{outbase}.f4.pdf"
+
+pop_names = ['CHB', 'JPT', 'CHS', 'CDX', 'KHV', 'CEU', 'TSI', 'FIN', 'GBR', 'IBS', 'YRI', 'LWK', 'GWD', 'MSL', 'ESN', 'ASW', 'ACB', 'MXL', 'PUR', 'CLM', 'PEL', 'GIH', 'PJL', 'BEB', 'STU', 'ITU']
+superpops = ['EAS', 'EAS', 'EAS', 'EAS', 'EAS', 'EUR', 'EUR', 'EUR', 'EUR', 'EUR', 'AFR', 'AFR', 'AFR', 'AFR', 'AFR', 'AFR', 'AFR', 'AMR', 'AMR', 'AMR', 'AMR', 'SAS', 'SAS', 'SAS', 'SAS', 'SAS']
+num_pops = len(pop_names)
+
+# which populations?
+f4_pops = [('PUR', 'TSI', 'GWD', 'CHB'),
+           ('ASW', 'CEU', 'MSL', 'CHB'),
+           ('GBR', 'IBS', 'ITU', 'JPT')]
+
+stat_names = [f"{a},{b};{c},{d}" for a, b, c, d in f4_pops]
+colors = plt.get_cmap("tab20").colors
+stat_colors = { k : colors[j] for j, k in enumerate(stat_names) }
+
+
+# read in already-computed divergences
+all_pairs = [(i, j) for i in range(num_pops) for j in range(i, num_pops)]
[…]

### 2019-08-09 figures/plot_1kg_f4.py

@@ -1,5 +1,5 @@
 #!/usr/bin/env python3
-import tskit, tszip, json
+import tskit, json
 import numpy as np
 import matplotlib
 matplotlib.use('Agg')
@@ -34,7 +34,8 @@ num_pops = len(pop_names)
 
 # which populations?
 f4_pops = [('PUR', 'TSI', 'GWD', 'CHB'),
-           ('ASW', 'CEU', 'MSL', 'CHB')]
+           ('ASW', 'CEU', 'MSL', 'CHB'),
+           ('GBR', 'IBS', 'ITU', 'JPT')]
 
 stat_names = [f"{a},{b};{c},{d}" for a, b, c, d in f4_pops]
 colors = plt.get_cmap("tab20").colors
@@ -123,7 +124,6 @@ except:
 
     refined_stat = ts.f4(pop_nodes, indexes=indexes,
                          windows=refined_windows, mode=mode, span_normalise=False)
-
     # combine the many little windows back to the big ones
     stat = np.zeros((len(windows) - 1, len(indexes)))
     for k in range(len(indexes)):
@@ -137,6 +137,7 @@ except:
     np.savetxt(statfile, stat, 
                header="\t".join(".".join(a) for a in f4_pops),
                delimiter="\t")
+
 
 #############
 # plot results
@@ -159,5 +160,3 @@ leg = ax.legend(
          borderpad=0)
 fig.savefig(plotfile, bbox_inches = "tight")
 plt.close(fig)
-
-

### 2019-08-09 figures/plot_intro.py

@@ -1,0 +1,285 @@
+#!/usr/bin/env python3
+import pyslim, tskit, msprime
+import numpy as np
+import os, sys
+
+import matplotlib
+matplotlib.use('Agg')
+import matplotlib.pyplot as plt
+
+usage = """
+Usage:
+    {} treefile Ne mut_rate window_width
+""".format(sys.argv[0])
+
+if len(sys.argv) != 5:
+    raise ValueError(usage)
+
+treefile = sys.argv[1]
+Ne = int(sys.argv[2])
+mut_rate = float(sys.argv[3])
+window_width = float(sys.argv[4])
+
+nreps = 20
+
+outbase = ".".join(treefile.split(".")[:-1])
+newtrees = outbase + ".recap.trees"
+
+plotfile = "{}.{}.{}.f4.pdf".format(outbase, int(window_width), mut_rate)
+sdplotfile = "{}.{}.{}.f4sd.pdf".format(outbase, int(window_width), mut_rate)
+
+try:
+    ts = pyslim.load(newtrees)
+except:
+    ots = pyslim.load(treefile)
+    rts = ots.recapitate(1e-8, Ne=Ne)
+    ts = rts.simplify()
+    ts.dump(newtrees)
+
+windows = np.arange(0, ts.sequence_length+1, window_width)
+windows[-1] = ts.sequence_length
+x = windows[:-1] + np.diff(windows)/2
+
+########################
+
+pop_names = ['p1', 'p2', 'p3', 'p4']
+num_pops = len(pop_names)
+
+# which populations?
+f4_pops = [('p1', 'p2', 'p3', 'p4'),
+           ('p1', 'p4', 'p2', 'p3')]
+f4_indexes = [tuple(pop_names.index(i) for i in a) for a in f4_pops]
+
+stat_names = [f"{a},{b};{c},{d}" for a, b, c, d in f4_pops]
+colors = plt.get_cmap("tab20").colors
+stat_colors = { k : colors[j] for j, k in enumerate(stat_names) }
+styles = ["-", "-.", ":"]
+stat_styles = { k : styles[j] for j, k in enumerate(stat_names) }
+
+alive = ts.individuals_alive_at(0)
+pop = np.array([ts.node(ts.individual(ind).nodes[0]).population for ind in alive])
+
+pop_nodes = [[] for _ in range(num_pops)]
+for i in alive:
+    ind = ts.individual(i)
[…]

### 2019-08-09 figures/plot_sweep.py

@@ -1,3 +1,4 @@
+#!/usr/bin/env python3
 import pyslim, tskit, msprime
 import numpy as np
 import os, sys

### 2019-08-12 treestats_paper.tex
fig improvement

@@
 \newcommand{\nodep}{\mbox{Node}_+} % polarised
+\newcommand{\given}{\;\vert\;}
 
@@
 
-\paragraph{What about polarisation?}
 In the definition above we simply sum over all alleles at each site.
@@
 This allows statistics in principle to differentiate ancestral from derived alleles,
-information which is available in practice (albeit noisily).
-One way to make use of this information is to sum over only derived alleles.
-We say a site statistic is \textbf{polarised} if we do this,
-and so define
+information which is available in practice (albeit noisily),
+and one way to make use of this information is to sum over only derived alleles.
+Analogously to the above,
+we say a site statistic is \textbf{polarised} if we do this,
+defining
 \begin{align} \label{eqn:site_polarised}
@@
 In fact, for any region of the genome between $a$ and $b$,
+with $\treeseq_{[a,b)}$ the trees spanning that interval,
 \begin{align}
@@
     =
-    \E\left[ \site(f, \iw)_{[a,b)} \right] ,
+    \E\left[ \site(f, \iw)_{[a,b)} \given \treeseq_{[a,b)} \right] ,
 \end{align}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{A selective sweep}
-
 In this view,
@@
 How big is the contribution of mutation to the overall sampling variance of a statistic?
-Figure~\ref{fig:sweep_duality} shows this in a particular situation:
-diversity along the chromosome following a few selective sweeps (two full sweeps and three partial).
+Concretely the law of total variance lets us partition the variance of a site statistic
+into the contributions of mutation and demographic noise:
+\begin{align*}
+    \var[\site(f, \iw)_{[a,b)}] 
+    &=
+        \E\left[ \var\left[\site(f, \iw)_{[a,b)} \given \treeseq_{[a,b)}\right] \right] 
+        + \var\left[ \E\left[\site(f, \iw)_{[a,b)} \given \treeseq_{[a,b)}\right] \right] \\
+    &=
+        \E\left[ \var\left[\site(f, \iw)_{[a,b)} \given \treeseq_{[a,b)}\right] \right] 
+        + \var\left[ \branch(f, \iw)_{[a,b)} \right] .
+\end{align*}
+The first term, the variance of the statistic given the trees,
+is the contribution to variance of mutation,
+while the second term, the variance of the branch statistic,
+is the contribution of demographic variation, including drift.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection*{Branch statistics for population genetics}
+
+We examined the contributions of mutation and demography to noise
+in two simulations where population genetic (site) statistics
+are important in making inferences:
+detecting recent selective sweeps,
+and detecting introgression.
+
+Figure~\ref{fig:sweep_duality} shows windowed
+diversity along the chromosome following a few selective sweeps.
 The top two plots compare ``branch'' diversity -- i.e., as computed only with tree shape --
@@
 The same things are shown in Supplementary Figure \ref{sfig:sweep_duality_10K}
-for a simulation with 10,000 individuals.
+for a simulation with 10,000 individuals, which shows similar patterns.
+Only the last of these plots is possible to directly observe in real data:
+in the top two plots,
+the spread of independent replicates of mutational noise (black lines)
+about their expectation based on the tree sequence (red line)
+is unobservable, although estimable.
 
@@
 
+As another example, we simulated an ``admixture'' scenario:
+a first population of size $N=1000$ splits into two of equal size,
+then after $N$ generations, the second population splits,
+after another $N$ generations, the third population splits again,
+and for the final $N$ generations populations 2 and 3
+have per-capita migration rates of $4/N$ to each other.
+In this situation, a significantly negative $f_4(1,2;3,4)$ is expected,
+which we indeed find, with a mean value of around -700.
+Using various sample sizes, mutation rates, and window sizes, we then
+calculated this $f_4$ statistic in windows along a 100Mb genome,
+and show the standard deviation of both ``site'' and ``branch'' statistics across windows
+in Figure \ref{fig:introf4}.
+Since the genome is uniform (no selection or variation in recombination or mutation),
+this standard deviation is a measure of noisiness.
+As expected, site statistics are noisier than branch statistics,
+by a factor that depends mostly on the ratio of mutation to recombination rates.
+The results suggest that branch statistics would provide substantially better resolution
+at small scales along the genome, especially if mutation rate is lower than recombination rate.
+However, in practice imperfect estimation of the tree sequence
+would introduce additional noise,
+so it remains to be seen if the improvement could be made in practice.
+
+\begin{figure}
+    \includegraphics{{figures/intro/introgressed.1000.23.10000.1e-09.f4sd}.pdf}
+    \caption{
+        Standard deviation of site (black) and branch (red) $f_4(1,2;3,4)$
+        in windows along a 100Mb genome, as a function of
+        \textbf{(left)} mutation rate,
+        \textbf{(middle)} sample size, and
+        \textbf{(right)} window width.
+        The recombination rate $10^{-8}$,
+        and unless otherwise stated, the mutation rate was $10^{-9}$,
+        the window width was 10Kb,
+        and the entire population (1000 diploids in each of four populations) were used.
+        \label{fig:introf4}
+    }
+\end{figure}
+
 %%%%%%%%%%%%%%%%%%
@@
 we calculated mean sequence diversity along chromosome 20 from the 1000 Genomes dataset
-in windows of 500Kb \plr{check this}
-separately in each of its 26 populations,
+in 1Mb windows separately in each of the five continental groupings,
 calculated using the site statistic described in Example~\ref{ex:site_diversity},
@@
 and (3) by Relate \citep{relate}, an independently developed method.
+Separate curves for each of the 26 populations are shown in Supplemental Figure \ref{fig:chr20_full}.
 
@@
 \begin{figure}
-    \includegraphics[width=0.5\textwidth]{{figures/1kg/relate_chr20.site.1000000.diversity}.pdf}
-    \includegraphics[width=0.5\textwidth]{{figures/1kg/1kg_chr20.branch.1000000.diversity}.pdf}
-    \includegraphics[width=0.5\textwidth]{{figures/1kg/tgp_geva_chr20.branch.1000000.diversity}.pdf}
-    \includegraphics[width=0.5\textwidth]{{figures/1kg/relate_chr20.branch.1000000.diversity}.pdf}
+    \includegraphics[width=0.5\textwidth]{{figures/1kg/relate_chr20.site.1000000.region.diversity}.pdf}
+    \includegraphics[width=0.5\textwidth]{{figures/1kg/1kg_chr20.branch.1000000.region.diversity}.pdf}
+    \includegraphics[width=0.5\textwidth]{{figures/1kg/relate_chr20.branch.1000000.region.diversity}.pdf}
+    \includegraphics[width=0.5\textwidth]{{figures/1kg/tgp_geva_chr20.branch.1000000.region.diversity}.pdf}
     \caption{
@@
         \textbf{(top right)} tsinfer,
-        \textbf{(bottom left)} GEVA, and 
-        \textbf{(bottom right)} Relate.
+        \textbf{(bottom left)} Relate, and
+        \textbf{(bottom right)} GEVA.
         \label{fig:chr20}
@@
 compare speed to scikit-allel
-
-%%%%%%%%%%%%%%%%%%
-\subsection*{Visualizing the time profile of PCA}
 
@@
 \subsection*{Acknowledgements}
-Thanks to Graham Coop for useful suggestions
-and to Wilder Wohns for providing the 1000 Genomes tree sequence.
+Thanks to Graham Coop, Andrew Kern, and Kelley Harris for useful suggestions
+and to Wilder Wohns and Leo Speidel for providing inferred tree sequences.
 
@@
 
+\begin{figure}
+    \includegraphics[width=\textwidth]{{figures/1kg/relate_chr20.site.1000000.diversity}.pdf}
+    \includegraphics[width=\textwidth]{{figures/1kg/1kg_chr20.branch.1000000.diversity}.pdf}
+    \includegraphics[width=\textwidth]{{figures/1kg/tgp_geva_chr20.branch.1000000.diversity}.pdf}
+    \includegraphics[width=\textwidth]{{figures/1kg/relate_chr20.branch.1000000.diversity}.pdf}
+    \caption{
+        \textbf{(top left)} Mean sequence diversity 
+        in 1Mb windows along human chromosome 20 of the 1000 Genomes Project,
+        and the dual branch statistic as calculated from tree sequences inferred by
+        \textbf{(top right)} tsinfer,
+        \textbf{(bottom left)} GEVA, and 
+        \textbf{(bottom right)} Relate.
+        \label{fig:chr20_full}
+    }
+\end{figure}
+
 \end{document}

### 2019-08-12 figures/plot_1kg_diversity.py

@@ -26,6 +26,13 @@ min_window_content = window_width / 3
 outbase = ".".join(treefile.split(".")[:-1]) + f".{mode}.{int(window_width)}"
 divfile = f"{outbase}.diversity.tsv"
 plotfile = f"{outbase}.diversity.pdf"
+regionplotfile = f"{outbase}.region.diversity.pdf"
+
+method_label = {
+        "1kg/relate_chr20.trees" : "Relate",
+        "1kg/tgp_geva_chr20.trees" : "GEVA",
+        "1kg/1kg_chr20.trees" : "tsinfer"
+        }[treefile]
 
 popstyles = {"EAS" : (0, ()),
              "EUR" : (0, (1, 2)),
@@ -38,43 +45,46 @@ superpops = ['EAS', 'EAS', 'EAS', 'EAS', 'EAS', 'EUR', 'EUR', 'EUR', 'EUR', 'EUR
 popcolors = { k : colors[j % len(colors)] for j, k in enumerate(pop_names) }
 num_pops = len(pop_names)
 
-ts = tskit.load(treefile)
-
-if ts.num_populations > 0:
-    pop_metadata = [json.loads(pop.metadata) for pop in ts.populations()]
-    for a, b, c in zip(pop_metadata, pop_names, superpops):
-        assert(a['name'] == b)
-        assert(a['super_population'] == c)
-
-    pop = [ts.node(ind.nodes[0]).population for ind in ts.individuals()]
-
-else:
-    # this should be a Relate tree sequence
-    assert(treefile.find("relate") >= 0)
-    # in which case the metadata are in this external file, in order,
-    # with one line per diploid
-    pop = []
-    with open("1kg/1000GP_Phase3_sub.sample", "r") as metafile:
-        header = metafile.readline()
-        assert(header == "ID POP GROUP SEX\n")
-        for line in metafile:
-            md = line.strip().split()
-            for _ in range(2): # diploids!
-                pop.append(pop_names.index(md[1]))
-
-    
-pop_nodes = [[] for _ in range(num_pops)]
-for ind in ts.individuals():
-    pop_nodes[pop[ind.id]].extend(ind.nodes)
-
-# define the big windows
-num_windows = int(ts.sequence_length / window_width)
-windows = np.linspace(0, ts.sequence_length, num_windows + 1).astype('int')
-
 try:
     bdiv = np.loadtxt(divfile, skiprows=1)
+    num_windows = bdiv.shape[0]
[…]

### 2019-08-12 figures/plot_intro.py

@@ -25,8 +25,10 @@ nreps = 20
 outbase = ".".join(treefile.split(".")[:-1])
 newtrees = outbase + ".recap.trees"
 
-plotfile = "{}.{}.{}.f4.pdf".format(outbase, int(window_width), mut_rate)
-sdplotfile = "{}.{}.{}.f4sd.pdf".format(outbase, int(window_width), mut_rate)
+f4plotfile = "{}.{}.{}.f4.pdf".format(outbase, int(window_width), mut_rate)
+f4sdplotfile = "{}.{}.{}.f4sd.pdf".format(outbase, int(window_width), mut_rate)
+divsdplotfile = "{}.{}.{}.divsd.pdf".format(outbase, int(window_width), mut_rate)
+windowsizeplotfile = "{}.{}.windowsd.pdf".format(outbase, mut_rate)
 
 try:
     ts = pyslim.load(newtrees)
@@ -211,7 +213,7 @@ leg = ax.legend(
          borderpad=0)
 
 # done
-fig.savefig(plotfile, bbox_inches = "tight")
+fig.savefig(f4plotfile, bbox_inches = "tight")
 plt.close(fig)
 
 
@@ -245,17 +247,166 @@ for j, sample_size in enumerate(sample_sizes):
 bf4 = ts.f4(pop_nodes, indexes=f4_indexes, mode='branch')
 sf4 = ts.f4(pop_nodes, indexes=f4_indexes, mode='site')
 
+def sd(array, axis):
+    n = array.shape[axis]
+    return np.std(array, axis=axis) * np.sqrt(n / (n - 1))
 
+mut_std = sd(mut_site[:, 0, :], axis=0)
+branch_std = sd(all_branch[:, 0], axis=0)
+sub_site_std = sd(sub_site[:, 0, :], axis=0)
+sub_branch_std = sd(sub_branch[:, 0, :], axis=0)
+ymax = max(np.max(mut_std), np.max(branch_std), np.max(sub_site_std), np.max(sub_branch_std))
+
+############# SD as a function of window size
+
+branchwin = []
+sitewin = []
+ww = np.linspace(1e4, 1e6, 20)
+mts = msprime.mutate(ts, rate=mut_rate, keep=True)
+for window_width in ww:
+    windows = np.arange(0, ts.sequence_length+1, window_width)
+    windows[-1] = ts.sequence_length
+    branchwin.append(ts.f4(pop_nodes, indexes=f4_indexes, windows=windows, mode='branch'))
[…]

### 2019-08-12 figures/bootstrap_1kg_D_f4.py
node timing plot script

@@ -1,3 +1,5 @@
+## WARNING: something is wrong with this script: DEBUG before using it.
+
 #!/usr/bin/env python3
 import tskit, json
 import numpy as np

### 2019-08-12 figures/plot_1kg_f4.py

@@ -1,162 +1,0 @@
-#!/usr/bin/env python3
-import tskit, json
-import numpy as np
-import matplotlib
-matplotlib.use('Agg')
-import matplotlib.pyplot as plt
-import os, sys
-
-usage = """
-Usage:
-    {} treefile maskfile mode window_width
-Here `maskfile` should be a fasta of 0s and 1s.
-""".format(sys.argv[0])
-
-
-if len(sys.argv) != 5:
-    raise ValueError(usage)
-
-treefile = sys.argv[1]
-maskfile = sys.argv[2]
-mode = sys.argv[3]
-window_width = float(sys.argv[4])
-
-# min amount of data allowed to still deal with the window
-min_window_content = window_width / 3
-
-outbase = ".".join(treefile.split(".")[:-1]) + f".{mode}.{int(window_width)}"
-statfile = f"{outbase}.f4.tsv"
-plotfile = f"{outbase}.f4.pdf"
-
-pop_names = ['CHB', 'JPT', 'CHS', 'CDX', 'KHV', 'CEU', 'TSI', 'FIN', 'GBR', 'IBS', 'YRI', 'LWK', 'GWD', 'MSL', 'ESN', 'ASW', 'ACB', 'MXL', 'PUR', 'CLM', 'PEL', 'GIH', 'PJL', 'BEB', 'STU', 'ITU']
-superpops = ['EAS', 'EAS', 'EAS', 'EAS', 'EAS', 'EUR', 'EUR', 'EUR', 'EUR', 'EUR', 'AFR', 'AFR', 'AFR', 'AFR', 'AFR', 'AFR', 'AFR', 'AMR', 'AMR', 'AMR', 'AMR', 'SAS', 'SAS', 'SAS', 'SAS', 'SAS']
-num_pops = len(pop_names)
-
-# which populations?
-f4_pops = [('PUR', 'TSI', 'GWD', 'CHB'),
-           ('ASW', 'CEU', 'MSL', 'CHB'),
-           ('GBR', 'IBS', 'ITU', 'JPT')]
-
-stat_names = [f"{a},{b};{c},{d}" for a, b, c, d in f4_pops]
-colors = plt.get_cmap("tab20").colors
-stat_colors = { k : colors[j] for j, k in enumerate(stat_names) }
-
-
-ts = tskit.load(treefile)
-
-if ts.num_populations > 0:
-    pop_metadata = [json.loads(pop.metadata) for pop in ts.populations()]
[…]

### 2019-08-12 figures/plot_1kg_node_f4.py

@@ -1,0 +1,104 @@
+#!/usr/bin/env python3
+import tskit, json
+import numpy as np
+import matplotlib
+matplotlib.use('Agg')
+import matplotlib.pyplot as plt
+import os, sys
+
+usage = """
+Usage:
+    {} treefile maskfile
+Here `maskfile` should be a fasta of 0s and 1s.
+""".format(sys.argv[0])
+
+
+if len(sys.argv) != 3:
+    raise ValueError(usage)
+
+treefile = sys.argv[1]
+maskfile = sys.argv[2]
+
+outbase = ".".join(treefile.split(".")[:-1]) + f".node"
+plotfile = f"{outbase}.f4.pdf"
+
+pop_names = ['CHB', 'JPT', 'CHS', 'CDX', 'KHV', 'CEU', 'TSI', 'FIN', 'GBR', 'IBS', 'YRI', 'LWK', 'GWD', 'MSL', 'ESN', 'ASW', 'ACB', 'MXL', 'PUR', 'CLM', 'PEL', 'GIH', 'PJL', 'BEB', 'STU', 'ITU']
+superpops = ['EAS', 'EAS', 'EAS', 'EAS', 'EAS', 'EUR', 'EUR', 'EUR', 'EUR', 'EUR', 'AFR', 'AFR', 'AFR', 'AFR', 'AFR', 'AFR', 'AFR', 'AMR', 'AMR', 'AMR', 'AMR', 'SAS', 'SAS', 'SAS', 'SAS', 'SAS']
+num_pops = len(pop_names)
+
+# which populations?
+f4_pops = [('PUR', 'TSI', 'GWD', 'CHB'),
+           ('ASW', 'CEU', 'MSL', 'CHB'),
+           ('GBR', 'IBS', 'ITU', 'JPT')]
+f4_indexes = [tuple(pop_names.index(i) for i in a) for a in f4_pops]
+
+stat_names = [f"{a},{b};{c},{d}" for a, b, c, d in f4_pops]
+colors = plt.get_cmap("tab20").colors
+stat_colors = { k : colors[j] for j, k in enumerate(stat_names) }
+
+
+ts = tskit.load(treefile)
+
+if ts.num_populations > 0:
+    pop_metadata = [json.loads(pop.metadata) for pop in ts.populations()]
+    for a, b, c in zip(pop_metadata, pop_names, superpops):
+        assert(a['name'] == b)
+        assert(a['super_population'] == c)
[…]

### 2019-08-12 treestats_paper.tex
node timing plot script
node f4

@@
 %%%%%%%%%%%%%%%%%%
-\subsection*{Empirical duality between branch and site statistics?}
+\section*{Application to 1000 Genomes tree sequences}
 
@@
 and (3) by Relate \citep{relate}, an independently developed method.
+All calculations were done only using the portions of the chromosome
+passing the ``strict'' mask for sequencing accessibility.
 Separate curves for each of the 26 populations are shown in Supplemental Figure \ref{fig:chr20_full}.
@@
 with remaining variation explained by selection and/or mutation rate variation.
+
+%%%%%%%%%%%%%%%%%%%%%
+\subsection*{Time stratification of statistics}
+
+Thus far we have examined how statistics vary along the genome.
+A tree sequence allows us access to another dimension: time.
+Next, we use node statistics to display the \emph{time} distribution of admixture signal
+in the tree sequences inferred by tsinfer/GEVA and by Relate.
+The ``node'' $f_4(A, B; C, D)$ statistic for four populations $A$, $B$, $C$, and $D$
+can be used to calculate, for each node, the average of $(p_A - p_B)(p_C - p_D)$
+across the portion of the genome that the node is present in the tree sequence,
+where $p_A$ is the proportion of samples in $A$ that lie below that node, etcetera.
+A node with a large value will contribute a large amount to the site or branch $f_4$.
+
+\begin{figure}
+    \includegraphics{{figures/1kg/relate_chr20.node.f4}.pdf}
+    \includegraphics{{figures/1kg/tgp_geva_chr20.node.f4}.pdf}
+    \caption{
+        Time distribution of signal for two $f_4$ statistics
+        in tree sequences inferred by \textbf{(top)} Relate and \textbf{(bottom)} GEVA.
+        Both 
+        \label{fig:node_f4}
+    }
+\end{figure}
 

### 2019-08-12 figures/plot_1kg_node_f4.py

@@ -22,14 +22,21 @@ maskfile = sys.argv[2]
 outbase = ".".join(treefile.split(".")[:-1]) + f".node"
 plotfile = f"{outbase}.f4.pdf"
 
+method_label = {
+        "1kg/relate_chr20.trees" : "Relate",
+        "1kg/tgp_geva_chr20.trees" : "GEVA",
+        "1kg/1kg_chr20.trees" : "tsinfer"
+        }[treefile]
+
 pop_names = ['CHB', 'JPT', 'CHS', 'CDX', 'KHV', 'CEU', 'TSI', 'FIN', 'GBR', 'IBS', 'YRI', 'LWK', 'GWD', 'MSL', 'ESN', 'ASW', 'ACB', 'MXL', 'PUR', 'CLM', 'PEL', 'GIH', 'PJL', 'BEB', 'STU', 'ITU']
 superpops = ['EAS', 'EAS', 'EAS', 'EAS', 'EAS', 'EUR', 'EUR', 'EUR', 'EUR', 'EUR', 'AFR', 'AFR', 'AFR', 'AFR', 'AFR', 'AFR', 'AFR', 'AMR', 'AMR', 'AMR', 'AMR', 'SAS', 'SAS', 'SAS', 'SAS', 'SAS']
 num_pops = len(pop_names)
 
 # which populations?
-f4_pops = [('PUR', 'TSI', 'GWD', 'CHB'),
+f4_pops = [('PUR', 'TSI', 'GWD', 'JPT'),
            ('ASW', 'CEU', 'MSL', 'CHB'),
-           ('GBR', 'IBS', 'ITU', 'JPT')]
+           ('IBS', 'GBR', 'FIN', 'JPT'),
+           ('TSI', 'CEU', 'FIN', 'CHB')]
 f4_indexes = [tuple(pop_names.index(i) for i in a) for a in f4_pops]
 
 stat_names = [f"{a},{b};{c},{d}" for a, b, c, d in f4_pops]
@@ -71,29 +78,39 @@ for ind in ts.individuals():
 nf4 = ts.f4(pop_nodes, indexes=f4_indexes,
            mode="node", span_normalise=False)[0, :, :]
 node_denom = ts.segregating_sites([ts.samples()], mode="node", span_normalise=False)[0, :, :]
-nf4 /= node_denom
-nf4[np.isnan(nf4)] = 0.0
+# nf4 /= node_denom
+# nf4[np.isnan(nf4)] = 0.0
 
[…]

### 2019-08-13 treestats_paper.tex
nodeish f4

--- 2019-08-12 treestats_paper.tex
@@
 with remaining variation explained by selection and/or mutation rate variation.
+
+%%%%%%%%%%%%%%%%%%%%%
+\subsection*{Time stratification of statistics}
+
+Thus far we have examined how statistics vary along the genome,
+but a tree sequence allows us access to another dimension: time.
+Next, we examine how the \emph{time} distribution of admixture signal
+in the tree sequence inferred by tsinfer/GEVA.
+The branch $f_4(A, B; C, D)$ statistic for four populations $A$, $B$, $C$, and $D$
+sums $(p_A - p_B)(p_C - p_D)$ across branches, multiplied by span and height,
+where $p_A$ is the proportion of samples in $A$ that lie below that node, etcetera.
+A branch with a large value is expected to contribute a large amount to the site $f_4$,
+which we calculate from data and is used to detect historical connections between populations.
+To examine the time signature, we assigned the contribution of each branch
+to its terminal node,
+and then summed these values across all nodes in 50 logarithmically-spaced bins by time.
+We then normalised these to sum to 1 by dividing by the total branch $f_4$ value.
+
+The four-tuple of populations for the first two statistics (PUR,TSI;GWD,JPT and ASW,CEU;MSL,CHB)
+are from similar regions of the world, 
+and show almost identical time signatures,
+even though the second set of populations has a value of $f_4$ nearly eight times higher.
+The second two statistics (IBS,GBR;FIN,JPT and TSI,CEU;FIN,CHB) are also comparable,
+and also show similar signals to each other.
+These curves are noisier, because the $f_4$ values are much smaller.
+For all four comparisons, the largest contribution to $f_4$ comes from
+ancestors living tens of thousands of generations ago.
+\plr{WHAT IS THE TIME UNIT?}
+This is surprising, because $f_4$ is expected to deviate from zero due to asymmetries in tree topology
+during recent history (e.g., more branches ancestral to ASW and MSL than CEU and MSL).
+Furthermore, we expected the different populations to have much more distinct time signatures.
+The shared, ancient time signatures may be because nonzero $f_4$ statistics
+are due to differentially sorting ancestral variation,
+or inacurracies in tree sequence inference.
+It is tempting but propably unwarranted to attribute the
+large downward spike between 2000 and 3000 generations in the TSI,CEU;FIN,CHB comparison
+to a historical connection between the ancestors of CEU and the Finnish not shared by the Tuscans
+during that time period.
+
+\begin{figure}
+    \includegraphics{{figures/1kg/tgp_geva_chr20.node.time.f4}.pdf}
+    \caption{
+        Time distribution of signal for four $f_4$ statistics
+        in the tree sequence for chromosome 20 inferred from 1000 genomes data by GEVA.
+        The branch $f_4$ statistics for these four are:
+        PUR,TSI;GWD,JPT: 1603,
+        ASW,CEU;MSL,CHB: 12587,
+        IBS,GBR;FIN,JPT: -129, and
+        TSI,CEU;FIN,CHB: -96.
+        Population codes are 
+        `PUR': Puerto Ricans from Puerto Rico;
+        `TSI': Toscani in Italia;
+        `GWD': Gambian in Western Divisions in the Gambia;
+        `JPT': Japanese in Tokyo, Japan;
+        `ASW': Americans of African Ancestry in SW USA;
+        `CEU': Utah Residents (CEPH) with Northern and Western European Ancestry,
+        `MSL': Mende in Sierra Leone;
+        `CHB': Han Chinese in Beijing, China;
+        `IBS': Iberian Population in Spain;
+        `GBR': British in England and Scotland;
+        and `FIN': Finnish in Finland.
+        \label{fig:node_f4}
+    }
+\end{figure}
 

### 2019-08-13 figures/calc_1kg_real_node_f4.py

@@ -1,0 +1,164 @@
+#!/usr/bin/env python3
+import tskit, json
+import numpy as np
+import matplotlib
+matplotlib.use('Agg')
+import matplotlib.pyplot as plt
+import os, sys
+
+usage = """
+Usage:
+    {} treefile maskfile
+Here `maskfile` should be a fasta of 0s and 1s.
+""".format(sys.argv[0])
+
+
+if len(sys.argv) != 3:
+    raise ValueError(usage)
+
+treefile = sys.argv[1]
+maskfile = sys.argv[2]
+
+outbase = ".".join(treefile.split(".")[:-1]) + f".node"
+
+method_label = {
+        "1kg/relate_chr20.trees" : "Relate",
+        "1kg/tgp_geva_chr20.trees" : "GEVA",
+        "1kg/1kg_chr20.trees" : "tsinfer"
+        }[treefile]
+
+pop_names = ['CHB', 'JPT', 'CHS', 'CDX', 'KHV', 'CEU', 'TSI', 'FIN', 'GBR', 'IBS', 'YRI', 'LWK', 'GWD', 'MSL', 'ESN', 'ASW', 'ACB', 'MXL', 'PUR', 'CLM', 'PEL', 'GIH', 'PJL', 'BEB', 'STU', 'ITU']
+superpops = ['EAS', 'EAS', 'EAS', 'EAS', 'EAS', 'EUR', 'EUR', 'EUR', 'EUR', 'EUR', 'AFR', 'AFR', 'AFR', 'AFR', 'AFR', 'AFR', 'AFR', 'AMR', 'AMR', 'AMR', 'AMR', 'SAS', 'SAS', 'SAS', 'SAS', 'SAS']
+num_pops = len(pop_names)
+
+# which populations?
+f4_pops = [('PUR', 'TSI', 'GWD', 'JPT'),
+           ('ASW', 'CEU', 'MSL', 'CHB'),
+           ('IBS', 'GBR', 'FIN', 'JPT'),
+           ('TSI', 'CEU', 'FIN', 'CHB')]
+f4_indexes = [tuple(pop_names.index(i) for i in a) for a in f4_pops]
+
+stat_names = [f"{a},{b};{c},{d}" for a, b, c, d in f4_pops]
+colors = plt.get_cmap("tab20").colors
+stat_colors = { k : colors[j] for j, k in enumerate(stat_names) }
+
+
+ts = tskit.load(treefile)
+
+if ts.num_populations > 0:
[…]

### 2019-08-13 figures/plot_1kg_node_f4.py

@@ -21,6 +21,7 @@ maskfile = sys.argv[2]
 
 outbase = ".".join(treefile.split(".")[:-1]) + f".node"
 plotfile = f"{outbase}.f4.pdf"
+timeplotfile = f"{outbase}.time.f4.pdf"
 
 method_label = {
         "1kg/relate_chr20.trees" : "Relate",
@@ -77,9 +78,6 @@ for ind in ts.individuals():
 
 nf4 = ts.f4(pop_nodes, indexes=f4_indexes,
            mode="node", span_normalise=False)[0, :, :]
-node_denom = ts.segregating_sites([ts.samples()], mode="node", span_normalise=False)[0, :, :]
-# nf4 /= node_denom
-# nf4[np.isnan(nf4)] = 0.0
 
 if method_label == "Relate":
     num_bins = 500
@@ -97,6 +95,7 @@ for j in range(nf4.shape[1]):
     #                    / np.bincount(node_bins, minlength=num_bins))
 
 f4_binned /= np.nansum(f4_binned, axis=0)
+
 
 #### plot it
 
@@ -119,3 +118,30 @@ ax.legend(
 
 fig.savefig(plotfile, bbox_inches = "tight")
 plt.close(fig)
+
+###### load and plot branch-length-weighted f4 values
+
+time_f4 = np.zeros((num_bins, len(f4_indexes)))
+for j, f4p in enumerate(f4_pops):
+    sn = ".".join(f4p)
+    datafile = f"{outbase}.{sn}.realf4.binned.txt"
+    rf4 = np.loadtxt(datafile)
+    time_f4[:, j] = rf4 / np.nansum(rf4)
+
+
+fig = plt.figure(figsize=(6, 3))
+ax = fig.add_subplot(111)
+ax.set_xlabel("time ago")
+ax.set_ylabel("branch f4")
+
+for j, sn in enumerate(stat_names):
+    fl = ax.semilogx(time_bins, time_f4[:, j], label=sn)
+
+ax.legend(
+        fontsize = "small",
+        frameon = False,
+        borderpad = 0)
+
+fig.savefig(timeplotfile, bbox_inches = "tight")
+plt.close(fig)
+

### 2019-08-30 algorithms.py
Add algorithms python file.
Also strip trailing spaces from tex file.

@@ -1,0 +1,144 @@
+"""
+Direct implementations of the algorithms in the paper.
+"""
+import msprime
+import tskit
+import numpy as np
+
+def diffs_branch_algorithm(ts, f, S, w):
+
+    tau = ts.tables.nodes.time
+    pi = np.zeros(ts.num_nodes, dtype=np.int32) - 1
+    beta = np.zeros(ts.num_nodes)
+    x = np.zeros(ts.num_nodes)
+    F = np.zeros(ts.num_nodes)
+    sigma = 0
+    s = 0
+
+    for j, u in enumerate(S):
+        x[u] = w[j]
+        F[u] = f(x[u])
+
+    for (t_left, t_right), edges_out, edges_in in ts.edge_diffs():
+        for edge in edges_out:
+            u, v = edge.child, edge.parent
+            s -= beta[u] * F[u]
+            pi[u] = -1
+            beta[u] = 0
+
+            while v != -1:
+                s -= beta[v] * F[v]
+                x[v] -= x[u]
+                F[v] = f(x[v])
+                s += beta[v] * F[v]
+                v = pi[v]
+
+        for edge in edges_in:
+            u, v = edge.child, edge.parent
+            pi[u] = v
+            beta[u] = tau[v] - tau[u]
+            s += beta[u] * F[u]
+
+            while v != -1:
+                s -= beta[v] * F[v]
+                x[v] += x[u]
+                F[v] = f(x[v])
+                s += beta[v] * F[v]
+                v = pi[v]
+
+        sigma += (t_right - t_left) * s
+
+    return sigma / ts.sequence_length
+
+
+def tables_branch_algorithm(ts, f, S, w):
+
+    N = ts.tables.nodes
+    E = ts.tables.edges
+    L = ts.sequence_length
+
+    tau = N.time
+    pi = np.zeros(len(N), dtype=np.int32) - 1
+    beta = np.zeros(len(N))
+    x = np.zeros(len(N))
+    F = np.zeros(len(N))
+    sigma = 0
+    s = 0
+
+    for j, u in enumerate(S):
+        x[u] = w[j]
+        F[u] = f(x[u])
+
+    I = sorted(
+        range(len(E)), key=lambda j: (
+            E[j].left, tau[E[j].parent], E[j].parent, E[j].child))
+    O = sorted(
+        range(len(E)), key=lambda j: (
+            E[j].right, -tau[E[j].parent], -E[j].parent, -E[j].child))
+    j = 0
+    k = 0
[…]

### 2019-08-30 treestats_paper.tex
Add algorithms python file.
Also strip trailing spaces from tex file.
Initial draft of the branch stats algorithm.

@@
 \newcommand{\aw}{{\bar x}} % allele weights
+
+\newcommand{\Nt}{\mathcal{N}}  % node table
+\newcommand{\Et}{\mathcal{E}}  % edge table
+\newcommand{\indexin}[0]{\ensuremath{\mathcal{I}}}
+\newcommand{\indexout}[0]{\ensuremath{\mathcal{O}}}
 
@@
 
+\clearpage
+\begin{taocpalg}{B}{General branch statistics}
+{Given a list of $n$ sample nodes $S$, corresponding weights $w$ and
+summary function $f: \R \rightarrow \R$
+compute the span-normalised statistic $\sigma$
+over a tree sequence with sequence length $L$
+defined by the node and edge tables $\Nt$ and $\Et$. We assume
+that the index vectors $\indexin$ and $\indexout$ have been precomputed.}
+
+\algstep{B1.}{Initialisation.}{
+    For $0 \leq u < |\Et|$ set $\beta_u \leftarrow x_u
+    \leftarrow F_u \leftarrow 0$ and $\pi_u \leftarrow -1$.
+    Then, for $0 \leq j < n$ set $u \leftarrow S_j$,
+    $x_u \leftarrow w_j$ and $F_u \leftarrow f(x_j)$. Finally,
+    set $s \leftarrow \sigma \leftarrow j \leftarrow k \leftarrow t_l \leftarrow 0$.
+}
+
+\algstep{B2.}{Terminate.}{If $j \geq |\Et|$ or $t_l \geq L$
+    set $\sigma \leftarrow \sigma / L$ and terminate.
+}
+
+\algstep{B3.}{Edge removal loop.}{If $k \geq |\Et|$ or $t_l \neq
+    \Et[\indexout_k]\mbox{.right}$ go to \algref{B6}}.
+
+\algstep{B4.}{Remove edge.}{Set
+    $u \leftarrow \Et[\indexout_k]\mbox{.child}$,
+    $v \leftarrow \Et[\indexout_k]\mbox{.parent}$ and $k \leftarrow k + 1$.
+    Then set $s \leftarrow s - \beta_u F_u$, $\pi_u \leftarrow -1$
+    and $\beta_u \leftarrow 0$.
+}
+
+\algstep{B5.}{Propagate loss.}{While $v \neq -1$, set
+    $s \leftarrow s - \beta_v F_v$, $x_v \leftarrow x_v - x_u$,
+    $F_v \leftarrow f(x_v)$, $s \leftarrow s - \beta_v F_v$
+    and $v \leftarrow \pi_v$. Afterwards, go to \algref{B3}.
+}
+
+\algstep{B6.}{Edge insertion loop.}{If $j \geq |\Et|$ or $t_l \neq
+    \Et[\indexin_j]\mbox{.left}$ go to \algref{B9}}.
+
+\algstep{B7.}{Insert edge.}{Set
+    $u \leftarrow \Et[\indexin_j]\mbox{.child}$,
+    $v \leftarrow \Et[\indexin_j]\mbox{.parent}$ and $j \leftarrow j + 1$.
+    Then set $\pi_u \leftarrow v$,
+    $\beta_u \leftarrow \Nt[v]\mbox{.time} - \Nt[u]\mbox{.time}$ and
+    $s \leftarrow s + \beta_u F_u$.
+}
+
+\algstep{B8.}{Propagate gain.}{While $v \neq -1$, set
+    $s \leftarrow s - \beta_v F_v$, $x_v \leftarrow x_v + x_u$,
+    $F_v \leftarrow f(x_v)$, $s \leftarrow s - \beta_v F_v$
+    and $v \leftarrow \pi_v$. Afterwards, go to \algref{B6}.
+}
+
+\algstep{B9.}{Tree loop tail.}{Set $t_r \leftarrow L$.
+    If $j < |\Et|$ set $t_r \leftarrow \min(t_r, \Et[\indexin_j]\mbox{.left})$.
+    Then, if $k < |\Et|$ set $t_r \leftarrow \min(t_r, \Et[\indexout_k]\mbox{.right})$
+    Finally, set $\sigma \leftarrow (t_r - t_l)s$,  $t_l \leftarrow t_r$
+    and go to \algref{B2}.
+}
+
+\end{taocpalg}
+
 %%%%%%%%%%%%%%%%%%%%%

### 2019-09-01 treestats_paper.tex
Draft figure showing algorithm state.

@@
 \subsection*{Algorithms}
-write-up of algorithm for maintaining branch statistic across trees
+In this section we provide describe and analyse the algorithm used
+to compute branch length statistics. This is a generalation
+of the algorithm used to maintain the number of samples from a
+given set that derive from each node in a tree sequence.
+\cite[Algorithm L]{kelleher2016efficient}.
+
+\jk{Plan for section}
+\begin{itemize}
+\item First paragraph is a general outline of the strategy.
+\item Explanation of notation and the index vectors (we should probably define
+these again, in the interest of completeness. I'm not sure the definition
+in the msprime paper really worked)
+\item Algorithm listing.
+\item Next para is an explanation of the algorithm in words, referring
+to the figure for what we mean by adding and removing edges.
+\item Next para is an analysis of the algorithm. We note here that
+we propagate stuff all the way up the tree close to optimally, because of the
+ordering of edge insertions and removals. For example, we always disconnect
+a subtree from the main tree first, meaning that the only time we go all the
+way to the root is the first disconnection.
+Discuss the figure perf figure.
+\end{itemize}
 
@@
 \end{taocpalg}
+
+\begin{figure}
+
+% Run the script illustrate_algorithm.py to generate these pdfs and
+% the latex tables below.
+\begin{tabular}{cccc}
+\includegraphics[width=3cm]{figures/tree_0_in_6.pdf} &
+\includegraphics[width=3cm]{figures/tree_0_out_0.pdf} &
+\includegraphics[width=3cm]{figures/tree_0_out_1.pdf} &
+\includegraphics[width=3cm]{figures/tree_1_in_0.pdf}
+\\
+
+\begin{tabular}{c|cccc}
+& $\pi$ & $\beta$ & $x$ & $F$\\
+0 & 7 & 2.0 & 1.0 & 0.2\\
+1 & 6 & 1.0 & 1.0 & 0.2\\
+2 & 6 & 1.0 & 1.0 & 0.2\\
+3 & 6 & 1.0 & 1.0 & 0.2\\
+4 & 8 & 3.0 & 1.0 & 0.2\\
+5 & -1 & 0.0 & 0.0 & 0.0\\
+6 & 7 & 1.0 & 3.0 & 0.3\\
+7 & 8 & 1.0 & 4.0 & 0.2\\
+8 & -1 & 0.0 & 5.0 & 0.0\\
+\end{tabular}
+&
+\begin{tabular}{c|cccc}
+& $\pi$ & $\beta$ & $x$ & $F$\\
+0 & -1 & 0.0 & 1.0 & 0.2\\
+1 & 6 & 1.0 & 1.0 & 0.2\\
+2 & 6 & 1.0 & 1.0 & 0.2\\
+3 & 6 & 1.0 & 1.0 & 0.2\\
+4 & 8 & 3.0 & 1.0 & 0.2\\
+5 & -1 & 0.0 & 0.0 & 0.0\\
+6 & 7 & 1.0 & 3.0 & 0.3\\
+7 & 8 & 1.0 & 3.0 & 0.3\\
+8 & -1 & 0.0 & 4.0 & 0.2\\
+\end{tabular}
+&
+\begin{tabular}{c|cccc}
+& $\pi$ & $\beta$ & $x$ & $F$\\
+0 & -1 & 0.0 & 1.0 & 0.2\\
+1 & 6 & 1.0 & 1.0 & 0.2\\
+2 & 6 & 1.0 & 1.0 & 0.2\\
+3 & -1 & 0.0 & 1.0 & 0.2\\
+4 & 8 & 3.0 & 1.0 & 0.2\\
+5 & -1 & 0.0 & 0.0 & 0.0\\
+6 & 7 & 1.0 & 2.0 & 0.3\\
+7 & 8 & 1.0 & 2.0 & 0.3\\
+8 & -1 & 0.0 & 3.0 & 0.3\\
+\end{tabular}
+&
+\begin{tabular}{c|cccc}
+& $\pi$ & $\beta$ & $x$ & $F$\\
+0 & 5 & 1.0 & 1.0 & 0.2\\
+1 & 6 & 1.0 & 1.0 & 0.2\\
+2 & 6 & 1.0 & 1.0 & 0.2\\
+3 & -1 & 0.0 & 1.0 & 0.2\\
+4 & 8 & 3.0 & 1.0 & 0.2\\
+5 & -1 & 0.0 & 1.0 & 0.2\\
+6 & 7 & 1.0 & 2.0 & 0.3\\
+7 & 8 & 1.0 & 2.0 & 0.3\\
+8 & -1 & 0.0 & 3.0 & 0.3\\
+\end{tabular}
+\end{tabular}
+
+\caption{Example of the algorithm state during the transition between two
+trees. \jk{If we keep this then we talk about what we see in the different
+examples. We want to give some insight into how the algorithm works by showing
+the intermediate states.}}
+\end{figure}
 

### 2019-09-01 illustrate_algorithm.py

@@ -1,0 +1,129 @@
+"""
+Draw an illustration of the branch stats algorithm.
+"""
+
+import msprime
+import tskit
+import numpy as np
+import subprocess
+
+
+def convert_svg(svg_file):
+    pdf_file = svg_file.replace(".svg", ".pdf")
+    print("converting", svg_file, pdf_file)
+    subprocess.check_call(f"rsvg-convert {svg_file} -o {pdf_file}", shell=True)
+
+def print_latex_table(pi, beta, x, F):
+    print("\\begin{tabular}{c|cccc}")
+    print("& $\pi$ & $\\beta$ & $x$ & $F$\\\\")
+    for u in range(len(pi)):
+        print(f"{u} & {pi[u]} & {beta[u]} & {x[u]} & {F[u]}\\\\")
+    print("\\end{tabular}")
+
+
+def branch_stat_algorithm(ts, f, S, w):
+
+    tau = ts.tables.nodes.time
+    pi = np.zeros(ts.num_nodes, dtype=np.int32) - 1
+    beta = np.zeros(ts.num_nodes)
+    x = np.zeros(ts.num_nodes)
+    F = np.zeros(ts.num_nodes)
+    sigma = 0
+    s = 0
+    for j, u in enumerate(S):
+        x[u] = w[j]
+        F[u] = f(x[u])
+
+    tree = tskit.Tree(ts)
+    # TODO Better SVG for not drawing this.
+    edge_attrs = {u: {"stroke": "white"} for u in range(ts.num_nodes)}
+
+    for (t_left, t_right), edges_out, edges_in in ts.edge_diffs():
+        for edge_num, edge in enumerate(edges_out):
+            u, v = edge.child, edge.parent
+            s -= beta[u] * F[u]
+            pi[u] = -1
+            beta[u] = 0
+
+            while v != -1:
+                s -= beta[v] * F[v]
+                x[v] -= x[u]
+                F[v] = f(x[v])
+                s += beta[v] * F[v]
+                v = pi[v]
+
+            print("out", edge_num)
+            print_latex_table(pi, beta, x, F)
+            edge_attrs[edge.child] = {"stroke": "white"}
+            svg_file = f"figures/tree_{tree.index}_out_{edge_num}.svg"
+            tree.draw_svg(
+                svg_file,
+                max_tree_height="ts",
+                edge_attrs=edge_attrs)
+            convert_svg(svg_file)
+
+        tree.next()
+
[…]

### 2019-09-02 illustrate_algorithm.py
Updated figure to use two states.

@@ -11,11 +11,14 @@ import subprocess
 def convert_svg(svg_file):
     pdf_file = svg_file.replace(".svg", ".pdf")
     print("converting", svg_file, pdf_file)
-    subprocess.check_call(f"rsvg-convert {svg_file} -o {pdf_file}", shell=True)
+    # subprocess.check_call(f"rsvg-convert {svg_file} -o {pdf_file}", shell=True)
+    # Inkscape does a better job of outputing vector graphics.
+    subprocess.check_call(f"inkscape {svg_file} --export-pdf={pdf_file}", shell=True)
 
 def print_latex_table(pi, beta, x, F):
     print("\\begin{tabular}{c|cccc}")
     print("& $\pi$ & $\\beta$ & $x$ & $F$\\\\")
+    print("\\hline")
     for u in range(len(pi)):
         print(f"{u} & {pi[u]} & {beta[u]} & {x[u]} & {F[u]}\\\\")
     print("\\end{tabular}")
@@ -35,8 +38,7 @@ def branch_stat_algorithm(ts, f, S, w):
         F[u] = f(x[u])
 
     tree = tskit.Tree(ts)
-    # TODO Better SVG for not drawing this.
-    edge_attrs = {u: {"stroke": "white"} for u in range(ts.num_nodes)}
+    edge_attrs = {}
 
     for (t_left, t_right), edges_out, edges_in in ts.edge_diffs():
         for edge_num, edge in enumerate(edges_out):
@@ -54,7 +56,7 @@ def branch_stat_algorithm(ts, f, S, w):
 
             print("out", edge_num)
             print_latex_table(pi, beta, x, F)
-            edge_attrs[edge.child] = {"stroke": "white"}
+            edge_attrs[edge.child] = {"stroke-opacity": "0"}
             svg_file = f"figures/tree_{tree.index}_out_{edge_num}.svg"
             tree.draw_svg(
                 svg_file,
@@ -77,16 +79,16 @@ def branch_stat_algorithm(ts, f, S, w):
                 s += beta[v] * F[v]
                 v = pi[v]
 
-            print("in", edge_num, edge)
-            print_latex_table(pi, beta, x, F)
-
             edge_attrs[edge.child] = {"stroke": "black"}
-            svg_file = f"figures/tree_{tree.index}_in_{edge_num}.svg"
-            tree.draw_svg(
[…]

### 2019-09-02 treestats_paper.tex

--- 2019-08-30 treestats_paper.tex
@@
 \end{taocpalg}
+
+\begin{figure}
+
+% Run the script illustrate_algorithm.py to generate these pdfs and
+% the latex tables below.
+\begin{center}
+\begin{tabular}{ccc}
+\includegraphics[width=4cm]{figures/tree_0_init.pdf} & &
+\includegraphics[width=4cm]{figures/tree_0_out_0.pdf}
+\\
+
+\begin{tabular}{c|cccc}
+& $\pi$ & $\beta$ & $x$ & $F$\\
+\hline
+0 & 8 & 4.0 & 1.0 & 0.2\\
+1 & 5 & 1.0 & 1.0 & 0.2\\
+2 & 5 & 1.0 & 1.0 & 0.2\\
+3 & 6 & 2.0 & 1.0 & 0.2\\
+4 & 6 & 2.0 & 1.0 & 0.2\\
+5 & 7 & 2.0 & 2.0 & 0.3\\
+6 & 7 & 1.0 & 2.0 & 0.3\\
+7 & 8 & 1.0 & 4.0 & 0.2\\
+8 & -1 & 0.0 & 5.0 & 0.0\\
+\end{tabular}
+& &
+\begin{tabular}{c|cccc}
+& $\pi$ & $\beta$ & $x$ & $F$\\
+\hline
+0 & 8 & 4.0 & 1.0 & 0.2\\
+1 & 5 & 1.0 & 1.0 & 0.2\\
+2 & 5 & 1.0 & 1.0 & 0.2\\
+3 & 6 & 2.0 & 1.0 & 0.2\\
+4 & 6 & 2.0 & 1.0 & 0.2\\
+5 & 7 & 2.0 & 2.0 & 0.3\\
+6 & -1 & 0.0 & 2.0 & 0.3\\
+7 & 8 & 1.0 & 2.0 & 0.3\\
+8 & -1 & 0.0 & 3.0 & 0.3\\
+\end{tabular}
+\\
+(A) && (B)
+\end{tabular}
+\end{center}
+
+\caption{
+\label{fig:algorithm_example}
+Illustration of the internal state for algorithm B for $\iw = \bone_S$ and
+$f(x) = x(5 - x) / 20$ as in Example~\ref{ex:site_diversity}. (A)
+Immediately before we remove the edge joining 6 to 7 and (B) immediately after.
+See the text for descriptions of the arrays encoding the state.}
+\end{figure}
 

### 2019-09-02 treestats_paper.tex
removed stuff (closes #7)

@@
 \subsection*{Acknowledgements}
-Thanks to Graham Coop, Andrew Kern, and Kelley Harris for useful suggestions
+Thanks to Graham Coop, Andrew Kern, Nick Patterson, and Kelley Harris for useful suggestions
 and to Wilder Wohns and Leo Speidel for providing inferred tree sequences.

### 2019-09-02 examples.py

@@ -1,22 +1,0 @@
-import msprime
-import numpy as np
-from treestats import *
-
-## Assign angles in [0,2*pi] to each sample, then compute the average angle
-## for each internal node of each tree of all samples below each node.
-
-ts = msprime.simulate(10, recombination_rate=1.0)
-
-sample_angles = np.random.uniform(0.0, 2*np.pi, ts.num_samples)
-
-def angle_mean(a):
-    # gives the angle of the average vector in the unit circle
-    # https://en.wikipedia.org/wiki/Mean_of_circular_quantities
-    return np.math.atan2(sum(np.sin(a)), sum(np.cos(a)))
-
-wt = weighted_trees(ts, [sample_angles], angle_mean)
-
-for t in wt:
-    print("Tree ", t.index)
-    for u, w in zip(t.nodes(), t.node_weights()):
-        print("   node ", u, ": angle ", w)

### 2019-09-02 figures/plot_1kg_node_f4.py

@@ -131,8 +131,8 @@ for j, f4p in enumerate(f4_pops):
 
 fig = plt.figure(figsize=(6, 3))
 ax = fig.add_subplot(111)
-ax.set_xlabel("time ago")
-ax.set_ylabel("branch f4")
+ax.set_xlabel("time ago (generations)")
+ax.set_ylabel("normalised branch f4")
 
 for j, sn in enumerate(stat_names):
     fl = ax.semilogx(time_bins, time_f4[:, j], label=sn)

### 2019-09-02 treestats/__init__.py

@@ -1,4 +1,0 @@
-from .weighted_trees import *
-from .sfs import *
-from .fst import *
-

### 2019-09-02 treestats/fst.py

@@ -1,92 +1,0 @@
-# Comparison of two methods for computing fst,
-# one with genotypes, one with trees.
-#
-# These assume an infinite-sites model.
-
-import msprime
-import numpy as np
-
-
-def variants_fst(tree_sequence, pop0, pop1, windows):
-    '''
-    Compute mean Fst across all variants in each window.
-    '''
-    n0 = len(pop0)
-    n1 = len(pop1)
-    num_windows = len(windows) - 1
-    if windows[0] != 0.0:
-        raise ValueError(
-            "Windows must start at the start of the sequence (at 0.0).")
-    if windows[-1] != tree_sequence.sequence_length:
-        raise ValueError("Windows must extend to the end of the sequence.")
-    for k in range(num_windows):
-        if windows[k + 1] <= windows[k]:
-            raise ValueError("Windows must be increasing.")
-    S = np.zeros(num_windows)
-    # index of *left-hand* end of the current window
-    window_num = 0
-    num_sites = 0
-    for v in tree_sequence.variants():
-        while v.position >= windows[window_num + 1]:
-            if num_sites > 0:
-                S[window_num] /= num_sites
-            window_num += 1
-            num_sites = 0
-        num_sites += 1
-        x0 = sum(v.genotypes[pop0])
-        x1 = sum(v.genotypes[pop1])
-        if (x0 + x1 > 0.0) and (x0 + x1 < n0 + n1):
-            p0 = x0 / n0
-            p1 = x1 / n1
-            fst = 1 - (p0 * (1 - p0) + p1 * (1 - p1)) / ((p0 + p1) * (1 - (p0 + p1) / 2))
-            S[window_num] += fst
-    # do the last window
-    if num_sites > 0:
-        S[window_num] /= num_sites
-    return S
-
-def mutations_fst(tree_sequence, pop0, pop1, windows):
-    '''
-    Compute mean Fst across all variants in each window.
-    Assumes an infinite-sites model.
-    '''
-    n0 = len(pop0)
-    n1 = len(pop1)
-    num_windows = len(windows) - 1
-    if windows[0] != 0.0:
-        raise ValueError(
-            "Windows must start at the start of the sequence (at 0.0).")
-    if windows[-1] != tree_sequence.sequence_length:
-        raise ValueError("Windows must extend to the end of the sequence.")
-    for k in range(num_windows):
-        if windows[k + 1] <= windows[k]:
-            raise ValueError("Windows must be increasing.")
-    S = np.zeros(num_windows)
[…]

### 2019-09-02 treestats/sfs.py

@@ -1,290 +1,0 @@
-# Experimental code for computing functions of the joint SFS along a sequence
-# and related things
-
-import msprime
-import numpy as np
-
-def get_derived_state(site, mut_id):
-    """
-    Find the derived state of the mutation with id `mut_id` at site `site`.
-    """
-    if mut_id == -1:
-        state = site.ancestral_state
-    else:
-        for m in site.mutations:
-            if m.id == mut_id:
-                state = m.derived_state
-    return state
-
-def tracked_samples_joint_branch_sfs(tree_sequence, pop0, pop1, begin=0.0, end=None):
-    if end is None:
-        end = tree_sequence.sequence_length
-    n0 = len(pop0)
-    n1 = len(pop1)
-    S = np.zeros((n0 + 1, n1 + 1))
-    tt0 = tree_sequence.trees(tracked_samples=pop0,
-                                   sample_counts=True)
-    tt1 = tree_sequence.trees(tracked_samples=pop1,
-                                   sample_counts=True)
-    for t0, t1 in zip(tt0, tt1):
-        root = t0.root
-        tr_len = min(end, t0.interval[1]) - max(begin, t0.interval[0])
-        if tr_len > 0:
-            for node in t0.nodes():
-                if node != root:
-                    x0 = t0.num_tracked_samples(node)
-                    x1 = t1.num_tracked_samples(node)
-                    if (x0 > 0) or (x1 > 0):
-                        S[x0, x1] += t0.branch_length(node) * tr_len
-    S /= (end-begin)
-    return S
-
-
-def variants_joint_sfs(tree_sequence, pop0, pop1, begin=0.0, end=None):
-    '''
-    '''
-    if end is None:
-        end = tree_sequence.sequence_length
-    n0 = len(pop0)
-    n1 = len(pop1)
-    S = np.zeros((n0 + 1, n1 + 1))
-    for v in tree_sequence.variants():
-        if v.position >= end:
-            break
-        if v.position >= begin:
-            for a in set(v.genotypes):
-                x0 = sum(v.genotypes[pop0] == a)
-                x1 = sum(v.genotypes[pop1] == a)
-                if (x0 > 0) or (x1 > 0):
-                    S[x0, x1] += 1.0
-    S /= (end - begin)
-    return S
-
-
-def joint_site_frequency_spectrum(tree_sequence, pop0, pop1, windows=None):
-    '''
-    Computes the joint folded site frequency spectrum between pop0 and pop1,
-    independently in windows.
-
[…]

### 2019-09-02 treestats/weighted_trees.py

@@ -1,57 +1,0 @@
-import msprime
-import itertools
-
-def weighted_trees(ts, sample_weight_list, node_fun=sum):
-    '''
-    Here ``sample_weight_list`` is a list of lists of weights, each of the same
-    length as the samples in the tree sequence ``ts``. This returns an iterator
-    over the trees in ``ts`` that is identical to ``ts.trees()`` except that
-    each tree ``t`` has the additional method `t.node_weights()` which returns
-    an iterator over the "weights" for each node in the tree, in the same order
-    as ``t.nodes()``.
-
-    Each node has one weight, computed separately for each set of weights in
-    ``sample_weight_list``. Each such weight is defined for a particular list
-    of ``sample_weights`` recursively:
-
-    1. First define ``all_weights[ts.samples()[j]] = sample_weights[j]``
-        and ``all_weights[k] = 0`` otherwise.
-    2. The weight for a node ``j`` with children ``u1, u2, ..., un`` is
-        ``node_fun([all_weights[j], weight[u1], ..., weight[un]])``.
-
-    For instance, if ``sample_weights`` is a vector of all ``1``s, and
-    ``node_fun`` is ``sum``, then the weight for each node in each tree
-    is the number of samples below it, equivalent to ``t.num_samples(j)``.
-
-    To do this, we need to only recurse upwards from the parent of each
-    added or removed edge, updating the weights.
-    '''
-    samples = ts.samples()
-    num_weights = len(sample_weight_list)
-    # initialize the weights
-    base_X = [[0.0 for _ in range(num_weights)] for _ in range(ts.num_nodes)]
-    X = [[0.0 for _ in range(num_weights)] for _ in range(ts.num_nodes)]
-    for j, u in enumerate(samples):
-        for k in range(num_weights):
-            X[u][k] = sample_weight_list[k][j]
-            base_X[u][k] = sample_weight_list[k][j]
-
-    for t, (interval, records_out, records_in) in zip(ts.trees(), ts.edge_diffs()):
-        for edge in itertools.chain(records_out, records_in):
-            u = edge.parent
-            while u != -1:
-                for k in range(num_weights):
-                    U = [base_X[u][k]] + [X[u][k] for u in t.children(u)]
-                    X[u][k] = node_fun(U)
-                u = t.parent(u)
-
-        def the_node_weights(self):
-            for u in self.nodes():
-                yield X[u]
-
-        # magic that uses "descriptor protocol"
[…]

### 2019-09-02 treestats_paper.tex
comments on algorithm

@@
 \newcommand{\Et}{\mathcal{E}}  % edge table
+\newcommand{\prop}[1]{.\mbox{\texttt{#1}}} % property of a thing
 \newcommand{\indexin}[0]{\ensuremath{\mathcal{I}}}
@@
 
+\plr{Question: Figures 1, 2 and 7 show the same sort of thing 
+    (fig 1 is a site stat, aiming to give the general overview;
+    figs 2 and 7 are branch stats). Should they be more similar?
+    Do we need all of these?}
+
+\plr{right now things like $\Et\mbox{.right}$ read like a funnily terminated sentence.
+     I've changed these with a macro, feel free to adjust.}
+
 \clearpage
@@
 defined by the node and edge tables $\Nt$ and $\Et$. We assume
-that the index vectors $\indexin$ and $\indexout$ have been precomputed.}
+that the index vectors $\indexin$ and $\indexout$ have been precomputed.
+\plr{I assume you plan to say in words what $\beta$, $x$, $F$, etc are}
+}
 
@@
     set $\sigma \leftarrow \sigma / L$ and terminate.
+    \plr{perhaps say "return $\sigma/L$" to make it more clear what the output is?}
 }
@@
 \algstep{B3.}{Edge removal loop.}{If $k \geq |\Et|$ or $t_l \neq
-    \Et[\indexout_k]\mbox{.right}$ go to \algref{B6}}.
+    \Et[\indexout_k]\prop{right}$ go to \algref{B6}.}
 
 \algstep{B4.}{Remove edge.}{Set
-    $u \leftarrow \Et[\indexout_k]\mbox{.child}$,
-    $v \leftarrow \Et[\indexout_k]\mbox{.parent}$ and $k \leftarrow k + 1$.
+    $u \leftarrow \Et[\indexout_k]\prop{child}$,
+    $v \leftarrow \Et[\indexout_k]\prop{parent}$ and $k \leftarrow k + 1$.
     Then set $s \leftarrow s - \beta_u F_u$, $\pi_u \leftarrow -1$
@@
 \algstep{B6.}{Edge insertion loop.}{If $j \geq |\Et|$ or $t_l \neq
-    \Et[\indexin_j]\mbox{.left}$ go to \algref{B9}}.
+    \Et[\indexin_j]\prop{left}$ go to \algref{B9}.}
 
 \algstep{B7.}{Insert edge.}{Set
-    $u \leftarrow \Et[\indexin_j]\mbox{.child}$,
-    $v \leftarrow \Et[\indexin_j]\mbox{.parent}$ and $j \leftarrow j + 1$.
+    $u \leftarrow \Et[\indexin_j]\prop{child}$,
+    $v \leftarrow \Et[\indexin_j]\prop{parent}$ and $j \leftarrow j + 1$.
     Then set $\pi_u \leftarrow v$,
-    $\beta_u \leftarrow \Nt[v]\mbox{.time} - \Nt[u]\mbox{.time}$ and
+    $\beta_u \leftarrow \Nt[v]\prop{time} - \Nt[u]\prop{time}$ and
     $s \leftarrow s + \beta_u F_u$.
@@
 \algstep{B9.}{Tree loop tail.}{Set $t_r \leftarrow L$.
-    If $j < |\Et|$ set $t_r \leftarrow \min(t_r, \Et[\indexin_j]\mbox{.left})$.
-    Then, if $k < |\Et|$ set $t_r \leftarrow \min(t_r, \Et[\indexout_k]\mbox{.right})$
+    If $j < |\Et|$ set $t_r \leftarrow \min(t_r, \Et[\indexin_j]\prop{left})$.
+    Then, if $k < |\Et|$ set $t_r \leftarrow \min(t_r, \Et[\indexout_k]\prop{right})$
     Finally, set $\sigma \leftarrow (t_r - t_l)s$,  $t_l \leftarrow t_r$
     and go to \algref{B2}.
+    \plr{I think you mean to add to $\sigma$, not replace it?}
 }

### 2019-09-02 tskit_stat_benchmarks/benchmarks_with_data_transfer.py
add benchmarks

@@ -1,0 +1,65 @@
+import msprime
+import libsequence
+import allel
+import timeit
+import numpy as np
+
+NE = 1e4
+#RECRATE = 1e-8
+RHO=1e4
+RECRATE=1e3/(4*NE)
+NU = [RECRATE]
+
+def make_tree_sequence(nsam):
+    """
+    Make a tree seq w/no mutations
+    """
+    return msprime.simulate(nsam, Ne=NE, recombination_rate=RECRATE)
+
+
+def mutate_tree_sequence(ts, Nu):
+    """ 
+    Copy a tree seq and mutate it, return
+    mutated copy
+    """
+    tscopy = ts.dump_tables().tree_sequence()
+    return msprime.mutate(tscopy, Nu)
+
+
+def tskit_tajd(ts):
+    return ts.Tajimas_D([[i for i in ts.samples()]])
+
+
+def libseq_tajd_via_genotype_matrix(ts, pos):
+    gm = ts.genotype_matrix()
+    vm = libsequence.VariantMatrix(gm, pos)
+    ac = vm.count_alleles()
+    return libsequence.tajd(ac)
+
+
+def allel_tajd(ts):
+    gm = ts.genotype_matrix()
+    ha = allel.HaplotypeArray(gm)
+    ac = ha.count_alleles()
+    return allel.tajima_d(ac)
+
+
+print("toolkit nsam nmutations Nu nbytes seconds")
+for nsam in np.logspace(np.log10(25), np.log10(1e6), 25).astype(np.int32):
+    ts = make_tree_sequence(nsam)
+    for Nu in NU:  # Skip no mutation as that is obviously dumb
+        tscopy = mutate_tree_sequence(ts, Nu)
+        # Conveniently, we can get the exact size
+        # of the genotype matrix w/o ever calculating it
+        nbytes = tscopy.num_sites * tscopy.num_samples
+        timer = timeit.Timer("tskit_tajd(tscopy)", globals=globals())
+        res = min(timer.repeat(repeat=5, number=1))
+        p = tscopy.tables.sites.position
+        print(f"tskit {nsam} {tscopy.num_mutations} {Nu} {nbytes} {res}")
+        timer = timeit.Timer(
+            "libseq_tajd_via_genotype_matrix(tscopy,p)", globals=globals())
+        res = min(timer.repeat(repeat=5, number=1))
+        print(f"libseq {nsam} {tscopy.num_mutations} {Nu} {nbytes} {res}")
[…]

### 2019-09-02 tskit_stat_benchmarks/plot_benchmarks.py

@@ -1,0 +1,18 @@
+import pandas as pd
+import matplotlib.pyplot as plt
+
+for infile in ["benchmarks_including_copy.txt", "benchmarks_without_copy.txt"]:
+    b = pd.read_csv(infile, sep=' ')
+    b = b.round({'Nu': 4})
+    b['seconds_snp'] = b.seconds/b.nmutations
+    grp = b.groupby(['toolkit'])
+    fig = plt.figure()
+    for n, g in grp:
+        plt.loglog(g.nsam, g.seconds_snp, 'o-', label=n)
+    plt.xlabel("Sample size (haploid)")
+    plt.ylabel("Run time per variant (seconds)")
+    of = infile.replace("txt", "pdf")
+    plt.legend(loc='upper left', fontsize='x-small')
+    plt.tight_layout()
+    plt.savefig(of)
+    plt.clf()

### 2019-09-02 tskit_stat_benchmarks/plot_benchmarks_new.py

@@ -1,0 +1,32 @@
+import pandas as pd
+import seaborn as sns
+import matplotlib.pyplot as plt
+
+yfactors=[0.4,0.3]
+for infile,yf in zip(["benchmarks_including_copy.txt", "benchmarks_without_copy.txt"], yfactors):
+    print(infile, yf)
+    b = pd.read_csv(infile, sep=' ')
+    b = b.round({'Nu':4})
+    b['seconds_snp'] = b.seconds/b.nmutations
+    grp = b.groupby(['toolkit'])
+    fig = plt.figure()
+    for n,g in grp:
+        plt.loglog(g.nsam, g.seconds_snp,'o-',label=n)
+    # p = sns.scatterplot(x='nsam', y='seconds_snp',
+    #p = sns.scatterplot(x='nsam', y='seconds_snp',
+    #                    hue='toolkit', style='Nu',
+    #                    data=b, alpha= 0.5)
+    # plot = p.get_figure()
+    # #plt.ylim(0.0, 0.0006)
+    # plt.xscale('log')
+    # ylim=plt.gca().get_ylim()
+    # ylim=(ylim[0]*1e-1,ylim[1]*yf)
+    # plt.ylim(*ylim)
+    plt.xlabel("Sample size (haploid)")
+    plt.ylabel("Run time per variant (seconds)")
+    #plt.legend(loc=2,bbox_to_anchor=(1,1))
+    of = infile.replace("txt","pdf")
+    plt.legend(loc='upper left',fontsize='x-small')
+    plt.tight_layout()
+    plt.savefig(of)
+    plt.clf()

### 2019-09-03 treestats_paper.tex
Tweaked plot & added to paper.

@@
 \caption{
-\label{fig:algorithm_example}
 Illustration of the internal state for algorithm B for $\iw = \bone_S$ and
@@
 Immediately before we remove the edge joining 6 to 7 and (B) immediately after.
-See the text for descriptions of the arrays encoding the state.}
+See the text for descriptions of the arrays encoding the state.
+\label{fig:algorithm_example}}
 \end{figure}
@@
 compare speed to scikit-allel
+
+\begin{figure}
+    \centering
+    \includegraphics{tskit_stat_benchmarks/benchmarks_without_copy_longer_genome.pdf}
+    \caption{Comparison of time required to compute site statistics
+        between matrix and tree based methods. [Details]
+        \label{fig:stats_performance}
+    }
+\end{figure}
 

### 2019-09-03 treestats_paper.tex
Algorithm description.

@@
 \newcommand{\prop}[1]{.\mbox{\texttt{#1}}} % property of a thing
-\newcommand{\indexin}[0]{\ensuremath{\mathcal{I}}}
-\newcommand{\indexout}[0]{\ensuremath{\mathcal{O}}}
-
+% Not sure about this notation. Want to use i and o, but using normal
+% math fonts would make things less clear, I think.
+\newcommand{\indexin}[0]{\ensuremath{\mathbf{i}}}
+\newcommand{\indexout}[0]{\ensuremath{\mathbf{o}}}
 \newcommand{\plr}[1]{{\color{blue}\textbf{plr:} \it #1}}
@@
 \subsection*{Algorithms}
-In this section we provide describe and analyse the algorithm used
-to compute branch length statistics. This is a generalation
+In this section we describe and analyse the algorithm used
+to compute branch length statistics. This is a generalization
 of the algorithm used to maintain the number of samples from a
-given set that derive from each node in a tree sequence.
-\cite[Algorithm L]{kelleher2016efficient}.
-
-\jk{Plan for section}
-\begin{itemize}
-\item First paragraph is a general outline of the strategy.
-\item Explanation of notation and the index vectors (we should probably define
-these again, in the interest of completeness. I'm not sure the definition
-in the msprime paper really worked)
-\item Algorithm listing.
-\item Next para is an explanation of the algorithm in words, referring
-to the figure for what we mean by adding and removing edges.
-\item Next para is an analysis of the algorithm. We note here that
-we propagate stuff all the way up the tree close to optimally, because of the
-ordering of edge insertions and removals. For example, we always disconnect
-a subtree from the main tree first, meaning that the only time we go all the
-way to the root is the first disconnection.
-Discuss the figure perf figure.
-\end{itemize}
-
-\plr{Question: Figures 1, 2 and 7 show the same sort of thing
-    (fig 1 is a site stat, aiming to give the general overview;
-    figs 2 and 7 are branch stats). Should they be more similar?
-    Do we need all of these?}
-
-\plr{right now things like $\Et\mbox{.right}$ read like a funnily terminated sentence.
-     I've changed these with a macro, feel free to adjust.}
-
-\clearpage
+given set that derive from each node in a tree
+sequence \cite[Algorithm L]{kelleher2016efficient}. The central
+problem which is shared across site, branch and node statistics
+is to efficiently maintain the ``state'' $x_u$ (where $x_u$ is the
+sum of the weights for all samples descending
+from, and including, node $u$; see Figure~\ref{fig:divergence_diagram}).
+As we move along the sequence, the trees change when branches are inserted
+and removed. By carefully ordering these insertions and removals, we
+ensure that this state is always correctly maintained and that we
+only need to make small adjustments for each tree. Computing the statistics
+is then a straightforward matter of applying the summary function $f$
+to the node states and aggregating the results appropiately for the
+particular statistics mode and windowing options.
+
+Formally, we define a tree sequence topology via a set of
+tables~\citep{kelleher2018efficient}. In a tree sequence, a node
+describes a specific ancestral haplotype, and details about these
+ancestors are stored in a node table, $\Nt$. The row $\Nt[u]$
+contains all information about the node $u$, and rows
+are indexed from zero such that $0 \leq u < |\Nt|$.
+In particular, the birth time for $u$ is given by
+$\Nt[u]\prop{time}$. Information about how nodes relate to each
+other along the genome is encoded in an edge table, $\Et$.
+For a given row $j$, $\Et[j]\prop{child}$ and $\Et[j]\prop{parent}$
+define a parent-child relationship between two nodes in a set of
+contiguous trees along the genome. $\Et[j]\prop{left}$ then denotes
+the left-most (inclusive) and $\Et[j]\prop{right}$ the right-most
+(exclusive) genome coordinates over which this branch exists.
+The order in which edges are inserted and removed is determined
+by the ``index vectors'' $\indexin$ and $\indexout$~\citep{kelleher2016efficient},
+which are defined as follows. \jk{TODO: succinct definition of these}.
+As we move along the tree sequence, the topology is recorded in the
+parent vector $\pi$, such that $\pi_u$ is the parent of node $u$
+and $\pi_u = -1$ if $u$ is a root. Similarly, we maintain the
+branch length $\beta$ such that
+$\beta_u = \Nt[\pi_u]\prop{time} - \Nt[u]\prop{time}$ and
+$\beta_u = 0$ if $u$ is a root. As evaluating the summary function
+$f$ is an expensive operation, we also maintain the vector $F$
+such that $F_u = f(x_u)$. For notational simplicity here we assume
+that weights and summary functions are scalars, but the extension to
+vectors is trivial. We also assume that we are computing the statistic
+for across the full span of the tree sequence; the extension to multiple
+windows is routine.
+
 \begin{taocpalg}{B}{General branch statistics}
@@
 that the index vectors $\indexin$ and $\indexout$ have been precomputed.
-\plr{I assume you plan to say in words what $\beta$, $x$, $F$, etc are}
 }
@@
 
-\algstep{B2.}{Terminate.}{If $j \geq |\Et|$ or $t_l \geq L$
-    set $\sigma \leftarrow \sigma / L$ and terminate.
-    \plr{perhaps say "return $\sigma/L$" to make it more clear what the output is?}
+\algstep{B2.}{Terminate.}{If $j \geq |\Et|$ or $t_l \geq L$,
+    return $\sigma / L$ and terminate.
 }
@@
     $s \leftarrow s - \beta_v F_v$, $x_v \leftarrow x_v - x_u$,
-    $F_v \leftarrow f(x_v)$, $s \leftarrow s - \beta_v F_v$
+    $F_v \leftarrow f(x_v)$, $s \leftarrow s + \beta_v F_v$
     and $v \leftarrow \pi_v$. Afterwards, go to \algref{B3}.
@@
     $s \leftarrow s - \beta_v F_v$, $x_v \leftarrow x_v + x_u$,
-    $F_v \leftarrow f(x_v)$, $s \leftarrow s - \beta_v F_v$
+    $F_v \leftarrow f(x_v)$, $s \leftarrow s + \beta_v F_v$
     and $v \leftarrow \pi_v$. Afterwards, go to \algref{B6}.
@@
     Then, if $k < |\Et|$ set $t_r \leftarrow \min(t_r, \Et[\indexout_k]\prop{right})$
-    Finally, set $\sigma \leftarrow (t_r - t_l)s$,  $t_l \leftarrow t_r$
+    Finally, set $\sigma \leftarrow \sigma + (t_r - t_l)s$,  $t_l \leftarrow t_r$
     and go to \algref{B2}.
-    \plr{I think you mean to add to $\sigma$, not replace it?}
 }
@@
 \end{taocpalg}
+
+Algorithm \algref{B} begins by setting the initial state for $\pi$, $\beta$,
+$x$ and $F$ for each node in the tree sequence, and then sets the values
+of $x_u$ and $F_u$ for each of the samples in $S$. Afterwards, we set
+our ``running sum'' $s$ and output statistic $\sigma$ to zero, along
+with the ``tree left'' variable $t_l$. The $j$ and $k$ variables are
+used to keep track of our position in the edge insertion and edge
+removal indexes, respectively. After completing initialisation in
+\algref{B1}, we enter the main tree loop in \algref{B2} which is
+run once for each tree in the sequence. As we are processing each
+tree, we keep track of the edges that need to be processed with
+the $t_l$ variable, which stores the left-most genome coordinate
+in the current tree. The first thing we do for a new tree is to
+remove any edges that were in the previous tree are not in the
+current tree. These must be the edges in which the right coordinate
+is equal to $t_l$, and so \algref{B3} loops over these edges using
+the edge removal index $\indexout$. Step \algref{B4} then removes
+the branch $u \mapsto v$ corresponding to a single edge, by
+subtracting its contribution to the running sum $s$, setting the parent of
+$u$ to $-1$ and its branch length to $0$. As shown in
+Figure~\ref{fig:algorithm_example}B, removing an edge will result
+in the subtree rooted at $u$ become disconnected from the rest of the
+tree. Step \algref{B5} ensures that the state of the rest of the tree
+is correctly maintained by propagating the loss of the state $x_u$
+up the tree from $v$. Thus, for each direct ancestor $v$ that was a
+direct ancestor of $u$ in the previous tree, we first subtract
+its contribution from the running sum $s$ and remove the contribution
+of $x_u$ from its state. Since the state of $v$ has changed, we must
+recompute the value of our summary cache $F_v$, and then finally
+add the new contribution of $v$ to $s$. (For example, note the changes in $x$
+and $F$ for nodes 7 and 8 in Figure~\ref{fig:algorithm_example}B.)
+
+Once all of the edges with right coordinate equal to $t_l$ have been removed in
+steps \algref{B3--B5}, we then insert edges that start in the current tree
+(i.e., with left coordinate $t_l$) in steps \algref{B6--B8}. We update the
+state to account for the new branch $u \mapsto v$ by setting the parent of
+$u$ to $v$, computing the branch length of $u$ and adding the new contribution
+of $u$ to $s$ in \algref{B7}. Then, as adding a new edge can connect the
+subtree rooted at $u$ to the larger tree at $v$, we propagate the gain of $x_u$
+up the tree in step \algref{B8} (note the symmetry with step \algref{B5}).
+Finally, once we have remove and inserted all of the relevant edges, we are
+ready to add the contribution of the current tree to the overall statistic,
+$\sigma$ and move on to the next tree in step \algref{B9}. To do this, we first
+compute the right-hand coordinate of the tree $t_r$ (a process slightly
+complicated by the possible presence of ``gaps'' in the tree sequence, spanned
+by no edges). Then, we add the running sum $s$ weighted by the span of the
+current tree $t_r - t_l$ to $\sigma$, and return to \algref{B2} to process the
+next tree. If we have reached the end of the sequence, we then divide $\sigma$
+by the sequence length to renormalise and exit.
+
+To analyse Algorithm~\algref{B} we will assume that the tree sequence has
+been simulated under the standard coalescent with a sample of size
+$n$ and population scaled recombination rate $\rho$. Under these conditions,
+we know that there are $O(n + \rho \ln n)$ edges in the tree
+sequence~\cite{kelleher2016efficient}. Clearly, each edge is examined
+exactly once in steps \algref{B6} and \algref{B7} in order to create
+the trees, and at most once in steps \algref{B3} and \algref{B4} (we do
+not remove the edges for the final tree). Additionally, each edge that
+we insert or remove may incur the cost of traversing up the tree as
+far as the root in steps \algref{B5} and \algref{B8}.
+As coalescent genealogies are asymptotically
+balanced~\citep{li2013coalescent}, the expected number of nodes
+on the path to root is $\log_2 n$. Therefore, the overall running time
+of Algorithms~\algref{B} is
+\[
+O(n \log_2 n + \rho \log_2 n \ln n)
+\]
+\jk{Slightly iffy on what this should be in $O$ terms. I'm pretty
+sure that the total number of edges of log base $e$ and the
+expected height is log base 2. Not sure what the asymptotics
+say here.}
+The asymptotic analysis here assumes that we traverse upwards to root
+for each edge, but this is not the case. Edges are removed in
+oldest-first order, guaranteeing that the deepest node any subtree being
+modified is processed first. Therefore, this subtree will be disconnected from the
+main tree, ensuring that we traverse upwards all the way to root once. Similarly,
+edges are inserted in youngest-first order, ensuring that subtrees are only
+reconnected to the main tree when they are fully built.
 
@@
 See the text for descriptions of the arrays encoding the state.
+\jk{TODO: the value of $s$ would be helpful here too.}
 \label{fig:algorithm_example}}

### 2019-09-03 treestats_paper.tex
initial text on the timings

@@
 \newcommand{\jk}[1]{{\color{red}\textbf{jk:} \it #1}}
+\newcommand{\krt}[1]{{\color{green}\textbf{krt:} \it #1}}
 
@@
 
-compare speed to scikit-allel
+\krt{need to sort out the big-O issue here}
+We used coalescent simulations to compare the performance of calculating Tajima's \citeyearpar{Tajima1989-de} $D$
+statistic from tree sequences to calculating it from integer matrices containing genotypes at all variable positions.
+\autoref{fig:stats_performance} shows that the tree sequence methods implemented in \texttt{tskit} process variant data
+three orders of magnitude faster than matrix-based approaches once the sample size is above $n \approx 500$ haploids.
+The expected number of mutations for each replicate is $10^4 \times \sum_{i=1}^{n-1}\frac{1}{i}$, and thus ranges from
+$\approx 32,000$ to $\approx 138,000$ \citep{Watterson1975-ej}.
+\autoref{fig:stats_performance} only considers the time required to calculate the statistic. It does not include the time
+required to build the genotype matrix, nor the memory required to store it.  For the largest samples size of $n=10^6$,
+the matrix is approximately 200 gigabytes, and thus not practical on many systems.
+
+\autoref{fig:stats_performance} shows that \texttt{tskit} requires more time per variant once the sample size is larger
+than $n \approx 10^4$.  For large sample sizes, the $O(n)$ complexity of building the first tree dominates the run time.
+This initial cost is amortized in longer genomes where more recombination leads to more trees.
 
@@
     \caption{Comparison of time required to compute site statistics
-        between matrix and tree based methods. [Details]
+        between matrix and tree based methods. For each sample size, a single replicate
+        was obtained using \texttt{msprime} with scaled neutral mutation rate $\theta = 4N_e\mu$ and
+        scaled recombination rate $\rho = 4Nr$ both equal to $10^4$. For each replicate, Tajima's $D$
+        \citep{Tajima1989-de} statistic was calculated using \texttt{tskit}, \texttt{libsequence}
+        \citep{Thornton2003-wj}, and \texttt{scikit-allel} \krt{need cite for this}.
+        The solid green line shows the result of fitting the timing data to a model based on the expected complexity of the 
+        algorithm used by \texttt{tskit}. \krt{need to revisit the correctness of this}
         \label{fig:stats_performance}

### 2019-09-04 treestats_paper.tex
O( ) adjustments

--- 2019-09-03 treestats_paper.tex
@@
 been simulated under the standard coalescent with a sample of size
-$n$ and population scaled recombination rate $\rho$. Under these conditions,
-we know that there are $O(n + \rho \ln n)$ edges in the tree
-sequence~\cite{kelleher2016efficient}. Clearly, each edge is examined
+$n$ and population scaled recombination rate $\rho = 4 N_e r$,
+where $r$ is the mean number of recombinations per chromosome per unit of time.
+Under these conditions, there are $O(n + \rho \ln n)$ edges in the tree sequence
+\citep{kelleher2016efficient}.  Clearly, each edge is examined
 exactly once in steps \algref{B6} and \algref{B7} in order to create
@@
 As coalescent genealogies are asymptotically
-balanced~\citep{li2013coalescent}, the expected number of nodes
+balanced \citep{li2013coalescent}, the expected number of nodes
 on the path to root is $\log_2 n$. Therefore, the overall running time
@@
 \[
-O(n \log_2 n + \rho \log_2 n \ln n)
+O(n + \rho \ln n) \log_2 n
 \]
-\jk{Slightly iffy on what this should be in $O$ terms. I'm pretty
-sure that the total number of edges of log base $e$ and the
-expected height is log base 2. Not sure what the asymptotics
-say here.}
+Computing a site statistic will have the same complexity,
+as the same internal state is maintained,
+and the number of segregating sites is proportional to the number of edges
+(with constant equal to the ratio of mutation to recombination rate).
 The asymptotic analysis here assumes that we traverse upwards to root

### 2019-09-04 tskit_stat_benchmarks/plot_benchmarks_longer_genome.py

@@ -4,17 +4,13 @@ import matplotlib.pyplot as plt
 import scipy.optimize as optimize
 
 
-def tskit_model(n, a, b, c):
+def tskit_model(logn, a, b, c):
     # The model is based on the proof that we have O(n + rho log n) edges
-    # and we look at each of these twice as we go through the trees.
-
-    # TODO not quite sure about this, need to think about it again.
-    # Excellent fit though.
+    # and we look at each of these twice as we go through the trees
+    # and that the number of mutations is also proportional to rho log n.
+    n = np.exp(logn)
     rho = 0.025  # Value from the file
-    z = rho * np.log(n)
-    # TODO: did KT mess this up?
-    return a*n + b*(z) * np.log(n) + c
-    # return a * (n+z) + b * np.log(n) + c
+    return np.log((a * n + b * rho * logn) * logn + c)
 
 
 symbols = {"tskit": "o", "allel": "^", "libseq": "s"}
@@ -30,8 +26,8 @@ for n, g in grp:
     if n == 'tskit':
         line, = plt.loglog(g.nsam, g.seconds_snp, symbols[n], label=n)
         fit_params, _ = optimize.curve_fit(
-            tskit_model, g.nsam, g.seconds_snp)
-        fit = tskit_model(g.nsam, *fit_params)
+            tskit_model, np.log(g.nsam), np.log(g.seconds))
+        fit = np.exp(tskit_model(np.log(g.nsam), *fit_params)) / g.nmutations
         plt.loglog(g.nsam, fit, color=line.get_color(),
                    label='__nolegend__')
     else:

### 2019-09-04 treestats_paper.tex
N -> Ne in \rho

--- 2019-09-03 treestats_paper.tex
@@
         was obtained using \texttt{msprime} with scaled neutral mutation rate $\theta = 4N_e\mu$ and
-        scaled recombination rate $\rho = 4Nr$ both equal to $10^4$. For each replicate, Tajima's $D$
+        scaled recombination rate $\rho = 4N_er$ both equal to $10^4$. For each replicate, Tajima's $D$
         \citep{Tajima1989-de} statistic was calculated using \texttt{tskit}, \texttt{libsequence}

### 2019-09-04 treestats_paper.tex
Describe the Ne value in fig caption

--- 2019-09-03 treestats_paper.tex
@@
         was obtained using \texttt{msprime} with scaled neutral mutation rate $\theta = 4N_e\mu$ and
-        scaled recombination rate $\rho = 4Nr$ both equal to $10^4$. For each replicate, Tajima's $D$
+        scaled recombination rate $\rho = 4N_er$ both equal to $10^4$ and $N_e = 10^4$. For each replicate, Tajima's $D$
         \citep{Tajima1989-de} statistic was calculated using \texttt{tskit}, \texttt{libsequence}

### 2019-09-04 treestats_paper.tex
Add crude estimate of tree seq RAM requirement

@@
 required to build the genotype matrix, nor the memory required to store it.  For the largest samples size of $n=10^6$,
-the matrix is approximately 200 gigabytes, and thus not practical on many systems.
+the matrix is approximately 200 gigabytes, and thus not practical on many systems.  The largest tree sequence simulated
+required less than one gigabyte of memory.
 

### 2019-09-06 treestats_paper.tex
Comment re: missing refs
Mention speed+RAM in abstract
Comment on 1st sentence

@@
 % \emph{Running head:} Genealogies and genomes
+
+\krt{several refs seem missing}
 
@@
 This formulation also makes it easy to see what statistic of the underlying genealogical trees
-each sequence-based statistic is estimating.
-
-\plr{add bits for speedup; data}
+each sequence-based statistic is estimating.  By using simulations to evaluate performance, we 
+show that calculating statistics from tree sequences is several orders of magnitude more efficient than
+methods traversing the entire data matrix in terms of both run time and memory requirements.
+
+\plr{add bits for data}
 
@@
 
+\krt{Why? I think this is an odd way of putting it.  The challenge back then was
+    genotyping.  Once that was done, summarizing the relationships was done on a 
+    calculator b/c the data sets were so small.}
 It was once a major undertaking to estimate a single summary

### 2019-09-06 treestats_paper.tex
Remove Indeed,

@@
 or iteratively refining a single computation.
-Indeed, many sorts of modern inference machinery, such as deep learning,
+Many sorts of modern inference machinery, such as deep learning,
 depend on being able to compute enough statistics that this becomes a serious bottleneck in practice.

### 2019-09-06 treestats_paper.tex
Attempt at segue sentence b/w paragraphs.
Remove superlatives.

@@
 depend on being able to compute enough statistics that this becomes a serious bottleneck in practice.
+To a large extent, this computational bottleneck is because existing approaches for computing statistics
+typically do not incorporate the relatedness amongst samples.
 
@@
 to store genome sequence and the associated genealogical relationships extremely compactly,
-as well as to compute several population genetics statistics highly efficiently.
+as well as to compute several population genetics statistics efficiently.
 This paper generalizes the strategy used by \citep{kelleher2016efficient} for statistics computation

### 2019-09-06 treestats_paper.tex
Say why trees only partially known
Add comment on older refs.

@@
 However, population genetics has the task of summarizing relationships across \emph{many} trees,
-each only partially known.
+each only partially known because mutations are sparse in the data.
 The basic relationship between tree shape and summaries of genetic variation
@@
 under the infinite-sites model of mutation \citep{ralph2019empirical}.
+\krt{The idea is much older than this: Taiima 83, 89(?), Fu 95 are all relevant here.  The latter
+derives a lot of properties of the expectations of "node weights" under Kingman's coalescent, including
+the polarised vs non- cases.}
 

### 2019-09-06 treestats_paper.tex
h(p) should be f(p) as it is written later in Example 2?

@@
 and suppose that for some function $f()$, we want to compute $\sum_p h(p) f(p)$.
-(For instance, mean genetic diversity is computed with $h(p) = p (1-p)$.)
+(For instance, mean genetic diversity is computed with $f(p) = p (1-p)$.)
 Since this is a simple sum over sites,

### 2019-09-06 treestats_paper.tex
Fix fragmented sentence.

@@
 \end{align*}
-(Note that since the ancestral allele is included, this is a function of the
-is invariant to relabeling of alleles.)
+(Note that since the ancestral allele is included, this function is invariant to relabeling of alleles.)
 If we know where each mutation has occurred on the genealogy,

### 2019-09-06 treestats_paper.tex
Introduce constraint on sample frequencies earlier.
Comment on fig 1
Question

@@
 occur below that mutation in the tree (and accounting for those that subsequently mutate).
+Here, we only consider statistics whose values are invariant to monomorphic sites. Thus, for a sample of $n$
+genomes, $0 < p \leq 1-1/n$.
 
@@
 % as we iterate through the tree sequence.
+\krt{Is this true?  If so, why isn't the current Tajima's D implentation done in one pass through the tree sequence?}
 In practice we allow summary functions to be vector-valued (so as to compute many statistics at once)
@@
         the total number of sites.
+        \krt{This is a nice fig, but the $2 \times$ and the $/10 sites$ parts may be easy to miss?}
         \label{fig:divergence_diagram}

### 2019-09-06 treestats_paper.tex
Total weight in words.

@@
 The requirement that $f(0) = f(\tiw) = 0$ is for consistency,
-so that statistics do not depend on portions of the tree not ancestral to any of the samples.
+so that statistics do not depend on portions of the tree either not ancestral to any of the samples or ancestral to all
+of them.
 % Below, in section XXX, we describe how to efficiently

### 2019-09-06 treestats_paper.tex
why trees differ

@@
 can cause the precise set of samples that lie below that edge to differ across
-trees.
+trees due to recombination.
 

### 2019-09-06 treestats_paper.tex
Attempt to streamline def'n of polarised node stat.

@@
 \paragraph{Polarisation}
-In this definition we have associated with each node the summary of both the weight below the node
-\emph{and} the weight not below the node.
-Would it not be more natural to only include the first term, $f(\nw_{T_k}(u))$?
-The reason for this definition is that it does not depend on the location of the root of the tree,
-which in practice may be unknown.
-Statistics of genotype sequence defined below in an analogous way
-will not depend on knowledge of which allele is ancestral, i.e., on \emph{polarisation} of alleles.
-For this reason, the node statistic above is the \emph{unpolarised} version,
-while a \emph{polarised} node statistic is defined without the second term:
+The node statistic defined above is appropriate for statistics such as nucleotide diversity, which is a function
+of all alleles at given variable site.  For other statistics, one is only concerned with the derived state(s)
+at a site, and we may defined a \emph{polarised} node statistic without the second term from our previous definition:
+
 \begin{align}

### 2019-09-06 treestats_paper.tex
Rework text around node stat examples.

@@
 
-Now we have defined a general, abstract, framework.
-Is it good for anything?
-Next we give some examples.
+Having defined a general framework, we may now give concrete examples of functions related to common summary statistics.
 
 \begin{example}[Nucleotide diversity] \label{ex:site_diversity}
-    Suppose we want to compute the mean density of nucleotide differences
-    in a group $S$ of $n$ sequences.
-    To do this,
+    The nucleotide diversity of a sample is the average number of differences between all haplotypes, which is
+    equivalent to the sum of heterozygosity at all positions.
+    To calculate this statistic,
     % let $\iw_u = 1$ for $u \in S$ and $\iw_v = 0$ otherwise,
@@
 \end{example}
+
+This example corresponds to the calculations shown in \autoref{fig:divergence_diagram}.
 

### 2019-09-06 treestats_paper.tex
note some possible terminology confusion.
Comment on alternate approach for sites with > 2 states.

@@
 % c(cor(g,z), sum(g*z)/sqrt(n*(n-1)*p*(1-p)))
-
+\krt{There is a possible point of confusion here.  So far, weight has been used to describe a property of nodes, and
+    then we have a function that operates on those weights.  In this example, the output of that function is also called
+a weight.  We may prefer to use a term like ``value" here?}
 \begin{example}[Patterson's $f_4$] \label{ex:site_f4}
@@
 
+\krt{There is another approach, introduced by Molly Prz.  I'll see if I can dig up a description.  The approach
+described here is preferable, though.}
 \paragraph{Polyallelic sites}

### 2019-09-06 treestats_paper.tex
comment on fig 2
Document annotations in fig 2

@@
     \textbf{(a)} Each mutation is given weight equal to $(p_A - p_B)(p_C - p_D)$,
-    where $p_X$ is the proportion of the samples inheriting that mutation in set $X$
+    where $p_X$ is the proportion of the samples inheriting that mutation (colored stars) in set $X$
     (shown in labels at the bottom).
@@
     \textbf{(b)} Each branch is assigned weight equal to the weight that would be assigned
-    to any mutation falling in it, and then the branch $f_4$ statistic
+    to any mutation falling in it (colored lined), and then the branch $f_4$ statistic
     is the sum over these weights, multiplied by the length of the branch
@@
     }
+    \krt{This figure is not as helpful as fig 1. It takes a lot of fiddling to work out the values in each panel. Is it
+    possible to fill it out as in fig 1, or is that just too much space?}
 \end{figure}

### 2019-09-06 treestats_paper.tex
does the bit on variance need further exposition?
comment re: results seemingly w/o a figure?

@@
 
+\krt{will this section go further?  It just kind of ends...}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 have per-capita migration rates of $4/N$ to each other.
+\krt{It is not clear what figure this bit of text refers to?  Is this out of date?}
 In this situation, a significantly negative $f_4(1,2;3,4)$ is expected,

### 2019-09-06 treestats_paper.tex
flag unclear reference to method

@@
 All calculations were done only using the portions of the chromosome
+\krt{unclear what this means--ref/citation is needed.}
 passing the ``strict'' mask for sequencing accessibility.

### 2019-09-06 treestats_paper.tex
flag interpretation that needs clarification

@@
 with Relate showing a significantly tighter agreement than GEVA.
+\krt{Does this make sense?  If geva is playing with node times based on ts inferred via tsinfer,
+    then a difference b/w it and relate is due to the difference in estimating the times, and not the
+details of the underlying ts, maybe?}
 Does this mean that Relate is doing a better job at inferring the tree sequence than GEVA?

### 2019-09-06 treestats_paper.tex
Flag section that needs some more work.

@@
 
+\krt{Hmmm...not sure about this paragraph.  Under strict BGS, some of what is stated here doesn't hold.  The tree will
+    still be neutral in shape, just shorter overall.  For the case of many weakly-deleterious segregating mutations, the
+    overall shape of the tree may not be that different from neutral expectations (Neuhaueser/Krone and the later
+Przeworksi paper, all in Genetics IIRC?) It also seems to me that the idea of what branch-vs-site should look like under
+these processes vs what they should look like in a ts inferred from data where the true model is these processes are
+getting conflated.}
 In practice, we do not expect branch and site diversity to agree exactly,

### 2019-09-09 treestats_paper.tex
Remove krt comment

@@
 % as we iterate through the tree sequence.
-\krt{Is this true?  If so, why isn't the current Tajima's D implentation done in one pass through the tree sequence?}
-In practice we allow summary functions to be vector-valued (so as to compute many statistics at once)
-but this will not feature in our examples.
-Next, we will describe how to use these subtree weights,
-passed through a ``summary function'' in various ways,
-to summarize genetic variation and tree shapes across the tree sequence.
 

### 2019-09-09 treestats_paper.tex
Clarifty geva vs relate and remove sentence starting w/unclarified This.

@@
 with Relate showing a significantly tighter agreement than GEVA.
-\krt{Does this make sense?  If geva is playing with node times based on ts inferred via tsinfer,
-    then a difference b/w it and relate is due to the difference in estimating the times, and not the
-details of the underlying ts, maybe?}
-Does this mean that Relate is doing a better job at inferring the tree sequence than GEVA?
-This is highly suggestive, but not entirely clear.
+Does this mean that Relate is doing a better job at inferring the tree sequence than \texttt{tsinfer} plus GEVA?
+The results presented here are highly suggestive, but not entirely clear.
 

### 2019-09-09 treestats_paper.tex
what is a branch stat from real data?

@@
 
+\krt{Is there a circularity here?  What does a branch statistic from real data mean
+when the branch lengths must be inferred somehow from the mutations? If this is not circular,the 
+reasons why need to be made clearer.}
+
 We do not know the true genealogies underlying real data,
@@
 with Relate showing a significantly tighter agreement than GEVA.
-Does this mean that Relate is doing a better job at inferring the tree sequence than \texttt{tsinfer} plus GEVA?
+Does this mean that Relate is doing a better job at inferring the tree sequence than \texttt{tsinfer} pluse GEVA?
 The results presented here are highly suggestive, but not entirely clear.
@@
         in 1Mb windows along human chromosome 20 of the 1000 Genomes Project,
+        \krt{the only appearances of "dual branch" are in figure legends.  need to define somewhere}
         and the dual branch statistic as calculated from tree sequences inferred by
@@
 
-\krt{Hmmm...not sure about this paragraph.  Under strict BGS, some of what is stated here doesn't hold.  The tree will
-    still be neutral in shape, just shorter overall.  For the case of many weakly-deleterious segregating mutations, the
-    overall shape of the tree may not be that different from neutral expectations (Neuhaueser/Krone and the later
-Przeworksi paper, all in Genetics IIRC?) It also seems to me that the idea of what branch-vs-site should look like under
-these processes vs what they should look like in a ts inferred from data where the true model is these processes are
-getting conflated.}
+\krt{Not sure exactly how to rewrite.}
 In practice, we do not expect branch and site diversity to agree exactly,

### 2019-09-10 treestats_paper.tex
above/below -> ancestral/descending: closes #8

@@
 we can easily find the frequency of the corresponding allele by counting how many samples
-occur below that mutation in the tree (and accounting for those that subsequently mutate).
+inherit from that mutation in the tree (and accounting for those that subsequently mutate).
 Here, we only consider statistics whose values are invariant to monomorphic sites. Thus, for a sample of $n$
@@
 For instance, we could calculate how many of a certain set $A$ of samples
-are below a particular edge in a tree
+inherit from a particular edge in a tree
 by (a) assigning each of these samples weight 1 (and other nodes weight zero), then
-(b) finding the total weight of all nodes in the subtree below that edge.
+(b) finding the total weight of all nodes in the subtree inheriting from that edge.
 It turns out to be useful to allow more general weights (including negative ones).
@@
 % Below, in section XXX, we describe how to efficiently
-% maintain a list of the weights of the subtrees below each node
+% maintain a list of the weights of the subtrees inherit from each node
 % as we iterate through the tree sequence.
@@
         and how many blue nodes ($b$) lie in that subtree.
-        \textbf{(c)} The summary function shown is applied to both the weight below
-        and the weight not below each mutation,
+        \textbf{(c)} The summary function shown is applied to both the weight of the subtree
+        and the remaining weight of each mutation,
         so for instance the mutation above the node with weights $(2,1)$
@@
 rearrangements of genealogical relationships
-can cause the precise set of samples that lie below that edge to differ across
+can cause the precise set of samples that inherit from that edge to differ across
 trees due to recombination.
@@
 associated with summary function $f()$ and sample weights $\iw$
-to be the sum of $f()$ applied to the weight of the subtree below node $u$
+to be the sum of $f()$ applied to the weight of the subtree inheriting from node $u$
 and to the remaining weight,
@@
 \end{align}
-The weight $\nw_{T_k}(u)$ is the subtree weight below $u$ in the tree $T_k$,
-and $\tiw - \nw_{T_k}(u)$ is the sum of all weights \emph{not} in the subtree below $u$ in $T_k$.
+The weight $\nw_{T_k}(u)$ is the subtree weight of node $u$ in the tree $T_k$,
+and $\tiw - \nw_{T_k}(u)$ is the sum of all weights \emph{not} in the subtree of $u$ in $T_k$.
 % i.e., the length of the segment of genome that the tree extends for.
@@
 
-We will use weights that count number of samples below each node
+We will use weights that count number of samples inheriting from each node
 sufficiently often that we give them a special name --
@@
     If $\iw$ are the indicator weights of the set $S$ of $n$ samples,
-    and then $\iw_u / n$ is the proportion of the samples in $S$ that lie below $u$.
+    and then $\iw_u / n$ is the proportion of the samples in $S$ that inherit from $u$.
     Therefore, if $f(x) = x / n$,
@@
 extend nicely to genetic variants
--- just as a node weight contains information about which samples are below the node and which are not,
+-- just as a node weight contains information about which samples inherit from the node and which are not,
 so we can summarise patterns of genetic variation by summing up weights of all samples
@@
 If there has been only one mutation at the site,
-then $\aw_j(a)$ is equal to the weight of the subtree below the mutation that produced $a$.
+then $\aw_j(a)$ is equal to the weight of the subtree inheriting from the mutation that produced $a$.
 More generally, $\aw$ is not necessarily equal to a subtree weight
@@
     let $\iw = \bone_S$,
-    so that $\nw(u)$ gives the number of nodes in $S$ below $u$,
+    so that $\nw(u)$ gives the number of nodes in $S$ inheriting from $u$,
     and define
@@
     let $\iw_j = \bone_{S_j}$,
-    so that $\nw_{j}(u)$ gives the number of nodes in $S_j$ below $u$.
+    so that $\nw_{j}(u)$ gives the number of nodes in $S_j$ inheriting from $u$.
     Then we define
@@
 \end{align}
-where $t_{u}$ is the length of the branch above node $u$ in tree $T$
-and $\nw_{T}(u)$ is the total weight of the subtree of $T$ below $u$ (as defined above).
-The term $\tiw - \nw_{T}(u)$ gives the total weight \emph{not} in the subtree of $T$ below $u$,
+where $t_{u}$ is the length of the branch ancestral to node $u$ in tree $T$
+and $\nw_{T}(u)$ is the total weight of the subtree of $T$ inheriting from $u$ (as defined above).
+The term $\tiw - \nw_{T}(u)$ gives the total weight \emph{not} in the subtree of $T$ inheriting from $u$,
 because the sum of all weights in the tree is always $\tiw$.
 The value $f(\nw(u))$ is the summary value that would be added to a site statistic
-if a single mutation occurred on the branch above $u$,
+if a single mutation occurred on the branch ancestral to $u$,
 and $f(\tiw - \nw_{T}(u))$ is the value that would be added due to its complementary allele,
@@
 showing how the $f_4$ site statistic assigns a weight to each mutation
-depending on the frequencies in each of the four sample sets below it,
+depending on its' frequency in each of the four sample sets,
 and how the corresponding branch statistic assigns the same weight to those branches.
@@
     If we take $\iw$ and $f$ exactly as in the ``Nucleotide diversity'' example above,
-    then $f(u)$ gives the probability that the branch above $u$
+    then $f(u)$ gives the probability that the branch ancestral to $u$
     lies on the path from two randomly chosen samples from $S$
@@
 sums $(p_A - p_B)(p_C - p_D)$ across branches, multiplied by span and height,
-where $p_A$ is the proportion of samples in $A$ that lie below that node, etcetera.
+where $p_A$ is the proportion of samples in $A$ that inherit from that node, etcetera.
 A branch with a large value is expected to contribute a large amount to the site $f_4$,
@@
 one would need to have an entry in the weight vector for \emph{each individual}.
-The genotype of a diploid at a site with a single mutation that occurred on the branch above node $n$
+The genotype of a diploid at a site with a single mutation that occurred on the branch ancestral to node $n$
 is determined by where that node falls relative to the diploid's two haplotypes in the tree:
@@
 $\iw_a = \delta_a$ and $\iw_b = \delta_b$
-(these weights mark nodes above $a$ and $b$ respectively)
+(these weights mark nodes ancestral to $a$ and $b$ respectively)
 and $\iw_t = \bone / n$
-(this weight gives the proportion of samples below each node).
+(this weight gives the proportion of samples inheriting from each node).
 Then, with
@@
     We could compute $Cz$ quickly if we could propagate down the tree:
-    we'd have $(Cz)_a$ equal to the sum of the weights at all nodes \emph{above} $a$. }
+    we'd have $(Cz)_a$ equal to the sum of the weights at all nodes \emph{ancestral to} $a$. }
 

### 2019-09-10 treestats_paper.tex
O() fixup: closes #12
added section on restrictions/properties: closes #17

@@
 %%%%%%%%%%%%%%%%%%
+\subsection*{General properties}
+
+The statistics defined here are all \emph{additive} along the genome:
+if we have computed a statistic in equal-size windows along a chromsome,
+then the value of the statistic for the entire chromosome is equal to
+the average of the values in those windows.
+Some commonly-used statistics (e.g., $F_{ST}$ or Tajima's $D$)
+are not additive, but are ratios of additive statistics, so can be easily computed in this way.
+Multilocus statistics that measure linkage disequilibrium or haplotype lengths
+do not fall in this category.
+We have also defined site statistics to be additive across \emph{alleles},
+to allow the definitions to apply to sites with more than two alleles.
+This decision has some nice properties, but is not the only reasonable choice.
+
+We also require that site statistics be computable from the genotype matrix
+(i.e., without knowledge of the trees),
+and, if the statistic is polarised, knowledge of the ancestral allele.
+We often would also like our statistics to be unaffected by the addition of monomorphic sites,
+which are usually omitted from genetic variation data.
+To ensure this we have required that the summary function returns zero when give both zero weight
+and the total weight: $f(0) = f(\tiw) = 0$.
+This also makes branch and node statistics unchanged by portions of the tree that do not lie on a path
+between any of the samples that have nonzero weight,
+so that in particular,
+\emph{simplification} of the tree sequence will not change the value of the statistic.
+However, in practice this requirement can be relaxed.
+
+%%%%%%%%%%%%%%%%%%
 \section*{Duality of site and branch statistics}
@@
 on the path to root is $\log_2 n$. Therefore, the overall running time
-of Algorithms~\algref{B} is
+of Algorithms~\algref{B} is of order $(n = \rho \ln n)\log_2 n$, which is
 \[
-O(n + \rho \ln n) \log_2 n
+    O\left(n \log n + \rho \log^2 n \right)
 \]

### 2019-09-11 treestats_paper.tex
introductory refs

@@
 thus providing a glimpse of the genealogical tree that relates the genomes to each other at that site.
-Summaries of genetic variation thus also aim to summarize the underlying genealogies.
+Summaries of genetic variation thus also describe the underlying genealogies.
 Here we exploit this duality to create a method that efficiently computes single-site statistics
@@
 (whose run time scales with the logarithm of the number of samples),
-and to suggest several statistics.
+and to suggest several new statistics.
 The general approach aggregates ``sample weights'' up each marginal tree,
@@
 by \emph{branch}, which gives the expected value of the site statistic
-with a large number of infinite-sites mutations,
+with a large number of mutations under the infinite-sites model,
 or by \emph{node}, which summarises the contribution of each node in the tree to the statistics.
@@
 methods traversing the entire data matrix in terms of both run time and memory requirements.
-
-\plr{add bits for data}
+We also use tree sequences inferred from the 1000 Genomes Project dataset,
+to evaluate how well this duality between site and branch statistics holds
+and to describe the temporal origin of some admixture statistics.
 
@@
 
-\krt{Why? I think this is an odd way of putting it.  The challenge back then was
-    genotyping.  Once that was done, summarizing the relationships was done on a 
-    calculator b/c the data sets were so small.}
-It was once a major undertaking to estimate a single summary
+It was once a major undertaking to collect data sufficent to estimate a single summary
 of the many genetic relationships between the individuals of a sample,
@@
 depend on being able to compute enough statistics that this becomes a serious bottleneck in practice.
-To a large extent, this computational bottleneck is because existing approaches for computing statistics
-typically do not incorporate the relatedness amongst samples.
+Bypassing this bottleneck will require new data structures appropriate for genomic data.
 
@@
 and refined for forwards-time simulations~\citep{kelleher2018efficient,haller2018tree}
-where similar efficiency gains were made. Recent
-work~\citep{kelleher2018inferring} has shown that tree sequence algorithms can
+where similar efficiency gains were made. 
+Recent work~\citep{kelleher2018inferring} has shown that tree sequence algorithms can
 also be used to massively increase the scalability of methods for inferring
-genome-wide genealogies, and shown that it is feasible to infer trees for
-millions of whole genome samples.
+genome-wide genealogies, and making it possible to infer trees for millions of whole genome samples.
 
@@
 algorithms is the way that shared structure in adjacent trees along the genome is encoded.
-As one looks across the genome, genealogies change because of the results of recombination in ancestors.
+As one looks across the genome, genealogies change at recombination breakpoints in ancestors.
 However, nearby trees tend to share a lot of common structure,
@@
 which manifest as shared \emph{edges} across many adjacent trees in the tree sequence.
-\citep{kelleher2016efficient} used this redundancy
+\citet{kelleher2016efficient} used this redundancy
 to store genome sequence and the associated genealogical relationships extremely compactly,
 as well as to compute several population genetics statistics efficiently.
-This paper generalizes the strategy used by \citep{kelleher2016efficient} for statistics computation
+This paper generalizes the strategy used by \citet{kelleher2016efficient} for statistics computation
 to a much broader class of statistics.
 
-The trees describing genealogical relationships as one moves along the genome
--- also described by the Ancestral Recombination Graph \citep[][or ARG,]{arg} --
-contain a good deal of information about the history and ongoing dynamics of a population.
-Given sufficiently many mutations, it would be possible to infer the genealogical tree
-at each point along the genome \citep{felsenstein_book},
-and each mutation gives a small piece of information about the tree at that site
-\citep{semple2003phylogenetics}.
-However, population genetics has the task of summarizing relationships across \emph{many} trees,
-each only partially known because mutations are sparse in the data.
+Naturally, the genealogical perspective on population genetics statistics has a long history.
+The sequence of trees describing genealogical relationships as one moves along the genome
+are also described by the Ancestral Recombination Graph \citep[or ARG,][]{arg},
+the basis for several inference methods \citep{argmethod}.
+These genealogical trees can be fairly precisely inferred in nonrecombining regions
+\citep{felsenstein_book},
+as for instance the mitochondria,
+and substantial theoretical work has been done in the context of phylogenetics \citep{semple2003phylogenetics}.
+% describing how each mutation gives a small piece of information about the tree at that site
+% However, population genetics has the task of summarizing relationships across \emph{many} trees,
+% each only partially known because mutations are sparse in the data.
 The basic relationship between tree shape and summaries of genetic variation
@@
 thus establishing a duality between genealogical and genetic summaries,
-under the infinite-sites model of mutation \citep{ralph2019empirical}.
-\krt{The idea is much older than this: Taiima 83, 89(?), Fu 95 are all relevant here.  The latter
-derives a lot of properties of the expectations of "node weights" under Kingman's coalescent, including
-the polarised vs non- cases.}
-
-\plr{previous work stuff here}
-Computing correlations with branches instead of SNPs is like stuff done in \citet{zollner2005coalescent} and \citet{minichiello2006mapping}, maybe.
-Cite dynamic graph algorithm something something.
+a fact used extensively in deriving properties of statistics in models of random mating
+\citep[e.g.][]{tajima1983evolutionary,tavare1984lineofdescent,fu1995statistical}.
+The fact that the duality applies regardless of the underlying demographic model was also used
+\citep[e.g.][]{gillespie1979evolutionary,hudson1983properties,slatkin1991inbreeding,mcvean2002genealogical,ralph2019empirical},
+but only recently have datasets made it possible to move beyond a few numerical summaries.
 
@@
     then $\site(\iw, f)_j = r_j^2$ is the squared correlation between $z$ and the allele at site $j$.
+    Appendix \ref{apx:regression} explains how to extend this to regression with covariates.
 \end{example}
@@
     aligns with similarity in phenotype.
+    Similar strategies for using local tree topology to improve associate mapping
+    were explored by \citet{zollner2005coalescent} and \citet{minichiello2006mapping}.
 \end{example}

### 2019-09-11 treestats_paper.tex
plr edits up until 1000G data

@@
 inherit from that mutation in the tree (and accounting for those that subsequently mutate).
-Here, we only consider statistics whose values are invariant to monomorphic sites. Thus, for a sample of $n$
-genomes, $0 < p \leq 1-1/n$.
-
-What we do here is a generalization of this idea.
-Roughly speaking, to do this we will
+
+This paper presents one way of generalizing of this idea.
+Roughly speaking, to compute a statistic we will
 propagate ``weights'' additively up each tree,
@@
 
-\begin{definition}[Sample and subtree weights]
+\begin{definition}[Sample and subtree weights] \label{defn:weights}
     The \emph{samples} of a tree sequence are the focal nodes,
@@
     \end{align*}
-    where $v \le_T u$ if $u$ is on the path from $v$ to root in $T$.
+    where $v \le_T u$ if $u$ is on the path from $v$ to root in the tree $T$.
+    The \emph{total weight} is the sum of the weights over all samples:
+    $\tiw = \sum_v \iw(v)$, which is the subtree weight of the root of any tree
+    that contains all samples.
 \end{definition}
+
+We allow vector-valued weights,
+i.e., $\iw(v)$ may be a vector $(\iw_1(v), \ldots, \iw_m(v))$,
+so that when summarising we have access to more than one aspect of each subtree.
 
@@
 
-We allow vector-valued weights,
-i.e., $\iw(v)$ may be a vector $(\iw_1(v), \ldots, \iw_m(v))$,
-so that when summarising we have access to more than one aspect of each subtree.
 The requirement that $f(0) = f(\tiw) = 0$ is for consistency,
@@
         the total number of sites.
-        \krt{This is a nice fig, but the $2 \times$ and the $/10 sites$ parts may be easy to miss?}
         \label{fig:divergence_diagram}
@@
 We refer to the branches in each tree using the most recent node,
-so for instance the branch describing a relationship where $u$ is the child of $v$
-is associated with $u$.
+so for instance a branch between an ancestor $v$ and descendant $u$
+is associated with the node $u$.
 The \emph{length} of this branch is the difference in birth times of $v$ and $u$,
@@
 (this is called an ``edge''),
-rearrangements of genealogical relationships
-can cause the precise set of samples that inherit from that edge to differ across
-trees due to recombination.
+rearrangements of genealogical relationships due to recombination
+can cause the precise set of samples that inherit from that edge to differ across trees.
 
@@
 Averaged across trees,
-this tells us what proportion of the sample's genomes were inherited from each node.
+this tells us, for each node, what proportion of the sample's genomes were inherited from that node.
 Motivated by this, we define the
@@
 of all alleles at given variable site.  For other statistics, one is only concerned with the derived state(s)
-at a site, and we may defined a \emph{polarised} node statistic without the second term from our previous definition:
+at a site, and we may define a \emph{polarised} node statistic without the second term from our previous definition:
 
@@
     If $\iw$ are the indicator weights of the set $S$ of $n$ samples,
-    and then $\iw_u / n$ is the proportion of the samples in $S$ that inherit from $u$.
+    then $\nw(u) / n$ is the proportion of the samples in $S$ that inherit from $u$.
     Therefore, if $f(x) = x / n$,
@@
 
-Now we arrive at our primary goal, which is to compute statistics from \emph{genomes}
-using this framework.
+Now we describe how to compute statistics from \emph{genomes} using this framework.
 To do this, we assume that the genetic variation data is embedded in a tree sequence,
@@
 
-Having defined a general framework, we may now give concrete examples of functions related to common summary statistics.
+Having defined a general framework, we may now give concrete examples
+of how to compute common summary statistics
 
 \begin{example}[Nucleotide diversity] \label{ex:site_diversity}
-    The nucleotide diversity of a sample is the average number of differences between all haplotypes, which is
-    equivalent to the sum of heterozygosity at all positions.
+    The nucleotide diversity of a group $S$ of $n$ samples
+    is the average density of differences between pairs of samples,
+    or equivalently, the average heterozygosity across positions.
     To calculate this statistic,
@@
     \end{align*}
-    Then $\site(f, \iw)_{[a,b)}$ is mean nucleotide diversity in the region between $a$ and $b$.
+    Then $\site(f, \iw)_{[a,b)}$ is mean nucleotide diversity of $S$ in the region between $a$ and $b$.
 \end{example}
@@
 \begin{example}[Nucleotide divergence] \label{ex:site_divergence}
-    Now suppose we want to compute the mean density of nucleotide differences
+    Now suppose we want to compute the mean pairwise density of nucleotide differences
     \emph{between} two nonoverlapping groups of samples, $S_1$ and $S_2$,
@@
     let $\iw_j = \bone_{S_j}$,
-    so that $\nw_{j}(u)$ gives the number of nodes in $S_j$ inheriting from $u$.
-    Then we define
+    so that $\nw_{j}(u)$ gives the number of nodes in $S_j$ inheriting from $u$,
+    and define
     \begin{align*}
@@
 \begin{example}[Segregating sites] \label{ex:segregating_sites}
-    Again, let $\iw = \bone_{S}$ for a group of $n$ sequences,
+    Again, let $\iw = \bone_{S}$ for a group of $n$ samples,
     and now define
@@
     then $\site(\iw, f)_j = r_j^2$ is the squared correlation between $z$ and the allele at site $j$.
-    Appendix \ref{apx:regression} explains how to extend this to regression with covariates.
 \end{example}
+
+Appendix \ref{apx:regression} explains how to extend the previous example
+to obtain the squared coefficient of genotype in the regression of phenotype
+against genotype with additional covariates.
 
@@
 % c(cor(g,z), sum(g*z)/sqrt(n*(n-1)*p*(1-p)))
-\krt{There is a possible point of confusion here.  So far, weight has been used to describe a property of nodes, and
-    then we have a function that operates on those weights.  In this example, the output of that function is also called
-a weight.  We may prefer to use a term like ``value" here?}
 \begin{example}[Patterson's $f_4$] \label{ex:site_f4}
@@
     $p_1 - p_2 = p_1 (1 - p_2) - (1 - p_1) p_2$,
-    and so the statistic counts with positive weight
+    and so the statistic counts with positive value
     alleles that split $S_1$ and $S_3$ from $S_2$ and $S_4$,
-    and negative weight ones that split $S_1$ and $S_4$ from $S_2$ and $S_3$.
+    and negative value ones that split $S_1$ and $S_4$ from $S_2$ and $S_3$.
     Therefore, if as before we
@@
         \frac{x_1}{n_1}
+        \left(1 - \frac{x_2}{n_2}\right)
         \frac{x_3}{n_3}
-        \left(1 - \frac{x_2}{n_2}\right)
         \left(1 - \frac{x_4}{n_4}\right)
@@
         \left(1 - \frac{x_1}{n_1}\right)
-        \left(1 - \frac{x_4}{n_4}\right)
         \frac{x_2}{n_2}
-        \frac{x_3}{n_3} ,
+        \frac{x_3}{n_3}
+        \left(1 - \frac{x_4}{n_4}\right),
     \end{align*}
@@
 described here is preferable, though.}
+
 \paragraph{Polyallelic sites}
@@
 would have site statistic $f(\aw(A)) + f(\aw(C)) + f(\aw(T))$.
-This is a natural defintion in that it agrees with what is obtained by
+This is not the only choice, but is natural because it agrees with what is obtained by
 looking at pairwise haplotype differences.
@@
 precisely because it tells us about the underlying patterns of genealogical relatedness.
-In other words, the genomes are useful in so far as they tell us about the trees.
+In other words, often the genomes are most useful in so far as they tell us about the trees.
 In the case where we actually have the trees, or a good proxy for them,
@@
 a statistic of tree shape by summing these expected contributions across its branches.
-We define the \textbf{branch statistic} of a single tree $T$ with span $L$
+We define the \textbf{branch statistic} of a single tree $T$
 obtained from summary function $f()$
 and sample weights $\iw$ with total weight $\tiw = \sum_u \iw(u)$
-to be the sum of the area of each branch in each tree
-multiplied by the summary function applied to its subtree weight:
+to be the sum of the length of each branch in each tree
+multiplied by the summary function applied to its subtree weight and its complement:
 \begin{align}
@@
     &=
-    \frac{1}{L} \sum_{u \in T} t_{u} \left( f(\nw_{T}(u)) + f(\tiw - \nw_{T}(u)) \right)  ,
+    \sum_{u \in T} t_{u} \left( f(\nw_{T}(u)) + f(\tiw - \nw_{T}(u)) \right)  ,
 \end{align}
@@
 and $f(\tiw - \nw_{T}(u))$ is the value that would be added due to its complementary allele,
-so $\branch(f, \iw)_T$ gives the expected contribution of this branch to $\site(f, \iw)$,
-per unit of sequence length that the tree $T$ covers.
+so $\branch(f, \iw)_T$ gives the expected contribution of the tree $T$ to $\site(f, \iw)$,
+per unit of sequence length that the tree covers.
 
@@
 This defines a statistic for a single window,
-but in practice it is probably most useful to average branch statistics
+but in practice it is more useful to average branch statistics
 over a region of the genome,
@@
 \end{align}
-where $\ell_k(c,d)$ is the length of the region in $[c,d)$ that the tree $T_k$ extends for
-(i.e., $T$ covers the half-open interval $[a_{k-1},a_k)$,
-so $\ell_k(c,d) = \max(0, \min(d,a_k) - \max(c,a_{k-1}))$).
-
-The \emph{polarised} version of this statistic
+where $\ell_k(c,d)$ is the length of the region in $[c,d)$ that the tree $T_k$ covers
+(i.e., if $T$ covers the half-open interval $[a_{k-1},a_k)$,
+then $\ell_k(c,d) = \max(0, \min(d,a_k) - \max(c,a_{k-1}))$).
+
+The \emph{polarised} version of a Branch statistic
 is defined analogously to the Node and Site versions:
@@
     then $f(u)$ gives the probability that the branch ancestral to $u$
-    lies on the path from two randomly chosen samples from $S$
+    lies on the path from one of two randomly chosen samples from $S$
     on the path up to their most recent common ancestor (MRCA).
-    Therefore, $\branch(f, \iw)$ for these choices
+    Therefore, $\branch(f, \iw)$
     gives the mean total distance in the tree between two samples from $S$,
@@
     aligns with similarity in phenotype.
-    Similar strategies for using local tree topology to improve associate mapping
-    were explored by \citet{zollner2005coalescent} and \citet{minichiello2006mapping}.
 \end{example}
+
+The previous example might be used to leverage local relatedness
+to improve the resolution of association studies.
+Similar strategies were explored by \citet{zollner2005coalescent} and \citet{minichiello2006mapping}.
 
@@
     The statistic $\branch(f, \iw)$ therefore
-    gives the difference in averages of these two types of branch,
+    gives the difference in average lengths of these two types of branch,
     averaged over choices of samples from the sample sets and averaged across the genome.
@@
 \subsection*{General properties}
+
+\plr{not sure if we need this section}
 
@@
 where the expected value averages over infinite-sites mutations with rate 1
-(and so lengths are measured in units of expected numbers of mutations).
-Let's unpack this statement a bit more: what exactly is the mutational model?
+(and so lengths are measured in units of expected numbers of mutations)
+but leaves the genealogies fixed.
+In other words, the Site statistic can be thought of an estimator of its corresponding Branch statistic,
+which is itself a summary of local tree shapes.
+Let's unpack the assumptions here: what exactly is the mutational model?
 First, we are taking the mutation rate to be 1, i.e.,
@@
 is the contribution of demographic variation, including drift.
-
-\krt{will this section go further?  It just kind of ends...}
+In the next section we examine this partitioning of variance using simulations.
 
@@
 have per-capita migration rates of $4/N$ to each other.
-\krt{It is not clear what figure this bit of text refers to?  Is this out of date?}
 In this situation, a significantly negative $f_4(1,2;3,4)$ is expected,
@@
 \section{Linear regression}
+\label{apx:regression}
 

### 2019-09-11 treestats_paper.tex
edits until disco

@@
 \newcommand{\tskit}{{\texttt{tskit}}}
+\newcommand{\tsinfer}{{\texttt{tsinfer}}}
 \newcommand{\branch}{\mbox{Branch}} % branch stat
@@
 We now begin to formalize these notions with definitions.
+In a tree sequence, each haplotype, modern or ancestral, is associated with a \emph{node},
+and the trees at each position along the genome have vertexes labeled by these nodes.
+A tree sequence describes the relationships between a special set of nodes, the \emph{samples},
+that appear in the trees at every point along the genome.
+The other (non-sample) nodes may not appear in every tree,
+as for instance if a portion of an ancestor's genome was not inherited by any of the samples.
 
 \begin{definition}[Sample and subtree weights] \label{defn:weights}
-    The \emph{samples} of a tree sequence are the focal nodes,
-    and appear in the trees at every point along the genome.
     A list of \emph{sample weights} $\iw$ assigns a numeric value $\iw(v)$
@@
 
-\krt{Is there a circularity here?  What does a branch statistic from real data mean
-when the branch lengths must be inferred somehow from the mutations? If this is not circular,the 
-reasons why need to be made clearer.}
-
 We do not know the true genealogies underlying real data,
 but recent methods are available to estimate them.
+Above, we showed that branch and site statistics matched well in simulated data,
+but these simulations made some simplifying assumptions, including constant mutation rates.
 To evaluate the correspondence between site and branch statistics
-in real data,
+in tree sequences inferred from real data,
 we calculated mean sequence diversity along chromosome 20 from the 1000 Genomes dataset
@@
 calculated using the site statistic described in Example~\ref{ex:site_diversity},
-and compared this to the corresponding branch statistic (of Example~\ref{ex:branch_diversity})
+and compared this to the dual branch statistic (of Example~\ref{ex:branch_diversity})
 as calculated from three inferred tree sequences:
-by tsinfer \citep{tsinfer}, which does not attenpt to estimate well-calibrated node times,
-(2) by GEVA \citep{geva}, which aims to calibrate the node times reported by tsinfer,
-and (3) by Relate \citep{relate}, an independently developed method.
+by \tsinfer{} \citep{kelleher2019inferring}, which only aims to estimate tree topologies,
+not well-calibrated node times,
+(2) by GEVA \citep{albers2019dating}, which aims to calibrate the node times reported by \tsinfer,
+and (3) by Relate \citep{speidel2019method}, an independently developed method.
 All calculations were done only using the portions of the chromosome
-\krt{unclear what this means--ref/citation is needed.}
-passing the ``strict'' mask for sequencing accessibility.
+passing the 1000 Genomes Phase 1 ``strict'' mask for sequencing accessibility.
+% from ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase1/analysis_results/supporting/accessible_genome_masks/README.20120515_phase1_pop_gen_mask
 Separate curves for each of the 26 populations are shown in Supplemental Figure \ref{fig:chr20_full}.
@@
 The comparisons are shown in Figure~\ref{fig:chr20}.
-Of these methods, branch diversity from tsinfer shows the worst agreement with site diversity,
-which is unsurprising as tsinfer makes no attempt to provide well-calibrated node times,
-and only aims to provide consistent tree topologies.
+Of these methods, branch diversity from \tsinfer{} shows the worst agreement with site diversity,
+which is unsurprising as \tsinfer{} uses allele frequency as a proxy for age.
 The two methods that infer node times differ substantially,
 with Relate showing a significantly tighter agreement than GEVA.
-Does this mean that Relate is doing a better job at inferring the tree sequence than \texttt{tsinfer} pluse GEVA?
-The results presented here are highly suggestive, but not entirely clear.
+Does this mean that Relate is doing a better job at inferring the tree sequence than \tsinfer/GEVA?
+Answering this question requires a deeper understanding of the processes that shape
+genetic diversity along the genome.
 
@@
         in 1Mb windows along human chromosome 20 of the 1000 Genomes Project,
-        \krt{the only appearances of "dual branch" are in figure legends.  need to define somewhere}
         and the dual branch statistic as calculated from tree sequences inferred by
-        \textbf{(top right)} tsinfer,
+        \textbf{(top right)} \tsinfer,
         \textbf{(bottom left)} Relate, and
-        \textbf{(bottom right)} GEVA.
+        \textbf{(bottom right)} GEVA/\tsinfer.
+        Separate curves are plotted for each sampling location.
         \label{fig:chr20}
@@
 
-\krt{Not sure exactly how to rewrite.}
 In practice, we do not expect branch and site diversity to agree exactly,
@@
 For instance, regions in which a high proportion of potential mutations are deleterious
-are expected to have a diversity for two reasons:
+are expected to have a lower diversity for two reasons:
 first, the deleterious mutations themselves are less likely to be found,
 and second, the indirect effect of deleterious mutations on nearby sites reduces typical tree height
-and thus diversity at even neutral, linked sites \citep{background_selection}.
+and thus diversity at even neutral, linked sites \citep{hudson1994levels,charlesworth1997effects}.
 The first effect would cause site and branch statistics to differ,
@@
 and violates the assumption of independence of mutations given the trees.
-However, the second effect does not affect the correspondence between site and branch statistics,
+However, the second effect does \emph{not} affect the correspondence between site and branch statistics,
 since it is mediated by tree shape.
 However, it is generally unknown how much mutation rate varies along the genome
-\citep[but see][]{mut_rate_varies}
-or how dense are targets of selection \citep{how_important_is_selection}.
+\citep[but see][]{supek2015differential}
+or how dense are targets of selection \citep{leffler2012revisiting}.
 For this reason, it is very interesting to see how close a correspondence between site and branch
@@
 as tight a match between site and branch statistics as possible,
-with remaining variation explained by selection and/or mutation rate variation.
+with remaining variation explained by direct selection and/or mutation rate variation.
 
@@
 Next, we examine how the \emph{time} distribution of admixture signal
-in the tree sequence inferred by tsinfer/GEVA.
-The branch $f_4(A, B; C, D)$ statistic for four populations $A$, $B$, $C$, and $D$
-sums $(p_A - p_B)(p_C - p_D)$ across branches, multiplied by span and height,
-where $p_A$ is the proportion of samples in $A$ that inherit from that node, etcetera.
-A branch with a large value is expected to contribute a large amount to the site $f_4$,
-which we calculate from data and is used to detect historical connections between populations.
-To examine the time signature, we assigned the contribution of each branch
+in the tree sequence inferred by \tsinfer/GEVA.
+The branch $f_4(A, B; C, D)$ statistic for four sample sets $A$, $B$, $C$, and $D$
+is the sum of $(p_A - p_B)(p_C - p_D)$ across branches, multiplied by span and height,
+where $p_X$ is the proportion of samples in sample set $X$ that inherit from that node.
+A branch with a large value is expected to contribute a large amount to the Site $f_4$,
+which is calculated from genotyping data
+and is used to infer historical connections between populations.
+To visualize the time signature, we assigned the contribution of each branch
 to its terminal node,
 and then summed these values across all nodes in 50 logarithmically-spaced bins by time.
-We then normalised these to sum to 1 by dividing by the total branch $f_4$ value.
-
-The four-tuple of populations for the first two statistics (PUR,TSI;GWD,JPT and ASW,CEU;MSL,CHB)
+We then normalised these to sum to 1 by dividing by the total Branch $f_4$ value.
+
+The four-tuple of sample sets for the first two statistics (PUR,TSI;GWD,JPT and ASW,CEU;MSL,CHB)
 are from similar regions of the world,
 and show almost identical time signatures,
-even though the second set of populations has a value of $f_4$ nearly eight times higher.
+even though the second set of sample sets has a value of $f_4$ nearly eight times higher.
 The second two statistics (IBS,GBR;FIN,JPT and TSI,CEU;FIN,CHB) are also comparable,
@@
 during recent history (e.g., more branches ancestral to ASW and MSL than CEU and MSL).
-Furthermore, we expected the different populations to have much more distinct time signatures.
+Furthermore, we expected the different sample sets to have much more distinct time signatures.
 The shared, ancient time signatures may be because nonzero $f_4$ statistics
@@
 during that time period.
+Making rigorous conclusions from this sort of visualization
+will require substantial validation work
+to understand the various sources of bias and uncertainty.
 
@@
 from, and including, node $u$; see Figure~\ref{fig:divergence_diagram}).
-As we move along the sequence, the trees change when branches are inserted
-and removed. By carefully ordering these insertions and removals, we
-ensure that this state is always correctly maintained and that we
-only need to make small adjustments for each tree. Computing the statistics
-is then a straightforward matter of applying the summary function $f$
+As we move along the sequence, the trees change when branches are inserted and removed.
+By carefully ordering these insertions and removals,
+we ensure that this state can be correctly maintained
+using only small adjustments for each tree.
+Computing the statistics is then a straightforward matter of applying the summary function $f$
 to the node states and aggregating the results appropiately for the
@@
 
-Formally, we define a tree sequence topology via a set of
-tables~\citep{kelleher2018efficient}. In a tree sequence, a node
-describes a specific ancestral haplotype, and details about these
-ancestors are stored in a node table, $\Nt$. The row $\Nt[u]$
-contains all information about the node $u$, and rows
+Formally, we represent a tree sequence using a set of tables~\citep{kelleher2018efficient}.
+Each node describes a particular haplotype (i.e., one of the two genomes of a diploid individual),
+and details about these haplotypes are stored in a node table, $\Nt$.
+The row $\Nt[u]$ contains all information about the node $u$, and rows
 are indexed from zero such that $0 \leq u < |\Nt|$.
-In particular, the birth time for $u$ is given by
-$\Nt[u]\prop{time}$. Information about how nodes relate to each
+In particular, the birth time for $u$ is given by $\Nt[u]\prop{time}$.
+Information about how nodes relate to each
 other along the genome is encoded in an edge table, $\Et$.
@@
 The order in which edges are inserted and removed is determined
-by the ``index vectors'' $\indexin$ and $\indexout$~\citep{kelleher2016efficient},
-which are defined as follows. \jk{TODO: succinct definition of these}.
-As we move along the tree sequence, the topology is recorded in the
-parent vector $\pi$, such that $\pi_u$ is the parent of node $u$
-and $\pi_u = -1$ if $u$ is a root. Similarly, we maintain the
-branch length $\beta$ such that
+by the ``index vectors'' $\indexin$ and $\indexout$~\citep{kelleher2016efficient}.
+The ``edge insertion'' vector $\indexin$ gives the ordering of edges
+sorted by left endpoint, and among edges with the same left endpoint,
+sorted so that edges closer to the root appear later.
+The ``edge removal'' vector $\indexout$ is similar, except gives the ordering of edges
+by right endpoint and with edges closer to the root appearing sooner.
+\plr{is that right?}
+As we move along the tree sequence, the topology of the current tree is recorded in the
+``parent vector'' $\pi$: the parent of each node $u$ is the node $\pi_u$,
+with $\pi_u = -1$ if $u$ is a root.
+Similarly, we maintain the branch length vector $\beta$ by setting
 $\beta_u = \Nt[\pi_u]\prop{time} - \Nt[u]\prop{time}$ and
@@
 vectors is trivial. We also assume that we are computing the statistic
-for across the full span of the tree sequence; the extension to multiple
-windows is routine.
+across the full span of the tree sequence; the extension to multiple windows is routine.
+See below for a description of the steps in words,
+and \autoref{fig:algorithm_example} for an example of the internal state of the algorithm.
 
@@
 $x$ and $F$ for each node in the tree sequence, and then sets the values
-of $x_u$ and $F_u$ for each of the samples in $S$. Afterwards, we set
-our ``running sum'' $s$ and output statistic $\sigma$ to zero, along
+of $x_u$ and $F_u$ for each of the samples $u$ in $S$.
+(The initial state is the ``empty forest'', where no nodes are connected to any others.)
+Afterwards, we set our ``running sum'' $s$ and output statistic $\sigma$ to zero, along
 with the ``tree left'' variable $t_l$. The $j$ and $k$ variables are
@@
 up the tree in step \algref{B8} (note the symmetry with step \algref{B5}).
-Finally, once we have remove and inserted all of the relevant edges, we are
+Finally, once we have removed and inserted all of the relevant edges, we are
 ready to add the contribution of the current tree to the overall statistic,
@@
 on the path to root is $\log_2 n$. Therefore, the overall running time
-of Algorithms~\algref{B} is of order $(n = \rho \ln n)\log_2 n$, which is
+of Algorithms~\algref{B} is of order $(n + \rho \log n)\log_2 n$, which is
 \[
@@
 modified is processed first. Therefore, this subtree will be disconnected from the
-main tree, ensuring that we traverse upwards all the way to root once. Similarly,
-edges are inserted in youngest-first order, ensuring that subtrees are only
+main tree, ensuring that we traverse upwards all the way to root only once.
+Similarly, edges are inserted in youngest-first order, ensuring that subtrees are only
 reconnected to the main tree when they are fully built.
@@
 statistic from tree sequences to calculating it from integer matrices containing genotypes at all variable positions.
-\autoref{fig:stats_performance} shows that the tree sequence methods implemented in \texttt{tskit} process variant data
+\autoref{fig:stats_performance} shows that the tree sequence methods implemented in \texttt{tskit} processes variant data
 three orders of magnitude faster than matrix-based approaches once the sample size is above $n \approx 500$ haploids.
@@
 $\approx 32,000$ to $\approx 138,000$ \citep{Watterson1975-ej}.
-\autoref{fig:stats_performance} only considers the time required to calculate the statistic. It does not include the time
-required to build the genotype matrix, nor the memory required to store it.  For the largest samples size of $n=10^6$,
-the matrix is approximately 200 gigabytes, and thus not practical on many systems.  The largest tree sequence simulated
-required less than one gigabyte of memory.
+\autoref{fig:stats_performance} only considers the time required to calculate the statistic,
+and does not include the time required to build the genotype matrix, nor the memory required to store it.
+For the largest samples size of $n=10^6$,
+the matrix is approximately 200 gigabytes, and thus not practical on many systems.
+The largest tree sequence simulated required less than one gigabyte of memory.
 
 \autoref{fig:stats_performance} shows that \texttt{tskit} requires more time per variant once the sample size is larger
-than $n \approx 10^4$.  For large sample sizes, the $O(n)$ complexity of building the first tree dominates the run time.
-This initial cost is amortized in longer genomes where more recombination leads to more trees.
+than $n \approx 10^4$.
+This occurs because for large sample sizes,
+the $O(n)$ complexity of building the first tree dominates the run time,
+a cost that is amortized in longer genomes where more recombination leads to more trees.
 
@@
         and the dual branch statistic as calculated from tree sequences inferred by
-        \textbf{(top right)} tsinfer,
+        \textbf{(top right)} \tsinfer,
         \textbf{(bottom left)} GEVA, and

### 2019-09-11 treestats_paper.tex
a few words in the disco

@@
 not remove the edges for the final tree). Additionally, each edge that
-we insert or remove may incur the cost of traversing up the tree as
+we insert or remove after the initial building of the tree
+may incur the cost of traversing up the tree as
 far as the root in steps \algref{B5} and \algref{B8}.
@@
 balanced \citep{li2013coalescent}, the expected number of nodes
-on the path to root is $\log_2 n$. Therefore, the overall running time
-of Algorithms~\algref{B} is of order $(n + \rho \log n)\log_2 n$, which is
+on the path to root is $\log_2 n$.
+Therefore, the overall running time
+of Algorithms~\algref{B} is of order $n + \rho \log(n) \log_2(n)$, which is
 \[
-    O\left(n \log n + \rho \log^2 n \right)
+    O\left(n + \rho \log(n)^2 \right)
 \]
@@
 \begin{figure}
-
     % Run the script illustrate_algorithm.py to generate these pdfs and
@@
 \autoref{fig:stats_performance} shows that the tree sequence methods implemented in \texttt{tskit} processes variant data
-three orders of magnitude faster than matrix-based approaches once the sample size is above $n \approx 500$ haploids.
+substantially faster than matrix-based approaches once the sample size is above $n \approx 500$ haploids --
+three times faster for one thousand haplotypes,
+growing to nearly three orders of magnitude faster for one million haplotypes.
 The expected number of mutations for each replicate is $10^4 \times \sum_{i=1}^{n-1}\frac{1}{i}$, and thus ranges from
@@
         scaled recombination rate $\rho = 4N_er$ both equal to $10^4$ and $N_e = 10^4$. For each replicate, Tajima's $D$
-        \citep{Tajima1989-de} statistic was calculated using \texttt{tskit}, \texttt{libsequence}
-        \citep{Thornton2003-wj}, and \texttt{scikit-allel} \krt{need cite for this}.
+        \citep{Tajima1989-de} statistic was calculated using \tskit, \texttt{libsequence}
+        \citep{Thornton2003-wj}, and \texttt{scikit-allel} \citep{miles2017scikit}.
         The solid green line shows the result of fitting the timing data to a model based on the expected complexity of the 
@@
 
+% this stuff is faster
+The most direct effect of the methods presented here might be on efficiency of computation
+with large genomic datasets --
+the python library \tskit{} provides fast, rigorously tested implementations
+of many standard population genetic statistics, with more on the way.
+The logarithmic scaling of storage and computation time with sample size
+could bring much more computation-intensive methods within reach on modern large genomic datasets
+of humans, 
+and helps make detailed investigation of whole populations in evolutionary simulations feasible.
+
+% but the duality thing is maybe more long-term exciting
+However, the focus of this paper was not on computation of standard population genetics statistics,
+but on the relationship of these statistics to the underlying trees.
+We have used essentially the same general framework
+to define two additional classes of statistics -- Branch and Node statistics --
+that well-defined relationships to the Site statistics computable from genome sequence.
+% describing trees is cool
+At least, this gives us descriptive tools to directly describe the genealogical trees
+of tree sequences obtained by simulation.
+% and we can do it with real data now
+However, new methods to infer tree sequences allow us to apply these methods to real data.
+% branch stats are less noisy!
+Given the true tree sequence,
+Branch statistics are less-noisy versions of the standard Site statistics,
+so methods that use Branch statistics may show increased power and precision
+over the corresponding genotype-only methods.
+We have shown that this is true in simulated data,
+but it is not yet known how much additional noise the process of tree sequence inference will add.
+% and we get the time dimension
+Furthermore, tree sequences add a \emph{time} dimension to genotype data,
+which these descriptive methods open up many possibilities for utilizing and visualizing of.
+
+\paragraph{Future work}
 In this paper,
-we consider only so-called \emph{single site} statistics
-(which we will define concretely shortly).
+we have considered only so-called additive, single site statistics.
 Extensions to statistics involving the pairwise joint distribution of genotypes across sites,
-such as linkage disequlibrium,
-are straightforward, and are planned for future work.
+such as linkage disequlibrium are planned for future work.
 Haplotype-based statistics may require a different class of algorithms.
-
-\paragraph{Diploid statistics}
-The statistics we develop here are in essence \emph{haploid} statistics.
-These tools can be used to compute a statistic about polyploid individuals,
-but they do not scale to large numbers of individuals.
-For instance, to compute the heterozygosity of a set of samples
--- this is $(1/N) \sum_{i=1}^N 2 p_i (1-p_i)$, where $p_i$ is the derived allele frequency
-in the $i^\text{th}$ individual --
-one would need to have an entry in the weight vector for \emph{each individual}.
-The genotype of a diploid at a site with a single mutation that occurred on the branch ancestral to node $n$
-is determined by where that node falls relative to the diploid's two haplotypes in the tree:
-if $n$ is on the path from either haplotype to their MRCA, the individual is heterozygous;
-if $n$ is on the path from the MRCA to the root, the individual is homozygous derived;
-and the individual is homozygous ancestral otherwise.
-This suggests the following generalization of our current scheme for diploids:
-for a list $(I_1, \ldots, I_N)$ of pairs of nodes,
-and two vectors $w^{(1)}$ and $w^{(2)}$ of weights of length $N$,
-add weight $w^{(1)}_i$ to each node on the path between the two nodes of $I_i$,
-and weight $w^{(2)}_i$ to each node on the path from the MRCA of $I_i$ to the root.
-With $k$ such weighting schemes, we would then need the summary function $F$
-to take $k \times 2$ values as input.
-Efficiently updating these ``diploid'' weights on the nodes should be possible,
-but is not a straightforward extension of our current method.
 
@@
 (This is true because of the singular value decomposition of $G$.)
+
+\paragraph{Diploid statistics}
+The statistics we develop here are in essence \emph{haploid} statistics.
+These tools can be used to compute a statistic about polyploid individuals,
+but they do not scale to large numbers of individuals.
+For instance, to compute the heterozygosity of a set of samples
+-- this is $(1/N) \sum_{i=1}^N 2 p_i (1-p_i)$, where $p_i$ is the derived allele frequency
+in the $i^\text{th}$ individual --
+one would need to have an entry in the weight vector for \emph{each individual}.
+The genotype of a diploid at a site with a single mutation that occurred on the branch ancestral to node $n$
+is determined by where that node falls relative to the diploid's two haplotypes in the tree:
+if $n$ is on the path from either haplotype to their MRCA, the individual is heterozygous;
+if $n$ is on the path from the MRCA to the root, the individual is homozygous derived;
+and the individual is homozygous ancestral otherwise.
+This suggests the following generalization of our current scheme for diploids:
+for a list $(I_1, \ldots, I_N)$ of pairs of nodes,
+and two vectors $w^{(1)}$ and $w^{(2)}$ of weights of length $N$,
+add weight $w^{(1)}_i$ to each node on the path between the two nodes of $I_i$,
+and weight $w^{(2)}_i$ to each node on the path from the MRCA of $I_i$ to the root.
+With $k$ such weighting schemes, we would then need the summary function $F$
+to take $k \times 2$ values as input.
+Efficiently updating these ``diploid'' weights on the nodes should be possible,
+but is not a straightforward extension of our current method.

### 2019-09-11 tskit_stat_benchmarks/plot_benchmarks_longer_genome.py

@@ -10,7 +10,7 @@ def tskit_model(logn, a, b, c):
     # and that the number of mutations is also proportional to rho log n.
     n = np.exp(logn)
     rho = 0.025  # Value from the file
-    return np.log((a * n + b * rho * logn) * logn + c)
+    return np.log(a * n + b * rho * logn * logn + c)
 
 
 symbols = {"tskit": "o", "allel": "^", "libseq": "s"}

### 2019-09-12 treestats_paper.tex
an actual disco now

@@
 
-It was once a major undertaking to collect data sufficent to estimate a single summary
+It was once a major undertaking to collect data sufficient to estimate a single summary
 of the many genetic relationships between the individuals of a sample,
@@
 where $T_k$ describes the genealogical relationships of the samples
-over the segment of genome betwen $a_{k-1}$ (inclusive) and $a_k$ (exclusive).
+over the segment of genome between $a_{k-1}$ (inclusive) and $a_k$ (exclusive).
 We say that tree $T_k$ \emph{covers} the (half-open) segment $[a_{k-1}, a_k)$,
@@
 % from ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase1/analysis_results/supporting/accessible_genome_masks/README.20120515_phase1_pop_gen_mask
-Separate curves for each of the 26 populations are shown in Supplemental Figure \ref{fig:chr20_full}.
+Separate curves for each of the 26 sampling locations are shown in Supplemental Figure \ref{fig:chr20_full}.
 
@@
 but a tree sequence allows us access to another dimension: time.
-Next, we examine how the \emph{time} distribution of admixture signal
+Next, we examine the \emph{time} distribution of admixture signal
 in the tree sequence inferred by \tsinfer/GEVA.
@@
         was obtained using \texttt{msprime} with scaled neutral mutation rate $\theta = 4N_e\mu$ and
-        scaled recombination rate $\rho = 4N_er$ both equal to $10^4$ and $N_e = 10^4$. For each replicate, Tajima's $D$
+        scaled recombination rate $\rho = 4N_er$ both equal to $10^4$ and $N_e = 10^4$,
+        where $\mu$ and $\rho$ are the total mutation and recombination rate across the simulated region
+        per generation, respectively.
+        For each replicate, Tajima's $D$
         \citep{Tajima1989-de} statistic was calculated using \tskit, \texttt{libsequence}
@@
 
-% this stuff is faster
-The most direct effect of the methods presented here might be on efficiency of computation
-with large genomic datasets --
-the python library \tskit{} provides fast, rigorously tested implementations
-of many standard population genetic statistics, with more on the way.
-The logarithmic scaling of storage and computation time with sample size
-could bring much more computation-intensive methods within reach on modern large genomic datasets
-of humans, 
-and helps make detailed investigation of whole populations in evolutionary simulations feasible.
-
-% but the duality thing is maybe more long-term exciting
-However, the focus of this paper was not on computation of standard population genetics statistics,
-but on the relationship of these statistics to the underlying trees.
-We have used essentially the same general framework
-to define two additional classes of statistics -- Branch and Node statistics --
-that well-defined relationships to the Site statistics computable from genome sequence.
-% describing trees is cool
-At least, this gives us descriptive tools to directly describe the genealogical trees
-of tree sequences obtained by simulation.
-% and we can do it with real data now
-However, new methods to infer tree sequences allow us to apply these methods to real data.
-% branch stats are less noisy!
-Given the true tree sequence,
-Branch statistics are less-noisy versions of the standard Site statistics,
-so methods that use Branch statistics may show increased power and precision
-over the corresponding genotype-only methods.
-We have shown that this is true in simulated data,
-but it is not yet known how much additional noise the process of tree sequence inference will add.
-% and we get the time dimension
-Furthermore, tree sequences add a \emph{time} dimension to genotype data,
-which these descriptive methods open up many possibilities for utilizing and visualizing of.
+% overview: general definition that is mostly functions of jAFS but includes others
+In this paper, we have described a general framework for summarising genetic variation
+and the underlying genealogies
+that encompasses many standard population genetic statistics
+and makes it easy to explore more.
+Many of these statistics are functions of the joint allele frequency spectrum,
+but the framework is more general,
+and can be used to quantify associations between genotypes and phenotypes.
+This generality was motivated in part by laziness --
+adding new statistics is now almost as easy as working out the appropriate summary function --
+but also makes it easy to computationally explore new classes of statistics.
+The range of statistics available to describe variation in a single exchangeable sample
+(i.e., a single population) are well-understood,
+but there are much larger and less well-explored classes of statistics
+describing genetic variation between many populations
+or across geographical space.
+
+% speed/efficiency: big data sets; large simulations
+The most obvious application of these methods on current practice
+is to improve efficiency of existing pipelines,
+as tree sequences allow storage and processing of large genomic data sets
+with orders of magnitude less space and time than most other methods.
+All statistics described here (and more) are implemented
+in the fast, rigorously tested python library \tskit{}
+that also provides tools for working with tree sequences.
+The largest efficiency gains are found for very large sample sizes
+only likely to be available in real data sets for humans in the near future,
+but the ability to analyze millions of whole genomes
+is useful for understanding evolutionary simulations of any species
+\citep[see][for a recent example]{galloway2019stickleback}.
+
+% duality: site stats ~ branch stat, but conversely branch more accurate version of site
+%  maybe analogous to imputation
+We hope a more broadly interesting and useful aspect of this paper
+is a more in-depth look at the correspondence between statistics calculated with genome sequence
+and properties of the underlying genealogies, i.e.,
+the duality between Site and Branch statistics under the infinite-sites model.
+This general correspondence is well-known,
+but the common framework implemented here makes it easy to translate between the two.
+One way to use this translation is to answer:
+what aspect of tree shape is this population genetics statistic summarising?
+This can help when interpreting results, especially if reality may not fit an idealised model
+of separate populations.
+However, methods to infer tree sequences from real data
+make it now possible to work in the other direction:
+instead of calculating (Site) statistics from the genotype matrix,
+we can calculate precisely analogous (Branch) statistics from the trees themselves,
+thus hopefully bypassing the extra layer of noise induced by mutation.
+How well this works will depend on the ability of inference methods
+to infer the true tree sequence.
+This might plausibly produce less noisy estimates because tree sequence inference
+essentially uses the signal from nearby patterns of variation
+to interpret relationships at a given site,
+thus tranforming the simple binary split induced by a SNP
+into a much richer source of information at every variant site.
+Furthermore,
+if tree sequence inference can be made insensitive to local variation in mutation rate,
+calculation of branch statistics would provide a summary method
+that is not affected by this potentially important confounding factor.
+
+% visualization and time decomposition adding a new dimension
+Genomic data are naturally distributed along two axes: along the genome, and across geography.
+The tree sequence extends this to a third axis: time.
+A great many methods in population genetics aim to describe aspects of history,
+and accurate (or at least unbiased) tree sequence inference
+may shift the focus of these problems from inference to descriptive analysis.
+In this paper, we have made an exploratory foray in this direction,
+by describing the time of origin of the ``admixture'' signal of nonzero $f_4$
+between several subsets of the 1000 Genomes dataset,
+in several inferred tree sequences.
+It is not yet clear whether these time profiles tell us more about history
+or about biases of the the inference methods,
+but either result would be useful.
+
+% moving past a few summaries in discrete groups to a genome/time/space view of relatedness
+The historical toolbox of population genetics
+is still mostly composed of statistics designed for analysis of a handful of loci
+in a small number of discrete, mostly separated populations.
+Both our data and our understanding of the world have moved beyond this.
+We hope that the tools we develop here will help make it possible
+to visualize and analyze genetic variation and genealogical realtedness
+along the genome, across space, and through time.
+\plr{okay rein me in here}
 
@@
 such as linkage disequlibrium are planned for future work.
-Haplotype-based statistics may require a different class of algorithms.
+Haplotype-based statistics may require different classes of algorithms.
 
@@
 \begin{figure}
-    \includegraphics[width=\textwidth]{{figures/1kg/relate_chr20.site.1000000.diversity}.pdf}
-    \includegraphics[width=\textwidth]{{figures/1kg/1kg_chr20.branch.1000000.diversity}.pdf}
-    \includegraphics[width=\textwidth]{{figures/1kg/tgp_geva_chr20.branch.1000000.diversity}.pdf}
-    \includegraphics[width=\textwidth]{{figures/1kg/relate_chr20.branch.1000000.diversity}.pdf}
+    \includegraphics[width=0.5\textwidth]{{figures/1kg/relate_chr20.site.1000000.diversity}.pdf}
+    \includegraphics[width=0.5\textwidth]{{figures/1kg/1kg_chr20.branch.1000000.diversity}.pdf}
+    \includegraphics[width=0.5\textwidth]{{figures/1kg/tgp_geva_chr20.branch.1000000.diversity}.pdf}
+    \includegraphics[width=0.5\textwidth]{{figures/1kg/relate_chr20.branch.1000000.diversity}.pdf}
     \caption{

### 2019-09-12 treestats_paper.tex
one more citation

@@
 we have considered only so-called additive, single site statistics.
-Extensions to statistics involving the pairwise joint distribution of genotypes across sites,
+Extensions to statistics involving the pairwise joint distribution of genotypes across sites
+\citep[as used by][]{hudson2001twolocus},
 such as linkage disequlibrium are planned for future work.

### 2019-09-12 treestats_paper.tex
Pass through Abstract and intro.

@@
 it distinguishes those genomes that have inherited it from those that have not,
-thus providing a glimpse of the genealogical tree that relates the genomes to each other at that site.
-Summaries of genetic variation thus also describe the underlying genealogies.
+providing a glimpse of the genealogical tree relating the genomes to each other at that site.
+Summaries of genetic variation therefore also describe the underlying genealogies.
 Here we exploit this duality to create a method that efficiently computes single-site statistics
 from population genetic data using the succinct tree sequence encoding
-of genome sequence and genealogies.
-We use the framework to provide fast implementations of many currently-defined statistics
-(whose run time scales with the logarithm of the number of samples),
-and to suggest several new statistics.
+of genealogies and genome sequence.
+We use the framework to provide efficient implementations of
+many currently-defined statistics, and to suggest several new statistics.
 The general approach aggregates ``sample weights'' up each marginal tree,
@@
 The results can be summarised in three different ways:
-by \emph{site}, which corresponds to statistics calculated from genome sequence,
+by \emph{site}, which corresponds to statistics calculated from genome
+sequence;
 by \emph{branch}, which gives the expected value of the site statistic
-with a large number of mutations under the infinite-sites model,
+with a large number of mutations under the infinite-sites model;
 or by \emph{node}, which summarises the contribution of each node in the tree to the statistics.
-This formulation also makes it easy to see what statistic of the underlying genealogical trees
+This formulation also makes it apparent what statistic of the underlying genealogical trees
 each sequence-based statistic is estimating.  By using simulations to evaluate performance, we
@@
 e.g., to estimate $F_{ST}$ between two populations using a dozen allozymes.
+\jk{Citation?}
 Today's vast quantity of whole-genome sequence
@@
 Computation is beginning to be a major problem:
-whole-genome sequence on the scale of the UK Biobank,
-for instance, might have 500K individuals genotyped at 10M loci \citep{ukbb},
+genetic data on the scale of the UK Biobank~\citep{bycroft2018genome},
+for instance, might have 500K individuals genotyped at 10M loci,
 resulting in a genotype matrix of $5 \times 10^{12}$ entries.
@@
 or iteratively refining a single computation.
-Many sorts of modern inference machinery, such as deep learning,
-depend on being able to compute enough statistics that this becomes a serious bottleneck in practice.
-Bypassing this bottleneck will require new data structures appropriate for genomic data.
-
+Furthermore, even larger datasets are being collected,
+and millions of whole genomes will soon be available. In this case the
+genotype matrix simply cannot fit into memory, and complex distributed
+computations are required to calculate population genetic statistics.
+
+Fortunately, the genotype matrix is not the only way to encode genetic data.
 The \emph{succinct tree sequence} (or \emph{tree sequence}, for brevity) is a data
 structure that efficiently encodes the genealogical trees describing how a set of
-individuals are related to each other at each point along their genome. The
-data structure was introduced in the context of coalescent
+individuals are related to each other at each point along their genome.
+It was introduced in the context of coalescent
 simulation~\citep{kelleher2016efficient}, and led to scalability increases of
-several orders of magnitude over existing methods. It was subsequently extended
-and refined for forwards-time simulations~\citep{kelleher2018efficient,haller2018tree}
-where similar efficiency gains were made. 
-Recent work~\citep{kelleher2018inferring} has shown that tree sequence algorithms can
+several orders of magnitude over existing methods. The methods were subsequently extended
+and refined for forwards-time
+simulations~\citep{kelleher2018efficient,haller2018tree}, with similar efficiency gains.
+Recent work~\citep{kelleher2019inferring} has shown that tree sequence algorithms can
 also be used to massively increase the scalability of methods for inferring
-genome-wide genealogies, and making it possible to infer trees for millions of whole genome samples.
-
+genome-wide genealogies, and making it possible to infer trees for millions of
+samples.
 The key to the remarkable scalability and efficiency of tree sequence
@@
 which manifest as shared \emph{edges} across many adjacent trees in the tree sequence.
-\citet{kelleher2016efficient} used this redundancy
-to store genome sequence and the associated genealogical relationships extremely compactly,
-as well as to compute several population genetics statistics efficiently.
-This paper generalizes the strategy used by \citet{kelleher2016efficient} for statistics computation
-to a much broader class of statistics.
+This redundancy can be exploited to store genome sequence and the
+associated genealogical relationships very compactly, while also
+allowing us to to compute several population genetics statistics
+efficiently~\citep{kelleher2016efficient}.
+This paper generalizes the approach to a much broader class of statistics.
 
@@
 The sequence of trees describing genealogical relationships as one moves along the genome
-are also described by the Ancestral Recombination Graph \citep[or ARG,][]{arg},
-the basis for several inference methods \citep{argmethod}.
+are also described by the Ancestral Recombination
+Graph~\citep{griffiths1991two,griffiths1996ancestral,rasmussen2014genome}.
 These genealogical trees can be fairly precisely inferred in nonrecombining regions
-\citep{felsenstein_book},
+\citep{felsenstein2004inferring},
 as for instance the mitochondria,

### 2019-09-12 treestats_paper.tex
Change & harmonise notation for branch length

@@
 
+\jk{These two intro paragraphs aren't working for me. There's too many
+parenthetical comments, mixing high-level overview with getting into
+the weeds. Does the first para actually help? Perhaps we're just confusing
+people by talking about the AFS here?}
 Many population genetic summary statistics can be defined using the allele frequency spectrum
@@
 and $h(p)$ for the allele frequency spectrum
--- i.e., the number of alleles whose frequency is $p$ in our sample --
+--- i.e., the number of alleles whose frequency is $p$ in our sample ---
 and suppose that for some function $f()$, we want to compute $\sum_p h(p) f(p)$.
@@
 
-This paper presents one way of generalizing of this idea.
+% 2-3 sentances of intro before this?
 Roughly speaking, to compute a statistic we will
@@
 summarize these weights on each branch with a ``summary function'',
-and aggregate these summaries to produce statistics
-(e.g., averaged in windows).
+and aggregate these summaries to produce statistics.
 For instance, we could calculate how many of a certain set $A$ of samples
@@
 (b) finding the total weight of all nodes in the subtree inheriting from that edge.
-It turns out to be useful to allow more general weights (including negative ones).
-Figure \ref{fig:divergence_diagram} shows a simple example of these procedures
+Figure~\ref{fig:divergence_diagram} shows a simple example of these procedures
 of propagating weights up a tree and summarising them,
@@
     the \emph{subtree weight} $\nw_T(u)$ for a node $u$ is the sum of all sample weights
-    of every sample node that is descended $u$ in the tree (including $u$, if it is a sample):
+    of every sample node that is descended from $u$ in the tree (including $u$, if it is a sample):
     \begin{align*}
@@
 The \emph{length} of this branch is the difference in birth times of $v$ and $u$,
-and is denoted $t_T(u)$ (dropping $T$ if it is clear from context).
+and is denoted $\beta_T(u)$.
 Note that although the same parent-child relationship may exist across many adjacent trees
@@
 If the mutation rate is constant,
-then this gives us an isometry --
+then this gives us an isometry ---
 measuring tree distances in branch lengths versus in numbers of mutations
@@
     &=
-    \sum_{u \in T} t_{u} \left( f(\nw_{T}(u)) + f(\tiw - \nw_{T}(u)) \right)  ,
+    \sum_{u \in T} \beta_{T}(u) \left( f(\nw_{T}(u)) + f(\tiw - \nw_{T}(u)) \right)  ,
 \end{align}
-where $t_{u}$ is the length of the branch ancestral to node $u$ in tree $T$
+where $\beta_{T}(u)$ is the length of the branch ancestral to node $u$ in tree $T$
 and $\nw_{T}(u)$ is the total weight of the subtree of $T$ inheriting from $u$ (as defined above).

### 2019-09-12 treestats_paper.tex
Resectioned a little & more edits.

@@
 efficiently~\citep{kelleher2016efficient}.
-This paper generalizes the approach to a much broader class of statistics.
+This paper generalizes the approach to a much broader class of statistics,
+supporting arbitrary tree topologies and patterns of allelic variation.
+% Want to emphasise generality here: people might assume we're only dealing
+% with the biallelic case.
 
@@
 or haplotype statistics) for later work.
-The methods we describe are implemented, tested, and validated in the \tskit{} library,
-which is available as an open-source Python package at \url{https://github.com/tskit-dev/tskit}.
+The methods we describe are implemented, tested, and validated in the \tskit{}
+Python and C libraries, available at \url{https://github.com/tskit-dev/tskit}
+under the terms of the MIT open source license.
 
@@
 We will use weights that count number of samples inheriting from each node
-sufficiently often that we give them a special name --
+sufficiently often that we give them a special name ---
 for a subset set $S$ of samples,
@@
 To do this, we assume that the genetic variation data is embedded in a tree sequence,
-but the trees are used only for efficiency --
-the results will not depend on the trees in any way
-(with the only exception being if the ancestral allele is assumed to be known, as discussed below).
-
-The summaries we defined above for each node in each tree
-extend nicely to genetic variants
--- just as a node weight contains information about which samples inherit from the node and which are not,
+but the trees are used only for efficiency --- the results will not depend on
+the trees in any way. The summaries we defined above nodes
+extend directly to genetic variants: just as a node weight contains information about
+which samples inherit from the node and which do not,
 so we can summarise patterns of genetic variation by summing up weights of all samples
 that carry a given allele.
-So, we define \emph{allele weights} to be
-the total weight of all sample nodes who have inherited that allele.
+Therefore, we define \emph{allele weights} to be
+the total weight of all sample nodes that have inherited that allele.
 
@@
 then $\aw_j(a)$ is equal to the weight of the subtree inheriting from the mutation that produced $a$.
-More generally, $\aw$ is not necessarily equal to a subtree weight
-but is easily computable from them.
-
-Using allele weights,
-we define the \textbf{site statistic} of site $j$ computed using a summary function $f()$
-and a set of sample weights $\iw$
-to be the sum of $f()$ applied to each of the allele weights:
+Similarly, in the more general case of recurrent and back mutations, $\aw$ can
+be computed from subtree weights in a straightforward way.
+We then define the \textbf{site statistic} of site $j$ for a summary function $f()$
+and sample weights $\iw$ to be the sum of $f()$ applied to the weight of
+each allele, $a$:
 \begin{align}
@@
     &=
-    \sum_{a} f(\aw_j(a)) ,
+    \sum_{a} f(\aw_j(a)).
 \end{align}
-where the sum is over all unique alleles at site $j$.
-
-Of course, we often want to summarize statistics across regions of the genome (``windows'').
+We often want to summarize statistics across regions of the genome (``windows'').
 To do this, we overload notation somewhat and use a subscript $[i,j)$ to denote an average
@@
 
-In the definition above we simply sum over all alleles at each site.
+In the definition above we sum over all alleles at each site.
 However, sometimes it is useful to distinguish the \emph{ancestral} allele
@@
 Having defined a general framework, we may now give concrete examples
-of how to compute common summary statistics
+of how to compute common summary statistics.
 
@@
 \end{example}
+
+\jk{I don't think there's anything to be gained by giving this example over
+an interval rather than saying ``Then $\site(f, \iw)_{j}$ is the nucleotide
+diversity within $S$ at site $j$". The windowing stuff is routine, I think, and is just
+distracting from the interesting stuff here. Ditto the next two examples.}
 
@@
     and we want to compute the correlation between this phenotype
-    and the genotype at each site in the tree sequence.
+    and the genotype at each site.
     For convenience, suppose $z$ is normalized
@@
 it is natural to summarize them directly, rather than working indirectly with the genotypes.
-If we assume that no two mutations occur at the same genomic position --
-i.e., the \emph{infinite sites model} --
+If we assume that no two mutations occur at the same genomic position ---
+i.e., the \emph{infinite sites model} ---
 then there is a natural correspondence between summaries of genotypes and summaries of tree shape.
@@
 is equal to the mutation rate multiplied by the length of the segment and by the length of the branch.
-In other words, the ``area'' of a branch in a tree of the tree sequence,
+In other words, the ``area'' of a branch in a tree,
 defined as its span (right minus left endpoint) multiplied by its length (parent time minus child time)
@@
 multiplied by the summary function applied to its subtree weight and its complement:
-\begin{align}
+\begin{align}\label{eqn:branch_stat_tree}
     \branch(f, \iw)_T
@@
 per unit of sequence length that the tree covers.
-
-An example is shown in Figure~\ref{fig:branch_site_diagram},
-showing how the $f_4$ site statistic assigns a weight to each mutation
-depending on its' frequency in each of the four sample sets,
+An example of this correspondence between site and branch statistics
+is shown in Figure~\ref{fig:branch_site_diagram}. Here, we
+see how the $f_4$ site statistic assigns a weight to each mutation
+depending on its frequency in each of the four sample sets,
 and how the corresponding branch statistic assigns the same weight to those branches.
@@
 
-This defines a statistic for a single window,
+Equation~\eqref{eqn:branch_stat_tree} defines a statistic for a single tree,
 but in practice it is more useful to average branch statistics
-over a region of the genome,
-which we do by averaging the tree statistics over the region
-with probabilities proportional to the trees' spans.
-We write this with similar notation:
+over a region of the genome. We do this by averaging the tree statistics over the region
+with probabilities proportional to the trees' spans:
 \begin{align}
@@
 then $\ell_k(c,d) = \max(0, \min(d,a_k) - \max(c,a_{k-1}))$).
-
 The \emph{polarised} version of a Branch statistic
@@
     averaged over choices of samples from the sample sets and averaged across the genome.
+    \jk{I don't follow this last bit: how can we averate over choices in the
+sample sets if there's only one element in each?}
 \end{example}
@@
 \plr{not sure if we need this section}
+\jk{Compress and move to the discussion?}
 
@@
 over $t$ units of time is equal to $\ell t$.
-This just amounts to a change of units --
+This just amounts to a change of units ---
 for instance, if genome length is measured in nucleotides,
@@
 that no two mutations occur at the same site
--- the fact that they do, occasionally, means that this is an approximation.
+--- the fact that they do, occasionally, means that this is an approximation.
 Third, we are assuming that mutation rates are constant through time and across the genome.
@@
 However, it is always possible to define a branch statistic that
-gives the expected value of the product.
-How to do this is described in \citet{ralph2019empirical}.
+gives the expected value of the product~\citep{ralph2019empirical}.
 
@@
 site statistics are noisy approximations to the corresponding branch statistic
--- but, how noisy?
+--- but, how noisy?
 How big is the contribution of mutation to the overall sampling variance of a statistic?
-Concretely the law of total variance lets us partition the variance of a site statistic
+The law of total variance lets us partition the variance of a site statistic
 into the contributions of mutation and demographic noise:
@@
 is the contribution of demographic variation, including drift.
-In the next section we examine this partitioning of variance using simulations.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Branch statistics for population genetics}
+
+% JK: deleted the section here as it seemed superfluous.
 
@@
 and detecting introgression.
-
 Figure~\ref{fig:sweep_duality} shows windowed
 diversity along the chromosome following a few selective sweeps.
-The top two plots compare ``branch'' diversity -- i.e., as computed only with tree shape --
+The top two plots compare ``branch'' diversity --- i.e., as computed only with tree shape ---
 to ``site'' diversity computed from sequence generated by 20 independent assignments of mutations to the same tree sequence,
@@
 how does sampling variance contribute?
-Not much, it turns out -- the bottom plot shows both site and branch diversity
+Not much, it turns out --- the bottom plot shows both site and branch diversity
 computed from 20 nonoverlapping groups of 100 samples.
@@
         in a population of size 1000 for 400 generations using SLiM,
-        followed by recapitation with $N_e=1000$.
+        followed by recapitation with $N_e=1000$~\citep{haller2018tree}.
         The selected alleles at the marked sites have selection coefficients between 0.08 and 0.25,
@@
         \textbf{(right)} window width.
-        The recombination rate $10^{-8}$,
+        The recombination rate was $10^{-8}$,
         and unless otherwise stated, the mutation rate was $10^{-9}$,

### 2019-09-12 treestats_paper.tex
Pass through final sections.

@@
 We do not know the true genealogies underlying real data,
-but recent methods are available to estimate them.
-Above, we showed that branch and site statistics matched well in simulated data,
-but these simulations made some simplifying assumptions, including constant mutation rates.
+but recent methods are available to estimate them at scale.
+In Figure~\ref{fig:sweep_duality}, we showed that branch and site statistics
+matched well in simulated data. However, these simulations make many
+simplifying assumptions, and, moreover,
+the underlying tree topologies and branch lengths are exactly
+correct, which no inference method can hope to achieve.
 To evaluate the correspondence between site and branch statistics
@@
 and compared this to the dual branch statistic (of Example~\ref{ex:branch_diversity})
-as calculated from three inferred tree sequences:
-by \tsinfer{} \citep{kelleher2019inferring}, which only aims to estimate tree topologies,
-not well-calibrated node times,
-(2) by GEVA \citep{albers2019dating}, which aims to calibrate the node times reported by \tsinfer,
-and (3) by Relate \citep{speidel2019method}, an independently developed method.
+as calculated from three inferred tree sequences.
+Firstly, we used the tree sequence estimated by \tsinfer{}
+\citep{kelleher2019inferring}. By default, \tsinfer{} uses the frequency of the
+derived allele at each site as a proxy for its age (and, therefore, the age
+corresponding ancestor and node), aiming only to estimate accurate
+tree topologies. We therefore expect the correspondence between branch
+length and site statistics in this case to be poor. Secondly,
+we used allele age estimates for each site supplied by GEVA~\citep{albers2019dating},
+as input to \tsinfer{}. In this case, ancestor ages were determined by
+the GEVA allele age estimates rather than the default frequency
+based heuristic used by \tsinfer{}; inference was run as usual
+otherwise. Thirdly, we used trees and node times estimated
+by Relate~\citep{speidel2019method}.
 All calculations were done only using the portions of the chromosome
@@
 The comparisons are shown in Figure~\ref{fig:chr20}.
-Of these methods, branch diversity from \tsinfer{} shows the worst agreement with site diversity,
-which is unsurprising as \tsinfer{} uses allele frequency as a proxy for age.
+As expected, branch diversity from \tsinfer{} shows the worst agreement with
+site diversity.
 The two methods that infer node times differ substantially,
-with Relate showing a significantly tighter agreement than GEVA.
-Does this mean that Relate is doing a better job at inferring the tree sequence than \tsinfer/GEVA?
+with Relate showing a significantly tighter agreement than
+GEVA+\tsinfer{}.
+Does this mean that Relate is doing a better job at inferring the tree sequence
+than GEVA+\tsinfer?
 Answering this question requires a deeper understanding of the processes that shape
 genetic diversity along the genome.
+
+\jk{I'm not particularly happy with this to be honest: it's all very
+hand-wavey. Can we be more precise about the correspondence between
+site and branch statistics? Relate looks a bit better, but I don't
+think we can say "significantly" (sounds like p-values or something).
+Is there some sort of time-series analysis we could do to show how
+similar sets of squiggly lines are to each other?}
 
@@
         \textbf{(bottom left)} Relate, and
-        \textbf{(bottom right)} GEVA/\tsinfer.
+        \textbf{(bottom right)} GEVA+\tsinfer.
         Separate curves are plotted for each sampling location.
@@
 Next, we examine the \emph{time} distribution of admixture signal
-in the tree sequence inferred by \tsinfer/GEVA.
-The branch $f_4(A, B; C, D)$ statistic for four sample sets $A$, $B$, $C$, and $D$
-is the sum of $(p_A - p_B)(p_C - p_D)$ across branches, multiplied by span and height,
+in the tree sequence inferred by GEVA+\tsinfer.
+The Branch $f_4(A, B; C, D)$ statistic for four sample sets $A$, $B$, $C$, and $D$
+is the sum of $(p_A - p_B)(p_C - p_D)$ across branches, multiplied by span and length,
 where $p_X$ is the proportion of samples in sample set $X$ that inherit from that node.
@@
         Time distribution of signal for four $f_4$ statistics
-        in the tree sequence for chromosome 20 inferred from 1000 genomes data by GEVA.
+        in the tree sequence for chromosome 20 inferred from 1000 genomes data
+        by GEVA+\tsinfer{}.
         The branch $f_4$ statistics for these four are:
@@
 sequence \cite[Algorithm L]{kelleher2016efficient}. The central
-problem which is shared across site, branch and node statistics
+problem shared across site, branch and node statistics
 is to efficiently maintain the ``state'' $x_u$ (where $x_u$ is the
@@
 Computing the statistics is then a straightforward matter of applying the summary function $f$
-to the node states and aggregating the results appropiately for the
+to the node states and aggregating the results appropriately for the
 particular statistics mode and windowing options.
@@
 by right endpoint and with edges closer to the root appearing sooner.
-\plr{is that right?}
 As we move along the tree sequence, the topology of the current tree is recorded in the
@@
 is correctly maintained by propagating the loss of the state $x_u$
-up the tree from $v$. Thus, for each direct ancestor $v$ that was a
+up the tree from $v$. Thus, for each node $v$ that was a
 direct ancestor of $u$ in the previous tree, we first subtract
@@
 where $r$ is the mean number of recombinations per chromosome per unit of time.
-Under these conditions, there are $O(n + \rho \ln n)$ edges in the tree sequence
+Under these conditions, there are $O(n + \rho \log n)$ edges in the tree sequence
 \citep{kelleher2016efficient}.  Clearly, each edge is examined
@@
 
-\autoref{fig:stats_performance} shows that \texttt{tskit} requires more time per variant once the sample size is larger
-than $n \approx 10^4$.
-This occurs because for large sample sizes,
-the $O(n)$ complexity of building the first tree dominates the run time,
+For larger sample sizes, there is an uptick in the time required per variant
+by \texttt{tskit} shown in \autoref{fig:stats_performance}.
+This occurs because the $O(n)$ complexity of building the first tree
+begins to dominate run time for large $n$,
 a cost that is amortized in longer genomes where more recombination leads to more trees.
@@
 and can be used to quantify associations between genotypes and phenotypes.
-This generality was motivated in part by laziness --
-adding new statistics is now almost as easy as working out the appropriate summary function --
+This generality was motivated in part by laziness ---
+adding new statistics is now almost as easy as working out the appropriate summary function ---
 but also makes it easy to computationally explore new classes of statistics.
@@
 All statistics described here (and more) are implemented
-in the fast, rigorously tested python library \tskit{}
-that also provides tools for working with tree sequences.
+in the fast, rigorously tested library \tskit{}
+that also provides tools for working with tree sequences in Python and C.
 The largest efficiency gains are found for very large sample sizes
@@
 How well this works will depend on the ability of inference methods
-to infer the true tree sequence.
+to estimate the true tree sequence.
 This might plausibly produce less noisy estimates because tree sequence inference
@@
 We hope that the tools we develop here will help make it possible
-to visualize and analyze genetic variation and genealogical realtedness
+to visualize and analyze genetic variation and genealogical relatedness
 along the genome, across space, and through time.
-\plr{okay rein me in here}
 
@@
 Thanks to Graham Coop, Andrew Kern, Nick Patterson, and Kelley Harris for useful suggestions
-and to Wilder Wohns and Leo Speidel for providing inferred tree sequences.
+and to Wilder Wohns, Yan Wong and Leo Speidel for providing inferred tree sequences.
 

### 2019-09-12 treestats_paper.tex

@@
 associated genealogical relationships very compactly, while also
-allowing us to to compute several population genetics statistics
-efficiently~\citep{kelleher2016efficient}.
+enabling efficient calculation of several population genetics
+statistics~\citep{kelleher2016efficient}.
 This paper generalizes the approach to a much broader class of statistics,
@@
 but the trees are used only for efficiency --- the results will not depend on
-the trees in any way. The summaries we defined above nodes
+the trees in any way. The summaries we defined above for nodes
 extend directly to genetic variants: just as a node weight contains information about
@@
 then $\aw_j(a)$ is equal to the weight of the subtree inheriting from the mutation that produced $a$.
-Similarly, in the more general case of recurrent and back mutations, $\aw$ can
+In the more general case of recurrent and back mutations, $\aw$ can
 be computed from subtree weights in a straightforward way.
 We then define the \textbf{site statistic} of site $j$ for a summary function $f()$
-and sample weights $\iw$ to be the sum of $f()$ applied to the weight of
-each allele, $a$:
+and sample weights $\iw$ to be the sum of $f()$
+applied to the weight of every allele, $a$, found at this site:
 \begin{align}
@@
 \end{example}
-
-\jk{I don't think there's anything to be gained by giving this example over
-an interval rather than saying ``Then $\site(f, \iw)_{j}$ is the nucleotide
-diversity within $S$ at site $j$". The windowing stuff is routine, I think, and is just
-distracting from the interesting stuff here. Ditto the next two examples.}
 
@@
     gives the difference in average lengths of these two types of branch,
-    averaged over choices of samples from the sample sets and averaged across the genome.
-    \jk{I don't follow this last bit: how can we averate over choices in the
-sample sets if there's only one element in each?}
+    averaged across the genome.
 \end{example}

### 2019-09-12 treestats_paper.tex
a few more bits

@@
 % \emph{Running head:} Genealogies and genomes
-
-\krt{several refs seem missing}
 
@@
 makes it possible to confidently estimate many more properties of genealogical relationships
-in local regions of genomes \citep{stankowski2019widespread,fst_landscapes},
-or between individuals rather than populations \citep{individual_stats}.
+in local regions of genomes \citep[e.g,][]{booker2018understanding,haenel2018metaanalysis,stankowski2019widespread},
+or between individuals rather than populations \citep[e.g.,][]{browning2010highresolution,speed2014relatedness,aguillon2017deconstructing}.
 Computation is beginning to be a major problem:
@@
 
-\jk{These two intro paragraphs aren't working for me. There's too many
-parenthetical comments, mixing high-level overview with getting into
-the weeds. Does the first para actually help? Perhaps we're just confusing
-people by talking about the AFS here?}
-Many population genetic summary statistics can be defined using the allele frequency spectrum
-of a collection of homologous sequences.
-Concretely, write $p_i(a)$ for the frequency of allele $a$ at a site $i$,
+Before formal definitions, we will give an overview of the general framework.
+Roughly speaking, to compute a statistic we will
+propagate ``weights'' additively up each tree,
+summarize these weights on each branch with a ``summary function'',
+and aggregate these summaries to produce statistics.
+The abstract idea of summarizing weights propagated up each tree
+is only a small generalization of a familiar operations in population genetics:
+summarizing the allele frequency spectrum.
+For instance, we could calculate how many of a certain set $A$ of samples
+inherit from a particular branch in a tree
+by (a) assigning each of these samples weight 1 (and other nodes weight zero), then
+(b) finding the total weight of all nodes in the subtree inheriting from that branch.
+Any mutation falling on this branch would be inherited by 
+This gives the number of samples that would inherit any mutation falling on this branch,
+so is analogous to the allele frequency spectrum.
+Figure~\ref{fig:divergence_diagram} shows a simple example of these procedures
+of propagating weights up a tree and summarising them,
+used to calculate mean sequence divergence over a short sequence spanned by a single tree.
+
+% Many population genetic summary statistics can be defined using the allele frequency spectrum of a collection of homologous sequences.
+If we only considered summaries of the allele frequency spectrum of a single sample,
+our statistics would be equivalent to that described by \citet{fu1995statistical}.
+To connect the framework to familiar notions,
+here is a reframing of \citeauthor{fu1995statistical}'s statistics.
+Write $p_i(a)$ for the frequency of allele $a$ at a site $i$,
 and $h(p)$ for the allele frequency spectrum
@@
 
-% 2-3 sentances of intro before this?
-Roughly speaking, to compute a statistic we will
-propagate ``weights'' additively up each tree,
-summarize these weights on each branch with a ``summary function'',
-and aggregate these summaries to produce statistics.
-For instance, we could calculate how many of a certain set $A$ of samples
-inherit from a particular edge in a tree
-by (a) assigning each of these samples weight 1 (and other nodes weight zero), then
-(b) finding the total weight of all nodes in the subtree inheriting from that edge.
-Figure~\ref{fig:divergence_diagram} shows a simple example of these procedures
-of propagating weights up a tree and summarising them,
-used to calculate mean sequence divergence over a short sequence spanned by a single tree.
-
-We now begin to formalize these notions with definitions.
+Before defining the statistics, we need a few definitions.
 In a tree sequence, each haplotype, modern or ancestral, is associated with a \emph{node},
@@
 %%%%%%%%%%%%%%%%%%
-\subsection*{General properties}
-
-\plr{not sure if we need this section}
-\jk{Compress and move to the discussion?}
-
-The statistics defined here are all \emph{additive} along the genome:
-if we have computed a statistic in equal-size windows along a chromsome,
-then the value of the statistic for the entire chromosome is equal to
-the average of the values in those windows.
-Some commonly-used statistics (e.g., $F_{ST}$ or Tajima's $D$)
-are not additive, but are ratios of additive statistics, so can be easily computed in this way.
-Multilocus statistics that measure linkage disequilibrium or haplotype lengths
-do not fall in this category.
-We have also defined site statistics to be additive across \emph{alleles},
-to allow the definitions to apply to sites with more than two alleles.
-This decision has some nice properties, but is not the only reasonable choice.
-
-We also require that site statistics be computable from the genotype matrix
-(i.e., without knowledge of the trees),
-and, if the statistic is polarised, knowledge of the ancestral allele.
-We often would also like our statistics to be unaffected by the addition of monomorphic sites,
-which are usually omitted from genetic variation data.
-To ensure this we have required that the summary function returns zero when give both zero weight
-and the total weight: $f(0) = f(\tiw) = 0$.
-This also makes branch and node statistics unchanged by portions of the tree that do not lie on a path
-between any of the samples that have nonzero weight,
-so that in particular,
-\emph{simplification} of the tree sequence will not change the value of the statistic.
-However, in practice this requirement can be relaxed.
-
-%%%%%%%%%%%%%%%%%%
 \section*{Duality of site and branch statistics}
@@
 
-\paragraph{Future work}
-In this paper,
-we have considered only so-called additive, single site statistics.
+%%%%%%%%%%%%%%%%%%
+\subsection*{Duality in practice?}
+
+The duality between Site and Branch statistics depends crucially on two assumptions:
+that each new mutation is distinguishable, as in the infinite-sites model,
+and that mutation rates are constant along the genome.
+Neither of these things are true, but are they good approximations?
+Multiple mutations at the same site, causing backmutations,
+are present in real datasets but are generally thought to be uncommon,
+especially if hypervariable sites are excluded.
+The effect of multiple mutations on Site statistics
+can be bounded by a function of the proportion of sites at which these occur.
+Large-scale variation in mutation rate (at, say, the megabase scale)
+might plausibly have a stronger effect on the correspondence,
+because a Branch and Site statistic are equal in expectation up to multiplication by the mutation rate.
+For instance, if we knew that the inferred tree sequences of \autoref{fig:chr20} were correct,
+then dividing the Site statistic by the Branch statistic
+would give us an estimate of relative average mutation rate in windows along the genome.
+Similarly, Branch statistics from well-calibrated tree sequences
+would provide a way to assess variation in population genetic summaries along the genome
+in a way independent of local mutation rate.
+
+A final crucial assumption underlying this duality is that mutations can be treated
+as indepenent of the trees themselves.
+This is clearly not always true --- for instance,
+a sweeping beneficial mutation strongly affects the local trees (as in \autoref{fig:sweep_duality}) ---
+but seems likely to be approximately true.
+\citet{ralph2019empirical} contains more discussion of variation in mutation rate,
+but more work needs to be done, particularly on models with large proprotions of sites under selection.
+
+%%%%%%%%%%%%%%%%%%
+\subsection*{General properties}
+
+The statistics defined here are all \emph{additive} along the genome:
+if we have computed a statistic in equal-size windows along a chromsome,
+then the value of the statistic for the entire chromosome is equal to
+the average of the values in those windows.
+Some commonly-used statistics (e.g., $F_{ST}$ or Tajima's $D$)
+are not additive, but are ratios of additive statistics, so can be easily computed in this way.
+We have also defined site statistics to be additive across \emph{alleles},
+to allow the definitions to apply to sites with more than two alleles.
+This decision has some nice properties, but is not the only reasonable choice.
+
+Multilocus statistics that measure linkage disequilibrium or haplotype lengths
+do not fall in this category.
 Extensions to statistics involving the pairwise joint distribution of genotypes across sites
@@
 Haplotype-based statistics may require different classes of algorithms.
+
+We also require that site statistics be computable from the genotype matrix
+(i.e., without knowledge of the trees),
+and, if the statistic is polarised, knowledge of the ancestral allele.
+We often would also like our statistics to be unaffected by the addition of monomorphic sites,
+which are usually omitted from genetic variation data.
+To ensure this we have required that the summary function returns zero when give both zero weight
+and the total weight: $f(0) = f(\tiw) = 0$.
+This also makes branch and node statistics unchanged by portions of the tree that do not lie on a path
+between any of the samples that have nonzero weight,
+so that in particular,
+\emph{simplification} of the tree sequence will not change the value of the statistic.
+However, in practice this requirement can be relaxed.
+
+%%%%%%%%%%%%%%%%%%
+\subsection*{Dynamic graph algorithms}
+
+The tree sequence describes trees that change as one moves along the genome,
+and so is a special case of a \emph{dynamic graph}
+\citep[also called a \emph{graph timeline}][]{lacki2013reachability}.
+Much of the work on dynamic graphs is focused on connectivity,
+e.g., maintaining a minimum spanning tree \citep{eppstein1994offline, eppstein1997sparsification, holm2001polylogarithmic},
+but development of parallel algorithms for more general operations on large dynamic graphs
+is an active area of research \citep[e.g.,][]{srinivasan2018sharedmemory}.
 

### 2019-09-12 treestats_paper.tex
Edits to the perf section.

@@
 of Algorithms~\algref{B} is of order $n + \rho \log(n) \log_2(n)$, which is
-\[
+\begin{equation}\label{eqn:alg_complexity}
     O\left(n + \rho \log(n)^2 \right)
-\]
+\end{equation}
 Computing a site statistic will have the same complexity,
@@
 %%%%%%%%%%%%%%%%%%%%%
-\subsection*{Speed}
-
-\krt{need to sort out the big-O issue here}
+\subsection*{Efficiency}
+
 We used coalescent simulations to compare the performance of calculating Tajima's \citeyearpar{Tajima1989-de} $D$
@@
 
-For larger sample sizes, there is an uptick in the time required per variant
-by \texttt{tskit} shown in \autoref{fig:stats_performance}.
-This occurs because the $O(n)$ complexity of building the first tree
-begins to dominate run time for large $n$,
-a cost that is amortized in longer genomes where more recombination leads to more trees.
+To determine how well our theoretical model of time complexity predicts the
+running time in \tskit{}'s implementation of the site statistics algorithm,
+we fitted a model based on Equation~\eqref{eqn:alg_complexity}.
+\autoref{fig:stats_performance} shows that theoretical predictions match
+the observed running time very well. As the overall complexity is
+$O(n + \rho \log(n)^2)$, for sufficiently large $n$ the initial term
+(representing the cost of building the first tree) will come to dominate.
+In this case, we can see this happening at around $n = 10^5$, where there
+is a noticeable uptick in the time required per variant. For
+longer genomes (i.e., larger values of $\rho$), this cost is amortized over more trees
+and is less apparent.
 
@@
     \centering
-    \includegraphics{tskit_stat_benchmarks/benchmarks_without_copy_longer_genome.pdf}
+    \includegraphics[width=9cm]{tskit_stat_benchmarks/benchmarks_without_copy_longer_genome.pdf}
     \caption{Comparison of time required to compute site statistics
@@
         \citep{Thornton2003-wj}, and \texttt{scikit-allel} \citep{miles2017scikit}.
-        The solid green line shows the result of fitting the timing data to a model based on the expected complexity of the
-        algorithm used by \texttt{tskit}. \krt{need to revisit the correctness of this}
+        The solid green line shows the result of fitting the timing data to a model
+        based on the theoretical complexity of the algorithm used by \texttt{tskit}.
         \label{fig:stats_performance}
@@
 \end{figure}
+
+It should be noted here that this example based on simulated data
+represents a best-case scenario in terms of the performance advantages
+of \tskit{} over the matrix-based methods. For real data, inferred
+tree sequences currently contain substantially more edges than we
+would expect based on simple neutral theory~\citep{kelleher2019inferring},
+and therefore computing statistics will not be as efficient as
+for an equivalent simulation. Tree sequence inference
+methods are in their infancy, however, and it is likely that
+as they improve the number of edges required to encode data will
+be reduced. For data simulated natively in tree sequence format
+via packages such as \texttt{msprime}~\citep{kelleher2016efficient},
+\texttt{SLiM}~\citep{haller2018tree,haller2019slim} and
+\texttt{fwdpy11}~\citep{thornton2014c++}, the advice is straightforward.
+Computing statistics using the algorithms in \tskit{}
+will always be more efficient than decoding the genotype
+matrix, importing it into another package and computing from the matrix.
 

### 2019-09-12 treestats_paper.tex
Tweaked alg-state figure.
fix typo in caption

--- 2019-09-12 treestats_paper.tex
@@
     \textbf{(b)} Each branch is assigned weight equal to the weight that would be assigned
-    to any mutation falling in it (colored lined), and then the branch $f_4$ statistic
+    to any mutation falling in it (colored lines), and then the branch $f_4$ statistic
     is the sum over these weights, multiplied by the length of the branch

### 2019-09-12 treestats_paper.tex
Kreitman 1983

@@
 of the many genetic relationships between the individuals of a sample,
-e.g., to estimate $F_{ST}$ between two populations using a dozen allozymes.
-\jk{Citation?}
+e.g., to estimate levels of sequence diversity from a few dozen alleles at an allozyme locus
+\citep{Kreitman1983-xr}.
 Today's vast quantity of whole-genome sequence

### 2019-09-12 treestats_paper.tex
Thursday PM editing pass.

@@
 
+\krt{I have related concerns. What does it mean to look at a branch stat and a site state
+    from an inferred tree sequence?  The input genotype data are informative about both 
+    the inferred edges and their branch lengths.   So it seems to me that comparing the two types of stats is comparing
+    non-independent quantities, and I am unclear how to interpret that? (I admit to not studying the GEVA paper closely
+yet, nor the Speidel).  Presumably, the site stat is going to give you the right value (e.g., match whatever the
+original paper reported for, say, pi), but the branch version of the stat could differ from this for perhaps a lot 
+of reasons that all contribute to "edge estimation error"?  Not calling variants below a certain frequency,
+preferentially calling variants you've seen in a previous study, or imputing in variants from other studies?
+Overall, I have to squint at Fig S2 to see differences, and most of the issue seems to be near what I assume to be a
+centromere?}
+
 \begin{figure}
@@
         The solid green line shows the result of fitting the timing data to a model based on the expected complexity of the
-        algorithm used by \texttt{tskit}. \krt{need to revisit the correctness of this}
+        algorithm used by \texttt{tskit}. 
         \label{fig:stats_performance}
@@
 to interpret relationships at a given site,
-thus tranforming the simple binary split induced by a SNP
+thus transforming the simple binary split induced by a SNP
 into a much richer source of information at every variant site.
@@
 It is not yet clear whether these time profiles tell us more about history
-or about biases of the the inference methods,
+or about biases of the inference methods,
 but either result would be useful.
@@
 Neither of these things are true, but are they good approximations?
-Multiple mutations at the same site, causing backmutations,
+Multiple mutations at the same site, causing back mutations,
 are present in real datasets but are generally thought to be uncommon,
@@
 A final crucial assumption underlying this duality is that mutations can be treated
-as indepenent of the trees themselves.
+as independent of the trees themselves.
 This is clearly not always true --- for instance,
@@
 \citet{ralph2019empirical} contains more discussion of variation in mutation rate,
-but more work needs to be done, particularly on models with large proprotions of sites under selection.
+but more work needs to be done, particularly on models with large proportions of sites under selection.
 
@@
 The statistics defined here are all \emph{additive} along the genome:
-if we have computed a statistic in equal-size windows along a chromsome,
+if we have computed a statistic in equal-size windows along a chromosome,
 then the value of the statistic for the entire chromosome is equal to
@@
 %%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Acknowledgements}
+\subsection*{Acknowledgments}
 Thanks to Graham Coop, Andrew Kern, Nick Patterson, and Kelley Harris for useful suggestions
@@
         \textbf{(bottom left)} GEVA, and
+        \krt{This should be tsinfer+GEVA, no?}
         \textbf{(bottom right)} Relate.

### 2019-09-12 treestats_paper.tex
Delete comment that is addressed in "Duality in practice" section

@@
 
-\krt{I have related concerns. What does it mean to look at a branch stat and a site state
-    from an inferred tree sequence?  The input genotype data are informative about both 
-    the inferred edges and their branch lengths.   So it seems to me that comparing the two types of stats is comparing
-    non-independent quantities, and I am unclear how to interpret that? (I admit to not studying the GEVA paper closely
-yet, nor the Speidel).  Presumably, the site stat is going to give you the right value (e.g., match whatever the
-original paper reported for, say, pi), but the branch version of the stat could differ from this for perhaps a lot 
-of reasons that all contribute to "edge estimation error"?  Not calling variants below a certain frequency,
-preferentially calling variants you've seen in a previous study, or imputing in variants from other studies?
+\krt{
 Overall, I have to squint at Fig S2 to see differences, and most of the issue seems to be near what I assume to be a

### 2019-09-12 treestats_paper.tex

@@
 \begin{figure}
+\footnotesize
     % Run the script illustrate_algorithm.py to generate these pdfs and
@@
     \\
+    $s = 3.1$ & & $s = 2.9$\\
 
@@
     Immediately before we remove the edge joining 6 to 7 and (B) immediately after.
-    See the text for descriptions of the arrays encoding the state.
-    \jk{TODO: the value of $s$ would be helpful here too.}
+    See the text for descriptions of the arrays encoding the state. Also shown
+    is the value of the running sum, $s = \sum_u \beta_u F_u$.
     \label{fig:algorithm_example}}
@@
         \citep{Thornton2003-wj}, and \texttt{scikit-allel} \citep{miles2017scikit}.
-        The solid green line shows the result of fitting the timing data to a model based on the expected complexity of the
-        algorithm used by \texttt{tskit}. 
+        The solid green line shows the result of fitting the timing data to a model based on the expected complexity of the algorithm used by \texttt{tskit}
+        (\autoref{eqn:alg_complexity}).
         \label{fig:stats_performance}

### 2019-09-13 treestats_paper.tex
removed tsinfer

@@
 \newcommand{\tskit}{{\texttt{tskit}}}
-\newcommand{\tsinfer}{{\texttt{tsinfer}}}
 \newcommand{\branch}{\mbox{Branch}} % branch stat
@@
 and compared this to the dual branch statistic (of Example~\ref{ex:branch_diversity})
-as calculated from three inferred tree sequences.
-Firstly, we used the tree sequence estimated by \tsinfer{}
-\citep{kelleher2019inferring}. By default, \tsinfer{} uses the frequency of the
-derived allele at each site as a proxy for its age (and, therefore, the age
-corresponding ancestor and node), aiming only to estimate accurate
-tree topologies. We therefore expect the correspondence between branch
-length and site statistics in this case to be poor. Secondly,
-we used allele age estimates for each site supplied by GEVA~\citep{albers2019dating},
-as input to \tsinfer{}. In this case, ancestor ages were determined by
-the GEVA allele age estimates rather than the default frequency
-based heuristic used by \tsinfer{}; inference was run as usual
-otherwise. Thirdly, we used trees and node times estimated
-by Relate~\citep{speidel2019method}.
+from the trees and node times estimated by Relate~\citep{speidel2019method}.
 All calculations were done only using the portions of the chromosome
@@
 % from ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase1/analysis_results/supporting/accessible_genome_masks/README.20120515_phase1_pop_gen_mask
-Separate curves for each of the 26 sampling locations are shown in Supplemental Figure \ref{fig:chr20_full}.
-
-The comparisons are shown in Figure~\ref{fig:chr20}.
-As expected, branch diversity from \tsinfer{} shows the worst agreement with
-site diversity.
-The two methods that infer node times differ substantially,
-with Relate showing a significantly tighter agreement than
-GEVA+\tsinfer{}.
+As shown in Figure~\ref{fig:chr20}, Branch and Site diversity along the genome are very similar.
 Does this mean that Relate is doing a better job at inferring the tree sequence
-than GEVA+\tsinfer?
+Is this an indication that Relate is doing a good job at inferring trees,
+or just a byproduct of the inference method (which, naturally, estimates branch lengths to be proportional to numbers of mutations)?
 Answering this question requires a deeper understanding of the processes that shape
@@
 
-\jk{I'm not particularly happy with this to be honest: it's all very
-hand-wavey. Can we be more precise about the correspondence between
-site and branch statistics? Relate looks a bit better, but I don't
-think we can say "significantly" (sounds like p-values or something).
-Is there some sort of time-series analysis we could do to show how
-similar sets of squiggly lines are to each other?}
-
-\krt{
-Overall, I have to squint at Fig S2 to see differences, and most of the issue seems to be near what I assume to be a
-centromere?}
-
 \begin{figure}
-    \includegraphics[width=0.5\textwidth]{{figures/1kg/relate_chr20.site.1000000.region.diversity}.pdf}
-    \includegraphics[width=0.5\textwidth]{{figures/1kg/1kg_chr20.branch.1000000.region.diversity}.pdf}
-    \includegraphics[width=0.5\textwidth]{{figures/1kg/relate_chr20.branch.1000000.region.diversity}.pdf}
-    \includegraphics[width=0.5\textwidth]{{figures/1kg/tgp_geva_chr20.branch.1000000.region.diversity}.pdf}
+    \includegraphics[width=0.8\textwidth]{{figures/1kg/relate_chr20.site.1000000.region.diversity}.pdf}
+    \includegraphics[width=0.8\textwidth]{{figures/1kg/relate_chr20.branch.1000000.region.diversity}.pdf}
     \caption{
-        \textbf{(top left)} Mean sequence diversity
+        \textbf{(top)} Mean sequence diversity
         in 1Mb windows along human chromosome 20 of the 1000 Genomes Project,
-        and the dual branch statistic as calculated from tree sequences inferred by
-        \textbf{(top right)} \tsinfer,
-        \textbf{(bottom left)} Relate, and
-        \textbf{(bottom right)} GEVA+\tsinfer.
+        and \textbf{(bottom)}
+        the dual branch statistic as calculated from tree sequences inferred by Relate~\citep{speidel2019method}.
         Separate curves are plotted for each sampling location.
@@
 since it is mediated by tree shape.
-However, it is generally unknown how much mutation rate varies along the genome
+It is generally unknown how much mutation rate varies along the genome
 \citep[but see][]{supek2015differential}
@@
 with remaining variation explained by direct selection and/or mutation rate variation.
-
-%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Time stratification of statistics}
-
-Thus far we have examined how statistics vary along the genome,
-but a tree sequence allows us access to another dimension: time.
-Next, we examine the \emph{time} distribution of admixture signal
-in the tree sequence inferred by GEVA+\tsinfer.
-The Branch $f_4(A, B; C, D)$ statistic for four sample sets $A$, $B$, $C$, and $D$
-is the sum of $(p_A - p_B)(p_C - p_D)$ across branches, multiplied by span and length,
-where $p_X$ is the proportion of samples in sample set $X$ that inherit from that node.
-A branch with a large value is expected to contribute a large amount to the Site $f_4$,
-which is calculated from genotyping data
-and is used to infer historical connections between populations.
-To visualize the time signature, we assigned the contribution of each branch
-to its terminal node,
-and then summed these values across all nodes in 50 logarithmically-spaced bins by time.
-We then normalised these to sum to 1 by dividing by the total Branch $f_4$ value.
-
-The four-tuple of sample sets for the first two statistics (PUR,TSI;GWD,JPT and ASW,CEU;MSL,CHB)
-are from similar regions of the world,
-and show almost identical time signatures,
-even though the second set of sample sets has a value of $f_4$ nearly eight times higher.
-The second two statistics (IBS,GBR;FIN,JPT and TSI,CEU;FIN,CHB) are also comparable,
-and also show similar signals to each other.
-These curves are noisier, because the $f_4$ values are much smaller.
-For all four comparisons, the largest contribution to $f_4$ comes from
-ancestors living tens of thousands of generations ago.
-\plr{WHAT IS THE TIME UNIT?}
-This is surprising, because $f_4$ is expected to deviate from zero due to asymmetries in tree topology
-during recent history (e.g., more branches ancestral to ASW and MSL than CEU and MSL).
-Furthermore, we expected the different sample sets to have much more distinct time signatures.
-The shared, ancient time signatures may be because nonzero $f_4$ statistics
-are due to differentially sorting ancestral variation,
-or inacurracies in tree sequence inference.
-It is tempting but propably unwarranted to attribute the
-large downward spike between 2000 and 3000 generations in the TSI,CEU;FIN,CHB comparison
-to a historical connection between the ancestors of CEU and the Finnish not shared by the Tuscans
-during that time period.
-Making rigorous conclusions from this sort of visualization
-will require substantial validation work
-to understand the various sources of bias and uncertainty.
-
-\begin{figure}
-    \includegraphics{{figures/1kg/tgp_geva_chr20.node.time.f4}.pdf}
-    \caption{
-        Time distribution of signal for four $f_4$ statistics
-        in the tree sequence for chromosome 20 inferred from 1000 genomes data
-        by GEVA+\tsinfer{}.
-        The branch $f_4$ statistics for these four are:
-        PUR,TSI;GWD,JPT: 1603,
-        ASW,CEU;MSL,CHB: 12587,
-        IBS,GBR;FIN,JPT: -129, and
-        TSI,CEU;FIN,CHB: -96.
-        Population codes are
-        `PUR': Puerto Ricans from Puerto Rico;
-        `TSI': Toscani in Italia;
-        `GWD': Gambian in Western Divisions in the Gambia;
-        `JPT': Japanese in Tokyo, Japan;
-        `ASW': Americans of African Ancestry in SW USA;
-        `CEU': Utah Residents (CEPH) with Northern and Western European Ancestry,
-        `MSL': Mende in Sierra Leone;
-        `CHB': Han Chinese in Beijing, China;
-        `IBS': Iberian Population in Spain;
-        `GBR': British in England and Scotland;
-        and `FIN': Finnish in Finland.
-        \label{fig:node_f4}
-    }
-\end{figure}
 
@@
 
+Many genomic datasets only contain genotype information at a subset of sites
+that do not represent an unbiased look at genome-wide variation,
+introducing an ascertainment bias that difficulty for many applications.
+If, however, unbiased tree sequences can be inferred using these data,
+then Branch statistics could provide unbiased estimates of genome-wide quantities.
+This is in principle possible, as \citet{kelleher2019inferring} inferred tree sequences
+with UK BioBank data, produced using genotyping arrays \citep{bycroft2018genome}.
+This would be similar to imputation, and would likely face similar challenges.
+
 % visualization and time decomposition adding a new dimension
@@
 then dividing the Site statistic by the Branch statistic
-would give us an estimate of relative average mutation rate in windows along the genome.
+could give us an estimate of relative average mutation rate in windows along the genome.
 Similarly, Branch statistics from well-calibrated tree sequences
@@
 
-\begin{figure}
-    \includegraphics[width=0.5\textwidth]{{figures/1kg/relate_chr20.site.1000000.diversity}.pdf}
-    \includegraphics[width=0.5\textwidth]{{figures/1kg/1kg_chr20.branch.1000000.diversity}.pdf}
-    \includegraphics[width=0.5\textwidth]{{figures/1kg/tgp_geva_chr20.branch.1000000.diversity}.pdf}
-    \includegraphics[width=0.5\textwidth]{{figures/1kg/relate_chr20.branch.1000000.diversity}.pdf}
-    \caption{
-        \textbf{(top left)} Mean sequence diversity
-        in 1Mb windows along human chromosome 20 of the 1000 Genomes Project,
-        and the dual branch statistic as calculated from tree sequences inferred by
-        \textbf{(top right)} \tsinfer,
-        \textbf{(bottom left)} GEVA, and
-        \krt{This should be tsinfer+GEVA, no?}
-        \textbf{(bottom right)} Relate.
-        \label{fig:chr20_full}
-    }
-\end{figure}
-
 \end{document}
-
-%%%%%%%%%%%%%%%%%%
-\subsection*{Genomic PCA}
-
-Let $G$ denote the genotype matrix, so that $G_{ja} \in \{0,1\}$
-is the genotype of the $a^\text{th}$ individual
-at the $j^\text{th}$ site, with `1` denoting the derived state
-(we assume here biallelic markers).
-Also let $P_j = (1/n) \sum_a G{ja}$ denote the vector of allele frequencies.
-Then the \emph{genetic covariance} between samples $a$ and $b$ is
-\begin{align*}
-    C_{ab}
-        &= \frac{1}{L} \sum_j (G_{ja} - P_j) (G_{jb} - P_j) . %  \\
-        % &= \frac{1}{L} \left( (G - P \bone^T)^T (G - P \bone^T) \right)_{ab} .
-\end{align*}
-This can be computed by letting
-$\iw_a = \delta_a$ and $\iw_b = \delta_b$
-(these weights mark nodes ancestral to $a$ and $b$ respectively)
-and $\iw_t = \bone / n$
-(this weight gives the proportion of samples inheriting from each node).
-Then, with
-$f(\nw_a, \nw_b, \nw_t) = (\nw_a - \nw_t) (\nw_b - \nw_t)$,
-we can compute the covariance as
-\begin{align*}
-    C_{ab} = \frac{1}{2} \bar \site(f, (\iw_a, \iw_b, \iw_t)) .
-\end{align*}
-This follows, including the factor of two,
-because the summary function applied to the ancestral allele
-at a site with derived allele frequencies $p_a$, $p_b$, and $p_t$
-in sample $a$, sample $b$, and total, respectively, is
-$f(1 - p_a, 1 - p_b, 1 - p_t) = f(p_a, p_b, p_t)$.
-
-Computing the full covariance matrix for a large number of samples may be infeasible.
-However, we can efficiently find $y^T C z$ for arbitrary vectors $y$ and $z$.
-This works because if we set the initial weights to $y$,
-then each allele's weight is equal to the sum of the entries of $y$
-corresponding to samples carrying that allele.
-Then, if we define
-$f_{yz}(u, v, t) = (u - (\sum_a y_a) t) (v - (\sum_b z_b) t)$,
-then
-\begin{align} \label{eqn:yCz}
-    \frac{1}{2} \bar \site(f_{yz}, (y, z, \iw_t))
-        &= \frac{1}{L} \sum_j (\sum_a y_a G_{ja} - (\sum_a y_a) P_j)
-                    (\sum_b z_b G_{jb} - (\sum_b z_b) P_j) \\
-        &= y^T C z .
-\end{align}
-
-\plr{Hmph: to actually compute $Cz$ we need one weight vector for each sample, again.
-    TODO: look up if you can do Krylov just by computing $y^t C z$ for a reasonable number of vectors.
-    We could compute $Cz$ quickly if we could propagate down the tree:
-    we'd have $(Cz)_a$ equal to the sum of the weights at all nodes \emph{ancestral to} $a$. }
-
-Where does the signal from an eigenvector come from?
-Let $y$ be an eigenvector of $C$ with eigenvalue $\lambda$, normalized so that $\|y\|^2 = 1$.
-Since $Cy = \lambda y$,
-we can write $1 = \sum_a y_a^2 = (1/\lambda) y^T C y$.
-However, equation \eqref{eqn:yCz} gives $y^T C y$ as a sum over \emph{sites},
-and thus distributes the variance in $G$ associated with $y$
-across the tree sequence.
-(This is true because of the singular value decomposition of $G$.)
 

### 2019-09-13 node_times.tex

%%%%%%%%%%%%%%%%%%%%%
\subsection*{Time stratification of statistics}

Thus far we have examined how statistics vary along the genome,
but a tree sequence allows us access to another dimension: time.
Next, we examine the \emph{time} distribution of admixture signal
in the tree sequence inferred by GEVA+\tsinfer.
The Branch $f_4(A, B; C, D)$ statistic for four sample sets $A$, $B$, $C$, and $D$
is the sum of $(p_A - p_B)(p_C - p_D)$ across branches, multiplied by span and length,
where $p_X$ is the proportion of samples in sample set $X$ that inherit from that node.
A branch with a large value is expected to contribute a large amount to the Site $f_4$,
which is calculated from genotyping data
and is used to infer historical connections between populations.
To visualize the time signature, we assigned the contribution of each branch
to its terminal node,
and then summed these values across all nodes in 50 logarithmically-spaced bins by time.
We then normalised these to sum to 1 by dividing by the total Branch $f_4$ value.

The four-tuple of sample sets for the first two statistics (PUR,TSI;GWD,JPT and ASW,CEU;MSL,CHB)
are from similar regions of the world,
and show almost identical time signatures,
even though the second set of sample sets has a value of $f_4$ nearly eight times higher.
The second two statistics (IBS,GBR;FIN,JPT and TSI,CEU;FIN,CHB) are also comparable,
and also show similar signals to each other.
These curves are noisier, because the $f_4$ values are much smaller.
For all four comparisons, the largest contribution to $f_4$ comes from
ancestors living tens of thousands of generations ago.
\plr{WHAT IS THE TIME UNIT?}
This is surprising, because $f_4$ is expected to deviate from zero due to asymmetries in tree topology
during recent history (e.g., more branches ancestral to ASW and MSL than CEU and MSL).
Furthermore, we expected the different sample sets to have much more distinct time signatures.
The shared, ancient time signatures may be because nonzero $f_4$ statistics
are due to differentially sorting ancestral variation,
or inacurracies in tree sequence inference.
It is tempting but propably unwarranted to attribute the
large downward spike between 2000 and 3000 generations in the TSI,CEU;FIN,CHB comparison
to a historical connection between the ancestors of CEU and the Finnish not shared by the Tuscans
during that time period.
Making rigorous conclusions from this sort of visualization
will require substantial validation work
to understand the various sources of bias and uncertainty.

\begin{figure}
    \includegraphics{{figures/1kg/tgp_geva_chr20.node.time.f4}.pdf}
    \caption{
        Time distribution of signal for four $f_4$ statistics
        in the tree sequence for chromosome 20 inferred from 1000 genomes data
        by GEVA+\tsinfer{}.
        The branch $f_4$ statistics for these four are:
        PUR,TSI;GWD,JPT: 1603,
        ASW,CEU;MSL,CHB: 12587,
        IBS,GBR;FIN,JPT: -129, and
        TSI,CEU;FIN,CHB: -96.
        Population codes are
        `PUR': Puerto Ricans from Puerto Rico;
        `TSI': Toscani in Italia;
        `GWD': Gambian in Western Divisions in the Gambia;
        `JPT': Japanese in Tokyo, Japan;
        `ASW': Americans of African Ancestry in SW USA;
        `CEU': Utah Residents (CEPH) with Northern and Western European Ancestry,
        `MSL': Mende in Sierra Leone;
        `CHB': Han Chinese in Beijing, China;
        `IBS': Iberian Population in Spain;
        `GBR': British in England and Scotland;
        and `FIN': Finnish in Finland.
        \label{fig:node_f4}
    }
\end{figure}

### 2019-09-13 treestats_paper.tex

@@
 \begin{figure}
-    \includegraphics[width=0.8\textwidth]{{figures/1kg/relate_chr20.site.1000000.region.diversity}.pdf}
-    \includegraphics[width=0.8\textwidth]{{figures/1kg/relate_chr20.branch.1000000.region.diversity}.pdf}
+    \includegraphics{{figures/1kg/relate_chr20.site.1000000.region.diversity}.pdf}
+    \includegraphics{{figures/1kg/relate_chr20.branch.1000000.region.diversity}.pdf}
     \caption{

### 2019-09-13 treestats_paper.tex
Minor edits

@@
 Graph~\citep{griffiths1991two,griffiths1996ancestral,rasmussen2014genome}.
-These genealogical trees can be fairly precisely inferred in nonrecombining regions
+These genealogical trees can be accurately inferred in nonrecombining regions
 \citep{felsenstein2004inferring},
@@
 % visualization and time decomposition adding a new dimension
+\krt{this para needs to go, right, now that the related results are offloaded
+into a new tex file? Perhaps say that we can do this sort of thing?}
 Genomic data are naturally distributed along two axes: along the genome, and across geography.
@@
 The tree sequence describes trees that change as one moves along the genome,
-and so is a special case of a \emph{dynamic graph}
-\citep[also called a \emph{graph timeline}][]{lacki2013reachability}.
+and so is a special case of a \emph{dynamic graph}, 
+also called a \emph{graph timeline} \citep{lacki2013reachability}.
 Much of the work on dynamic graphs is focused on connectivity,
@@
 
+\clearpage
+\newpage
 \section*{Supplementary figures}
 
-\begin{figure}
+\begin{figure}[!h]
     \includegraphics{{figures/swept.10000.765.1e-09.diversity}.pdf}

### 2019-09-13 treestats_paper.tex
Fix "this what"
remove old comment

@@
 
-\krt{There is another approach, introduced by Molly Prz.  I'll see if I can dig up a description.  The approach
-described here is preferable, though.}
-
 \paragraph{Polyallelic sites}
@@
 with UK BioBank data, produced using genotyping arrays \citep{bycroft2018genome}.
-This would be similar to imputation, and would likely face similar challenges.
+This procedure would be similar to imputation, and would likely face similar challenges.
 

### 2019-09-13 treestats_paper.tex
fixed last paragraph about time

@@
 % visualization and time decomposition adding a new dimension
-\krt{this para needs to go, right, now that the related results are offloaded
-into a new tex file? Perhaps say that we can do this sort of thing?}
 Genomic data are naturally distributed along two axes: along the genome, and across geography.
@@
 may shift the focus of these problems from inference to descriptive analysis.
-In this paper, we have made an exploratory foray in this direction,
-by describing the time of origin of the ``admixture'' signal of nonzero $f_4$
-between several subsets of the 1000 Genomes dataset,
-in several inferred tree sequences.
-It is not yet clear whether these time profiles tell us more about history
-or about biases of the inference methods,
-but either result would be useful.
+The methods we develop here distribute the contribution to various statistics across each tree,
+and so could also be used to summarise contributions to various statistics across \emph{time}.
+This could provide, for instance, the time distribution of mutations or branches
+contributing positively or negatively to $f_4$ statistics of introgression,
+which could help historical interpretations of these signals.
+This additional summarization mode, however, is postponed for future work.
 

### 2019-09-15 treestats_paper.tex
added f4 scale explanation

@@
 have per-capita migration rates of $4/N$ to each other.
-In this situation, a significantly negative $f_4(1,2;3,4)$ is expected,
-which we indeed find, with a mean value of around -700.
+We expect a significantly negative $f_4(1,2;3,4)$ in this situation,
+which we indeed find
+(the genome-wide mean Branch $f_4$ is around -700,
+as is the Site $f_4$ divided by mutation rate;
+see Supplemental Figure \ref{fig:introf4} for a plot along the genome).
 Using various sample sizes, mutation rates, and window sizes, we then
@@
 and show the standard deviation of both ``site'' and ``branch'' statistics across windows
-in Figure \ref{fig:introf4}.
+in Figure \ref{fig:introf4sd}.
 Since the genome is uniform (no selection or variation in recombination or mutation),
@@
         and the entire population (1000 diploids in each of four populations) were used.
-        \label{fig:introf4}
+        \label{fig:introf4sd}
     }
@@
 
+\begin{figure}[!h]
+    \includegraphics{{figures/intro/introgressed.1000.23.1000000.1e-09.f4}.pdf}
+    \caption{
+        Values of the $f_4$ statistic in megabase windows along the genome
+        from the simulations of introgression described in the text.
+        Statistics are shown in units of generations:
+        Branch statistics are computed in units of time, which is in generations for these simulations;
+        and Site statistics are reported in units of mutations per unit of sequence length, 
+        which we divide by mutation rate per unit of sequence length and per generation
+        to obtain units of generations.
+        \label{fig:introf4}
+    }
+\end{figure}
+
 \end{document}

### 2019-09-16 treestats_paper.tex
fixed error in branch defn
clarifying units of duality

@@
     &=
-    \sum_{u \in T} f(\nw_{T}(u)) ,
+    \sum_{u \in T} \beta_T(u) f(\nw_{T}(u)) ,
 \end{align}
@@
     =
+    \frac{1}{\mu}
     \E\left[ \site(f, \iw)_{[a,b)} \given \treeseq_{[a,b)} \right] ,
 \end{align}
-where the expected value averages over infinite-sites mutations with rate 1
-(and so lengths are measured in units of expected numbers of mutations)
-but leaves the genealogies fixed.
+where the expected value leaves the genealogies fixed,
+averaging over infinite-sites mutations
+with rate $\mu$ per unit time and per unit of sequence length.
 In other words, the Site statistic can be thought of an estimator of its corresponding Branch statistic,
 which is itself a summary of local tree shapes.
+Because we have normalised these statistics by the length of the genome under consideration,
+both sides of this equation are in units of \emph{time}:
+Branch statistics give mean weighted branch lengths;
+Site statistics give mean densities of mutations per unit of sequence length,
+which divided by $\mu$ is converted to time.
 Let's unpack the assumptions here: what exactly is the mutational model?
-First, we are taking the mutation rate to be 1, i.e.,
+First, we are taking the mutation rate to be $\mu$, i.e.,
 the expected number of mutations that occur on a region of the genome of length $\ell$
-over $t$ units of time is equal to $\ell t$.
-This just amounts to a change of units ---
-for instance, if genome length is measured in nucleotides,
-and the probability of mutation is $\mu$ per nucleotide and per generation,
-then we are measuring times in units of $1/\mu$ generations.
+over $t$ units of time is equal to $\mu \ell t$.
 Second, we are assuming that the probability of per-site mutation is low enough

### 2019-09-16 treestats_paper.tex
branch length units?

@@
         in 1Mb windows along human chromosome 20 of the 1000 Genomes Project,
+        in units of mean number of pairwise differences per nucleotide,
         and \textbf{(bottom)}
-        the dual branch statistic as calculated from tree sequences inferred by Relate~\citep{speidel2019method}.
+        the dual branch statistic as calculated from tree sequences inferred by Relate~\citep{speidel2019method},
+        in units of generations \plr{XXX confirm this}.
         Separate curves are plotted for each sampling location.
+        Note that the Site statistic divided by the Branch statistic is close to $2 \times 10^{-8}$,
+        the mutation rate used by Relate \plr{XXX total speculation here}.
         \label{fig:chr20}

### 2019-09-16 treestats_paper.tex
Minor text tweaks.

@@
     }
-    \krt{This figure is not as helpful as fig 1. It takes a lot of fiddling to work out the values in each panel. Is it
-    possible to fill it out as in fig 1, or is that just too much space?}
 \end{figure}
@@
         and \textbf{(bottom)}
-        the dual branch statistic as calculated from tree sequences inferred by Relate~\citep{speidel2019method},
-        in units of generations \plr{XXX confirm this}.
+        the dual branch statistic as calculated from tree sequences inferred by
+        Relate~\citep{speidel2019method},
+        in units of generations.
         Separate curves are plotted for each sampling location.
@@
 \begin{figure}[!h]
-    \includegraphics{{figures/swept.10000.765.1e-09.diversity}.pdf}
+    \includegraphics[width=\textwidth]{{figures/swept.10000.765.1e-09.diversity}.pdf}
     \caption{
@@
 \begin{figure}[!h]
-    \includegraphics{{figures/intro/introgressed.1000.23.1000000.1e-09.f4}.pdf}
+    \includegraphics[width=\textwidth]{{figures/intro/introgressed.1000.23.1000000.1e-09.f4}.pdf}
     \caption{

### 2019-09-16 treestats_paper.tex
Rejigged the abstract.

@@
 providing a glimpse of the genealogical tree relating the genomes to each other at that site.
-Summaries of genetic variation therefore also describe the underlying genealogies.
-Here we exploit this duality to create a method that efficiently computes single-site statistics
+Statistical summaries of genetic variation therefore also describe the underlying genealogies.
+We exploit this duality here to create a method that efficiently computes single-site statistics
 from population genetic data using the succinct tree sequence encoding
 of genealogies and genome sequence.
-We use the framework to provide efficient implementations of
-many currently-defined statistics, and to suggest several new statistics.
-The general approach aggregates ``sample weights'' up each marginal tree,
-which are then summarised using a ``summary function'';
-different statistics result from different choices of weight or function.
-The results can be summarised in three different ways:
+The general approach accumulates ``sample weights'' within each marginal tree,
+which are then combined using a ``summary function'';
+different statistics result from different choices of weight and function.
+Results can be reported in two (dual) ways:
 by \emph{site}, which corresponds to statistics calculated from genome
-sequence;
-by \emph{branch}, which gives the expected value of the site statistic
-with a large number of mutations under the infinite-sites model;
-or by \emph{node}, which summarises the contribution of each node in the tree to the statistics.
-This formulation also makes it apparent what statistic of the underlying genealogical trees
-each sequence-based statistic is estimating.  By using simulations to evaluate performance, we
-show that calculating statistics from tree sequences is several orders of magnitude more efficient than
-methods traversing the entire data matrix in terms of both run time and memory requirements.
-We also use tree sequences inferred from the 1000 Genomes Project dataset,
-to evaluate how well this duality between site and branch statistics holds
-and to describe the temporal origin of some admixture statistics.
+sequence; and by \emph{branch}, which gives the expected value of the site statistic
+with a large number of mutations under the infinite-sites model.
+We use the framework to implement many currently-defined statistics of genome
+sequence (making the statistics' relationship to the underlying genealogical trees
+concrete and explicit) and suggest several new statistics.
+We evaluate computational performance using simulated data, and
+show that calculating statistics from tree sequences using this general
+framework is several orders of magnitude more efficient than
+bespoke matrix-based methods in terms of both run time and memory requirements.
+We also explore how well the duality between site and branch statistics holds
+in practice on trees inferred from the 1000 Genomes Project dataset,
+and discuss ways in which deviations may hold interesting biological
+signals.
+
+\jk{Not sure if the last bit is expressed the right way}
 

### 2019-09-16 treestats_paper.tex
Change summarise to summarize.

@@
 \title{
-    Efficiently summarising relationships in large samples:
+    Efficiently summarizing relationships in large samples:
     a general duality between statistics of genealogies and genomes}
@@
 and millions of whole genomes will soon be available. In this case the
-genotype matrix simply cannot fit into memory, and complex distributed
-computations are required to calculate population genetic statistics.
+genotype matrix simply cannot fit into memory, and complex distributed computations
+are required.
 
@@
 simulations~\citep{kelleher2018efficient,haller2018tree}, with similar efficiency gains.
-Recent work~\citep{kelleher2019inferring} has shown that tree sequence algorithms can
+Recent work has shown that tree sequence algorithms can
 also be used to massively increase the scalability of methods for inferring
 genome-wide genealogies, and making it possible to infer trees for millions of
-samples.
-The key to the remarkable scalability and efficiency of tree sequence
+samples~\citep{kelleher2019inferring}.
+The key to the remarkable efficiency of tree sequence
 algorithms is the way that shared structure in adjacent trees along the genome is encoded.
@@
 Figure~\ref{fig:divergence_diagram} shows a simple example of these procedures
-of propagating weights up a tree and summarising them,
+of propagating weights up a tree and summarizing them,
 used to calculate mean sequence divergence over a short sequence spanned by a single tree.
@@
 i.e., $\iw(v)$ may be a vector $(\iw_1(v), \ldots, \iw_m(v))$,
-so that when summarising we have access to more than one aspect of each subtree.
+so that when summarizing we have access to more than one aspect of each subtree.
 
@@
 
-Perhaps the most natural way to summarise weights on the tree
+Perhaps the most natural way to summarize weights on the tree
 is simply to examine the values at each node.
@@
 which samples inherit from the node and which do not,
-so we can summarise patterns of genetic variation by summing up weights of all samples
+so we can summarize patterns of genetic variation by summing up weights of all samples
 that carry a given allele.
@@
 % overview: general definition that is mostly functions of jAFS but includes others
-In this paper, we have described a general framework for summarising genetic variation
+In this paper, we have described a general framework for summarizing genetic variation
 and the underlying genealogies
@@
 One way to use this translation is to answer:
-what aspect of tree shape is this population genetics statistic summarising?
+what aspect of tree shape is this population genetics statistic summarizing?
 This can help when interpreting results, especially if reality may not fit an idealised model
@@
 The methods we develop here distribute the contribution to various statistics across each tree,
-and so could also be used to summarise contributions to various statistics across \emph{time}.
+and so could also be used to summarize contributions to various statistics across \emph{time}.
 This could provide, for instance, the time distribution of mutations or branches

### 2019-09-16 treestats_paper.tex
Editing pass.

@@
 
-Before formal definitions, we will give an overview of the general framework.
-Roughly speaking, to compute a statistic we will
-propagate ``weights'' additively up each tree,
-summarize these weights on each branch with a ``summary function'',
-and aggregate these summaries to produce statistics.
-The abstract idea of summarizing weights propagated up each tree
-is only a small generalization of a familiar operations in population genetics:
-summarizing the allele frequency spectrum.
-For instance, we could calculate how many of a certain set $A$ of samples
-inherit from a particular branch in a tree
-by (a) assigning each of these samples weight 1 (and other nodes weight zero), then
-(b) finding the total weight of all nodes in the subtree inheriting from that branch.
-Any mutation falling on this branch would be inherited by
-This gives the number of samples that would inherit any mutation falling on this branch,
-so is analogous to the allele frequency spectrum.
-Figure~\ref{fig:divergence_diagram} shows a simple example of these procedures
-of propagating weights up a tree and summarizing them,
-used to calculate mean sequence divergence over a short sequence spanned by a single tree.
-
-% Many population genetic summary statistics can be defined using the allele frequency spectrum of a collection of homologous sequences.
-If we only considered summaries of the allele frequency spectrum of a single sample,
-our statistics would be equivalent to that described by \citet{fu1995statistical}.
-To connect the framework to familiar notions,
-here is a reframing of \citeauthor{fu1995statistical}'s statistics.
-Write $p_i(a)$ for the frequency of allele $a$ at a site $i$,
-and $h(p)$ for the allele frequency spectrum
---- i.e., the number of alleles whose frequency is $p$ in our sample ---
-and suppose that for some function $f()$, we want to compute $\sum_p h(p) f(p)$.
-(For instance, mean genetic diversity is computed with $f(p) = p (1-p)$.)
-Since this is a simple sum over sites,
-we can compute this by summing over the $L$ loci, as
-\begin{align*}
-    \sum_p h(p) f(p) = \sum_{i=1}^L \sum_a f(p_i(a)).
-\end{align*}
-(Note that since the ancestral allele is included, this function is invariant to relabeling of alleles.)
-If we know where each mutation has occurred on the genealogy,
-we can easily find the frequency of the corresponding allele by counting how many samples
-inherit from that mutation in the tree (and accounting for those that subsequently mutate).
-
-Before defining the statistics, we need a few definitions.
-In a tree sequence, each haplotype, modern or ancestral, is associated with a \emph{node},
-and the trees at each position along the genome have vertexes labeled by these nodes.
-A tree sequence describes the relationships between a special set of nodes, the \emph{samples},
-that appear in the trees at every point along the genome.
-The other (non-sample) nodes may not appear in every tree,
-as for instance if a portion of an ancestor's genome was not inherited by any of the samples.
-
-\begin{definition}[Sample and subtree weights] \label{defn:weights}
-    A list of \emph{sample weights} $\iw$ assigns a numeric value $\iw(v)$
-    to every sample node.
-    Given these and a tree $T$,
-    the \emph{subtree weight} $\nw_T(u)$ for a node $u$ is the sum of all sample weights
-    of every sample node that is descended from $u$ in the tree (including $u$, if it is a sample):
-    \begin{align*}
-        \nw_T(u) = \sum_{v \,:\, v \le_T u} \iw(v) ,
-    \end{align*}
-    where $v \le_T u$ if $u$ is on the path from $v$ to root in the tree $T$.
-    The \emph{total weight} is the sum of the weights over all samples:
-    $\tiw = \sum_v \iw(v)$, which is the subtree weight of the root of any tree
-    that contains all samples.
-\end{definition}
-
-We allow vector-valued weights,
-i.e., $\iw(v)$ may be a vector $(\iw_1(v), \ldots, \iw_m(v))$,
-so that when summarizing we have access to more than one aspect of each subtree.
-
-\begin{definition}[Summary function]
-    For a set of $k$-dimensional weights with total weight $\tiw$,
-    a summary function is a real-valued function $f(w_1, \ldots, w_k)$
-    with the property that $f(0) = f(\tiw) = 0$.
-\end{definition}
-
-The requirement that $f(0) = f(\tiw) = 0$ is for consistency,
-so that statistics do not depend on portions of the tree either not ancestral to any of the samples or ancestral to all
-of them.
-% Below, in section XXX, we describe how to efficiently
-% maintain a list of the weights of the subtrees inherit from each node
-% as we iterate through the tree sequence.
-
 \begin{figure}
@@
 \end{figure}
+
+Before formal definitions, we will give an overview of the general framework.
+Roughly speaking, to compute a statistic we will
+propagate ``weights'' additively up each tree,
+summarize these weights on each branch with a ``summary function'',
+and aggregate these summaries to produce statistics.
+The abstract idea of summarizing weights propagated up each tree
+is only a small generalization of a familiar operations in population genetics:
+summarizing the allele frequency spectrum.
+For instance, we could calculate how many of a certain set $A$ of samples
+inherit from a particular branch in a tree
+by (a) assigning each of these samples weight 1 (and other samples weight zero), then
+(b) finding the total weight of all samples in the subtree inheriting from that branch.
+This gives the number of samples that would inherit any mutation falling on this branch,
+so is analogous to the allele frequency spectrum.
+Figure~\ref{fig:divergence_diagram} shows a simple example of these procedures
+of propagating weights up a tree and summarizing them,
+used to calculate mean sequence divergence over a short sequence spanned by a single tree.
+
+% Many population genetic summary statistics can be defined using the allele frequency spectrum of a collection of homologous sequences.
+If we only considered summaries of the allele frequency spectrum of a single sample,
+our statistics would be equivalent to that described by \citet{fu1995statistical}.
+To connect the framework to familiar notions,
+here is a reframing of \citeauthor{fu1995statistical}'s statistics.
+Write $p_i(a)$ for the frequency of allele $a$ at a site $i$,
+and $h(p)$ for the allele frequency spectrum
+--- i.e., the number of alleles whose frequency is $p$ in our sample ---
+and suppose that for some function $f()$, we want to compute $\sum_p h(p) f(p)$.
+(For instance, mean genetic diversity is computed with $f(p) = p (1-p)$.)
+Since this is a simple sum over sites,
+we can compute this by summing over the $L$ loci, as
+\begin{align*}
+    \sum_p h(p) f(p) = \sum_{i=1}^L \sum_a f(p_i(a)).
+\end{align*}
+% (Note that since the ancestral allele is included, this function is invariant to relabeling of alleles.)
+If we know where each mutation has occurred on the genealogy,
+we can easily find the frequency of the corresponding allele by counting how many samples
+inherit from that mutation in the tree (and accounting for those that subsequently mutate).
+
+Before defining the statistics, we need a few definitions.
+In a tree sequence, each haplotype, modern or ancestral, is associated with a \emph{node},
+and the trees at each position along the genome have vertexes labeled by these nodes.
+A tree sequence describes the relationships between a special set of nodes, the \emph{samples},
+that appear in the trees at every point along the genome.
+The other (non-sample) nodes may not appear in every tree,
+as for instance if a portion of an ancestor's genome was not inherited by any of the samples.
+
+\begin{definition}[Sample and subtree weights] \label{defn:weights}
+    A list of \emph{sample weights} $\iw$ assigns a numeric value $\iw(v)$
+    to every sample node.
+    Given these and a tree $T$,
+    the \emph{subtree weight} $\nw_T(u)$ for a node $u$ is the sum of all sample weights
+    of every sample node that is descended from $u$ in the tree (including $u$, if it is a sample):
+    \begin{align*}
+        \nw_T(u) = \sum_{v \,:\, v \le_T u} \iw(v) ,
+    \end{align*}
+    where $v \le_T u$ if $u$ is on the path from $v$ to root in the tree $T$.
+    The \emph{total weight} is the sum of the weights over all samples:
+    $\tiw = \sum_v \iw(v)$, which is the subtree weight of the root of any tree
+    that contains all samples.
+\end{definition}
+
+We allow vector-valued weights,
+i.e., $\iw(v)$ may be a vector $(\iw_1(v), \ldots, \iw_m(v))$,
+so that when summarizing we have access to more than one aspect of each subtree.
+We will often be interested in statistics defined in terms of different
+subsets of our samples (corresponding to, e.g., different populations); it
+is useful to have some additional notation to define these statistics.
+Given a subset set $S$ of samples,
+the \emph{indicator weights of $S$} are the sample weights $\bone_S$ with
+$\bone_S(u) = 1$ if $u \in S$ and $\bone_S(u) = 0$ otherwise.
+
+\begin{definition}[Summary function]
+    For a set of $k$-dimensional weights with total weight $\tiw$,
+    a summary function is a real-valued function $f(w_1, \ldots, w_k)$
+    with the property that $f(0) = f(\tiw) = 0$.
+\end{definition}
+
+The requirement that $f(0) = f(\tiw) = 0$ is for consistency,
+so that statistics do not depend on portions of the tree either not ancestral to any of the samples or ancestral to all
+of them.
+% Below, in section XXX, we describe how to efficiently
+% maintain a list of the weights of the subtrees inherit from each node
+% as we iterate through the tree sequence.
 
@@
 \end{align}
-
-We will use weights that count number of samples inheriting from each node
-sufficiently often that we give them a special name ---
-for a subset set $S$ of samples,
-the \emph{indicator weights of $S$} are the sample weights $\bone_S$ with
-$\bone_S(u) = 1$ if $u \in S$ and $\bone_S(u) = 0$ otherwise.
 
@@
 
+\jk{What does this last sentence mean?}
+
 \begin{example}[Mean TMRCA] \label{ex:branch_diversity}
@@
 \begin{figure}
-    \includegraphics{{figures/swept.1000.999.1e-09.diversity}.pdf}
+\centering
+    \includegraphics[width=0.8\textwidth]{{figures/swept.1000.999.1e-09.diversity}.pdf}
     \caption{
@@
 We do not know the true genealogies underlying real data,
-but recent methods are available to estimate them at scale.
+but recent methods are available to estimate them at
+scale~\citep{harris2019database,kelleher2019inferring,speidel2019method}.
 In Figure~\ref{fig:sweep_duality}, we showed that branch and site statistics
@@
 the underlying tree topologies and branch lengths are exactly
-correct, which no inference method can hope to achieve.
+correct (which inference methods can only hope to approximate).
 To evaluate the correspondence between site and branch statistics
 in tree sequences inferred from real data,
-we calculated mean sequence diversity along chromosome 20 from the 1000 Genomes dataset
-in 1Mb windows separately in each of the five continental groupings,
-calculated using the site statistic described in Example~\ref{ex:site_diversity},
-and compared this to the dual branch statistic (of Example~\ref{ex:branch_diversity})
-from the trees and node times estimated by Relate~\citep{speidel2019method}.
+we calculated statistics for chromosome 20
+of the 1000 Genomes dataset~\citep{1k2015global} using dated trees estimated
+by Relate~\citep{speidel2019method}.
+Specifically, we calculated diveristy using the site statistic
+described in Example~\ref{ex:site_diversity},
+and compared this to the dual branch statistic (Example~\ref{ex:branch_diversity})
+in 1Mb windows in each of the five continental groupings.
 All calculations were done only using the portions of the chromosome
@@
 As shown in Figure~\ref{fig:chr20}, Branch and Site diversity along the genome are very similar.
-Does this mean that Relate is doing a better job at inferring the tree sequence
 Is this an indication that Relate is doing a good job at inferring trees,
@@
 For this reason, it is very interesting to see how close a correspondence between site and branch
-statistics it is \emph{possible} to obtain --
+statistics it is \emph{possible} to obtain ---
 it is tempting to say that the best tree sequence would obtain
@@
 In this section we describe and analyse the algorithm used
-to compute branch length statistics. This is a generalization
+to compute Branch statistics. This is a generalization
 of the algorithm used to maintain the number of samples from a
@@
 
-\algstep{B2.}{Terminate.}{If $j \geq |\Et|$ or $t_l \geq L$,
-    return $\sigma / L$ and terminate.
+\algstep{B2.}{Terminate.}{If $j = |\Et|$ return $\sigma / L$ and terminate.
 }
 
-\algstep{B3.}{Edge removal loop.}{If $k \geq |\Et|$ or $t_l \neq
+\algstep{B3.}{Edge removal loop.}{If $k = |\Et|$ or $t_l \neq
     \Et[\indexout_k]\prop{right}$ go to \algref{B6}.}
@@
 
-\algstep{B6.}{Edge insertion loop.}{If $j \geq |\Et|$ or $t_l \neq
+\algstep{B6.}{Edge insertion loop.}{If $j = |\Et|$ or $t_l \neq
     \Et[\indexin_j]\prop{left}$ go to \algref{B9}.}
@@
 in the current tree. The first thing we do for a new tree is to
-remove any edges that were in the previous tree are not in the
+remove any edges that were in the previous and tree are not in the
 current tree. These must be the edges in which the right coordinate
@@
 next tree. If we have reached the end of the sequence, we then divide $\sigma$
-by the sequence length to renormalise and exit.
+by the sequence length to normalise and exit.
 
@@
 \begin{equation}\label{eqn:alg_complexity}
-    O\left(n + \rho \log(n)^2 \right)
+    O\left(n + \rho \log(n)^2 \right).
 \end{equation}
@@
 \autoref{fig:stats_performance} shows that the tree sequence methods implemented in \texttt{tskit} processes variant data
-substantially faster than matrix-based approaches once the sample size is above $n \approx 500$ haploids --
+substantially faster than matrix-based approaches once the sample size is above $n \approx 500$ haploids ---
 three times faster for one thousand haplotypes,
@@
 $\approx 32,000$ to $\approx 138,000$ \citep{Watterson1975-ej}.
-\autoref{fig:stats_performance} only considers the time required to calculate the statistic,
-and does not include the time required to build the genotype matrix, nor the memory required to store it.
+\autoref{fig:stats_performance} only considers the time required to calculate the statistic
+once the data is present in each library's native format.
 For the largest samples size of $n=10^6$,
@@
 All statistics described here (and more) are implemented
-in the fast, rigorously tested library \tskit{}
-that also provides tools for working with tree sequences in Python and C.
+in the rigorously tested \tskit{} library,
+which provides a suite tools for efficiently working with tree sequences in Python and C.
 The largest efficiency gains are found for very large sample sizes
@@
 is useful for understanding evolutionary simulations of any species
-\citep[see][for a recent example]{galloway2019stickleback}.
+\citep[see, e.g.,][]{martin2017human,galloway2019stickleback}.
 
@@
 One way to use this translation is to answer:
-what aspect of tree shape is this population genetics statistic summarizing?
+what aspect of tree shape is a particular population genetics statistic summarizing?
 This can help when interpreting results, especially if reality may not fit an idealised model
@@
 This might plausibly produce less noisy estimates because tree sequence inference
-essentially uses the signal from nearby patterns of variation
+may use the signal from nearby patterns of variation
 to interpret relationships at a given site,
@@
 Thanks to Graham Coop, Andrew Kern, Nick Patterson, and Kelley Harris for useful suggestions
-and to Wilder Wohns, Yan Wong and Leo Speidel for providing inferred tree sequences.
+and Leo Speidel for providing the inferred 1000 Genomes tree sequence.
 

### 2019-09-16 treestats_paper.tex
Spell check with EN_us.

@@
 
-\paragraph{Polarisation}
+\paragraph{Polarization}
 The node statistic defined above is appropriate for statistics such as nucleotide diversity, which is a function
 of all alleles at given variable site.  For other statistics, one is only concerned with the derived state(s)
-at a site, and we may define a \emph{polarised} node statistic without the second term from our previous definition:
+at a site, and we may define a \emph{polarized} node statistic without the second term from our previous definition:
 
 \begin{align}
-    \text{(polarised)} \qquad
+    \text{(polarized)} \qquad
     \nodep(f, \iw)_u
@@
 Analogously to the above,
-we say a site statistic is \textbf{polarised} if we do this,
+we say a site statistic is \textbf{polarized} if we do this,
 defining
 \begin{align} \label{eqn:site_polarised}
-    \text{(polarised)} \qquad
+    \text{(polarized)} \qquad
     \sitep(f, \iw)_j
@@
     (shown in labels at the bottom).
-    Note that each of the mutations shown occurs at a distinct site, so there is no backmutation.
+    Note that each of the mutations shown occurs at a distinct site, so there is no back mutation.
     \textbf{(b)} Each branch is assigned weight equal to the weight that would be assigned
@@
 then $\ell_k(c,d) = \max(0, \min(d,a_k) - \max(c,a_{k-1}))$).
-The \emph{polarised} version of a Branch statistic
+The \emph{polarized} version of a Branch statistic
 is defined analogously to the Node and Site versions:
 \begin{align} \label{eqn:branch_polarised}
-    \text{(polarised)} \qquad
+    \text{(polarized)} \qquad
     \branchp(f, \iw)_T
@@
 and $\branchp(f, \iw)_{[c,d)}$ is defined using $\branchp(f, \iw)_T$ as before.
-However, the correspondence between the polarised statistics is less tight
-than for the unpolarised ones.
+However, the correspondence between the polarized statistics is less tight
+than for the unpolished ones.
 
@@
 which is itself a summary of local tree shapes.
-Because we have normalised these statistics by the length of the genome under consideration,
+Because we have normalized these statistics by the length of the genome under consideration,
 both sides of this equation are in units of \emph{time}:
@@
 by Relate~\citep{speidel2019method}.
-Specifically, we calculated diveristy using the site statistic
+Specifically, we calculated diversity using the site statistic
 described in Example~\ref{ex:site_diversity},
@@
 \subsection*{Algorithms}
-In this section we describe and analyse the algorithm used
+In this section we describe and analyze the algorithm used
 to compute Branch statistics. This is a generalization
@@
 summary function $f: \R \rightarrow \R$
-compute the span-normalised statistic $\sigma$
+compute the span-normalized statistic $\sigma$
 over a tree sequence with sequence length $L$
@@
 
-\algstep{B1.}{Initialisation.}{
+\algstep{B1.}{Initialization.}{
     For $0 \leq u < |\Et|$ set $\beta_u \leftarrow x_u
@@
 used to keep track of our position in the edge insertion and edge
-removal indexes, respectively. After completing initialisation in
+removal indexes, respectively. After completing initialization in
 \algref{B1}, we enter the main tree loop in \algref{B2} which is
@@
 next tree. If we have reached the end of the sequence, we then divide $\sigma$
-by the sequence length to normalise and exit.
-
-To analyse Algorithm~\algref{B} we will assume that the tree sequence has
+by the sequence length to normalize and exit.
+
+To analyze Algorithm~\algref{B} we will assume that the tree sequence has
 been simulated under the standard coalescent with a sample of size
@@
 what aspect of tree shape is a particular population genetics statistic summarizing?
-This can help when interpreting results, especially if reality may not fit an idealised model
+This can help when interpreting results, especially if reality may not fit an idealized model
 of separate populations.
@@
 \citep[as used by][]{hudson2001twolocus},
-such as linkage disequlibrium are planned for future work.
+such as linkage disequilibrium are planned for future work.
 Haplotype-based statistics may require different classes of algorithms.
@@
 (i.e., without knowledge of the trees),
-and, if the statistic is polarised, knowledge of the ancestral allele.
+and, if the statistic is polarized, knowledge of the ancestral allele.
 We often would also like our statistics to be unaffected by the addition of monomorphic sites,
@@
 \end{align*}
-Writing this in block matrix notation, $a$ and $b$ minimise
+Writing this in block matrix notation, $a$ and $b$ minimize
 \begin{align*}
@@
 then decorrelate the resulting matrix, so that now $Z^T Z = I$.
-Then, put this normalised version of $Z$
+Then, put this normalized version of $Z$
 into the first $k$ columns of the weight matrix (so that $w_j(u) = Z_{uj}$),

### 2019-09-16 treestats_paper.tex
Tweaks.

@@
 that do not represent an unbiased look at genome-wide variation,
-introducing an ascertainment bias that difficulty for many applications.
+introducing an ascertainment bias that is difficult to account for in for many applications.
 If, however, unbiased tree sequences can be inferred using these data,
@@
 which could help historical interpretations of these signals.
-This additional summarization mode, however, is postponed for future work.
+This additional mode of reporting statistics, however, is postponed for future work.
 
@@
 Both our data and our understanding of the world have moved beyond this.
-We hope that the tools we develop here will help make it possible
+We hope that the tools developed here will help make it possible
 to visualize and analyze genetic variation and genealogical relatedness

### 2019-09-16 treestats_paper.tex
Review comments from Peter.

@@
 sequence; and by \emph{branch}, which gives the expected value of the site statistic
-with a large number of mutations under the infinite-sites model.
+under the infinite-sites model.
 We use the framework to implement many currently-defined statistics of genome
@@
 framework is several orders of magnitude more efficient than
-bespoke matrix-based methods in terms of both run time and memory requirements.
+optimised matrix-based methods in terms of both run time and memory requirements.
 We also explore how well the duality between site and branch statistics holds
@@
 signals.
-
-\jk{Not sure if the last bit is expressed the right way}
 
@@
 However, the correspondence between the polarized statistics is less tight
-than for the unpolished ones.
+than for the unpolarized ones.
 
@@
 in the current tree. The first thing we do for a new tree is to
-remove any edges that were in the previous and tree are not in the
+remove any edges that were in the previous tree and are not in the
 current tree. These must be the edges in which the right coordinate
@@
 \subsection*{Acknowledgments}
-Thanks to Graham Coop, Andrew Kern, Nick Patterson, and Kelley Harris for useful suggestions
+Thanks to Graham Coop, Andrew Kern, Nick Patterson, Kelley Harris and
+Aaron Ragsdale for useful suggestions
 and Leo Speidel for providing the inferred 1000 Genomes tree sequence.

### 2019-09-17 treestats_paper.tex
Minor updates and notes.

@@
 Statistical summaries of genetic variation therefore also describe the underlying genealogies.
-We exploit this duality here to create a method that efficiently computes single-site statistics
-from population genetic data using the succinct tree sequence encoding
-of genealogies and genome sequence.
+We exploit this correspondence here to define a general framework that
+efficiently compute single-site population genetic statistics
+using the succinct tree sequence encoding of genealogies and genome sequence.
 The general approach accumulates ``sample weights'' within each marginal tree,
@@
 different statistics result from different choices of weight and function.
-Results can be reported in two (dual) ways:
+Results can be reported in two dual ways:
 by \emph{site}, which corresponds to statistics calculated from genome
@@
 framework is several orders of magnitude more efficient than
-optimised matrix-based methods in terms of both run time and memory requirements.
+optimized matrix-based methods in terms of both run time and memory requirements.
 We also explore how well the duality between site and branch statistics holds
 in practice on trees inferred from the 1000 Genomes Project dataset,
-and discuss ways in which deviations may hold interesting biological
+and discuss ways in which deviations may encode interesting biological
 signals.
@@
 It was introduced in the context of coalescent
-simulation~\citep{kelleher2016efficient}, and led to scalability increases of
+simulation~\citep{kelleher2016efficient}, and lead to scalability increases of
 several orders of magnitude over existing methods. The methods were subsequently extended
 and refined for forwards-time
-simulations~\citep{kelleher2018efficient,haller2018tree}, with similar efficiency gains.
+simulations~\citep{kelleher2018efficient,haller2018tree}, with similarly large efficiency gains.
 Recent work has shown that tree sequence algorithms can
@@
 is useful for understanding evolutionary simulations of any species
-\citep[see, e.g.,][]{martin2017human,galloway2019stickleback}.
+\citep[see, e.g.,][]{martin2017human,browning2018one,galloway2019stickleback}.
+% TODO rephrase this: large scale simulations are useful for lots of things.
+% No need for the slightly negative note about large datasets only being
+% relevant for humans, I think.
 
@@
 %  maybe analogous to imputation
+
+% TODO: too negative.
 We hope a more broadly interesting and useful aspect of this paper
@@
 
+% Not sure about this: tangent? Perhaps combine this point with some others
+% in a compressed form?
 Many genomic datasets only contain genotype information at a subset of sites
@@
 which could help historical interpretations of these signals.
-This additional mode of reporting statistics, however, is postponed for future work.
+This additional mode of reporting statistics is an important subject for future work.
 

### 2019-09-17 treestats_paper.tex
Moved Duality in Practise and Dynamic Graph sections out of disco.

@@
 
+% This paragraph is essentially saying the same thing as the previous one,
+% so am cutting it. Perhaps we could merge some of the content into the
+% the paragtraph above?
+
+% The duality between Site and Branch statistics depends crucially on two assumptions:
+% that each new mutation is distinguishable, as in the infinite-sites model,
+% and that mutation rates are constant along the genome.
+% Neither of these things are true, but are they good approximations?
+% Multiple mutations at the same site, causing back mutations,
+% are present in real datasets but are generally thought to be uncommon,
+% especially if hypervariable sites are excluded.
+% The effect of multiple mutations on Site statistics
+% can be bounded by a function of the proportion of sites at which these occur.
+% Large-scale variation in mutation rate (at, say, the megabase scale)
+% might plausibly have a stronger effect on the correspondence,
+% because a Branch and Site statistic are equal in expectation up to multiplication by the mutation rate.
+% For instance, if we knew that the inferred tree sequences of \autoref{fig:chr20} were correct,
+% then dividing the Site statistic by the Branch statistic
+% could give us an estimate of relative average mutation rate in windows along the genome.
+% Similarly, Branch statistics from well-calibrated tree sequences
+% would provide a way to assess variation in population genetic summaries along the genome
+% in a way independent of local mutation rate.
+
+A final crucial assumption underlying the duality between Site and Branch
+statistics is that mutations can be treated as independent of the trees themselves.
+This is clearly not always true --- for instance,
+a sweeping beneficial mutation strongly affects the local trees (as in \autoref{fig:sweep_duality}) ---
+but seems likely to be approximately true.
+\citet{ralph2019empirical} contains more discussion of variation in mutation rate,
+but more work needs to be done, particularly on models with large proportions of sites under selection.
+
 %%%%%%%%%%%%%%%%%%
@@
 
+The tree sequence describes trees that change as one moves along the genome,
+and so is a special case of a \emph{dynamic graph},
+also called a \emph{graph timeline} \citep{lacki2013reachability}.
+Much of the work on dynamic graphs is focused on connectivity,
+e.g., maintaining a minimum spanning tree \citep{eppstein1994offline, eppstein1997sparsification, holm2001polylogarithmic},
+but development of parallel algorithms for more general operations on large dynamic graphs
+is an active area of research \citep[e.g.,][]{srinivasan2018sharedmemory}.
+An interesting direction for future work is to develop
+parallel versions of tree sequence algorithms such as Algorithm B.
+
 %%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%
-\subsection*{Duality in practice?}
-
-The duality between Site and Branch statistics depends crucially on two assumptions:
-that each new mutation is distinguishable, as in the infinite-sites model,
-and that mutation rates are constant along the genome.
-Neither of these things are true, but are they good approximations?
-Multiple mutations at the same site, causing back mutations,
-are present in real datasets but are generally thought to be uncommon,
-especially if hypervariable sites are excluded.
-The effect of multiple mutations on Site statistics
-can be bounded by a function of the proportion of sites at which these occur.
-Large-scale variation in mutation rate (at, say, the megabase scale)
-might plausibly have a stronger effect on the correspondence,
-because a Branch and Site statistic are equal in expectation up to multiplication by the mutation rate.
-For instance, if we knew that the inferred tree sequences of \autoref{fig:chr20} were correct,
-then dividing the Site statistic by the Branch statistic
-could give us an estimate of relative average mutation rate in windows along the genome.
-Similarly, Branch statistics from well-calibrated tree sequences
-would provide a way to assess variation in population genetic summaries along the genome
-in a way independent of local mutation rate.
-
-A final crucial assumption underlying this duality is that mutations can be treated
-as independent of the trees themselves.
-This is clearly not always true --- for instance,
-a sweeping beneficial mutation strongly affects the local trees (as in \autoref{fig:sweep_duality}) ---
-but seems likely to be approximately true.
-\citet{ralph2019empirical} contains more discussion of variation in mutation rate,
-but more work needs to be done, particularly on models with large proportions of sites under selection.
+% \subsection*{Duality in practice?}
+% Moved this to the 1000G section where we're already discussing it.
 
@@
 %%%%%%%%%%%%%%%%%%
-\subsection*{Dynamic graph algorithms}
-
-The tree sequence describes trees that change as one moves along the genome,
-and so is a special case of a \emph{dynamic graph},
-also called a \emph{graph timeline} \citep{lacki2013reachability}.
-Much of the work on dynamic graphs is focused on connectivity,
-e.g., maintaining a minimum spanning tree \citep{eppstein1994offline, eppstein1997sparsification, holm2001polylogarithmic},
-but development of parallel algorithms for more general operations on large dynamic graphs
-is an active area of research \citep[e.g.,][]{srinivasan2018sharedmemory}.
+% Moved this content into the "Algorithms" section
+% \subsection*{Dynamic graph algorithms}
 

### 2019-09-17 treestats_paper.tex
Partial refactor of disco.

@@
 and can be used to quantify associations between genotypes and phenotypes.
-This generality was motivated in part by laziness ---
-adding new statistics is now almost as easy as working out the appropriate summary function ---
-but also makes it easy to computationally explore new classes of statistics.
+This generality was partially motivated by laziness ---
+implementing new statistics is now simply a matter of devising the appropriate summary function ---
+but it also allows us to easily explore new classes of statistics.
 The range of statistics available to describe variation in a single exchangeable sample
@@
 or across geographical space.
+The statistics defined here are all \emph{additive} along the genome:
+if we have computed a statistic in equal-size windows along a chromosome,
+then the value of the statistic for the entire chromosome is equal to
+the average of the values in those windows.
+Some commonly-used statistics (e.g., $F_{ST}$ or Tajima's $D$)
+are not additive, but are ratios of additive statistics, so can be easily computed in this way.
+% We have also defined site statistics to be additive across \emph{alleles},
+% to allow the definitions to apply to sites with more than two alleles.
+% This decision has some nice properties, but is not the only reasonable choice.
+% Multilocus statistics that measure linkage disequilibrium or haplotype lengths
+% do not fall in this category.
+Extensions to statistics involving the pairwise joint distribution of genotypes across
+sites~\citep{hudson2001twolocus} such as linkage disequilibrium are planned for future work.
+Haplotype-based statistics may require different classes of algorithms.
+
+% jk: Cut this: too technical for the discussion
+% We also require that site statistics be computable from the genotype matrix
+% (i.e., without knowledge of the trees),
+% and, if the statistic is polarized, knowledge of the ancestral allele.
+% We often would also like our statistics to be unaffected by the addition of monomorphic sites,
+% which are usually omitted from genetic variation data.
+% To ensure this we have required that the summary function returns zero when give both zero weight
+% and the total weight: $f(0) = f(\tiw) = 0$.
+% This also makes branch and node statistics unchanged by portions of the tree that do not lie on a path
+% between any of the samples that have nonzero weight,
+% so that in particular,
+% \emph{simplification} of the tree sequence will not change the value of the statistic.
+% However, in practice this requirement can be relaxed.
 
@@
 %%%%%%%%%%%%%%%%%%
-\subsection*{General properties}
-
-The statistics defined here are all \emph{additive} along the genome:
-if we have computed a statistic in equal-size windows along a chromosome,
-then the value of the statistic for the entire chromosome is equal to
-the average of the values in those windows.
-Some commonly-used statistics (e.g., $F_{ST}$ or Tajima's $D$)
-are not additive, but are ratios of additive statistics, so can be easily computed in this way.
-We have also defined site statistics to be additive across \emph{alleles},
-to allow the definitions to apply to sites with more than two alleles.
-This decision has some nice properties, but is not the only reasonable choice.
-
-Multilocus statistics that measure linkage disequilibrium or haplotype lengths
-do not fall in this category.
-Extensions to statistics involving the pairwise joint distribution of genotypes across sites
-\citep[as used by][]{hudson2001twolocus},
-such as linkage disequilibrium are planned for future work.
-Haplotype-based statistics may require different classes of algorithms.
-
-We also require that site statistics be computable from the genotype matrix
-(i.e., without knowledge of the trees),
-and, if the statistic is polarized, knowledge of the ancestral allele.
-We often would also like our statistics to be unaffected by the addition of monomorphic sites,
-which are usually omitted from genetic variation data.
-To ensure this we have required that the summary function returns zero when give both zero weight
-and the total weight: $f(0) = f(\tiw) = 0$.
-This also makes branch and node statistics unchanged by portions of the tree that do not lie on a path
-between any of the samples that have nonzero weight,
-so that in particular,
-\emph{simplification} of the tree sequence will not change the value of the statistic.
-However, in practice this requirement can be relaxed.
+% \subsection*{General properties}
+
+% Merged in with disco content above.
 

### 2019-09-18 treestats_paper.tex
Strike confusing sentence.
Tightened up discussion.

@@
 and $\branchp(f, \iw)_{[c,d)}$ is defined using $\branchp(f, \iw)_T$ as before.
-However, the correspondence between the polarized statistics is less tight
-than for the unpolarized ones.
-
-\jk{What does this last sentence mean?}
 
@@
 % overview: general definition that is mostly functions of jAFS but includes others
-In this paper, we have described a general framework for summarizing genetic variation
+In this paper we have described a general framework for summarizing genetic variation
 and the underlying genealogies
-that encompasses many standard population genetic statistics
-and makes it easy to explore more.
+that encompasses many standard population genetic statistics.
 Many of these statistics are functions of the joint allele frequency spectrum,
 but the framework is more general,
-and can be used to quantify associations between genotypes and phenotypes.
-This generality was partially motivated by laziness ---
-implementing new statistics is now simply a matter of devising the appropriate summary function ---
-but it also allows us to easily explore new classes of statistics.
+and can be used, for example, to quantify associations between genotypes and phenotypes.
+This generality greatly reduces the software development effort in
+implementing statistics efficiently,
+and it also allows users to easily explore new classes of statistics.
 The range of statistics available to describe variation in a single exchangeable sample
@@
 or across geographical space.
-The statistics defined here are all \emph{additive} along the genome:
+The statistics defined here are all additive along the genome:
 if we have computed a statistic in equal-size windows along a chromosome,
@@
 
-% duality: site stats ~ branch stat, but conversely branch more accurate version of site
-%  maybe analogous to imputation
-
 % TODO: too negative.
-We hope a more broadly interesting and useful aspect of this paper
-is a more in-depth look at the correspondence between statistics calculated with genome sequence
-and properties of the underlying genealogies, i.e.,
-the duality between Site and Branch statistics under the infinite-sites model.
-This general correspondence is well-known,
-but the common framework implemented here makes it easy to translate between the two.
-One way to use this translation is to answer:
+% We hope a more broadly interesting and useful aspect of this paper
+% is a more in-depth look at the correspondence between statistics calculated with genome sequence
+% and properties of the underlying genealogies, i.e.,
+% the duality between Site and Branch statistics under the infinite-sites model.
+The correspondence between genome sequence and properties of the
+underlying genealogies we have used here is well-known. The general framework
+that we have defined, however, codifies this relationship in a
+mathematically elegant and computationally efficient way, and may lead to
+new perspectives on well-studied problems.
+One way to use the duality between Site and Branch statistics is to answer:
 what aspect of tree shape is a particular population genetics statistic summarizing?
@@
 This might plausibly produce less noisy estimates because tree sequence inference
-may use the signal from nearby patterns of variation
+should use the signal from nearby patterns of variation
 to interpret relationships at a given site,
 thus transforming the simple binary split induced by a SNP
-into a much richer source of information at every variant site.
+into a much richer source of information.
 Furthermore,
@@
 that is not affected by this potentially important confounding factor.
-
+Similarly, if tree sequences inferred from
+genotype array data~\citep{kelleher2019inferring} are unbiased,
+then Branch statistics could provide a way
+to estimate genome-wide quantities without ascertainment bias.
+This procedure would be similar to imputation, and would likely face similar challenges.
+
+% duality: site stats ~ branch stat, but conversely branch more accurate version of site
+%  maybe analogous to imputation
 % Not sure about this: tangent? Perhaps combine this point with some others
 % in a compressed form?
-Many genomic datasets only contain genotype information at a subset of sites
-that do not represent an unbiased look at genome-wide variation,
-introducing an ascertainment bias that is difficult to account for in for many applications.
-If, however, unbiased tree sequences can be inferred using these data,
-then Branch statistics could provide unbiased estimates of genome-wide quantities.
-This is in principle possible, as \citet{kelleher2019inferring} inferred tree sequences
-with UK BioBank data, produced using genotyping arrays \citep{bycroft2018genome}.
-This procedure would be similar to imputation, and would likely face similar challenges.
+% Many genomic datasets only contain genotype information at a subset of sites
+% that do not represent an unbiased look at genome-wide variation,
+% introducing an ascertainment bias that is difficult to account for in for many applications.
+% If, however, unbiased tree sequences can be inferred using these data,
+% then Branch statistics could provide unbiased estimates of genome-wide quantities.
+% This is in principle possible, as \citet{kelleher2019inferring} inferred tree sequences
+% with UK BioBank data, produced using genotyping arrays \citep{bycroft2018genome}.
+% This procedure would be similar to imputation, and would likely face similar challenges.
 
@@
 may shift the focus of these problems from inference to descriptive analysis.
-The methods we develop here distribute the contribution to various statistics across each tree,
-and so could also be used to summarize contributions to various statistics across \emph{time}.
+The methods developed here distribute the contribution to various statistics across each tree,
+and so could also be used to summarize contributions to various statistics across time.
 This could provide, for instance, the time distribution of mutations or branches
 contributing positively or negatively to $f_4$ statistics of introgression,
-which could help historical interpretations of these signals.
-This additional mode of reporting statistics is an important subject for future work.
-
+enabling historical interpretations of these signals.
+% This additional mode of reporting statistics is an important subject for future work.
 % moving past a few summaries in discrete groups to a genome/time/space view of relatedness
-The historical toolbox of population genetics
+The computational toolbox of population genetics
 is still mostly composed of statistics designed for analysis of a handful of loci

### 2019-09-18 treestats_paper.tex
Updated intro and discussion perf aspects.

@@
 Statistical summaries of genetic variation therefore also describe the underlying genealogies.
-We exploit this correspondence here to define a general framework that
-efficiently compute single-site population genetic statistics
+We use this correspondence to define a general framework that
+efficiently computes single-site population genetic statistics
 using the succinct tree sequence encoding of genealogies and genome sequence.
@@
 It was once a major undertaking to collect data sufficient to estimate a single summary
-of the many genetic relationships between the individuals of a sample,
-e.g., to estimate levels of sequence diversity from a few dozen alleles at an allozyme locus
-\citep{Kreitman1983-xr}.
+of the genetic relationships between the individuals of a
+sample~\citep[e.g.][]{Kreitman1983-xr}.
+% e.g., to estimate levels of sequence diversity from a few dozen alleles at an allozyme locus
+% \citep{Kreitman1983-xr}.
 Today's vast quantity of whole-genome sequence
 makes it possible to confidently estimate many more properties of genealogical relationships
-in local regions of genomes \citep[e.g,][]{booker2018understanding,haenel2018metaanalysis,stankowski2019widespread},
-or between individuals rather than populations \citep[e.g.,][]{browning2010highresolution,speed2014relatedness,aguillon2017deconstructing}.
+in local regions of genomes, both between
+individuals~\citep[e.g.,][]{browning2010highresolution,speed2014relatedness,aguillon2017deconstructing},
+and within
+populations~\citep[e.g,][]{booker2018understanding,haenel2018metaanalysis,stankowski2019widespread}.
 Computation is beginning to be a major problem:
-genetic data on the scale of the UK Biobank~\citep{bycroft2018genome},
-for instance, might have 500K individuals genotyped at 10M loci,
-resulting in a genotype matrix of $5 \times 10^{12}$ entries.
-Computing a single summary from this matrix is still relatively quick,
-once loaded into memory,
-but for many applications we are now interested in computing a large number of statistics,
-or iteratively refining a single computation.
-Furthermore, even larger datasets are being collected,
-and millions of whole genomes will soon be available. In this case the
-genotype matrix simply cannot fit into memory, and complex distributed computations
-are required.
-
-Fortunately, the genotype matrix is not the only way to encode genetic data.
+projects on the scale of the UK Biobank~\citep{bycroft2018genome}
+and gnomAD~\citep{karczewski2019variation} hold genetic data for
+hundreds of thousands of samples at tens to hundreds of millions of variant sites.
+Such large genotype matrices are extremely unwieldy and difficult to process.
+Tools such as \texttt{hail} (\url{https://hail.is}) allow computations to be sharded
+across many machines in parallel, thereby calculating statistics far more
+quickly than is possible using single-computer methods such as
+\texttt{plink}~\citep{purcell2007plink}. Computer time must still be
+paid for, however, and running calculations concurrently on thousands of cores quickly
+becomes expensive. Larger datasets still, consisting of millions of whole
+genomes, are currently being collected, and it is clear that processing
+these genotype matrices will be very costly.
+
+The genotype matrix, fortunately, is not the only way to encode genetic data.
 The \emph{succinct tree sequence} (or \emph{tree sequence}, for brevity) is a data
@@
 Many of these statistics are functions of the joint allele frequency spectrum,
-but the framework is more general,
-and can be used, for example, to quantify associations between genotypes and phenotypes.
+but the framework is more general and can be used,
+for example, to quantify associations between genotypes and phenotypes.
 This generality greatly reduces the software development effort in
@@
 as tree sequences allow storage and processing of large genomic data sets
-with orders of magnitude less space and time than most other methods.
+with orders of magnitude less space and time than standard matrix-based methods.
 All statistics described here (and more) are implemented
@@
 which provides a suite tools for efficiently working with tree sequences in Python and C.
-The largest efficiency gains are found for very large sample sizes
-only likely to be available in real data sets for humans in the near future,
-but the ability to analyze millions of whole genomes
-is useful for understanding evolutionary simulations of any species
-\citep[see, e.g.,][]{martin2017human,browning2018one,galloway2019stickleback}.
-% TODO rephrase this: large scale simulations are useful for lots of things.
-% No need for the slightly negative note about large datasets only being
-% relevant for humans, I think.
-
-% TODO: too negative.
-% We hope a more broadly interesting and useful aspect of this paper
-% is a more in-depth look at the correspondence between statistics calculated with genome sequence
-% and properties of the underlying genealogies, i.e.,
-% the duality between Site and Branch statistics under the infinite-sites model.
+Large-scale simulations are useful in many contexts
+\citep[e.g.,][]{martin2017human,browning2018one,galloway2019stickleback}
+and the ability to quickly compute a wide range of statistics from
+these (previously prohibitively large) datasets opens up new possibilities.
+At smaller scales, the statistics are still highly efficient and avoiding
+these cost of exporting data to a genotype matrix will in practice greatly
+speed up applications such as deep learning \jk{cite} and ABC~\jk{cite}
+that compute many statistics from simulated data. The ability to
+efficiently compute statistics from real data is also welcome, of course.
+
 The correspondence between genome sequence and properties of the

### 2019-09-18 treestats_paper.tex
Smoother segue into intro para 2.  We hadn't explicitly mentioned the matrix idea before this.

@@
 in local regions of genomes, both between
-individuals~\citep[e.g.,][]{browning2010highresolution,speed2014relatedness,aguillon2017deconstructing},
+individuals~\citep[e.g.,][]{browning2010highresolution,speed2014relatedness,aguillon2017deconstructing}
 and within
@@
 Computation is beginning to be a major problem:
-projects on the scale of the UK Biobank~\citep{bycroft2018genome}
+projects such as UK Biobank~\citep{bycroft2018genome}
 and gnomAD~\citep{karczewski2019variation} hold genetic data for
@@
 genomes, are currently being collected, and it is clear that processing
-these genotype matrices will be very costly.
-
-The genotype matrix, fortunately, is not the only way to encode genetic data.
-The \emph{succinct tree sequence} (or \emph{tree sequence}, for brevity) is a data
-structure that efficiently encodes the genealogical trees describing how a set of
-individuals are related to each other at each point along their genome.
+and storing these genotype matrices will be very costly.
+
+A key area of inefficiency in existing methods is that they focus on representing
+the genotype data as a \emph{matrix}.  As such, computational complexity if often linear
+in the matrix size. An alternative would be to take advantage of the relatedness amongst
+individuals in the sample and represent the data as trees.
+Trees describe the ancestral relationships between
+samples~\citep{felsenstein2004inferring,semple2003phylogenetics},
+and provide a natural and elegant way of encode genetic data,
+not only recording the history of a sample but also massively compressing
+the genotype matrix~\citep{ane2005missing}. Until recently, however,
+tree based approaches could not be used to encode data from sexually
+reproducing organisms as recombination results in many trees along the
+genome~\citep{griffiths1991two,griffiths1996ancestral,rasmussen2014genome},
+and there was no efficient way of representing this information.
+The \emph{succinct tree sequence} (or \emph{tree sequence}, for brevity)
+is a recently introduced data structure that
+encodes these genealogical trees along a genome concisely.
 It was introduced in the context of coalescent
-simulation~\citep{kelleher2016efficient}, and lead to scalability increases of
+simulation~\citep{kelleher2016efficient}, leading to scalability increases of
 several orders of magnitude over existing methods. The methods were subsequently extended
@@
 samples~\citep{kelleher2019inferring}.
+
 The key to the remarkable efficiency of tree sequence
@@
 
-Naturally, the genealogical perspective on population genetics statistics has a long history.
-The sequence of trees describing genealogical relationships as one moves along the genome
-are also described by the Ancestral Recombination
-Graph~\citep{griffiths1991two,griffiths1996ancestral,rasmussen2014genome}.
-These genealogical trees can be accurately inferred in nonrecombining regions
-\citep{felsenstein2004inferring},
-as for instance the mitochondria,
-and substantial theoretical work has been done in the context of phylogenetics \citep{semple2003phylogenetics}.
-% describing how each mutation gives a small piece of information about the tree at that site
-% However, population genetics has the task of summarizing relationships across \emph{many} trees,
-% each only partially known because mutations are sparse in the data.
-The basic relationship between tree shape and summaries of genetic variation
-is that if mutations are neutral then the expected number that occur on a given branch
-is proportional to the length of that branch,
-thus establishing a duality between genealogical and genetic summaries,
-a fact used extensively in deriving properties of statistics in models of random mating
-\citep[e.g.][]{tajima1983evolutionary,tavare1984lineofdescent,fu1995statistical}.
-The fact that the duality applies regardless of the underlying demographic model was also used
-\citep[e.g.][]{gillespie1979evolutionary,hudson1983properties,slatkin1991inbreeding,mcvean2002genealogical,ralph2019empirical},
-but only recently have datasets made it possible to move beyond a few numerical summaries.
-
 % summary
-In this paper, we study only ``single-site'' genetic statistics,
+In this paper, we study ``single-site'' genetic statistics,
 i.e., statistics of aligned genome sequence that can be expressed as averages over values computed
-separately for each site,
-leaving other statistics (e.g., pairwise statistics such as linkage disequilibrium
-or haplotype statistics) for later work.
+separately for each site. We develop a theoretical and computational framework
+that encompasses a large class of population genetic statistics, generalising
+many classical summaries of genetic variation. We define
+and explore a duality between ``Site'' and ``Branch'' statistics, and
+provide a detailed description (and analysis) of an efficient algorithm
+to compute statistics in this general setting.
 The methods we describe are implemented, tested, and validated in the \tskit{}
@@
 \section*{Framework and statistics}
+
+The basic relationship between tree shape and summaries of genetic variation
+is this: if mutations are neutral then the expected number that occur on a
+given branch of a tree is proportional to the length of that branch.
+This ``duality'' (we will use the term more precisely later) has been
+used extensively in deriving properties of statistics in models of random mating
+\citep[e.g.][]{tajima1983evolutionary,tavare1984lineofdescent,fu1995statistical},
+as has the fact that it applies regardless of the underlying demographic model
+\citep[e.g.][]{gillespie1979evolutionary,hudson1983properties,slatkin1991inbreeding,
+mcvean2002genealogical,ralph2019empirical}. We build on this central
+insight to define a flexible and computationally efficient framework for
+population genetic statistics on tree sequences.
+Roughly speaking, to compute a statistic we will
+propagate ``weights'' additively up each tree,
+summarize these weights on each branch with a ``summary function'',
+and aggregate these summaries to produce statistics.
+Figure~\ref{fig:divergence_diagram} shows a simple example of these procedures
+of propagating weights up a tree and summarizing them,
+used to calculate mean sequence divergence over a short sequence spanned by a single tree.
 
@@
 
-Before formal definitions, we will give an overview of the general framework.
-Roughly speaking, to compute a statistic we will
-propagate ``weights'' additively up each tree,
-summarize these weights on each branch with a ``summary function'',
-and aggregate these summaries to produce statistics.
 The abstract idea of summarizing weights propagated up each tree
-is only a small generalization of a familiar operations in population genetics:
+is only a small generalization of a familiar operation in population genetics:
 summarizing the allele frequency spectrum.
@@
 so is analogous to the allele frequency spectrum.
-Figure~\ref{fig:divergence_diagram} shows a simple example of these procedures
-of propagating weights up a tree and summarizing them,
-used to calculate mean sequence divergence over a short sequence spanned by a single tree.
-
-% Many population genetic summary statistics can be defined using the allele frequency spectrum of a collection of homologous sequences.
 If we only considered summaries of the allele frequency spectrum of a single sample,
-our statistics would be equivalent to that described by \citet{fu1995statistical}.
-To connect the framework to familiar notions,
+our statistics would be equivalent to the framework described by \citet{fu1995statistical}.
+To connect our  generalization to familiar ideas,
 here is a reframing of \citeauthor{fu1995statistical}'s statistics.
@@
 Since this is a simple sum over sites,
-we can compute this by summing over the $L$ loci, as
+we can compute this by summing over the $L$ loci and all alleles
+(including the ancestral allele),
 \begin{align*}
@@
 \end{align*}
-% (Note that since the ancestral allele is included, this function is invariant to relabeling of alleles.)
 If we know where each mutation has occurred on the genealogy,
@@
 
-Before defining the statistics, we need a few definitions.
-In a tree sequence, each haplotype, modern or ancestral, is associated with a \emph{node},
+Our framework is more general as we can also define statistics that are
+\emph{not} functions of the allele frequency spectrum.
+Before we define the framework, we need some notation and definitions.
+A tree sequence describes how a set of $n$ sampled chromosomes
+are related to each other along a (linear) genome of length $L$ \citep{kelleher2016efficient,kelleher2018efficient}.
+Each haplotype, modern or ancestral, is associated with a \emph{node},
 and the trees at each position along the genome have vertexes labeled by these nodes.
@@
 as for instance if a portion of an ancestor's genome was not inherited by any of the samples.
+From the tree sequence can be extracted a sequence of trees,
+$\treeseq = (T_1, T_2, \ldots, T_{n(\treeseq)})$
+and a sequence of breakpoints $0 = a_0 < a_1 < \cdots < a_{n(\treeseq)} = L$,
+where $T_k$ describes the genealogical relationships of the samples
+over the segment of genome between $a_{k-1}$ (inclusive) and $a_k$ (exclusive).
+We say that tree $T_k$ \emph{covers} the (half-open) segment $[a_{k-1}, a_k)$,
+and call the length of this segment its \emph{span}, denoted $L_k = a_k - a_{k-1}$.
+We refer to the branches in each tree using the most recent node,
+so for instance a branch between an ancestor $v$ and descendant $u$
+is associated with the node $u$.
+Note that although the same parent-child relationship may exist across many adjacent trees
+(this is called an ``edge''),
+rearrangements of genealogical relationships due to recombination
+can cause the precise set of samples that inherit from that edge to differ across trees.
 
@@
 %%%%%%%%%%%%%%%%%%
-\subsection*{Tree sequences}
-
-Before defining the different classes of statistic,
-it will help to establish a little more notation.
-A tree sequence efficiently describes how a set of $n$ sampled chromosomes
-are related to each other along a (linear) genome of length $L$ \citep{kelleher2016efficient,kelleher2018efficient}.
-% These trees are highly correlated,
-% a fact which allows them to be stored and processed very efficiently,
-% as described in \citet{kelleher2018efficient}.
-From the tree sequence can be extracted a sequence of trees,
-$\treeseq = (T_1, T_2, \ldots, T_{n(\treeseq)})$
-and a sequence of breakpoints $0 = a_0 < a_1 < \cdots < a_{n(T)} = L$,
-where $T_k$ describes the genealogical relationships of the samples
-over the segment of genome between $a_{k-1}$ (inclusive) and $a_k$ (exclusive).
-We say that tree $T_k$ \emph{covers} the (half-open) segment $[a_{k-1}, a_k)$,
-and call the length of this segment its \emph{span}, denoted $L_k = a_k - a_{k-1}$.
-We refer to the branches in each tree using the most recent node,
-so for instance a branch between an ancestor $v$ and descendant $u$
-is associated with the node $u$.
-The \emph{length} of this branch is the difference in birth times of $v$ and $u$,
-and is denoted $\beta_T(u)$.
-Note that although the same parent-child relationship may exist across many adjacent trees
-(this is called an ``edge''),
-rearrangements of genealogical relationships due to recombination
-can cause the precise set of samples that inherit from that edge to differ across trees.
+% \subsection*{Tree sequences}
 
@@
 in the rigorously tested \tskit{} library,
-which provides a suite tools for efficiently working with tree sequences in Python and C.
+which provides a suite of tools for working with tree sequences in Python and C.
 Large-scale simulations are useful in many contexts
@@
 these (previously prohibitively large) datasets opens up new possibilities.
-At smaller scales, the statistics are still highly efficient and avoiding
-these cost of exporting data to a genotype matrix will in practice greatly
-speed up applications such as deep learning \jk{cite} and ABC~\jk{cite}
-that compute many statistics from simulated data. The ability to
-efficiently compute statistics from real data is also welcome, of course.
+At smaller scales the statistics are still highly efficient, and avoiding
+the cost of exporting simulated data to a genotype matrix will in practice greatly
+speed up inference based on summaries computed from simulated
+data~\citep{beaumont2002approximate,csillery2010approximate,schrider2018supervised,flagel2018unreasonable}.
+The ability to efficiently compute statistics from real data is, of course, also welcome.
 

### 2019-09-19 treestats_paper.tex
Give example of a "non-AFS" summary to give reader idea.

@@
 
-A key area of inefficiency in existing methods is that they focus on representing
-the genotype data as a \emph{matrix}.  As such, computational complexity if often linear
-in the matrix size. An alternative would be to take advantage of the relatedness amongst
-individuals in the sample and represent the data as trees.
+Fortunately, the genotype matrix is not the only way to encode genetic data.
 Trees describe the ancestral relationships between

### 2019-09-19 treestats_paper.tex
edits until duality

@@
 sequence; and by \emph{branch}, which gives the expected value of the site statistic
-under the infinite-sites model.
+under the infinite-sites model of mutation.
 We use the framework to implement many currently-defined statistics of genome
@@
 of the genetic relationships between the individuals of a
-sample~\citep[e.g.][]{Kreitman1983-xr}.
+sample~\citep[e.g.,][]{Kreitman1983-xr}.
 % e.g., to estimate levels of sequence diversity from a few dozen alleles at an allozyme locus
@@
 in local regions of genomes, both between
-individuals~\citep[e.g.,][]{browning2010highresolution,speed2014relatedness,aguillon2017deconstructing}
+individuals~\citep[e.g.,][]{browning2010highresolution,aguillon2017deconstructing}
 and within
-populations~\citep[e.g,][]{booker2018understanding,haenel2018metaanalysis,stankowski2019widespread}.
+populations~\citep[e.g.,][]{booker2018understanding,haenel2018metaanalysis,stankowski2019widespread}.
 Computation is beginning to be a major problem:
@@
 hundreds of thousands of samples at tens to hundreds of millions of variant sites.
-Such large genotype matrices are extremely unwieldy and difficult to process.
+Such large genotype matrices are extremely unwieldy and difficult to process,
+and computational complexity is often linear in the size of the matrix. 
 Tools such as \texttt{hail} (\url{https://hail.is}) allow computations to be sharded
@@
 
-Fortunately, the genotype matrix is not the only way to encode genetic data.
+Fortunately, the genotype matrix is not the only way to encode genetic data:
+an alternative would be to take advantage of the relatedness amongst
+individuals and represent the data using trees.
 Trees describe the ancestral relationships between
@@
 As one looks across the genome, genealogies change at recombination breakpoints in ancestors.
-However, nearby trees tend to share a lot of common structure,
+However, nearby trees tend to share much of their structure,
 and single genealogical relationships (e.g., individual $x$ inherits from individual $y$)
@@
 This redundancy can be exploited to store genome sequence and the
-associated genealogical relationships very compactly, while also
-enabling efficient calculation of several population genetics
+associated genealogical relationships very compactly, and has formed the basis for
+efficient algorithms to calculate several population genetics
 statistics~\citep{kelleher2016efficient}.
@@
 used extensively in deriving properties of statistics in models of random mating
-\citep[e.g.][]{tajima1983evolutionary,tavare1984lineofdescent,fu1995statistical},
+\citep[e.g.,][]{tajima1983evolutionary,tavare1984lineofdescent,fu1995statistical},
 as has the fact that it applies regardless of the underlying demographic model
-\citep[e.g.][]{gillespie1979evolutionary,hudson1983properties,slatkin1991inbreeding,
+\citep[e.g.,][]{gillespie1979evolutionary,hudson1983properties,slatkin1991inbreeding,
 mcvean2002genealogical,ralph2019empirical}. We build on this central
@@
 and aggregate these summaries to produce statistics.
-Figure~\ref{fig:divergence_diagram} shows a simple example of these procedures
+\autoref{fig:divergence_diagram} shows a simple example of these procedures
 of propagating weights up a tree and summarizing them,
@@
 \begin{figure}
-    \centering
+    \begin{center}
     \includegraphics{figures/divergence_diagram}
+    \end{center}
     \caption{
@@
 and suppose that for some function $f()$, we want to compute $\sum_p h(p) f(p)$.
-(For instance, mean genetic diversity is computed with $f(p) = p (1-p)$.)
 Since this is a simple sum over sites,
@@
 \end{align*}
+For instance, mean genetic diversity across biallelic loci is computed with $f(p) = p (1-p)$,
+omitting the usual factor of two because the sum is over \emph{both} alleles at each site.
 If we know where each mutation has occurred on the genealogy,
@@
 Our framework is more general as we can also define statistics that are
-\emph{not} functions of the allele frequency spectrum.
+\emph{not} functions of the allele frequency spectrum, such as the correlation between genotype and phenotype.
 Before we define the framework, we need some notation and definitions.
@@
 so that when summarizing we have access to more than one aspect of each subtree.
-We will often be interested in statistics defined in terms of different
-subsets of our samples (corresponding to, e.g., different populations); it
-is useful to have some additional notation to define these statistics.
+We will often be interested in statistics defined in terms of different subsets of our samples;
+it is useful to have some additional notation to define these statistics.
 Given a subset set $S$ of samples,
@@
 \end{align}
-The weight $\nw_{T_k}(u)$ is the subtree weight of node $u$ in the tree $T_k$,
-and $\tiw - \nw_{T_k}(u)$ is the sum of all weights \emph{not} in the subtree of $u$ in $T_k$.
-% i.e., the length of the segment of genome that the tree extends for.
-% If this is computed over only a ``window'' of the genome
-% then $L_k$ is replaced by the length of the portion of that window
-% that $T$ extends for.
+The weight $\nw_{T_k}(u)$ is the sum of the weights of all samples descending from node $u$
+in the tree $T_k$,
+and $\tiw - \nw_{T_k}(u)$ is the sum of all remaining weights.
+If this is computed over only a ``window'' of the genome
+then $L_k$ is replaced by the length of the portion of that window that the tree $T_k$ extends for,
+and $L$ is replaced by the length of the window.
 
 \paragraph{Polarization}
-The node statistic defined above is appropriate for statistics such as nucleotide diversity, which is a function
-of all alleles at given variable site.  For other statistics, one is only concerned with the derived state(s)
-at a site, and we may define a \emph{polarized} node statistic without the second term from our previous definition:
+The node statistic defined above is appropriate for statistics of all alleles at given variable site.
+For other statistics, one is only concerned with the derived state(s) at a site,
+and we define a \emph{polarized} node statistic without the second term from our previous definition:
 
@@
 \end{example}
+
+An alternative and possibly more familiar way to compute heterozygosity
+would be to use a polarized Site statistic with summary function $f(x) = 2 x(n-x)/n(n-1)$.
+However, this approach fails with more than two alleles per site.
 
@@
 measuring tree distances in branch lengths versus in numbers of mutations
-are equal in expectation, up to a constant.
+are equal in expectation, up to a multiplicative factor of the mutation rate.
 
@@
 and sample weights $\iw$ with total weight $\tiw = \sum_u \iw(u)$
-to be the sum of the length of each branch in each tree
+to be the sum of the length of each branch
 multiplied by the summary function applied to its subtree weight and its complement:
@@
 and $\nw_{T}(u)$ is the total weight of the subtree of $T$ inheriting from $u$ (as defined above).
-The term $\tiw - \nw_{T}(u)$ gives the total weight \emph{not} in the subtree of $T$ inheriting from $u$,
-because the sum of all weights in the tree is always $\tiw$.
+The term $\tiw - \nw_{T}(u)$ gives the total weight \emph{not} in the subtree of $T$ inheriting from $u$.
 The value $f(\nw(u))$ is the summary value that would be added to a site statistic
@@
     to any mutation falling in it (colored lines), and then the branch $f_4$ statistic
-    is the sum over these weights, multiplied by the length of the branch
-    (and the span of the tree, not depicted).
+    is the sum over these weights, multiplied by the length of the branch.
         \label{fig:branch_site_diagram}
@@
 but in practice it is more useful to average branch statistics
-over a region of the genome. We do this by averaging the tree statistics over the region
+over a region of the genome, as we did for Site statistics.
+We do this by averaging the tree statistics over the region
 with probabilities proportional to the trees' spans:
 \begin{align}
-    \branch(f, \iw)_{[c,d)}
+    \branch(f, \iw)_{[i,j)}
     &=
-    \frac{1}{d-c} \sum_{k=1}^{n(\treeseq)} \ell_k(c,d) \branch(f, \iw)_{T_k} ,
+    \frac{1}{j-i} \sum_{k=1}^{n(\treeseq)} \ell_k(i,j) \branch(f, \iw)_{T_k} ,
 \end{align}
-where $\ell_k(c,d)$ is the length of the region in $[c,d)$ that the tree $T_k$ covers
+where $\ell_k(i,j)$ is the length of the region in $[i,j)$ that the tree $T_k$ covers
 (i.e., if $T$ covers the half-open interval $[a_{k-1},a_k)$,
-then $\ell_k(c,d) = \max(0, \min(d,a_k) - \max(c,a_{k-1}))$).
+then $\ell_k(i,j) = \max(0, \min(j,a_k) - \max(i,a_{k-1}))$).
 The \emph{polarized} version of a Branch statistic
@@
 \end{align}
-and $\branchp(f, \iw)_{[c,d)}$ is defined using $\branchp(f, \iw)_T$ as before.
+and $\branchp(f, \iw)_{[i,j)}$ is defined using $\branchp(f, \iw)_T$ as before.
 
@@
     averaged across the sequence.
-    This is twice the mean TMRCA if the samples are all from the same time.
+    This is twice the mean time back to the MRCA if the samples are all from the same time.
 \end{example}
@@
     If we take $\iw$ and $f$ as in the ``Phenotypic correlations'' example above,
-    what does $\branch(f, \iw)$ tell us?
-    The statistic tells us the \emph{expected} correlation between phenotype and any mutations
-    appearing on the tree, which is a summary of how much local relatedness
+    then $\branch(f, \iw)$ gives the \emph{expected} correlation between phenotype and any mutations
+    appearing on the tree. This is a summary of how much local relatedness
     aligns with similarity in phenotype.
@@
 This relationship is true more generally.
-In fact, for any region of the genome between $a$ and $b$,
-with $\treeseq_{[a,b)}$ the trees spanning that interval,
+In fact, for any region of the genome between $i$ and $j$,
+with $\treeseq_{[i,j)}$ the trees spanning that interval,
 \begin{align}
-    \branch(f, \iw)_{[a,b)}
+    \branch(f, \iw)_{[i,j)}
     =
     \frac{1}{\mu}
-    \E\left[ \site(f, \iw)_{[a,b)} \given \treeseq_{[a,b)} \right] ,
+    \E\left[ \site(f, \iw)_{[i,j)} \given \treeseq_{[i,j)} \right] ,
 \end{align}

### 2019-09-19 treestats_paper.tex
Partial editing pass.

@@
 Such large genotype matrices are extremely unwieldy and difficult to process,
-and computational complexity is often linear in the size of the matrix. 
-Tools such as \texttt{hail} (\url{https://hail.is}) allow computations to be sharded
+and computational complexity is usually linear in the size of the matrix.
+Tools such as \texttt{Hail} (\url{https://hail.is}) allow computations to be sharded
 across many machines in parallel, thereby calculating statistics far more
@@
 becomes expensive. Larger datasets still, consisting of millions of whole
-genomes, are currently being collected, and it is clear that processing
+genomes, are currently being collected and it is clear that processing
 and storing these genotype matrices will be very costly.
 
-Fortunately, the genotype matrix is not the only way to encode genetic data:
-an alternative would be to take advantage of the relatedness amongst
-individuals and represent the data using trees.
-Trees describe the ancestral relationships between
-samples~\citep{felsenstein2004inferring,semple2003phylogenetics},
-and provide a natural and elegant way of encode genetic data,
+% Fortunately, the genotype matrix is not the only way to encode genetic data:
+% an alternative would be to take advantage of the relatedness amongst
+% individuals and represent the data using trees.
+Genotype matrices are massively redundant, however, and several compression methods
+have been proposed~\citep{durbin2014efficient} \jk{TODO more refs}.
+The fundamental reason for this redundancy is that samples share ancestry:
+related individuals will tend to share the same state at a particular variant site.
+Trees describe these ancestral
+relationships~\citep{felsenstein2004inferring,semple2003phylogenetics},
+and provide a natural and elegant way of encoding genetic data,
 not only recording the history of a sample but also massively compressing
@@
 samples~\citep{kelleher2019inferring}.
-
 The key to the remarkable efficiency of tree sequence
@@
 The methods we describe are implemented, tested, and validated in the \tskit{}
-Python and C libraries, available at \url{https://github.com/tskit-dev/tskit}
-under the terms of the MIT open source license.
+Python and C libraries, freely available at \url{https://github.com/tskit-dev/tskit}
+under the terms of the MIT license.
 
@@
 as has the fact that it applies regardless of the underlying demographic model
+\jk{CITE Konrad/Nick here?}
 \citep[e.g.,][]{gillespie1979evolutionary,hudson1983properties,slatkin1991inbreeding,
@@
 Roughly speaking, to compute a statistic we will
-propagate ``weights'' additively up each tree,
+propagate ``weights'' additively up each marginal tree,
 summarize these weights on each branch with a ``summary function'',
@@
 An alternative and possibly more familiar way to compute heterozygosity
-would be to use a polarized Site statistic with summary function $f(x) = 2 x(n-x)/n(n-1)$.
+would be to use a polarized Site statistic with summary function $f(x) = 2
+x(n-x)/\left(n(n-1)\right)$.
 However, this approach fails with more than two alleles per site.
@@
     Patterson's $f_4(S_1, S_2; S_3, S_4)$ statistic for an allele with frequency $p_i$ in group $S_i$
-    is $(p_1 - p_2)(p_3 - p_4)$ (and then this is typically averaged over sites).
+    is $(p_1 - p_2)(p_3 - p_4)$ \jk{CITATION NEEDED}.
     To rewrite this as a sum over alleles, note that
@@
 In the case where we actually have the trees, or a good proxy for them,
-it is natural to summarize them directly, rather than working indirectly with the genotypes.
+it is natural to summarize them directly rather than working indirectly with the
+genotypes~\citep{harris2019database}.
 If we assume that no two mutations occur at the same genomic position ---
@@
 If the mutation rate is constant,
-then this gives us an isometry ---
-measuring tree distances in branch lengths versus in numbers of mutations
+then this gives us an isometry:
+tree distances measured in branch lengths versus in numbers of mutations
 are equal in expectation, up to a multiplicative factor of the mutation rate.
@@
 a statistic of tree shape by summing these expected contributions across its branches.
-We define the \textbf{branch statistic} of a single tree $T$
-obtained from summary function $f()$
-and sample weights $\iw$ with total weight $\tiw = \sum_u \iw(u)$
+We define the \textbf{branch statistic} of a single tree $T$,
+obtained from summary function $f()$ and sample weights $\iw$,
 to be the sum of the length of each branch
@@
 \end{align}
-where $\beta_{T}(u)$ is the length of the branch ancestral to node $u$ in tree $T$
+where $\beta_{T}(u)$ is the length of the branch ancestral to node $u$ in tree $T$,
+$\tiw = \sum_u \iw(u)$ is the total weight,
 and $\nw_{T}(u)$ is the total weight of the subtree of $T$ inheriting from $u$ (as defined above).
@@
 the expected number of mutations per branch is proportional to its length.
-This implies an isormorphism between ``site'' and ``branch'' statistics defined above,
+This implies an isormorphism between ``Site'' and ``Branch'' statistics defined above,
 which is discussed in more detail in \citet{ralph2019empirical}.
@@
 and the branch statistic of Example \ref{ex:branch_diversity} (mean TMRCA)
-use the same summary function $f(x) = x(n-x)/n(n-1)$.
+use the same summary function $f(x) = x(n-x)/\left(n(n-1)\right)$.
 These are closely related because under an infinite-sites model of mutation,
@@
 the expected value of genetic diversity,
-averaging over mutational noise give the tree sequence,
+averaging over mutational noise given the tree sequence,
 is equal to the mutation rate multiplied by the average distance between the two in the trees.
@@
 with rate $\mu$ per unit time and per unit of sequence length.
-In other words, the Site statistic can be thought of an estimator of its corresponding Branch statistic,
+Note that the expected product of two site statistics,
+$\E[\site(f, \iw) \site(g, \iw')]$
+is not equal to the product of the two branch statistics, $\branch(f, \iw) \branch(g, \iw')$,
+because they are not independent.
+However, it is always possible to define a branch statistic that
+gives the expected value of the product~\citep{ralph2019empirical}.
+
+A Site statistic can therefore be thought of as an estimator of its corresponding Branch statistic,
 which is itself a summary of local tree shapes.
@@
 in a way that mutation rates are constant, but how these vary is generally unknown.
-
-Note that the expected product of two site statistics,
-$\E[\site(f, \iw) \site(g, \iw')]$
-is not equal to the product of the two branch statistics, $\branch(f, \iw) \branch(g, \iw')$,
-because they are not independent.
-However, it is always possible to define a branch statistic that
-gives the expected value of the product~\citep{ralph2019empirical}.
 

### 2019-09-19 treestats_paper.tex

@@
 % individuals and represent the data using trees.
-Genotype matrices are massively redundant, however, and several compression methods
-have been proposed~\citep{durbin2014efficient} \jk{TODO more refs}.
+Genotype matrices are massively redundant, however, and several specialised
+compression methods
+%JK Note for later: I think there's an earlier citation I'm missing here, but
+% I can't find it right now. Look it up when doing later corrections.
+have been proposed~\citep{durbin2014efficient,layer2016efficient,danek2018gtc,lin2019sparse}.
 The fundamental reason for this redundancy is that samples share ancestry:
@@
 as has the fact that it applies regardless of the underlying demographic model
-\jk{CITE Konrad/Nick here?}
 \citep[e.g.,][]{gillespie1979evolutionary,hudson1983properties,slatkin1991inbreeding,
-mcvean2002genealogical,ralph2019empirical}. We build on this central
+mcvean2002genealogical,lohse2016efficient,ralph2019empirical}. We build on this central
 insight to define a flexible and computationally efficient framework for
@@
     Given four disjoint groups of samples, $S_1$, $S_2$, $S_3$, and $S_4$,
-    Patterson's $f_4(S_1, S_2; S_3, S_4)$ statistic for an allele with frequency $p_i$ in group $S_i$
-    is $(p_1 - p_2)(p_3 - p_4)$ \jk{CITATION NEEDED}.
-    To rewrite this as a sum over alleles, note that
+    Patterson's $f_4(S_1, S_2; S_3, S_4)$
+    statistic~\citep{reich2009reconstructing,patterson2012ancient}
+    for an allele with frequency $p_i$ in group $S_i$
+    is $(p_1 - p_2)(p_3 - p_4)$. To rewrite this as a sum over alleles, note that
     $p_1 - p_2 = p_1 (1 - p_2) - (1 - p_1) p_2$,
@@
 diversity along the chromosome following a few selective sweeps.
-The top two plots compare ``branch'' diversity --- i.e., as computed only with tree shape ---
-to ``site'' diversity computed from sequence generated by 20 independent assignments of mutations to the same tree sequence,
+The top two plots compare Branch diversity --- i.e., as computed only with tree shape ---
+to Site diversity computed from sequence generated by 20 independent assignments of mutations to the same tree sequence,
 with mutation rates $10^{-9}$ and $10^{-8}$, respectively.
 We see that as the mutation rate increases, the signal of decrease in diversity around swept loci becomes more clear,
-and approaches ``branch'' diversity.
+and approaches Branch diversity.
 These were computed using the entire population of 1,000 individuals;
@@
 as is the Site $f_4$ divided by mutation rate;
-see Supplemental Figure \ref{fig:introf4} for a plot along the genome).
+see Supplemental Figure~\ref{fig:introf4} for a plot along the genome).
 Using various sample sizes, mutation rates, and window sizes, we then
 calculated this $f_4$ statistic in windows along a 100Mb genome,
-and show the standard deviation of both ``site'' and ``branch'' statistics across windows
-in Figure \ref{fig:introf4sd}.
+and show the standard deviation of both Site and Branch statistics across windows
+in Figure~\ref{fig:introf4sd}.
 Since the genome is uniform (no selection or variation in recombination or mutation),
@@
 but recent methods are available to estimate them at
-scale~\citep{harris2019database,kelleher2019inferring,speidel2019method}.
+scale~\citep{kelleher2019inferring,speidel2019method}.
 In Figure~\ref{fig:sweep_duality}, we showed that branch and site statistics
@@
 
-In practice, we do not expect branch and site diversity to agree exactly,
+In practice, we do not expect Branch and Site diversity to agree exactly,
 because of local differences in mutation rate and intensities of selected mutations.
@@
 and thus diversity at even neutral, linked sites \citep{hudson1994levels,charlesworth1997effects}.
-The first effect would cause site and branch statistics to differ,
+The first effect would cause Site and Branch statistics to differ,
 because it effectively reduces the mutation rate in the region
 and violates the assumption of independence of mutations given the trees.
-However, the second effect does \emph{not} affect the correspondence between site and branch statistics,
+However, the second effect does \emph{not} affect the correspondence between Site and Branch statistics,
 since it is mediated by tree shape.
@@
 \citep[but see][]{supek2015differential}
-or how dense are targets of selection \citep{leffler2012revisiting}.
-For this reason, it is very interesting to see how close a correspondence between site and branch
+or how dense targets of selection are~\citep{leffler2012revisiting}.
+For this reason, it is very interesting to see how close a correspondence between Site and Branch
 statistics it is \emph{possible} to obtain ---
 it is tempting to say that the best tree sequence would obtain
-as tight a match between site and branch statistics as possible,
+as tight a match between Site and Branch statistics as possible,
 with remaining variation explained by direct selection and/or mutation rate variation.
@@
 sequence \cite[Algorithm L]{kelleher2016efficient}. The central
-problem shared across site, branch and node statistics
+problem shared across Site, Branch and Node statistics
 is to efficiently maintain the ``state'' $x_u$ (where $x_u$ is the
@@
 
-\begin{taocpalg}{B}{General branch statistics}
+\begin{taocpalg}{B}{General Branch statistics}
 {Given a list of $n$ sample nodes $S$, corresponding weights $w$ and
-summary function $f: \R \rightarrow \R$
+summary function $f: \R \rightarrow \R$,
 compute the span-normalized statistic $\sigma$
@@
 Figure~\ref{fig:algorithm_example}B, removing an edge will result
-in the subtree rooted at $u$ become disconnected from the rest of the
+in the subtree rooted at $u$ becoming disconnected from the rest of the
 tree. Step \algref{B5} ensures that the state of the rest of the tree
@@
 \end{equation}
-Computing a site statistic will have the same complexity,
+Computing a Site statistic will have the same complexity,
 as the same internal state is maintained,
@@
 To determine how well our theoretical model of time complexity predicts the
-running time in \tskit{}'s implementation of the site statistics algorithm,
+running time in \tskit{}'s implementation of the Site statistics algorithm,
 we fitted a model based on Equation~\eqref{eqn:alg_complexity}.
@@
 \texttt{SLiM}~\citep{haller2018tree,haller2019slim} and
-\texttt{fwdpy11}~\citep{thornton2014c++}, the advice is straightforward.
+\texttt{fwdpy11}~\citep{thornton2014c++}, however, the advice is straightforward.
 Computing statistics using the algorithms in \tskit{}
@@
 % visualization and time decomposition adding a new dimension
-Genomic data are naturally distributed along two axes: along the genome, and across geography.
+Genomic data are naturally distributed on two axes: along the genome, and across geography.
 The tree sequence extends this to a third axis: time.

### 2019-09-19 treestats_paper.tex
variance is a branch stat

@@
 How big is the contribution of mutation to the overall sampling variance of a statistic?
-The law of total variance lets us partition the variance of a site statistic
-into the contributions of mutation and demographic noise:
+The law of total variance partitions the variance of a site statistic
+into the contributions of noise from demography and mutation:
 \begin{align*}
-    \var[\site(f, \iw)_{[a,b)}]
+    \var[\site(f, \iw)_{[i,j)}]
     &=
-        \E\left[ \var\left[\site(f, \iw)_{[a,b)} \given \treeseq_{[a,b)}\right] \right]
-        + \var\left[ \E\left[\site(f, \iw)_{[a,b)} \given \treeseq_{[a,b)}\right] \right] \\
+        \var\left[ \E\left[\site(f, \iw)_{[i,j)} \given \treeseq_{[i,j)}\right] \right] 
+            +
+        \E\left[ \var\left[\site(f, \iw)_{[i,j)} \given \treeseq_{[i,j)}\right] \right]
+        \\
     &=
-        \E\left[ \var\left[\site(f, \iw)_{[a,b)} \given \treeseq_{[a,b)}\right] \right]
-        + \var\left[ \branch(f, \iw)_{[a,b)} \right] .
+        \var\left[ \branch(f, \iw)_{[i,j)} \right]
+            +
+        \frac{1}{j-i} \E\left[ \branch(f^2, \iw)_{[i,j)} \right] .
 \end{align*}
-The first term, the variance of the statistic given the trees,
+The first term is the variance of the expected Site statistic given the trees,
+which by duality is the variance of the Branch statistic,
+i.e., the contribution of demographic noise, including genetic drift.
+The second term, the expected variance of the statistic given the trees,
 is the contribution to variance of mutation,
-while the second term, the variance of the branch statistic,
-is the contribution of demographic variation, including drift.
-
-% JK: deleted the section here as it seemed superfluous.
+and can itself be written as a Branch statistic,
+using the same weights and the summary function $f(x)^2$
+\citep[Lemma 2 in][]{ralph2019empirical}.
+The reason for this is straightforward ---
+the Site statistic is a sum across branches of the number of mutations on the branch
+multiplied by a summary, $f(x)$.
+Given the trees, these numbers of mutations are independent and Poisson,
+and the variance of $f(x)$ multiplied by a Poisson random variable with mean $m$ is $m f(x)^2$.
+The sum is divided by the length of the region, $j-i$,
+and so the variance is divided by $(j-i)^2$.
 

### 2019-09-19 treestats_paper.tex
done with pass through

@@
 We examined the contributions of mutation and demography to noise
-in two simulations where population genetic (site) statistics
+in two simulations where population genetic (Site) statistics
 are important in making inferences:
@@
 and detecting introgression.
-Figure~\ref{fig:sweep_duality} shows windowed
+\autoref{fig:sweep_duality} shows windowed
 diversity along the chromosome following a few selective sweeps.
@@
 how does sampling variance contribute?
-Not much, it turns out --- the bottom plot shows both site and branch diversity
+Not much, it turns out --- the bottom plot shows both Site and Branch diversity
 computed from 20 nonoverlapping groups of 100 samples.
-Neither site or branch diversity vary much between these samples,
+Neither Site or Branch diversity vary much between these samples,
 implying that the subsample gives us a good estimate of the whole-population values of each.
 However, as we see in the top figure,
-whole-population site diversity is itself only a quite noisy estimator of branch diversity.
+whole-population Site diversity is itself only a quite noisy estimator of Branch diversity.
 The same things are shown in Supplementary Figure \ref{sfig:sweep_duality_10K}
@@
         In each case, ``site'' is mean genetic diversity (Tajima's $\pi$) divided by mutation rate,
-        and ``branch'' is the corresponding branch statistic described above.
+        and ``branch'' is the corresponding Branch statistic.
         The tree sequence was produced by simulating mutations under positive selection with mutation rate $10^{-12}$
@@
         The selected alleles at the marked sites have selection coefficients between 0.08 and 0.25,
-        and are at frequencies 96.8\%, 100\%, 16.25\%, 100\%, and 82.6\% in the final generation, respectively.
+        and are at frequencies 96.8\%, 100\%, 100\%, and 82.6\% in the final generation, respectively.
         All curves use the same tree sequence, including selected mutations,
-        and with additional neutral mutations added.
+        but with additional neutral mutations added.
         \textbf{(top)}
         Diversity within the entire population,
-        computed using for mode ``site'' from 20 independent assignments of mutations to the same tree sequence with mutation rate $\mu = 10^{-9}$.
+        computed as Site statistic from 20 independent assignments of mutations to the same tree sequence with mutation rate $\mu = 10^{-9}$.
         \textbf{(middle)}
         As in the top panel, but with mutation rate $\mu=10^{-8}$,
-        showing that as mutation rate increases, mode ``site'' (divided by $\mu$) converges to mode ``branch''.
+        showing that as mutation rate increases, the Site statistic (divided by $\mu$)
+        converges to the Branch statistic.
         \textbf{(bottom)}
-        ``Site'' and ``branch'' diversity within 20 disjoint samples of size 100 each,
-        computed on a single tree sequence with mutation rate $\mu = 10^{-9}$.
+        Site and Branch diversity within 20 disjoint samples of size 100 each,
+        with mutation rate $\mu = 10^{-9}$.
         \label{fig:sweep_duality}
@@
 this standard deviation is a measure of noisiness.
-As expected, site statistics are noisier than branch statistics,
+As expected, Site statistics are noisier than Branch statistics,
 by a factor that depends mostly on the ratio of mutation to recombination rates.
-The results suggest that branch statistics would provide substantially better resolution
+The results suggest that Branch statistics would provide substantially better resolution
 at small scales along the genome, especially if mutation rate is lower than recombination rate.
@@
     \caption{
-        Standard deviation of site (black) and branch (red) $f_4(1,2;3,4)$
+        Standard deviation of Site (black) and Branch (red) $f_4(1,2;3,4)$
         in windows along a 100Mb genome, as a function of
@@
 scale~\citep{kelleher2019inferring,speidel2019method}.
-In Figure~\ref{fig:sweep_duality}, we showed that branch and site statistics
+In Figure~\ref{fig:sweep_duality}, we showed that Branch and Site statistics
 matched well in simulated data. However, these simulations make many
@@
 correct (which inference methods can only hope to approximate).
-To evaluate the correspondence between site and branch statistics
+To evaluate the correspondence between Site and Branch statistics
 in tree sequences inferred from real data,
@@
 by Relate~\citep{speidel2019method}.
-Specifically, we calculated diversity using the site statistic
+Specifically, we calculated diversity using the Site statistic
 described in Example~\ref{ex:site_diversity},
-and compared this to the dual branch statistic (Example~\ref{ex:branch_diversity})
+and compared this to the dual Branch statistic (Example~\ref{ex:branch_diversity})
 in 1Mb windows in each of the five continental groupings.
@@
 Is this an indication that Relate is doing a good job at inferring trees,
-or just a byproduct of the inference method (which, naturally, estimates branch lengths to be proportional to numbers of mutations)?
+or just a byproduct of the inference method
+(which, naturally, estimates branch lengths to be proportional to numbers of mutations)?
 Answering this question requires a deeper understanding of the processes that shape
@@
         and \textbf{(bottom)}
-        the dual branch statistic as calculated from tree sequences inferred by
+        the dual Branch statistic as calculated from tree sequences inferred by
         Relate~\citep{speidel2019method},
@@
         Separate curves are plotted for each sampling location.
-        Note that the Site statistic divided by the Branch statistic is close to $2 \times 10^{-8}$,
-        the mutation rate used by Relate \plr{XXX total speculation here}.
         \label{fig:chr20}
@@
 running time in \tskit{}'s implementation of the Site statistics algorithm,
-we fitted a model based on Equation~\eqref{eqn:alg_complexity}.
+we fit a model based on Equation~\eqref{eqn:alg_complexity}.
 \autoref{fig:stats_performance} shows that theoretical predictions match
@@
 (representing the cost of building the first tree) will come to dominate.
-In this case, we can see this happening at around $n = 10^5$, where there
+In our simulations, we can see this happening at around $n = 10^5$, where there
 is a noticeable uptick in the time required per variant. For
@@
     \includegraphics[width=9cm]{tskit_stat_benchmarks/benchmarks_without_copy_longer_genome.pdf}
-    \caption{Comparison of time required to compute site statistics
+    \caption{Comparison of time required to compute Site statistics
         between matrix and tree based methods. For each sample size, a single replicate
-        was obtained using \texttt{msprime} with scaled neutral mutation rate $\theta = 4N_e\mu$ and
+        was obtained using \texttt{msprime} with scaled neutral mutation rate $\theta = 4 N_e L \mu$ and
         scaled recombination rate $\rho = 4N_er$ both equal to $10^4$ and $N_e = 10^4$,
-        where $\mu$ and $\rho$ are the total mutation and recombination rate across the simulated region
+        where $L\mu$ and $\rho$ are the total mutation and recombination rate across the simulated region
         per generation, respectively.
@@
 The range of statistics available to describe variation in a single exchangeable sample
-(i.e., a single population) are well-understood,
+(e.g., an isolated population) are well-understood,
 but there are much larger and less well-explored classes of statistics
@@
 % jk: Cut this: too technical for the discussion
-% We also require that site statistics be computable from the genotype matrix
+% We also require that Site statistics be computable from the genotype matrix
 % (i.e., without knowledge of the trees),
@@
 % and the total weight: $f(0) = f(\tiw) = 0$.
-% This also makes branch and node statistics unchanged by portions of the tree that do not lie on a path
+% This also makes Branch and Node statistics unchanged by portions of the tree that do not lie on a path
 % between any of the samples that have nonzero weight,
@@
 if tree sequence inference can be made insensitive to local variation in mutation rate,
-calculation of branch statistics would provide a summary method
+calculation of Branch statistics would provide a summary method
 that is not affected by this potentially important confounding factor.
@@
 The computational toolbox of population genetics
-is still mostly composed of statistics designed for analysis of a handful of loci
+is still mostly composed of statistics orignally designed for analysis of a handful of loci
 in a small number of discrete, mostly separated populations.
@@
 Aaron Ragsdale for useful suggestions
-and Leo Speidel for providing the inferred 1000 Genomes tree sequence.
+and to Leo Speidel for providing the inferred 1000 Genomes tree sequence.
 

### 2019-09-19 figures/plot_sweep.py
updated ylims on S1

@@ -78,7 +78,7 @@ mts = msprime.mutate(ts, rate=mut_rate, keep=True)
 sub_branch_div = ts.diversity(sample_sets, windows=windows, mode='branch')
 sub_site_div = (1/mut_rate) * mts.diversity(sample_sets, windows=windows, mode='site')
 
-maxy = 0.7 * max(np.max(np.abs(all_site_div)), np.max(np.abs(all_branch_div)),
+maxy = 1.1 * max(np.max(np.abs(all_site_div)), np.max(np.abs(all_branch_div)),
                  np.max(np.abs(sub_site_div)), np.max(np.abs(sub_branch_div)))
 
 

### 2019-09-19 treestats_paper.tex
added ratio to relate

@@
 \begin{figure}
-    \includegraphics{{figures/1kg/relate_chr20.site.1000000.region.diversity}.pdf}
-    \includegraphics{{figures/1kg/relate_chr20.branch.1000000.region.diversity}.pdf}
+    \includegraphics{{figures/1kg/relate_chr20_site_div_branch.1000000.diversity}.pdf}
     \caption{
@@
         in 1Mb windows along human chromosome 20 of the 1000 Genomes Project,
-        in units of mean number of pairwise differences per nucleotide,
-        and \textbf{(bottom)}
-        the dual Branch statistic as calculated from tree sequences inferred by
+        in units of mean number of pairwise differences per nucleotide.
+        \textbf{(middle)}
+        The dual Branch statistic as calculated from tree sequences inferred by
         Relate~\citep{speidel2019method},
         in units of generations.
+        \textbf{(bottom)} The ratio of the Site statistic to the Branch statistic,
+        which note hovers around typical per-generation estimates of the human mutation rate.
         Separate curves are plotted for each sampling location.
@@
     \caption{
-        As in Figure~\ref{fig:sweep_duality}, but in a population of size 10,000.
+        As in \autoref{fig:sweep_duality}, but in a population of size 10,000.
         \label{sfig:sweep_duality_10K}

### 2019-09-19 treestats_paper.tex
.

@@
 The computational toolbox of population genetics
-is still mostly composed of statistics orignally designed for analysis of a handful of loci
+is still mostly composed of statistics orginally designed for analysis of a handful of loci
 in a small number of discrete, mostly separated populations.

### 2019-09-20 treestats_paper.tex
Added some compression refs.

@@
 sample~\citep[e.g.,][]{Kreitman1983-xr}.
-% e.g., to estimate levels of sequence diversity from a few dozen alleles at an allozyme locus
-% \citep{Kreitman1983-xr}.
 Today's vast quantity of whole-genome sequence
@@
 
-% Fortunately, the genotype matrix is not the only way to encode genetic data:
-% an alternative would be to take advantage of the relatedness amongst
-% individuals and represent the data using trees.
 Genotype matrices are massively redundant, however, and several specialised
-compression methods
-%JK Note for later: I think there's an earlier citation I'm missing here, but
-% I can't find it right now. Look it up when doing later corrections.
-have been proposed~\citep{durbin2014efficient,layer2016efficient,danek2018gtc,lin2019sparse}.
+compression methods have been proposed~\citep{
+christley2008human,qiao2012handling,durbin2014efficient,sambo2014compression,layer2016efficient,
+danek2018gtc,lin2019sparse}.
 The fundamental reason for this redundancy is that samples share ancestry:

### 2019-09-20 treestats_paper.tex
spell check with en_US.

@@
 
-Genotype matrices are massively redundant, however, and several specialised
+Genotype matrices are massively redundant, however, and several specialized
 compression methods have been proposed~\citep{
@@
 separately for each site. We develop a theoretical and computational framework
-that encompasses a large class of population genetic statistics, generalising
+that encompasses a large class of population genetic statistics, generalizing
 many classical summaries of genetic variation. We define
@@
 The computational toolbox of population genetics
-is still mostly composed of statistics orginally designed for analysis of a handful of loci
+is still mostly composed of statistics originally designed for analysis of a handful of loci
 in a small number of discrete, mostly separated populations.

### 2019-09-20 treestats_paper.tex
Update figure 5.

@@
         \textbf{(bottom)} The ratio of the Site statistic to the Branch statistic,
-        which note hovers around typical per-generation estimates of the human mutation rate.
-        Separate curves are plotted for each sampling location.
+        which, we note, hovers around typical per-generation estimates of the human mutation
+        rate~\citep{narasimhan2017estimating}.
+        Separate curves are plotted for continental groupings:
+        AFR, Africa; AMR, America; EAS, east Asia; EUR, Europe; SAS, south Asia.
         \label{fig:chr20}

### 2019-09-20 figures/compare_relate_diversity.py

@@ -3,7 +3,17 @@ import numpy as np
 import matplotlib
 matplotlib.use('Agg')
 import matplotlib.pyplot as plt
+import seaborn as sns
 import os, sys
+
+# Roughly match the standard TGP color pallettes
+tgp_region_colors = {
+    "EAS": sns.color_palette("Greens", 2)[1],
+    "EUR": sns.color_palette("Blues", 1)[0],
+    "AFR": sns.color_palette("Wistia", 3)[0],
+    "AMR": sns.color_palette("Reds", 2)[1],
+    "SAS": sns.color_palette("Purples", 2)[1],
+}
 
 divfiles = { 'branch' : "1kg/relate_chr20.branch.1000000.diversity.tsv",
         'site' : "1kg/relate_chr20.site.1000000.diversity.tsv" }
@@ -14,15 +24,16 @@ bdivs = { a : np.loadtxt(divfiles[a], skiprows=1) for a in divfiles }
 num_windows = bdivs['branch'].shape[0]
 windows = np.linspace(0, 63025522, num_windows + 1).astype('int')
 
+
 colors = plt.get_cmap("tab20").colors
+
 pop_names = ['CHB', 'JPT', 'CHS', 'CDX', 'KHV', 'CEU', 'TSI', 'FIN', 'GBR', 'IBS', 'YRI', 'LWK', 'GWD', 'MSL', 'ESN', 'ASW', 'ACB', 'MXL', 'PUR', 'CLM', 'PEL', 'GIH', 'PJL', 'BEB', 'STU', 'ITU']
 superpops = ['EAS', 'EAS', 'EAS', 'EAS', 'EAS', 'EUR', 'EUR', 'EUR', 'EUR', 'EUR', 'AFR', 'AFR', 'AFR', 'AFR', 'AFR', 'AFR', 'AFR', 'AMR', 'AMR', 'AMR', 'AMR', 'SAS', 'SAS', 'SAS', 'SAS', 'SAS']
-popcolors = { k : colors[j % len(colors)] for j, k in enumerate(pop_names) }
 num_pops = len(pop_names)
 
 
 supernames = ['EAS', 'EUR', 'AFR', 'AMR', 'SAS']
-superpopcolors = { k : popcolors[pop_names[superpops.index(k)]] for k in supernames }
+superpopcolors = tgp_region_colors
 sbdivs = { a : np.zeros((bdivs[a].shape[0], len(supernames))) for a in bdivs }
[…]

### 2019-09-20 treestats_paper.tex
Get rid of empty section to stop Alg B splitting over page.
updated relate disc

@@
 If we only considered summaries of the allele frequency spectrum of a single sample,
-our statistics would be equivalent to the framework described by \citet{fu1995statistical}.
+our statistics would be equivalent to the linear functions of the frequency spectrum
+discussed by \citet{fu1995statistical} and \citet{achaz2009frequency}.
 To connect our generalization to familiar ideas,
-here is a reframing of \citeauthor{fu1995statistical}'s statistics.
+here is a reframing of these statistics of the frequency spectrum.
 Write $p_i(a)$ for the frequency of allele $a$ at a site $i$,
@@
 Given the trees, these numbers of mutations are independent and Poisson,
-and the variance of $f(x)$ multiplied by a Poisson random variable with mean $m$ is $m f(x)^2$.
+and the variance of the product of the constant $f(x)$
+and a Poisson random variable with mean $m$ is $m f(x)^2$.
 The sum is divided by the length of the region, $j-i$,
@@
 % from ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase1/analysis_results/supporting/accessible_genome_masks/README.20120515_phase1_pop_gen_mask
-As shown in Figure~\ref{fig:chr20}, Branch and Site diversity along the genome are very similar.
-Is this an indication that Relate is doing a good job at inferring trees,
-or just a byproduct of the inference method
-(which, naturally, estimates branch lengths to be proportional to numbers of mutations)?
+As shown in Figure~\ref{fig:chr20}, the ratio of Site and Branch diversity
+is relatively constant, hovering around typical estimates for the average genome-wide human mutation rate~\citep{narasimhan2017estimating}.
+On the one hand, this is expected ---
+Relate estimates branch lengths in part assuming that mutations fall at a constant rate through time
+(although \citeauthor{speidel2019method} also showed signal of changing mutation rates,
+pointing the way towards a more general method).
+On the other hand,
+since estimated node ages are shared across long regions of the genome,
+a tight agreement may not be possible with poorly inferred trees.
+So, the relative constancy of the ratio of Site to Branch diversity
+suggests that Relate is doing a good job at inferring trees,
+but what to make of the twofold variation in this ratio?
 Answering this question requires a deeper understanding of the processes that shape
@@
 %%%%%%%%%%%%%%%%%%
-\section*{Efficient algorithms and implementation}
-
-%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Algorithms}
+\section*{Algorithm and implementation}
+
 In this section we describe and analyze the algorithm used

### 2019-09-20 treestats_paper.tex
andy comments

@@
 reproducing organisms as recombination results in many trees along the
-genome~\citep{griffiths1991two,griffiths1996ancestral,rasmussen2014genome},
+genome~\citep{hudson1983properties,griffiths1991two,griffiths1996ancestral,rasmussen2014genome},
 and there was no efficient way of representing this information.
@@
 From the tree sequence can be extracted a sequence of trees,
-$\treeseq = (T_1, T_2, \ldots, T_{n(\treeseq)})$
-and a sequence of breakpoints $0 = a_0 < a_1 < \cdots < a_{n(\treeseq)} = L$,
+$\treeseq = (T_1, T_2, \ldots, T_{|\treeseq|})$
+and a sequence of breakpoints $0 = a_0 < a_1 < \cdots < a_{|\treeseq|} = L$,
 where $T_k$ describes the genealogical relationships of the samples
-over the segment of genome between $a_{k-1}$ (inclusive) and $a_k$ (exclusive).
+over the segment of genome between $a_{k-1}$ (inclusive) and $a_k$ (exclusive),
+and $|\treeseq|$ is the number of trees.
 We say that tree $T_k$ \emph{covers} the (half-open) segment $[a_{k-1}, a_k)$,
@@
     The \emph{total weight} is the sum of the weights over all samples:
-    $\tiw = \sum_v \iw(v)$, which is the subtree weight of the root of any tree
-    that contains all samples.
+    $\tiw = \sum_v \iw(v)$.
 \end{definition}
@@
 The requirement that $f(0) = f(\tiw) = 0$ is for consistency,
-so that statistics do not depend on portions of the tree either not ancestral to any of the samples or ancestral to all
-of them.
+so that statistics do not depend on portions of the tree either not ancestral to any of the samples
+or ancestral to all of them.
 % Below, in section XXX, we describe how to efficiently
@@
     =
-    \frac{1}{L} \sum_{k=1}^{n(\treeseq)} L_k \left( f(\nw_{T_k}(u)) + f(\tiw - \nw_{T_k}(u)) \right).
+    \frac{1}{L} \sum_{k=1}^{|\treeseq|} L_k \left( f(\nw_{T_k}(u)) + f(\tiw - \nw_{T_k}(u)) \right).
 \end{align}
@@
     =
-    \frac{1}{L} \sum_{k=1}^{n(\treeseq)} L_k f(\nw_{T_k}(u)) .
+    \frac{1}{L} \sum_{k=1}^{|\treeseq|} L_k f(\nw_{T_k}(u)) .
 \end{align}
@@
     which is the density of segregating sites if all sites are biallelic.
+    (The actual number of mutations per unit length will be greater if there have been backmutations.)
 \end{example}
+
+\paragraph{Polyallelic sites}
+The summary function in Example~\ref{ex:segregating_sites} might be surprising:
+the reason for its particular form is so that the statistic returns a sensible answer
+even when sites may have more than two alleles.
+There is not a consensus in the field
+on how to use sites with more than two alleles, but the most common practice is
+to reduce to biallelic data, by either discarding other sites
+or by marking all non-ancestral alleles as (the same) ``derived'' allele.
+We choose to define statistics in a way that still makes formal sense for polyallelic sites,
+so that for instance a site with ancestral allele $A$ and derived alleles $C$ and $T$
+would have site statistic $f(\aw(A)) + f(\aw(C)) + f(\aw(T))$.
+This is not the only choice, but is natural because it agrees with what is obtained by
+looking at pairwise haplotype differences.
+For instance, the definition of nucleotide divergence in Example \ref{ex:site_divergence}
+gives the probability that a randomly chosen sample from each set differ,
+a definition that makes sense even with polyallelic sites.
 
@@
 
-\paragraph{Polyallelic sites}
-The statistics defined above map exactly onto those standard in population genetics
-when all sites are biallelic.
-There is not a consensus in the field
-on how to use sites with more than two alleles, but the most common practice is
-to reduce to biallelic data, by either discarding other sites
-or by marking all non-ancestral alleles as (the same) ``derived'' allele.
-We choose to define statistics in a way that still makes formal sense for polyallelic sites,
-so that for instance a site with ancestral allele $A$ and derived alleles $C$ and $T$
-would have site statistic $f(\aw(A)) + f(\aw(C)) + f(\aw(T))$.
-This is not the only choice, but is natural because it agrees with what is obtained by
-looking at pairwise haplotype differences.
-For instance, the definition of nucleotide divergence in Example \ref{ex:site_divergence}
-gives the probability that a randomly chosen sample from each set differ.
-This definition makes sense even at polyallelic sites,
-and (as we discuss below) has a natural correspondence to statistics related to branch lengths.
-
 %%%%%%%%%%%%%%%%%%
@@
     &=
-    \frac{1}{j-i} \sum_{k=1}^{n(\treeseq)} \ell_k(i,j) \branch(f, \iw)_{T_k} ,
+    \frac{1}{j-i} \sum_{k=1}^{|\treeseq|} \ell_k(i,j) \branch(f, \iw)_{T_k} ,
 \end{align}
@@
 whole-population Site diversity is itself only a quite noisy estimator of Branch diversity.
+(Note, however, that statistics other than diversity may vary much more between subsamples.)
 The same things are shown in Supplementary Figure \ref{sfig:sweep_duality_10K}
@@
 speed up inference based on summaries computed from simulated
-data~\citep{beaumont2002approximate,csillery2010approximate,schrider2018supervised,flagel2018unreasonable}.
+data~\citep{beaumont2002approximate,csillery2010approximate,schrider2018supervised}.
 The ability to efficiently compute statistics from real data is, of course, also welcome.

### 2019-09-23 submit/cover_letter.tex
cover letter and line numbers

\documentclass[12pt,letterpaper]{article}
\input{plrletter-commands}
\usepackage[hidelinks]{hyperref}

\signature{
    Peter Ralph, on behalf of Kevin Thornton and Jerome Kelleher.}
\name{Peter Ralph}
\address{Peter Ralph\\ Institute for Ecology and Evolution\\ and Departments of Mathematics and Biology,\\ University of Oregon\\
plr@uoregon.edu}
\telephone{(707) 502-5854}
\email{plr@uoregon.edu}

\begin{document}

\pagestyle{plain}

\noindent

\opening{To the editor(s) --}
\hspace{\stretch{1}}
\today

    We are writing to submit our manuscript, entitled
    \emph{Efficiently summarizing relationships in large samples: a general
        duality between statistics of genealogies and genomes},
    for consideration for publication in \textit{Genetics}.

    An important part of the paper is that it introduces fast methods for computation
    of population genetic statistics from large numbers of genomes using tree sequences,
    available through the tested and validated python package \texttt{tskit}.
    However, that's not the main point ---
    if it was, we'd be submitting to a more bioinformatics-focused journal. 
    It is well-known that population genetics statistics say something about the underlying genealogies,
    although it is not common for researchers to explicitly discuss this.
    The bigger point of our manuscript is to describe a general framework that makes it clear
    precisely what aspect of the underlying genealogies is described by each statistic,
    under the infinite-sites model of mutation,
    and to provide methods to compute these statistics of tree shape as well.
    We think this makes population genetics statistics conceptually cleaner,
    but has some other possible benefits, besides speed:
    (1) we make it easy to invent and implement new statistics;
    (2) the framework allows separation of mutational from demographic noise;
    and (3) if unbiased tree sequences can be inferred from genomic data,
    summarizing the trees themselves provides a descriptive method
    insensitive to variation in substitution rate or perhaps even marker ascertainment.
    The last is dependent on testing and improvement of tree sequence inference,
    but would be tremendously useful, if it works out.

    We have put a lot of effort into making the paper accessible to the readership of \textit{Genetics}.
    The paper begins with a careful treatment of the theoretical framework and many examples,
    then illustrates points with simulation and a tree sequence inferred from 1000 Genomes data
    (by Speidel et al, 2019),
    and finally describes the fundamental algorithm and compares computational efficiency
    to other methods.
    The definitions and framework might seem lengthy,
    but in our assessment they are necessarily so:
    there are many false paths that intuition can lead even experienced theoreticians down,
    and choices we have made in our framework that might seem strange at first
    turn out to be natural once carefully considered.

    We are enthusiastic about the potential impact of this paper
    and its interest to the readers of \textit{Genetics}.
    We hope that you agree.

\closing{Sincerely,}

\end{document}

### 2019-09-23 treestats_paper.tex
funding

@@
 \include{macros}
+\newif\ifreviewresponses
+\reviewresponsestrue  % include them
+% \reviewresponsesfalse  % don't include them
+\newcommand{\responsefile}{review-responses.tex}  % name of the review reponses file
+\input{review-response-commands} % for line numbers only for now
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%
+\section*{Data and code accessibility}
+
+All methods described here are implemented in python and C in the package \tskit{},
+available from \url{https://github.com/tskit-dev/tskit}
+under the terms of the MIT license.
+All code used to produce the figures in this paper
+is available from \url{https://github.com/petrelharp/treestats_ms}.
+
+%%%%%%%%%%%%%%%%%%%%%%%
 \section*{Discussion}
@@
 and to Leo Speidel for providing the inferred 1000 Genomes tree sequence.
+This work was supported by NSF grant \#1262645 (DBI) to PLR
+and by NIH grant R01-GM115564 to KRT. JK was supported by
+the Robertson Foundation and Wellcome Trust grant 100956/Z/13/Z
+to Gil McVean.
 

### 2019-09-23 treestats_paper.tex
Added some references.

@@
 sequence (making the statistics' relationship to the underlying genealogical trees
-concrete and explicit) and suggest several new statistics.
+concrete and explicit),
+as well as the analogous ``branch'' statistics of tree shape. \revpoint{2}{10}
 We evaluate computational performance using simulated data, and
@@
 
-\begin{example}[Ancestry proportions] \label{ex:ancestry_props}
+\begin{example}[Ancestry proportions] \label{ex:ancestry_props} \revpoint{2}{5}
     If $\iw$ are the indicator weights of the set $S$ of $n$ samples,
@@
     that inherit each allele.
-    A site with $k$ distinct alleles that segregate in $S$ with proportions $p_1 + \cdots + p_k = 1$
-    contributes $f(p_1) + \cdots + f(p_k) = (1 - p_1) + \cdots + (1 - p_k) = k - 1$ to the statistic.
+    A site with $k$ distinct alleles that segregate in $S$ with counts $x_1 + \cdots + x_k = n$
+    contributes $f(x_1) + \cdots + f(x_k) = (1 - x_1/n) + \cdots + (1 - x_k/n) = k - 1$
+    to the statistic. \revpoint{2}{6}
     Therefore, $\site(f, \iw)_{[a,b)}$, is the minimum number of derived mutations per unit length,
@@
 
+Note that we have taken care in all of these definitions
+so that they make sense even with more than two alleles per site.
+For instance, the $f_4$ statistic is defined in \autoref{ex:site_f4} as follows:
+averaged across a random choice of individuals from the four groups of samples,
+let ``BABA'' be the proportion of sites at which samples 1 and 3 agree,
+but differ from samples 2 and 4 (which may differ from each other as well);
+and let ``ABBA'' be the proportion of sites at which 2 and 3 agree, but differ from 1 and 4;
+then $f_4$ is BABA minus ABBA.
+In this mnemnonic, B is standing in for a \emph{specific} allele, that must match;
+but A is standing in for ``any allele that is not B''.
+\revpoint{2}{7}
+
 %%%%%%%%%%%%%%%%%%
@@
 
-\begin{example}[Patterson's $f_4$] \label{ex:branch_f4}
+\begin{example}[Patterson's $f_4$] \label{ex:branch_f4} \revpoint{2}{7}
     Suppose that the four subsets each consist of only a single sample.
@@
     &=
-        \var\left[ \branch(f, \iw)_{[i,j)} \right]
+        \mu^2 \var\left[ \branch(f, \iw)_{[i,j)} \right]
             +
-        \frac{1}{j-i} \E\left[ \branch(f^2, \iw)_{[i,j)} \right] .
+        \frac{\mu}{j-i} \E\left[ \branch(f^2, \iw)_{[i,j)} \right] .
 \end{align*}
@@
 which we indeed find
-(the genome-wide mean Branch $f_4$ is around -700,
+(the genome-wide mean Branch $f_4$ is around -700 generations, \revpoint{2}{12}
 as is the Site $f_4$ divided by mutation rate;
@@
 of the 1000 Genomes dataset~\citep{1k2015global} using dated trees estimated
-by Relate~\citep{speidel2019method}.
+by Relate~\citep{speidel2019method}. \revpoint{2}{9}
 Specifically, we calculated diversity using the Site statistic
@@
 data~\citep{beaumont2002approximate,csillery2010approximate,schrider2018supervised}.
+Efficient simulations coupled with the framework developed here allow us to
+explore the full distribution of summaries along the genome, which has
+important applications in (e.g.) speciation genomics~\citep{lohse2017come}.
 The ability to efficiently compute statistics from real data is, of course, also welcome.
@@
 The correspondence between genome sequence and properties of the
-underlying genealogies we have used here is well-known. The general framework
+underlying genealogies we have used here is well-known,
+and is the basis for several inference methods~\citep{becquet2007new,beeravolu2018able}.
+The general framework
 that we have defined, however, codifies this relationship in a
@@
 \subsection*{Acknowledgments}
-Thanks to Graham Coop, Andrew Kern, Nick Patterson, Kelley Harris and
-Aaron Ragsdale for useful suggestions
+Thanks to Graham Coop, Andrew Kern, Nick Patterson, Kelley Harris,
+Aaron Ragsdale and Konrad Lohse for useful suggestions
 and to Leo Speidel for providing the inferred 1000 Genomes tree sequence.
@@
 to Gil McVean.
+
+\paragraph{Author contributions}
+Designed the study: PR, JK. Wrote the code: PR, JK. Carried out the experiments: PR, KT, JK. Wrote the paper: PR, KT, JK.
 
@@
 
+\includereviews
+
 \end{document}

### 2019-09-23 treestats_paper.tex
added missing tex
contributions

--- 2019-09-23 treestats_paper.tex
@@
 \newif\ifreviewresponses
-\reviewresponsestrue  % include them
-% \reviewresponsesfalse  % don't include them
+% \reviewresponsestrue  % include them
+\reviewresponsesfalse  % don't include them
 \newcommand{\responsefile}{review-responses.tex}  % name of the review reponses file
@@
 %%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Acknowledgments}
+\subsection*{Acknowledgments and author contributions}
 Thanks to Graham Coop, Andrew Kern, Nick Patterson, Kelley Harris and

### 2019-11-04 review-responses.tex
review responses

%%%%%%
%%
%%  Don't reorder the reviewer points; that'll mess up the automatic referencing!
%%
%%%%%

\begin{minipage}[b]{2.5in}
  Resubmission Cover Letter \\
    {\it Genetics}
\end{minipage}
\hfill
\begin{minipage}[b]{2.5in}
    Peter Ralph, \\
    Kevin Thornton, \\
    \emph{and} Jerome Kelleher \\
  \today
\end{minipage}

\vskip 2em

\noindent
{\bf To the Editor(s) -- }

\vskip 1em

We are writing to submit a revised version of our manuscript, 
``Efficiently summarizing relationships in large samples: a general duality between statistics of genealogies and genomes''.

\vspace{5em}

\noindent \hspace{4em}
\begin{minipage}{3in}
\noindent
{\bf Sincerely,}

\vskip 2em

{\bf 
Peter Ralph, Kevin Thornton, and Jerome Kelleher
}\\
\end{minipage}

\vskip 4em

\pagebreak
\setcounter{page}{1}

%%%%%%%%%%%%%%
\reviewersection{AE}

\begin{quote}
It is most important that you address the following in a revised manuscript: thorough revisions of Figures 1 and 2, and the text in lines 80-101.
\end{quote}

%%%%%%%%%%%%%%
\reviewersection{1}

\begin{quote}
Ralph et al.\ describe a duality between summary statistics obtained from marginal trees in ancestral recombination graphs (ARG) and site-based summary statistics.
They use simulated data to demonstrate this leads to computational speed-ups and apply their methods using the ARG of 1000 Genomes data recently inferred by Speidel et al.
As the authors point out, the idea of a duality between genotypic and genealogical data is not new and has been extensively exploited in the past both theoretically and in practice,
but this paper presents a unified view that will likely be of interest to the readers of Genetics.
\end{quote}

\begin{point}{Figure 1:}
    This is an important figure in the paper. I think the figure (and/or the legend) can be improved, mainly by adding a definition of all quantities being used (at the beginning of the legend or in a box within the figure).
    \begin{itemize}
        \item when first reading the paper, it takes some effort to understand what "a" and "b" represent in $f(a,b)$
            (and it doesn't help that this is a multi-panel figure with panels named the same way).
            It would also be good to make clear what the numbers "3" and "2" reflect, which requires some interpretation (this could be just stated explicitly, as for other parameters).
        \item "seven mutations", I believe the authors mean "five mutations".
        \item "plus the weight of the node itself, if it is a sample", this does not apply to this example and is confusing, it should probably be mentioned in Def 1 but not Fig 1.
        \item "remaining weight of each mutation" has not been defined in the figure, which is referred to before Def 1 is encountered.
    \end{itemize}
\end{point}

\reply{
}

\begin{point}{}
    Perhaps "u" would work best as first argument rather than subscript in Eq 1's LHS. Similar for "j" in Eq 2, Eq 4, and related notation throughout.
\end{point}

\reply{
}

%%%%%%%%%%%%%%
\reviewersection{2}

\begin{quote}
This is an important paper that explores summary statistics at the level of inferred genealogical trees along the genome, as opposed to the standard of summary statistics of SNPs (i.e.\ sites).
The authors present a new framework of thinking about and generalising summary statistics such as pairwise nucleotide diversity or Patterson's f4 statistics.
They use simulations and data analysis based on 1000 Genomes data to show how their newly introduced "branch statistics" differ from traditional "site" statistics.
I think overall results are convincing, the methods are sound, and this is definitely impressive and important work!
\end{quote}

\begin{point}{}
The entire section from around line 80 (starting with "Figure 1 shows a simple example of these procedures") up until line 101 (ending with "correlation between genotype and phenotype.") was utterly confusing to me. Sentences such as the one in L 84:
    \begin{quote}
"For instance, we could calculate how many of a certain set A of samples inherit from a particular branch in a tree by (a) assigning each of these samples weight 1 (and other samples weight zero), then (b) finding the total weight of all samples in the subtree inheriting from that branch."
    \end{quote}
were not at all understandable to me at that early point in the manuscript. And Figure 1 didn't help (see below). At this point, the reader doesn't even know what weights, branches and subtrees are, and what is being attempted here. Why are we talking about trees and not sequence data, which is usually the starting point in genomic analyses? Why are we summarising things along branches? If it were me, I would really start with site statistics, as it's something all readers can relate to, and which is already non-trivial given the generalisations using weights and summary functions and in particular the generalisation to for k-alleles. But Site statistics were literally the first thing in the paper I really understood. From there it was fairly easy for me to move on to branch statistics (as determining the expectation of site stats) and then understanding the essence of comparing the two. I still don't actually understand why "Node statistics" are even introduced. As far as I can see the paper is all about Site vs. Branch statistics, with Node statistics being introduced and then ignored for the rest of the paper. It's weird to start off with something that can never be computed from real data as such, since trees are fundamentally unobserved. At the least, these Node Statistics should be very carefully motivated and explained better.
\end{point}

\reply{
}

\begin{point}{}
Figure 1 is not understandable. It's not even understandable to me now, having read the entire paper. Some feedback: I'm seeing trees with 5 samples, which are gray in the first, and blue/red in the second tree. I'm seeing five mutations as yellow stars (do the mutations relate to the red/blue?). I'm reading in the legend something about 7 mutations and 10 sites, and neither tof these two numbers are found somewhere in the figure. I see bracketed pairs of numbers next to nodes that I don't understand, and I see confusing equations. And the legend doesn't help.

I absolutely share a sense with the authors that there should be such a figure, but I think the choice of what information to show in the graphic is simply not working (for me). I don't really know what to do about it, since - as written above - I don't even understand why "Node statistics" need to be introduced at that point in the text. I have two comments that might help improving or replacing that figure:
- Consider using a different summary function. The sequence divergence is among the more complex statistics, which require two groups of samples. Why not taking something much simpler, such as pairwise nucleotide diversity?
- Consider replacing this figure with one that explains how Site statistics are calculated using weights and a summary function.

I don't know, but at the very least, if you want to keep that figure more or less as is, you need to spend an entire paragraph explaining it properly to the reader. What's written in main text and legend about that figure is simply not sufficient!
\end{point}

\reply{
}

\begin{point}{}
Figure 2 is equally hard to understand as Figure 1, but in that case I could at least understand it after significant contemplation so hopefully can give more constructive feedback for improvement. I think it's great to show a figure that shows that correspondence between site and branch statistics that the whole paper is about. But I think in case of panel a) it would help enormously to also see the traditional representation via a genotype site matrix to pick up readers from a more familiar point. For example, in order to confirm the weights contributed from each mutation, I had to go through every sub-group (A,B,C,D) and see what the frequency of that mutation is in that subgroup, which is very indirect using the tree-representation. I then understand that the red mutation, for example, has frequency 1/2 in A, 1 in B, and 0 in both C and D. So then (pA-pC)(pB-pD) give -1/2 which is written next to the mutation. This kind of calculation is of course trivial, but the connection is very hard to make for a reader who tries to understand that figure. And by the way, the colouring is not explained. I get that you want to create the correspondence to the coloured branches in B), but that is not explained. With a genotype-site-matrix, with individual genomes ordered in A, B, C and D, one could somehow visualise better the allele frequency in each subgroup and how that statistics comes about.
\end{point}

\reply{
}

\begin{point}{Figure 2:}
There is a problem, by the way, since the figure makes it look as if f4 statistics were extensive, i.e. summed up along the genome. But they are actually normalised by the number of sites. So the f4 statistics in panel a) shouldn't be 1.0, but 1 divided by the sequence length. Or am I missing something?
\end{point}

\reply{
}

\begin{point}{L 145}
        The term "Ancestry Proportions" is in my view a poor choice of naming. This term usually refers to proportions of contributions from populations in mixture models rather than node-specific contributions. For example, the ancestry of African-Americans can be decomposed into an ancestry proportion from Europe and one from West-Africa.
\end{point}

\reply{
}

\begin{point}{L 198.}
    I think there is a small bug in the notation. You introduce $p_1+\ldots+p_k=1$ as proportions, and then write $f(p_1)+\ldots+f(p_k)$. But $f(x)$ is defined in terms of allele counts $x$, not allele frequencies $p$, or am I misunderstanding? I think it would be more consistent to avoid $p_1,\ldots,p_k$ and instead work with $\bar{x}$ as you already had further above, or so.
\end{point}

\reply{
}

\begin{point}{L 224}
    Patterson's f4: Is this definition of f4 generalised to k alleles, as is Example 4 (segregating sites)? If so, could this be briefly commented on in a sentence, as I think that would be new over previous definitions.
\end{point}

\reply{
}

\begin{point}{References}
    should be alphabetically ordered in a author-year referencing style. It's hard to lookup a reference.
\end{point}

\reply{
}

\begin{point}{L 353:}
I was surprised that the authors didn't use "tsinfer" on 1000 Genomes to get to the trees instead of Relate. I'm sure both methods are good, but given the overlap in authorship I would have expected "tsinfer"-ed trees to be analysed with "tskit".
\end{point}

\reply{
}

\begin{point}{Abstract L 16:}
    "several new statistics"? I didn't see newly introduced statistics that aren't already known.
\end{point}

\reply{
}

\begin{point}{}
    Some notes about how to deal with missing data might be useful to understand how this framework could be applied to incomplete real data.
\end{point}

\reply{
}

\begin{point}{L 343:}
    "the genome-wide mean Branch f4 is around -700". What does that number mean? Shouldn't that be normalised by the number of sites considered? Usually, f4 statistics are reported with a Z-score that gives statistical support for how far away from 0 the statistics is. Is there any way to do that here?
\end{point}

\reply{
}

### 2020-01-18 review-responses.tex
minor revisions

@@
 
-\begin{point}{L 145}
+\begin{point}{\revref}
         The term "Ancestry Proportions" is in my view a poor choice of naming.
@@
 
-\begin{point}{L 198.}
+\begin{point}{\revref}
     I think there is a small bug in the notation.
@@
 \reply{
+    Thanks for catching that! Fixed. \revref
 }
 
-\begin{point}{L 224}
+\begin{point}{\revref}
     Patterson's f4: Is this definition of f4 generalised to k alleles, as is Example 4 (segregating sites)?
@@
 \reply{
+    This does generalize the usual definition -- we have added some explanation of this (that ended up being longer than a sentence). \revref
 }
@@
 \reply{
+    We think you must have gotten confused by the first names --
+    the references actually \emph{are} ordered alphabetically, by last name of first author (as usual).
 }
 
-\begin{point}{L 353:}
+\begin{point}{\revref}
 I was surprised that the authors didn't use "tsinfer" on 1000 Genomes to get to the trees instead of Relate.
@@
 \reply{
+    We would love to also! However, at time of writing,
+    tsinfer does not attempt to infer trees with well-calibrated branch lengths.
 }
 
-\begin{point}{Abstract L 16:}
+\begin{point}{Abstract \revref}
     "several new statistics"? I didn't see newly introduced statistics that aren't already known.
@@
 \reply{
+    Good point: we think the tree statistics and the definitions allowing for multiple alleles
+    are in some ways new, but we don't introduce new genotype statistics.
+    We've changed this phrase. \revref
 }
@@
 \reply{
+    We agree, but it's not something we've implemented yet in \tskit,
+    so we don't want to discuss it here (as people would get the wrong idea it was implemented already).
 }
 
-\begin{point}{L 343:}
+\begin{point}{\revref}
     "the genome-wide mean Branch f4 is around -700".
@@
 \reply{
+    The Branch $f_4$ statistic is in units of generations, or really, net generations
+    (generations of BABA minus generations of ABBA).
+    We've added the units \revref, although this might be confusing.
+    We could $z$-normalize the scores, but then we couldn't
+    demonstrate that the Branch and Site values match when the right units are used.
 }

### 2020-01-18 treestats_paper.tex
review responses
minor revisions

--- 2019-09-23 treestats_paper.tex
@@
 data~\citep{beaumont2002approximate,csillery2010approximate,schrider2018supervised}.
-Efficient simulations coupled with the framework developed here allow us to
-explore the full distribution of summaries along the genome, which has
-important applications in (e.g.) speciation genomics~\citep{lohse2017come}.
 The ability to efficiently compute statistics from real data is, of course, also welcome.
@@
 The correspondence between genome sequence and properties of the
-underlying genealogies we have used here is well-known,
-and is the basis for several inference methods~\citep{becquet2007new,beeravolu2018able}.
-The general framework
+underlying genealogies we have used here is well-known. The general framework
 that we have defined, however, codifies this relationship in a
@@
 %%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Acknowledgments}
-Thanks to Graham Coop, Andrew Kern, Nick Patterson, Kelley Harris,
-Aaron Ragsdale and Konrad Lohse for useful suggestions
+\subsection*{Acknowledgments and author contributions}
+Thanks to Graham Coop, Andrew Kern, Nick Patterson, Kelley Harris and
+Aaron Ragsdale for useful suggestions
 and to Leo Speidel for providing the inferred 1000 Genomes tree sequence.

### 2020-01-18 treestats_paper.tex

--- 2019-09-23 treestats_paper.tex
@@
 
-All methods described here are implemented in python and C in the package \tskit{},
+All methods described here are implemented in Python and C in the package \tskit{},
 available from \url{https://github.com/tskit-dev/tskit}

### 2020-01-18 treestats_paper.tex
fig 1 draft

@@
     \begin{center}
-    \includegraphics{figures/divergence_diagram}
+    \includegraphics{figures/divergence_diagram_simple}
     \end{center}
     \caption{
+        \textbf{FIXME:}
         An example of computing sequence divergence between red nodes and blue nodes

### 2020-04-01 treestats_paper.tex
fig 1 tweaks, closes #38

@@
         \textbf{FIXME:}
-        An example of computing sequence divergence between red nodes and blue nodes
-        from seven mutations at distinct sites in a sequence of length 10 using the tree.
-        The summary function, $f(a,b) = (a/3)(1-b/2)$ gives the probability that
-        an allele found in $a$ of the red nodes and $b$ of the blue nodes is present in
-        a randomly chosen red node but not in a randomly chosen blue node.
+        An example of computing sequence divergence
+        between red samples (triangles) and blue samples (squares)
+        at a single site with a T$\to$G mutation, using the tree.
+        The derived mutation is found in $a=2$ out of the $n_\text{red} = 3$ red samples
+        and in $b=1$ of the $n_\text{blue}=2$ blue samples, and so the probability that a randomly chosen red node carries the derived allele
+        but a randomly chosen blue node does not
+        is given by the summary function, $f(a,b) = f(2, 1) = 2/6$.
+        The complementary value, $f(1, 1) = 1/6$ is the probability that the mutation separates
+        a randomly chosen red and blue pair, but with the blue node carrying the derived allele,
+        so the total probability that a randomly chosen pair
+        differs at this site is $f(2,1) + f(1,1) = 1/2$.
         \textbf{(a)} Weights are assigned to each subtree by adding together the weights of any children,
         plus the weight of the node itself, if it is a sample.
-        \textbf{(b)} In this example, subtree weights $(a,b)$ record how many red nodes ($a$)
-        and how many blue nodes ($b$) lie in that subtree.
-        \textbf{(c)} The summary function shown is applied to both the weight of the subtree
+        \textbf{(b)} In this example, subtree weights $(a,b)$ record how many red samples ($a$)
+        and how many blue samples ($b$) lie in that subtree.
+        The summary function shown is applied to both the weight of the subtree
         and the remaining weight of each mutation,
@@
         contributes $f(2,1) + f(1,1) = 1/3 + 1/6 = 1/2$.
-        The final value for divergence is the sum of these values divided by 10,
-        the total number of sites.
         \label{fig:divergence_diagram}

### 2020-04-01 treestats_paper.tex
cite ferretti; closes #42

@@
     \caption{
-        \textbf{FIXME:}
         An example of computing sequence divergence
@@
 The range of statistics available to describe variation in a single exchangeable sample
-(e.g., an isolated population) are well-understood,
+(e.g., an isolated population) are well-understood
+\citep{achaz2009frequency,ferretti2017decomposing},
 but there are much larger and less well-explored classes of statistics

### 2020-04-01 treestats_paper.tex

@@
 of propagating weights up a tree and summarizing them,
-used to calculate mean sequence divergence over a short sequence spanned by a single tree.
+used to calculate mean sequence divergence at a segregating site.
 
@@
         and how many blue samples ($b$) lie in that subtree.
-        The summary function shown is applied to both the weight of the subtree
-        and the remaining weight of each mutation,
-        so for instance the mutation above the node with weights $(2,1)$
-        contributes $f(2,1) + f(1,1) = 1/3 + 1/6 = 1/2$.
         \label{fig:divergence_diagram}
@@
     to every sample node.
-    Given these and a tree $T$,
-    the \emph{subtree weight} $\nw_T(u)$ for a node $u$ is the sum of all sample weights
+    Given these weights, on tree $T_k$
+    the \emph{subtree weight} $\nw_k(u)$
+    for a node $u$ is the sum of all sample weights
     of every sample node that is descended from $u$ in the tree (including $u$, if it is a sample):
     \begin{align*}
-        \nw_T(u) = \sum_{v \,:\, v \le_T u} \iw(v) ,
+        \nw_k(u) = \sum_{v \,:\, v \le_{T_k} u} \iw(v) ,
     \end{align*}
-    where $v \le_T u$ if $u$ is on the path from $v$ to root in the tree $T$.
+    where $v \le_{T_k} u$ if $u$ is on the path from $v$ to root in the tree $T_k$.
     The \emph{total weight} is the sum of the weights over all samples:
@@
 \end{definition}
+
+If a given node $u$ is a sample and has no offspring (i.e., it is a leaf),
+then its ``sample weight'' and ``subtree weight'' are the same: $\nw_k(u) = \iw(u)$.
+However, 
+these differ for any samples that are internal nodes in a tree,
+and so we use different notation to distinguish the two concepts.
+Also note that $\tiw$ is the subtree weight of the root of any tree,
+as long as the tree contains all the samples.
 
@@
     For a set of $k$-dimensional weights with total weight $\tiw$,
-    a summary function is a real-valued function $f(w_1, \ldots, w_k)$
-    with the property that $f(0) = f(\tiw) = 0$.
+    a summary function is any real-valued function $f(w_1, \ldots, w_k)$.
+    We call the summary function \emph{strict} if $f(0) = f(\tiw) = 0$.
 \end{definition}
 
-The requirement that $f(0) = f(\tiw) = 0$ is for consistency,
-so that statistics do not depend on portions of the tree either not ancestral to any of the samples
+The requirement that $f(0) = f(\tiw) = 0$ ensures
+that statistics do not depend on portions of the tree either not ancestral to any of the samples
 or ancestral to all of them.
+This is desireable because genetic variation within the samples
+cannot inform us about such parts of the tree.
+However, as we will see below it is sometimes useful to use non-strict summary functions.
 % Below, in section XXX, we describe how to efficiently
@@
 to be the sum of $f()$ applied to the weight of the subtree inheriting from node $u$
-and to the remaining weight,
+and to the remaining weight of the rest of the tree,
 averaged across the genome:
@@
 
-Node statistics only make sense in the context of the tree sequence,
-but provide a useful bridge to the next type of statistic,
+Note that the summary function in this last example was not \emph{strict}, since $f(\tiw) \neq 0$.
+Strictness is less important for node statistics than for the remaining classes of statistic,
+because Node statistics only make sense in the context of the tree sequence.
+However, Node statistics provide a useful bridge to the next type of statistic,
 which are defined directly in terms of the genotypes.
@@
 \end{align}
+Here, we think of a ``site'' as a single nucleotide position on the genome.
 We often want to summarize statistics across regions of the genome (``windows'').
@@
 \end{align}
-Since we have required $f(0) = f(\tiw) = 0$,
-the sum will only be affected by polymorphic sites,
-although the normalization is always by total number of sites, to allow comparison between regions.
+The requirement on strict summary functions that $f(0) = f(\tiw) = 0$
+ensures that the sum is only affected by polymorphic sites,
+although we normalize by total number of sites, to allow comparison between regions.
 
@@
 a statistic of tree shape by summing these expected contributions across its branches.
-We define the \textbf{branch statistic} of a single tree $T$,
+We define the \textbf{branch statistic} of the $k^\text{th}$ tree $T_k$,
 obtained from summary function $f()$ and sample weights $\iw$,
 to be the sum of the length of each branch
-multiplied by the summary function applied to its subtree weight and its complement:
+multiplied by the summary function applied to its subtree weight
+and the remaining weight not in the subtree:
 \begin{align}\label{eqn:branch_stat_tree}
-    \branch(f, \iw)_T
+    \branch(f, \iw)_k
     &=
-    \sum_{u \in T} \beta_{T}(u) \left( f(\nw_{T}(u)) + f(\tiw - \nw_{T}(u)) \right)  ,
+    \sum_{u \in T_k} \beta_{T_k}(u) \left( f(\nw_{k}(u)) + f(\tiw - \nw_{k}(u)) \right)  .
 \end{align}
-where $\beta_{T}(u)$ is the length of the branch ancestral to node $u$ in tree $T$,
+Here, $\beta_{T_k}(u)$ is the length of the branch between $u$ and its immediate ancestor in tree $T_k$,
 $\tiw = \sum_u \iw(u)$ is the total weight,
-and $\nw_{T}(u)$ is the total weight of the subtree of $T$ inheriting from $u$ (as defined above).
-The term $\tiw - \nw_{T}(u)$ gives the total weight \emph{not} in the subtree of $T$ inheriting from $u$.
+and $\nw_{k}(u)$ is the total weight of the subtree of $T_k$ inheriting from $u$ (as defined above).
+The term $\tiw - \nw_{k}(u)$ gives the total weight \emph{not} in the subtree of $T_k$ inheriting from $u$.
 The value $f(\nw(u))$ is the summary value that would be added to a site statistic
 if a single mutation occurred on the branch ancestral to $u$,
-and $f(\tiw - \nw_{T}(u))$ is the value that would be added due to its complementary allele,
-so $\branch(f, \iw)_T$ gives the expected contribution of the tree $T$ to $\site(f, \iw)$,
+and $f(\tiw - \nw_{k}(u))$ is the value that would be added due to its complementary allele,
+so $\branch(f, \iw)_k$ gives the expected contribution of the tree $T_k$ to $\site(f, \iw)$,
 per unit of sequence length that the tree covers.
@@
     \caption{
-    Computing the $f_4(A,B;C,D)$ statistic on a tree.
+    Computing the $f_4(A,B;C,D)$ statistic on a tree,
+    for groups of genomes labeled $A$, $B$, $C$, and $D$
+    (only group $A$ has more than one sample).
     \textbf{(a)} Each mutation is given weight equal to $(p_A - p_B)(p_C - p_D)$,
@@
     is the sum over these weights, multiplied by the length of the branch.
+    Remaining branches colored in grey do not contribute,
+    because any mutation falling on these will have either $p_A - p_B = 0$ or $p_C - p_D = 0$.
         \label{fig:branch_site_diagram}
@@
     &=
-    \frac{1}{j-i} \sum_{k=1}^{|\treeseq|} \ell_k(i,j) \branch(f, \iw)_{T_k} ,
+    \frac{1}{j-i} \sum_{k=1}^{|\treeseq|} \ell_k(i,j) \branch(f, \iw)_{k} ,
 \end{align}
 where $\ell_k(i,j)$ is the length of the region in $[i,j)$ that the tree $T_k$ covers
-(i.e., if $T$ covers the half-open interval $[a_{k-1},a_k)$,
+(i.e., if $T_k$ covers the half-open interval $[a_{k-1},a_k)$,
 then $\ell_k(i,j) = \max(0, \min(j,a_k) - \max(i,a_{k-1}))$).
@@
     \text{(polarized)} \qquad
-    \branchp(f, \iw)_T
+    \branchp(f, \iw)_k
     &=
-    \sum_{u \in T} \beta_T(u) f(\nw_{T}(u)) ,
+    \sum_{u \in T_k} \beta_{T_k}(u) f(\nw_{k}(u)) ,
 \end{align}
-and $\branchp(f, \iw)_{[i,j)}$ is defined using $\branchp(f, \iw)_T$ as before.
+and $\branchp(f, \iw)_{[i,j)}$ is defined using $\branchp(f, \iw)_k$ as before.
 
@@
     If we take $\iw$ and $f$ exactly as in the ``Nucleotide diversity'' example above,
-    then $f(u)$ gives the probability that the branch ancestral to $u$
+    then $f(u)$ gives the probability that the branch between $u$ and its ancestor in the tree
     lies on the path from one of two randomly chosen samples from $S$
@@
 \algstep{B1.}{Initialization.}{
-    For $0 \leq u < |\Et|$ set $\beta_u \leftarrow x_u
+    For $0 \leq u < |\Nt|$ set $\beta_u \leftarrow x_u
     \leftarrow F_u \leftarrow 0$ and $\pi_u \leftarrow -1$.
@@
 
-Algorithm \algref{B} begins by setting the initial state for $\pi$, $\beta$,
+Algorithm \algref{B} (named ``B'' for ``Branch'')
+begins by setting the initial state for $\pi$, $\beta$,
 $x$ and $F$ for each node in the tree sequence, and then sets the values
@@
 \subsection*{Acknowledgments}
-Thanks to Graham Coop, Andrew Kern, Nick Patterson, Kelley Harris,
-Aaron Ragsdale and Konrad Lohse for useful suggestions
+Thanks to Georgia Tsambos, Graham Coop, Andrew Kern, Nick Patterson, Kelley Harris,
+Aaron Ragsdale, Konrad Lohse, and Wilder Wohns for useful suggestions
 and to Leo Speidel for providing the inferred 1000 Genomes tree sequence.

### 2020-04-01 review-responses.tex
various edits, closes #43

@@
 
+Thanks -- we're glad you think it will be useful.
+
 \begin{point}{Figure 1:}
@@
     \begin{itemize}
-        \item when first reading the paper, it takes some effort to understand what "a" and "b" represent in $f(a,b)$
+        \item when first reading the paper, it takes some effort to understand what ``a'' and ``b'' represent in $f(a,b)$
             (and it doesn't help that this is a multi-panel figure with panels named the same way).
-            It would also be good to make clear what the numbers "3" and "2" reflect, which requires some interpretation (this could be just stated explicitly, as for other parameters).
-        \item "seven mutations", I believe the authors mean "five mutations".
-        \item "plus the weight of the node itself, if it is a sample", this does not apply to this example and is confusing, it should probably be mentioned in Def 1 but not Fig 1.
-        \item "remaining weight of each mutation" has not been defined in the figure, which is referred to before Def 1 is encountered.
+            It would also be good to make clear what the numbers ``3'' and ``2'' reflect, which requires some interpretation (this could be just stated explicitly, as for other parameters).
+        \item ``seven mutations'', I believe the authors mean ``five mutations''.
+        \item ``plus the weight of the node itself, if it is a sample'', this does not apply to this example and is confusing, it should probably be mentioned in Def 1 but not Fig 1.
+        \item ``remaining weight of each mutation'' has not been defined in the figure, which is referred to before Def 1 is encountered.
     \end{itemize}
@@
 \reply{
+    We've substantially reworked this figure,
+    simplifying it and providing more description in words.
 }
@@
 \begin{point}{}
-    Perhaps "u" would work best as first argument rather than subscript in Eq 1's LHS. Similar for "j" in Eq 2, Eq 4, and related notation throughout.
+    Perhaps ``u'' would work best as first argument rather than subscript in Eq 1's LHS. Similar for ``j'' in Eq 2, Eq 4, and related notation throughout.
 \end{point}
@@
 \reply{
+    Good suggestion, and we're sympathetic,
+    but after consideration we'd like to keep the conceptually quite different things separate
+    in the notation: $f$ and $w$ define the statistic (abstractly),
+    while $u$, $j$, and $k$ simply denote which portion of the genome it is calculated on.
 }
@@
 The authors present a new framework of thinking about and generalising summary statistics such as pairwise nucleotide diversity or Patterson's f4 statistics.
-They use simulations and data analysis based on 1000 Genomes data to show how their newly introduced "branch statistics" differ from traditional "site" statistics.
+They use simulations and data analysis based on 1000 Genomes data to show how their newly introduced ``branch statistics'' differ from traditional ``site'' statistics.
 I think overall results are convincing, the methods are sound, and this is definitely impressive and important work!
@@
 
+Thanks for the encouraging words!
+
 \begin{point}{}
-The entire section from around line 80 (starting with "Figure 1 shows a simple example of these procedures") up until line 101 (ending with "correlation between genotype and phenotype.") was utterly confusing to me.
+The entire section from around line 80 (starting with ``Figure 1 shows a simple example of these procedures'') up until line 101 (ending with ``correlation between genotype and phenotype.'') was utterly confusing to me.
 Sentences such as the one in L 84:
     \begin{quote}
-"For instance, we could calculate how many of a certain set A of samples inherit from a particular branch in a tree by (a) assigning each of these samples weight 1 (and other samples weight zero), then (b) finding the total weight of all samples in the subtree inheriting from that branch."
+``For instance, we could calculate how many of a certain set A of samples inherit from a particular branch in a tree by (a) assigning each of these samples weight 1 (and other samples weight zero), then (b) finding the total weight of all samples in the subtree inheriting from that branch.''
     \end{quote}
@@
 From there it was fairly easy for me to move on to branch statistics (as determining the expectation of site stats) and then understanding the essence of comparing the two.
-I still don't actually understand why "Node statistics" are even introduced.
+I still don't actually understand why ``Node statistics'' are even introduced.
 As far as I can see the paper is all about Site vs.
@@
 I absolutely share a sense with the authors that there should be such a figure, but I think the choice of what information to show in the graphic is simply not working (for me).
-I don't really know what to do about it, since - as written above - I don't even understand why "Node statistics" need to be introduced at that point in the text.
+I don't really know what to do about it, since - as written above - I don't even understand why ``Node statistics'' need to be introduced at that point in the text.
 I have two comments that might help improving or replacing that figure:
-- Consider using a different summary function.
+    \begin{itemize}
+        \item Consider using a different summary function.
 The sequence divergence is among the more complex statistics, which require two groups of samples.
 Why not taking something much simpler, such as pairwise nucleotide diversity?
-- Consider replacing this figure with one that explains how Site statistics are calculated using weights and a summary function.
+        \item Consider replacing this figure with one that explains how Site statistics are calculated using weights and a summary function.
+    \end{itemize}
 
@@
 \reply{
+    Apologies that this figure was so confusing!
+    We have substantially simplified this figure,
+    and have rewritten the explanation. Hopefully it is more clear now!
+    A different summary function is a good suggestion, but we actually think that divergence
+    is easier to understand than diversity, as it doesn't require sampling without replacement.
 }
@@
 \reply{
+    Good suggestion - we've added a genotype matrix and included more explanation
+    of what's going on here.
 }
@@
 \begin{point}{\revref}
-        The term "Ancestry Proportions" is in my view a poor choice of naming.
+        The term ``Ancestry Proportions'' is in my view a poor choice of naming.
 This term usually refers to proportions of contributions from populations in mixture models rather than node-specific contributions.
@@
 \begin{point}{\revref}
-I was surprised that the authors didn't use "tsinfer" on 1000 Genomes to get to the trees instead of Relate.
-I'm sure both methods are good, but given the overlap in authorship I would have expected "tsinfer"-ed trees to be analysed with "tskit".
+I was surprised that the authors didn't use ``tsinfer'' on 1000 Genomes to get to the trees instead of Relate.
+I'm sure both methods are good, but given the overlap in authorship I would have expected ``tsinfer''-ed trees to be analysed with ``tskit''.
 \end{point}
@@
 \begin{point}{Abstract \revref}
-    "several new statistics"? I didn't see newly introduced statistics that aren't already known.
+    ``several new statistics''? I didn't see newly introduced statistics that aren't already known.
 \end{point}
@@
 \begin{point}{\revref}
-    "the genome-wide mean Branch f4 is around -700".
+    ``the genome-wide mean Branch f4 is around -700''.
 What does that number mean?

### 2020-04-01 treestats_paper.tex
various edits, closes #43

@@
 
+Dividing this statistic by the total amount of the genome each node is ancestral to any samples,
+we would obtain the ``genomic descent'' statistic defined by \citet{scheib2019anglian}.
+
 Note that the summary function in this last example was not \emph{strict}, since $f(\tiw) \neq 0$.
@@
     (only group $A$ has more than one sample).
-    \textbf{(a)} Each mutation is given weight equal to $(p_A - p_B)(p_C - p_D)$,
-    where $p_X$ is the proportion of the samples inheriting that mutation (colored stars) in set $X$
-    (shown in labels at the bottom).
-    Note that each of the mutations shown occurs at a distinct site, so there is no back mutation.
-    \textbf{(b)} Each branch is assigned weight equal to the weight that would be assigned
-    to any mutation falling in it (colored lines), and then the branch $f_4$ statistic
+    \textbf{(a)} The genotype matrix, for five sampled genomes at five sites.
+    Also shown is the derived allele frequencies in each of the four groups,
+    denoted by $p_A$, $p_B$, $p_C$, and $p_D$ respectively.
+    \textbf{(b)} Each mutation is given weight equal to $(p_A - p_B)(p_C - p_D)$,
+    shown as a label next to each mutation (depicted as a star).
+    Note that each of the mutations shown occurs at a distinct site, so there is no back mutation,
+    and that all mutations on the same branch have the same weight.
+    \textbf{(c)} Each branch is assigned weight equal to the weight that would be assigned
+    to any mutation falling in it (colored lines, with colors matching the mutations of (b)),
+    and then the branch $f_4$ statistic
     is the sum over these weights, multiplied by the length of the branch.
     Remaining branches colored in grey do not contribute,
-    because any mutation falling on these will have either $p_A - p_B = 0$ or $p_C - p_D = 0$.
+    because any mutation falling on these would have either $p_A - p_B = 0$ or $p_C - p_D = 0$.
         \label{fig:branch_site_diagram}
@@
 by Relate~\citep{speidel2019method}. \revpoint{2}{9}
+(Although Relate does not directly return a tree sequence, its output can be converted to a tree sequence.)
 Specifically, we calculated diversity using the Site statistic

### 2020-04-02 review-responses.tex
responses to other reviews

@@
 
+In response to reviewer comments (and feedback from other sources)
+we have substantially rewritten the introductory portions of the manuscript,
+hopefully making it more accessible to a wider audience.
+We think the manuscript is improved, and hope you agree.
+
+In addition to the usual revised manuscript and responses to reviewers (in which page and line numbers
+refer to the revised manuscript),
+we have also uploaded a ``diff'' highlighting changes since the previous submission.
+
 \vspace{5em}
@@
 
+We have thoroughly revised both figures, and added an additional figure,
+to ease the conceptual burden,
+as well as reorganized the introductory section,
+adding a more gentle introduction to the concepts.
+
 %%%%%%%%%%%%%%
@@
 
-Thanks -- we're glad you think it will be useful.
+Thanks -- we're glad you think it will be useful!
 
-\begin{point}{Figure 1:}
+\begin{point}{Figure \ref{fig:divergence_diagram}}
     This is an important figure in the paper.
@@
     We've substantially reworked this figure,
-    simplifying it and providing more description in words.
+    simplifying it and providing more description in words,
+    as well as adding another figure before it
+    to reinforce the same basic intuition.
+    We've also rephrased the undefined phrase ``remaining weight''.
 }
@@
 \reply{
+    Our goal here was to begin with something more familiar to readers --
+    the allele frequency spectrum.
+    We seem to have missed the mark, based on these and other comments we received.
+    (Apologies.)
+    In response, we've rearranged the text, moving the description of Figure \ref{fig:divergence_diagram}
+    later,
+    introducing the concepts with an additional figure (Figure \ref{fig:allele_freqs}),
+    and adding substantial additional text.
+    We hope it is is better motivated now?
 }
@@
     We have substantially simplified this figure,
-    and have rewritten the explanation. Hopefully it is more clear now!
+    have rewritten the explanation,
+    and begun with a different, simpler figure.
+    Hopefully it is more clear now!
     A different summary function is a good suggestion, but we actually think that divergence
@@
     Good suggestion - we've added a genotype matrix and included more explanation
-    of what's going on here.
+    of what's going on here;
+    and now this example also ties in to Figure \ref{fig:allele_freqs},
+    providing additional context.
 }
@@
 \reply{
+    Good point; we've added division by sequence length ($L$) to the figure and the caption.
 }
@@
 \reply{
+    This is true, but aren't these actually the same concept?
+    For instance, the proportion of a group of samples that inherit from West Africa,
+    is equal to the sum of the statistic of Example \ref{ex:ancestry_props}
+    over all ``West African'' ancestors.
+    We'd prefer to keep the name.
 }
@@
     I think there is a small bug in the notation.
-You introduce $p_1+\ldots+p_k=1$ as proportions, and then write $f(p_1)+\ldots+f(p_k)$.
+You introduce $p_1+\cdots+p_k=1$ as proportions, and then write $f(p_1)+\ldots+f(p_k)$.
 But $f(x)$ is defined in terms of allele counts $x$, not allele frequencies $p$, or am I misunderstanding?
@@
 \begin{point}{\revref}
-    Patterson's f4: Is this definition of f4 generalised to k alleles, as is Example 4 (segregating sites)?
+    Patterson's $f_4$: Is this definition of $f_4$ generalised to $k$ alleles, as is Example \ref{ex:segregating_sites} (segregating sites)?
 If so, could this be briefly commented on in a sentence, as I think that would be new over previous definitions.
@@
 \reply{
-    We think you must have gotten confused by the first names --
-    the references actually \emph{are} ordered alphabetically, by last name of first author (as usual).
+    Unless we've gotten pdfs mixed up,
+    the references actually \emph{are} ordered alphabetically by last name of first author;
+    this might not be obvious because the authors are listed with first name first?
 }
@@
     We would love to also! However, at time of writing,
-    tsinfer does not attempt to infer trees with well-calibrated branch lengths.
+    tsinfer does not attempt to infer trees with well-calibrated branch lengths,
+    so we thought it best to not get into this.
 }
@@
 \begin{point}{\revref}
-    ``the genome-wide mean Branch f4 is around -700''.
+    ``the genome-wide mean Branch $f_4$ is around -700''.
 What does that number mean?
 Shouldn't that be normalised by the number of sites considered?
-Usually, f4 statistics are reported with a Z-score that gives statistical support for how far away from 0 the statistics is.
+Usually, $f_4$ statistics are reported with a $Z$-score that gives statistical support for how far away from 0 the statistic is.
 Is there any way to do that here?

### 2020-04-02 treestats_paper.tex
responses to other reviews

@@
 
-%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Framework and statistics}
-
 The basic relationship between tree shape and summaries of genetic variation
-is this: if mutations are neutral then the expected number that occur on a
+that we will explain and utilize below is this:
+if mutations are neutral, then the expected number that occur on a
 given branch of a tree is proportional to the length of that branch.
@@
 and aggregate these summaries to produce statistics.
-\autoref{fig:divergence_diagram} shows a simple example of these procedures
+
+%%%%%%%%%%%%%%%%%%%%%%%
+\section*{Framework and statistics}
+
+Population genetic data is most directly represented as a matrix of genotypes,
+and so summaries of popluation genetic data usually begin with this data structure.
+However, since the genetic variation represented in a genotype matrix
+was generated by mutations inherited down genealogical trees,
+it is often informative to think about these statistics in terms of these trees.
+Most population genetic summary statistics can be thought of in terms of the \emph{allele frequencies},
+i.e., what proportion of the sampled genotypes carry each allele.
+The frequency of any particular allele
+can easily be found from the location of the mutation(s) that produced it
+on the marginal genealogical tree,
+by counting the number of sampled genotypes below it in the tree.
+If many mutations have occurred on a single genealogical tree,
+then these mutations start to give us an idea of the shape of that tree,
+and the frequencies of all of them can be easily found by propagating sample counts up the tree.
+This relationship is depicted in \autoref{fig:allele_freqs},
+which shows the relationship between the genealogical tree
+and the frequencies of alleles at five different sites across a region of the genome
+described by that single tree.
+Our goal will be to generalize this sort of summary of the genotype matrix,
+using the underlying genealogical trees.
+Working with the trees serves two purposes.
+First, the trees are closer to the demographic processes that we are really interested in,
+and so allow us to reason more directly about population genetic signal and noise.
+Second, trees are highly efficient data structures,
+and allow us to compute with genetic data far more efficiently than with the genotype data itself.
+
+\begin{figure}
+    \begin{center}
+        \includegraphics{figures/allele_frequency_diagram}
+    \end{center}
+    \caption{
+        The relationship between allele frequencies and the underlying genealogical trees.
+        \textbf{(a)} The genotype matrix, showing genotypes of five sequences at five separate sites,
+        as well as the derived allele frequencies (derived alleles are shown in color).
+        \textbf{(b)} The genealogical tree describing how those five samples are related to each other
+        over this segment of genome. The samples are leaves,
+        and each mutation is represented as a star on the genealogy,
+        with color matching the genotype matrix, and
+        labeled with its index in the genotype matrix
+        and the ancestral and derived states at that site.
+        \textbf{(c)} Each node in the tree is labeled with the number of samples
+        at or below that node, which we think of as a \emph{weight}.
+        Note that this is \emph{additive}:
+        the weight of each node is the sum of the weights of its child nodes.
+        Since the frequency of a derived allele is the number of samples
+        that have inherited that mutation, the allele frequency of each mutation
+        is equal to the weight of the node directly below the mutation,
+        divided by the total sample size.
+        \label{fig:allele_freqs}
+    }
+\end{figure}
+
+The operation of finding allele frequencies on a tree,
+depicted in \autoref{fig:allele_freqs}(c)
+is the basis of a familiar operation in population genetics:
+summarizing the allele frequency spectrum.
+Somewhat more generally, we could calculate how many of a certain set $A$ of samples
+inherit from a particular branch in a tree
+by (a) assigning each of these samples weight 1 (and other samples weight zero), then
+(b) finding the total weight of all samples in the subtree inheriting from that branch.
+This gives the number of samples that would inherit any mutation falling on this branch.
+Suppose we want to calculate one of the linear functions of the frequency spectrum
+discussed by \citet{fu1995statistical} and \citet{achaz2009frequency},
+which can be described as follows.
+Write $p_i(a)$ for the frequency of allele $a$ at a site $i$,
+and $h(p)$ for the allele frequency spectrum
+--- i.e., the number of alleles whose frequency is $p$ in our sample ---
+and suppose that for some function $f()$, we want to compute $\sum_p h(p) f(p)$.
+Since this is a simple sum over sites,
+we can compute this by summing over the $L$ loci and all alleles
+(including the ancestral allele),
+\begin{align*}
+    \sum_p h(p) f(p) = \sum_{i=1}^L \sum_a f(p_i(a)).
+\end{align*}
+For instance, mean genetic diversity across biallelic loci is computed using this formula
+with $f(p) = p (1-p)$,
+omitting the usual factor of two because the sum is over \emph{both} alleles at each site.
+This is because genetic diversity is defined to be the probability that two randomly chosen sequences
+differ at a randomly chosen site,
+and the probability that two randomly chosen genomes differ at a site with allele frequency $p$
+is $p (1-p) + (1-p) p$.
+
+If we know where each mutation has occurred on each marginal genealogical tree,
+we can easily compute statistics of this form
+by finding allele frequencies, applying the function $f( )$ to each,
+and summing up the results.
+\autoref{fig:divergence_diagram} shows a simple example of this procedure
 of propagating weights up a tree and summarizing them,
-used to calculate mean sequence divergence at a segregating site.
+used to calculate mean sequence divergence between two groups of samples at a single segregating site.
 
@@
 
-The abstract idea of summarizing weights propagated up each tree
-is only a small generalization of a familiar operation in population genetics:
-summarizing the allele frequency spectrum.
-For instance, we could calculate how many of a certain set $A$ of samples
-inherit from a particular branch in a tree
-by (a) assigning each of these samples weight 1 (and other samples weight zero), then
-(b) finding the total weight of all samples in the subtree inheriting from that branch.
-This gives the number of samples that would inherit any mutation falling on this branch,
-so is analogous to the allele frequency spectrum.
-If we only considered summaries of the allele frequency spectrum of a single sample,
-our statistics would be equivalent to the linear functions of the frequency spectrum
-discussed by \citet{fu1995statistical} and \citet{achaz2009frequency}.
-To connect our generalization to familiar ideas,
-here is a reframing of these statistics of the frequency spectrum.
-Write $p_i(a)$ for the frequency of allele $a$ at a site $i$,
-and $h(p)$ for the allele frequency spectrum
---- i.e., the number of alleles whose frequency is $p$ in our sample ---
-and suppose that for some function $f()$, we want to compute $\sum_p h(p) f(p)$.
-Since this is a simple sum over sites,
-we can compute this by summing over the $L$ loci and all alleles
-(including the ancestral allele),
-\begin{align*}
-    \sum_p h(p) f(p) = \sum_{i=1}^L \sum_a f(p_i(a)).
-\end{align*}
-For instance, mean genetic diversity across biallelic loci is computed with $f(p) = p (1-p)$,
-omitting the usual factor of two because the sum is over \emph{both} alleles at each site.
-If we know where each mutation has occurred on the genealogy,
-we can easily find the frequency of the corresponding allele by counting how many samples
-inherit from that mutation in the tree (and accounting for those that subsequently mutate).
-
-Our framework is more general as we can also define statistics that are
-\emph{not} functions of the allele frequency spectrum, such as the correlation between genotype and phenotype.
-Before we define the framework, we need some notation and definitions.
+Our framework is somewhat more general than this, since we also define statistics that are
+\emph{not} functions of the allele frequency spectrum,
+such as the correlation between genotype and phenotype.
+Before we formally describe the statistics, we need some notation and definitions.
 A tree sequence describes how a set of $n$ sampled chromosomes
@@
     for groups of genomes labeled $A$, $B$, $C$, and $D$
-    (only group $A$ has more than one sample).
+    (only group $A$ has more than one sample),
+    on the same basic example as Figure \ref{fig:allele_freqs}.
     \textbf{(a)} The genotype matrix, for five sampled genomes at five sites.
@@
     \textbf{(b)} Each mutation is given weight equal to $(p_A - p_B)(p_C - p_D)$,
-    shown as a label next to each mutation (depicted as a star).
+    shown as a label next to each mutation (depicted as a star),
+    and the Site $f_4(A,B;C,D)$ statistic is the sum of these weights
+    divided by $L$, the sequence length.
     Note that each of the mutations shown occurs at a distinct site, so there is no back mutation,
@@
     and then the branch $f_4$ statistic
-    is the sum over these weights, multiplied by the length of the branch.
+    is the sum over these weights, multiplied by the length of the branch,
+    and divided by $L$, the sequence length.
     Remaining branches colored in grey do not contribute,

### 2020-04-02 treestats_paper.tex
edits for JK comments

@@
 Population genetic data is most directly represented as a matrix of genotypes,
-and so summaries of popluation genetic data usually begin with this data structure.
+and so summaries of population genetic data usually begin with this data structure.
 However, since the genetic variation represented in a genotype matrix
-was generated by mutations inherited down genealogical trees,
+was generated by mutations inherited through genealogical trees,
 it is often informative to think about these statistics in terms of these trees.
@@
 If many mutations have occurred on a single genealogical tree,
-then these mutations start to give us an idea of the shape of that tree,
-and the frequencies of all of them can be easily found by propagating sample counts up the tree.
+then these mutations start to give us an idea of the shape of that tree.
 This relationship is depicted in \autoref{fig:allele_freqs},
@@
 Second, trees are highly efficient data structures,
-and allow us to compute with genetic data far more efficiently than with the genotype data itself.
+and allow us to compute with genetic data far more efficiently than with the genotype matrix itself.
 
@@
         \textbf{(a)} The genotype matrix, showing genotypes of five sequences at five separate sites,
-        as well as the derived allele frequencies (derived alleles are shown in color).
+        as well as the derived allele frequencies (derived alleles are shown in color,
+        with separate colors for each genotype pattern).
         \textbf{(b)} The genealogical tree describing how those five samples are related to each other
@@
 Since this is a simple sum over sites,
-we can compute this by summing over the $L$ loci and all alleles
+we can compute this by summing over the $L$ sites in the genome and all alleles
 (including the ancestral allele),

### 2020-04-02 figures/plot_1kg_diversity.py
onepop scripts

@@ -137,6 +137,7 @@ except:
 
     # unmasked_bdiv = ts.diversity(pop_nodes, windows=windows, mode=mode)
 
+    print(f"Saving to {divfile}.")
     np.savetxt(divfile, bdiv, 
                header="\t".join(pop_names), delimiter="\t")
 

### 2020-04-03 treestats_paper.tex

@@
     which is the density of segregating sites if all sites are biallelic.
-    (The actual number of mutations per unit length will be greater if there have been backmutations.)
+    (The actual number of mutations per unit length will be greater if there have been back mutations.)
 \end{example}
@@
 so that they make sense even with more than two alleles per site.
-For instance, the $f_4$ statistic is defined in \autoref{ex:site_f4} as follows:
+For instance, the $f_4$ statistic is defined in Example~\ref{ex:site_f4} as follows:
 averaged across a random choice of individuals from the four groups of samples,
@@
     and divided by $L$, the sequence length.
-    Remaining branches colored in grey do not contribute,
-    because any mutation falling on these would have either $p_A - p_B = 0$ or $p_C - p_D = 0$.
+    Remaining branches do not contribute,
+    because any mutation falling on these have either $p_A - p_B = 0$ or $p_C - p_D = 0$.
         \label{fig:branch_site_diagram}
@@
 two sequences differ at a site only if there has been a mutation somewhere on the branches going back
-to their most recent common ancestor,
-and so if mutations occur with constant rate,
-the expected value of genetic diversity,
-averaging over mutational noise given the tree sequence,
-is equal to the mutation rate multiplied by the average distance between the two in the trees.
+to their most recent common ancestor.
+Therefore, if mutations occur with constant rate
+the expected value of genetic diversity
+% averaging over mutational noise given the tree sequence,
+is equal to the mutation rate multiplied by the average path distance between
+the two samples in the trees.
 
@@
 In fact, for any region of the genome between $i$ and $j$,
-with $\treeseq_{[i,j)}$ the trees spanning that interval,
+% with $\treeseq_{[i,j)}$ the trees spanning that interval,
 \begin{align}
@@
     \frac{1}{\mu}
-    \E\left[ \site(f, \iw)_{[i,j)} \given \treeseq_{[i,j)} \right] ,
+    \E\left[ \site(f, \iw)_{[i,j)} \given \treeseq_{[i,j)} \right].
 \end{align}
-where the expected value leaves the genealogies fixed,
-averaging over infinite-sites mutations
+Here,  the expectation keeps the genealogies fixed
+and averages over infinite-sites mutations
 with rate $\mu$ per unit time and per unit of sequence length.
@@
 To evaluate the correspondence between Site and Branch statistics
-in tree sequences inferred from real data,
+in trees inferred from real data,
 we calculated statistics for chromosome 20
@@
 by Relate~\citep{speidel2019method}. \revpoint{2}{9}
-(Although Relate does not directly return a tree sequence, its output can be converted to a tree sequence.)
+(Although Relate estimates a succession of essentially independent marginal trees rather
+than a succinct tree sequence,
+the output can be converted to a tree sequence, and since the statistics considered
+here are ``single site'' the distinction is not important.)
 Specifically, we calculated diversity using the Site statistic
@@
 Aaron Ragsdale, Konrad Lohse, and Wilder Wohns for useful suggestions
-and to Leo Speidel for providing the inferred 1000 Genomes tree sequence.
+and to Leo Speidel for providing the inferred 1000 Genomes trees.
 This work was supported by NSF grant \#1262645 (DBI) to PLR

### 2020-04-03 treestats_paper.tex
figure rearrangement and relate update

@@
 is $p (1-p) + (1-p) p$.
-
 If we know where each mutation has occurred on each marginal genealogical tree,
@@
 and summing up the results.
-\autoref{fig:divergence_diagram} shows a simple example of this procedure
-of propagating weights up a tree and summarizing them,
-used to calculate mean sequence divergence between two groups of samples at a single segregating site.
+
+Our framework is somewhat more general than this, since we also define statistics that are
+\emph{not} functions of the allele frequency spectrum.
+The small but useful generalization we make is to allow the ``weights'' we propagate up the tree
+to be numbers other than zero or one.
+The general procedure is depicted in \autoref{fig:adding_weights}.
+If weights \emph{are} all zero or one, then these weights will count numbers of samples
+in each part of the tree;
+but below we use other weights to compute the correlation between genotype and phenotype.
 
@@
     \begin{center}
-    \includegraphics{figures/divergence_diagram_simple}
+        \includegraphics{figures/adding_weights}
     \end{center}
     \caption{
-        An example of computing sequence divergence
-        between red samples (triangles) and blue samples (squares)
-        at a single site with a T$\to$G mutation, using the tree.
-        The derived mutation is found in $a=2$ out of the $n_\text{red} = 3$ red samples
-        and in $b=1$ of the $n_\text{blue}=2$ blue samples, and so the probability that a randomly chosen red node carries the derived allele
-        but a randomly chosen blue node does not
-        is given by the summary function, $f(a,b) = f(2, 1) = 2/6$.
-        The complementary value, $f(1, 1) = 1/6$ is the probability that the mutation separates
-        a randomly chosen red and blue pair, but with the blue node carrying the derived allele,
-        so the total probability that a randomly chosen pair
-        differs at this site is $f(2,1) + f(1,1) = 1/2$.
-        \textbf{(a)} Weights are assigned to each subtree by adding together the weights of any children,
+        \emph{Subtree weights} are assigned to each node by adding together the weights of any children,
         plus the weight of the node itself, if it is a sample.
-        \textbf{(b)} In this example, subtree weights $(a,b)$ record how many red samples ($a$)
-        and how many blue samples ($b$) lie in that subtree.
-        \label{fig:divergence_diagram}
+        In this example, the samples are the five leaves, and have weights
+        $\iw_1, \ldots, \iw_5$; the total weight -- also the subtree weight of the root --
+        is $\tiw = \iw_1 + \cdots + \iw_5$.
+        This can summarize the tree in many ways:
+        for instance, if all weights were equal to one, then subtree weights would count the number of
+        samples below that node.
+        On the other hand, if $\iw_1 = 1$ but other weights were zero,
+        then subtree weights would indicate whether a node was an ancestor of sample 1.
+        If $\iw_k$ was the value of a phenotype of sample $k$,
+        then the subtree weight of a node
+        would equal the sum of all phenotypes inheriting from that node.
+        \label{fig:adding_weights}
     }
@@
 
-Our framework is somewhat more general than this, since we also define statistics that are
-\emph{not} functions of the allele frequency spectrum,
-such as the correlation between genotype and phenotype.
 Before we formally describe the statistics, we need some notation and definitions.
@@
     \begin{align*}
-        f(x_1, x_2) = \frac{x_1 (n_2 - x_2)}{n_1 n_2} .
+        f(x_1, x_2) = \frac{x_1}{n_1} \left(1 - \frac{x_2}{n_2}\right) .
     \end{align*}
@@
 
-This example corresponds to the calculations shown in \autoref{fig:divergence_diagram}.
+The procedure of computing divergence at a single site is shown in \autoref{fig:divergence_diagram};
+divergence over a region of the genome would sum these values across all polymorphic sites
+and divide by the length of the region.
+
+\begin{figure}
+    \begin{center}
+    \includegraphics{figures/divergence_diagram_simple}
+    \end{center}
+    \caption{
+        An example of computing sequence divergence
+        between red samples (triangles) and blue samples (squares)
+        at a single site with a T$\to$G mutation, using the tree.
+        The derived mutation is found in $a=2$ out of the $n_\text{red} = 3$ red samples
+        and in $b=1$ of the $n_\text{blue}=2$ blue samples, and so the probability that a randomly chosen red node carries the derived allele
+        but a randomly chosen blue node does not
+        is given by the summary function, $f(a,b) = f(2, 1) = 2/6$.
+        The complementary value, $f(1, 1) = 1/6$ is the probability that the mutation separates
+        a randomly chosen red and blue pair, but with the blue node carrying the derived allele,
+        so the total probability that a randomly chosen pair
+        differs at this site is $f(2,1) + f(1,1) = 1/2$.
+        In this example, subtree weights $(a,b)$ record how many red samples ($a$)
+        and how many blue samples ($b$) lie in that subtree,
+        and the weights at each node are equal to the sum of the weights of the node's children,
+        as in \autoref{fig:adding_weights}.
+        \label{fig:divergence_diagram}
+    }
+\end{figure}
 
@@
 \end{align}
-Here,  the expectation keeps the genealogies fixed
+Here, $\treeseq_{[i,j)}$ denotes the tree sequence on the interval $[i,j)$,
+and so the expectation keeps the genealogies fixed
 and averages over infinite-sites mutations
@@
 As shown in Figure~\ref{fig:chr20}, the ratio of Site and Branch diversity
-is relatively constant, hovering around typical estimates for the average genome-wide human mutation rate~\citep{narasimhan2017estimating}.
-On the one hand, this is expected ---
-Relate estimates branch lengths in part assuming that mutations fall at a constant rate through time
-(although \citeauthor{speidel2019method} also showed signal of changing mutation rates,
-pointing the way towards a more general method).
+is relatively constant, hovering around $2.5 \times 10^{-8}$.
+This is somewhat higher than
+typical estimates for the average genome-wide human mutation rate~\citep[e.g., $1.45\times 10^{-8}$,][]{narasimhan2017estimating},
+which suggests a mis-calibration in inferred tree times.
+These trees were inferred using Relate with the Ne fixed to 20,000;
+if trees are instead jointly inferred in Relate along with an Ne, with mutation rate fixed,
+then indeed the ratio of Site to Branch diversity averages to the (externally specified) mutation rate
+(\autoref{sfig:chr20_GBR}).
+On the one hand, some amount of constancy of this ratio is expected, since
+Relate estimates branch lengths in part assuming that mutations fall at a constant rate through time.
+(But, note that \citeauthor{speidel2019method} also showed signal of changing mutation rates,
+pointing the way towards a more general method.)
 On the other hand,
@@
         \textbf{(middle)}
-        The dual Branch statistic as calculated from tree sequences inferred by
+        The dual Branch statistic as calculated from trees inferred by
         Relate~\citep{speidel2019method},
         in units of generations.
-        \textbf{(bottom)} The ratio of the Site statistic to the Branch statistic,
-        which, we note, hovers around typical per-generation estimates of the human mutation
-        rate~\citep{narasimhan2017estimating}.
+        \textbf{(bottom)} The ratio of the Site statistic to the Branch statistic.
         Separate curves are plotted for continental groupings:
@@
 
+\begin{figure}[!h]
+    \includegraphics[width=\textwidth]{{figures/intro/introgressed.1000.23.1000000.1e-09.f4}.pdf}
+    \caption{
+        \textbf{(top)} Mean sequence diversity
+        in 1Mb windows along human chromosome 20 of the GBR (Great Britain)
+        sample of the 1000 Genomes Project,
+        in units of mean number of pairwise differences per nucleotide.
+        \textbf{(middle)}
+        The dual Branch statistic as calculated from trees inferred by
+        Relate~\citep{speidel2019method},
+        in units of generations.
+        \textbf{(bottom)} The ratio of the Site statistic to the Branch statistic.
+        This figure differs from \autoref{fig:chr20} in that the trees were inferred 
+        using only GBR samples, and Relate was allowed to co-estimate population size and branch lengths,
+        with mutation rate fixed at $1.45 \times 10^{-8}$ mutations per generation;
+        the trees shown in \autoref{fig:chr20} were inferred with a fixed effective population size
+        of $N_e = 20,000$.
+        \label{fig:fig:chr20_GBR}
+    }
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%
 \includereviews

### 2020-04-06 treestats_paper.tex
correction from Leo

@@
 which suggests a mis-calibration in inferred tree times.
-These trees were inferred using Relate with the Ne fixed to 20,000;
+These trees were inferred using Relate with the Ne fixed to 15,000;
 if trees are instead jointly inferred in Relate along with an Ne, with mutation rate fixed,

### 2020-04-06 treestats_paper.tex
N_e not Ne

@@
 which suggests a mis-calibration in inferred tree times.
-These trees were inferred using Relate with the Ne fixed to 15,000;
-if trees are instead jointly inferred in Relate along with an Ne, with mutation rate fixed,
+These trees were inferred using Relate with the $N_e$ fixed to 15,000;
+if trees are instead jointly inferred in Relate along with an $N_e$, with mutation rate fixed,
 then indeed the ratio of Site to Branch diversity averages to the (externally specified) mutation rate

### 2020-04-06 review-responses.tex

@@
 In response to reviewer comments (and feedback from other sources)
-we have substantially rewritten the introductory portions of the manuscript,
+we have substantially rewritten the introductory portions of the manuscript
+and reworked the figures,
 hopefully making it more accessible to a wider audience.
@@
 
-We have thoroughly revised both figures, and added an additional figure,
+We have thoroughly revised both figures, split Figure 1 into two figures,
+and added an additional figure,
 to ease the conceptual burden,
@@
     We've substantially reworked this figure,
-    simplifying it and providing more description in words,
-    as well as adding another figure before it
+    splitting out part (a) into \autoref{fig:adding_weights},
+    simplifying it and providing more description in words.
+    We have also added another figure before it
     to reinforce the same basic intuition.
@@
     We have substantially simplified this figure,
-    have rewritten the explanation,
+    split out \autoref{fig:adding_weights} into a separate figure,
+    rewritten the explanation,
     and begun with a different, simpler figure.

### 2020-04-07 treestats_paper.tex

@@
 or ancestral to all of them.
-This is desireable because genetic variation within the samples
+This is desirable because genetic variation within the samples
 cannot inform us about such parts of the tree.
@@
 the expected number of mutations per branch is proportional to its length.
-This implies an isormorphism between ``Site'' and ``Branch'' statistics defined above,
+This implies an isomorphism between ``Site'' and ``Branch'' statistics defined above,
 which is discussed in more detail in \citet{ralph2019empirical}.

### 2020-04-10 treestats_paper.tex
for resubmission

@@
 using the succinct tree sequence encoding of genealogies and genome sequence.
-The general approach accumulates ``sample weights'' within each marginal tree,
+The general approach accumulates ``sample weights''
+within the genealogical tree at each position on the genome,
 which are then combined using a ``summary function'';
 different statistics result from different choices of weight and function.
-Results can be reported in two dual ways:
-by \emph{site}, which corresponds to statistics calculated from genome
-sequence; and by \emph{branch}, which gives the expected value of the site statistic
-under the infinite-sites model of mutation.
-We use the framework to implement many currently-defined statistics of genome
-sequence (making the statistics' relationship to the underlying genealogical trees
-concrete and explicit),
-as well as the analogous ``branch'' statistics of tree shape. \revpoint{2}{10}
+Results can be reported in three ways:
+by \emph{site}, which corresponds to statistics calculated as usual from genome sequence;
+by \emph{branch}, which gives the expected value of the dual site statistic
+under the infinite-sites model of mutation,
+and by \emph{node}, which summarizes the contribution of each ancestor to these statistics.
+We use the framework to implement many currently-defined statistics of genome sequence
+(making the statistics' relationship to the underlying genealogical trees concrete and explicit),
+as well as the corresponding ``branch'' statistics of tree shape. \revpoint{2}{10}
 We evaluate computational performance using simulated data, and
@@
 in practice on trees inferred from the 1000 Genomes Project dataset,
-and discuss ways in which deviations may encode interesting biological
-signals.
+and discuss ways in which deviations may encode interesting biological signals.
 
@@
 the genotype matrix~\citep{ane2005missing}. Until recently, however,
-tree based approaches could not be used to encode data from sexually
-reproducing organisms as recombination results in many trees along the
+tree based approaches could not be used to encode data from sexually reproducing organisms, because
+recombination results in many trees along the
 genome~\citep{hudson1983properties,griffiths1991two,griffiths1996ancestral,rasmussen2014genome},
@@
 This redundancy can be exploited to store genome sequence and the
-associated genealogical relationships very compactly, and has formed the basis for
+associated genealogical relationships very compactly, and formed the basis for
 efficient algorithms to calculate several population genetics
 statistics~\citep{kelleher2016efficient}.
-This paper generalizes the approach to a much broader class of statistics,
-supporting arbitrary tree topologies and patterns of allelic variation.
+This paper generalizes the approach of \citet{kelleher2016efficient} to a much broader class of statistics,
+supporting arbitrary tree topologies and patterns of allelic variation,
+including polyallelic sites and back mutations.
 % Want to emphasise generality here: people might assume we're only dealing
@@
 and aggregate these summaries to produce statistics.
+Different choices of weights and summary functions
+will then produce both familiar and novel statistics
+of both genotypes and genealogies.
 
@@
 Most population genetic summary statistics can be thought of in terms of the \emph{allele frequencies},
-i.e., what proportion of the sampled genotypes carry each allele.
+i.e., what proportion of some subset of the sampled genotypes carry each allele.
 The frequency of any particular allele
@@
 on the marginal genealogical tree,
-by counting the number of sampled genotypes below it in the tree.
-If many mutations have occurred on a single genealogical tree,
+by counting the number of sampled genotypes below the mutation(s) in the tree.
+If many mutations have occurred at distinct sites that share a single genealogical tree,
 then these mutations start to give us an idea of the shape of that tree.
@@
         The relationship between allele frequencies and the underlying genealogical trees.
-        \textbf{(a)} The genotype matrix, showing genotypes of five sequences at five separate sites,
+        \textbf{(a)} The genotype matrix, showing genotypes of five sequences (``samples'')
+        at five separate sites,
         as well as the derived allele frequencies (derived alleles are shown in color,
@@
 is $p (1-p) + (1-p) p$.
-If we know where each mutation has occurred on each marginal genealogical tree,
+If we know where each mutation has occurred on the genealogical tree at each site in the genome,
 we can easily compute statistics of this form
@@
     to every sample node.
-    Given these weights, on tree $T_k$
-    the \emph{subtree weight} $\nw_k(u)$
+    Given these weights,
+    the \emph{subtree weight} $\nw_k(u)$ on tree $T_k$
     for a node $u$ is the sum of all sample weights
@@
 and $\tiw - \nw_{T_k}(u)$ is the sum of all remaining weights.
-If this is computed over only a ``window'' of the genome
+This is an average over the genome because each tree is weighted by its span, $L_k$,
+and the sum is divided by the total genome length,
+which is the sum of all the spans: $L = L_1 + \cdots + L_{|\treeseq|}$.
+If the node statistic is computed over only a ``window'' of the genome
 then $L_k$ is replaced by the length of the portion of that window that the tree $T_k$ extends for,
@@
 \begin{example}[Ancestry proportions] \label{ex:ancestry_props} \revpoint{2}{5}
-    If $\iw$ are the indicator weights of the set $S$ of $n$ samples,
+    If $\iw = \bone_S$ are the indicator weights of the set $S$ of $n$ samples,
     then $\nw(u) / n$ is the proportion of the samples in $S$ that inherit from $u$.
@@
 ensures that the sum is only affected by polymorphic sites,
-although we normalize by total number of sites, to allow comparison between regions.
+although we normalize by total number of sites,
+so that the values are comparable between different regions of the genome.
 
@@
     \end{align*}
-    Then $\site(f, \iw)_{[a,b)}$ is mean nucleotide diversity of $S$ in the region between $a$ and $b$.
+    Then $\site(f, \iw)_{[a,b)}$ is mean nucleotide diversity of $S$
+    in the region of the genome between $a$ and $b$.
 \end{example}
@@
     Then $\site(f, \iw)_{[a,b)}$ is mean nucleotide divergence between the two groups
-    in the region between $a$ and $b$.
+    in the region of the genome between $a$ and $b$.
 \end{example}
@@
         at a single site with a T$\to$G mutation, using the tree.
-        The derived mutation is found in $a=2$ out of the $n_\text{red} = 3$ red samples
-        and in $b=1$ of the $n_\text{blue}=2$ blue samples, and so the probability that a randomly chosen red node carries the derived allele
+        Each node is labeled with subtree weights $(x_\text{red},x_\text{blue})$,
+        recording how many red samples ($x_\text{red}$)
+        and how many blue samples ($x_\text{blue}$) lie in that subtree.
+        The derived mutation is found in $x_\text{red}=2$ out of the $n_\text{red} = 3$ red samples
+        and in $x_\text{blue}=1$ of the $n_\text{blue}=2$ blue samples, and so the probability that a randomly chosen red node carries the derived allele
         but a randomly chosen blue node does not
-        is given by the summary function, $f(a,b) = f(2, 1) = 2/6$.
+        is given by the summary function, $f(x_\text{red}, x_\text{blue}) = f(2, 1) = 2/6$.
         The complementary value, $f(1, 1) = 1/6$ is the probability that the mutation separates
@@
         differs at this site is $f(2,1) + f(1,1) = 1/2$.
-        In this example, subtree weights $(a,b)$ record how many red samples ($a$)
-        and how many blue samples ($b$) lie in that subtree,
-        and the weights at each node are equal to the sum of the weights of the node's children,
-        as in \autoref{fig:adding_weights}.
         \label{fig:divergence_diagram}
@@
     \end{align*}
-    In computing $\site(f, \iw)_{[a,b)}$, the function $f()$ is given the number of samples in $S$
-    that inherit each allele.
+    In computing $\site(f, \iw)_{[a,b)}$, the argument $x$ to the function $f(x)$
+    is the number of samples in $S$ that inherit each allele.
     A site with $k$ distinct alleles that segregate in $S$ with counts $x_1 + \cdots + x_k = n$
@@
 or by marking all non-ancestral alleles as (the same) ``derived'' allele.
-We choose to define statistics in a way that still makes formal sense for polyallelic sites,
+In contrast, we choose to define statistics in a way that still makes formal sense for polyallelic sites,
 so that for instance a site with ancestral allele $A$ and derived alleles $C$ and $T$
@@
     % let $\iw_{ju} = 1$ if $u \in S_j$
-    (so that $\iw_{ju}$ tells us the number of samples in $S_j$ descended from $u$),
+    (so that $\iw_{j}(u)$ tells us the number of samples in $S_j$ descended from $u$),
     and write $n_i$ for the number of samples in $S_i$, and
@@
 
-Note that we have taken care in all of these definitions
-so that they make sense even with more than two alleles per site.
-For instance, the $f_4$ statistic is defined in Example~\ref{ex:site_f4} as follows:
+Again, we have taken care so that this definition
+makes sense even with more than two alleles per site.
+This definition of $f_4$ can be restated as follows:
 averaged across a random choice of individuals from the four groups of samples,
@@
 but A is standing in for ``any allele that is not B''.
+This is more general than previous definitions, but agrees with them for biallelic sites.
 \revpoint{2}{7}
@@
     &=
-    \sum_{u \in T_k} \beta_{T_k}(u) \left( f(\nw_{k}(u)) + f(\tiw - \nw_{k}(u)) \right)  .
+    \sum_{u \in T_k} \beta_{k}(u) \left( f(\nw_{k}(u)) + f(\tiw - \nw_{k}(u)) \right)  .
 \end{align}
-Here, $\beta_{T_k}(u)$ is the length of the branch between $u$ and its immediate ancestor in tree $T_k$,
+Here, $\beta_{k}(u)$ is the length of the branch between $u$ and its immediate ancestor in tree $T_k$,
 $\tiw = \sum_u \iw(u)$ is the total weight,
@@
     \centering
-    \includegraphics{figures/branch_site_diagram}
+    \includegraphics[width=\textwidth]{figures/branch_site_diagram}
     \caption{
@@
     \textbf{(a)} The genotype matrix, for five sampled genomes at five sites.
-    Also shown is the derived allele frequencies in each of the four groups,
+    Also shown are the derived allele frequencies in each of the four groups,
     denoted by $p_A$, $p_B$, $p_C$, and $p_D$ respectively.
@@
     &=
-    \sum_{u \in T_k} \beta_{T_k}(u) f(\nw_{k}(u)) ,
+    \sum_{u \in T_k} \beta_{k}(u) f(\nw_{k}(u)) ,
 \end{align}
@@
 Note that the expected product of two site statistics,
-$\E[\site(f, \iw) \site(g, \iw')]$
-is not equal to the product of the two branch statistics, $\branch(f, \iw) \branch(g, \iw')$,
+$\E[\site(f, \iw) \site(g, \iw') \given \treeseq]$
+is not equal to the product of the two branch statistics, $\mu^2 \branch(f, \iw) \branch(g, \iw')$,
 because they are not independent.
 However, it is always possible to define a branch statistic that
-gives the expected value of the product~\citep{ralph2019empirical}.
+gives the expected value of the product, as described in \citet{ralph2019empirical}.
 
@@
 Site statistics give mean densities of mutations per unit of sequence length,
-which divided by $\mu$ is converted to time.
+which divided by the mutation rate $\mu$ is converted to time.
 Let's unpack the assumptions here: what exactly is the mutational model?
@@
 The sum is divided by the length of the region, $j-i$,
-and so the variance is divided by $(j-i)^2$.
+and so the variance is divided by $(j-i)^2$,
+but one factor of $(j-i)$ is absorbed by the definition of the branch statistic.
 
@@
 The top two plots compare Branch diversity --- i.e., as computed only with tree shape ---
-to Site diversity computed from sequence generated by 20 independent assignments of mutations to the same tree sequence,
+to Site diversity computed from sequences generated by 20 independent assignments of mutations to the same tree sequence,
 with mutation rates $10^{-9}$ and $10^{-8}$, respectively.
 We see that as the mutation rate increases, the signal of decrease in diversity around swept loci becomes more clear,
-and approaches Branch diversity.
+and Site diversity approaches Branch diversity.
 These were computed using the entire population of 1,000 individuals;
@@
         Diversity within the entire population,
-        computed as Site statistic from 20 independent assignments of mutations to the same tree sequence with mutation rate $\mu = 10^{-9}$.
+        computed as a Site statistic from 20 independent assignments of mutations to the same tree sequence with mutation rate $\mu = 10^{-9}$.
         \textbf{(middle)}
@@
         \textbf{(bottom)}
-        Site and Branch diversity within 20 disjoint samples of size 100 each,
+        Site and Branch diversity within 20 disjoint samples of size 100 each
+        from a \emph{single} assignment of mutations to the tree sequence
         with mutation rate $\mu = 10^{-9}$.
@@
         the window width was 10Kb,
-        and the entire population (1000 diploids in each of four populations) were used.
+        and the entire population (1000 diploids in each of four populations) was used.
         \label{fig:introf4sd}
@@
 then indeed the ratio of Site to Branch diversity averages to the (externally specified) mutation rate
-(\autoref{sfig:chr20_GBR}).
+(\autoref{fig:chr20_GBR}).
 On the one hand, some amount of constancy of this ratio is expected, since
@@
 problem shared across Site, Branch and Node statistics
-is to efficiently maintain the ``state'' $x_u$ (where $x_u$ is the
+is to efficiently maintain the ``state'' $x(u)$ (where $x(u)$ is the
 sum of the weights for all samples descending
-from, and including, node $u$; see Figure~\ref{fig:divergence_diagram}).
+from, and including, node $u$; see Figure~\ref{fig:adding_weights}).
 As we move along the sequence, the trees change when branches are inserted and removed.
@@
 \end{align*}
+Our goal is to compute $a$.
 Writing this in block matrix notation, $a$ and $b$ minimize
@@
 \begin{figure}[!h]
-    \includegraphics[width=\textwidth]{{figures/intro/introgressed.1000.23.1000000.1e-09.f4}.pdf}
+    \includegraphics[width=\textwidth]{{figures/1kg/relate_chr20_GBR.branch.1000000.site.ratio}.pdf}
     \caption{
@@
         the trees shown in \autoref{fig:chr20} were inferred with a fixed effective population size
-        of $N_e = 20,000$.
-        \label{fig:fig:chr20_GBR}
+        of $N_e = 15,000$.
+        \label{fig:chr20_GBR}
     }

### 2020-04-29 treestats_paper.tex
final submission version

@@
         as well as the derived allele frequencies (derived alleles are shown in color,
-        with separate colors for each genotype pattern).
+        with separate colors for each genotype pattern:
+        for example, all green mutations separate samples $s_1$, $s_3$ and $s_5$ from $s_2$ and $s_4$.
         \textbf{(b)} The genealogical tree describing how those five samples are related to each other
@@
 have per-capita migration rates of $4/N$ to each other.
-We expect a significantly negative $f_4(1,2;3,4)$ in this situation,
+We expect a negative $f_4(1,2;3,4)$ in this situation,
 which we indeed find

### 2020-04-29 treestats_paper.tex
genetics formatting

@@
-\documentclass{article}
-\usepackage{fullpage}
-\usepackage{graphicx}
-\usepackage[hidelinks]{hyperref}
-
-\usepackage{color}
+\documentclass[9pt,twocolumn,twoside,lineno]{gsajnl}
+\articletype{inv} % article type
+
 \usepackage{amsmath,amssymb,amsthm}
-\usepackage{natbib}
-
-\bibliographystyle{plainnat}
-
-\include{macros}
-\newif\ifreviewresponses
-% \reviewresponsestrue  % include them
-\reviewresponsesfalse  % don't include them
-\newcommand{\responsefile}{review-responses.tex}  % name of the review reponses file
-\input{review-response-commands} % for line numbers only for now
+
+\newcommand{\R}{\mathbb{R}}
+\renewcommand{\P}{\mathbb{P}}
+\newcommand{\E}{\mathbb{E}}
+\newcommand{\var}{\mathop{\mbox{Var}}}
+\newcommand{\cov}{\mathop{\mbox{cov}}}
+\newcommand{\bone}{\mathbf{1}}
+\newcommand{\st}{\,\colon\,}
+
+% These macros are borrowed from TAOCPMAC.tex
+\newcommand{\slug}{\hbox{\kern1.5pt\vrule width2.5pt height6pt depth1.5pt\kern1.5pt}}
+\def\xskip{\hskip 7pt plus 3pt minus 4pt}
+\newdimen\algindent
+\newif\ifitempar \itempartrue % normally true unless briefly set false
+\def\algindentset#1{\setbox0\hbox{{\bf #1.\kern.25em}}\algindent=\wd0\relax}
+\def\algbegin #1 #2{\algindentset{#21}\alg #1 #2} % when steps all have 1 digit
+\def\aalgbegin #1 #2{\algindentset{#211}\alg #1 #2} % when 10 or more steps
+\def\alg#1(#2). {\medbreak % Usage: \algbegin Algorithm A (algname). This...
+  \noindent{\bf#1}({\it#2\/}).\xskip\ignorespaces}
+\def\kalgstep#1.{\ifitempar\smallskip\noindent\else\itempartrue
+   \hskip-\parindent\fi
+   \hbox to\algindent{\bf\hfil #1.\kern.25em}%
+   \hangindent=\algindent\hangafter=1\ignorespaces}
+
+\newcommand{\algstep}[3]{\kalgstep #1 [#2] #3 }
+\newenvironment{taocpalg}[3]{%
+\vspace{1em}%
+\algbegin Algorithm #1. ({#2}). #3 }
+{\vspace{1em}}
+
+\newcommand{\algorithmref}[1]{#1}
+
+\newtheorem{definition}{Definition}
+\newtheorem{lemma}{Lemma}
+\newtheorem{cor}{Corollary}
+\newtheorem{theorem}{Theorem}
+\newtheorem{example}{Example}
 
@@
 
-\begin{document}
-
 \title{
@@
     a general duality between statistics of genealogies and genomes}
-\author{Peter Ralph, Kevin Thornton, and Jerome Kelleher}
-\maketitle
-
-% \emph{Running head:} Genealogies and genomes
+
+\author[$\ast$,1]{Peter Ralph}
+\author[$\dagger$]{Kevin Thornton}
+\author[$\ddagger$]{Jerome Kelleher}
+
+\affil[$\ast$]{Institute of Evolution and Ecology, Departments of Mathematics and Biology, University of Oregon, Eugene, Oregon}
+\affil[$\dagger$]{Department of Ecology and Evolutionary Biology, University of California, Irvine, California}
+\affil[$\ddagger$]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, University of Oxford}
+
+\keywords{genealogy, tree sequence, genotype statistics}
+
+\runningtitle{Genealogies and genomes}
+\runningauthor{Ralph \textit{et al.}}
 
 %%%%%%%%%%
-\section*{Abstract}
-
+\begin{abstract}
 As a genetic mutation is passed down across generations,
@@
 (making the statistics' relationship to the underlying genealogical trees concrete and explicit),
-as well as the corresponding ``branch'' statistics of tree shape. \revpoint{2}{10}
+as well as the corresponding ``branch'' statistics of tree shape.
 We evaluate computational performance using simulated data, and
@@
 and discuss ways in which deviations may encode interesting biological signals.
-
-%%% OUTLINE
-% 1. Framework and algorithms for computing general stats
-%
-%     * Framework:
-%
-%         - single-site stats are functions of the genotype vector
-%         - in infinite sites, the genotype vector of a mutation is determined by the nodes below it
-%         - ex: divergence as a sum over trees and branches
-%         - ex: ancestry proportions
-%         - general definition, using summary function of weights propagated up the tree
-%             * recursion relating weights on nodes of two forests that differ by a branch
-%
-%     * Review of tree sequences (quick)
-%
-%     * Algorithms:
-%
-%         - weight propagation across tree differences
-%         - count by node length ("Node Statistics")
-%         - count by branch area ("Branch Statistics")
-%         - count by site ("Site Statistics")
-%         - note on outputs (by node, by site, in windows)
-%
-% 2. Examples: in sections, one of each type of output
-%
-%     - ancestry proportions
-%     - list of standard summary statistics: divergence, Fst, f4, covariance
-%         * performance for one of these
-%     - PCA: Krylov method
-%         * performance
-%
-% 3. Correspondence to tree statistics
-%
-%     - math: in infinite sites, neutral model, E[site] = branch
-%     - plot from simulation showing branch stats are less noisy versions of the site stats
-%     - show time profile of PC1 as computed from rare variants and common variants
+\end{abstract}
+
+\begin{document}
+
+\maketitle
+\thispagestyle{firststyle}
+\marginmark
+\firstpagefootnote
+
+\correspondingauthoraffiliation{1}{Corresponding author: {plr@uoregon.edu}}
+\vspace{-33pt}% Only used for adjusting extra space in the left column of the first page
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%
-\section*{Framework and statistics}
+\section*{Materials and Methods}
+
+\subsection*{Framework and statistics}
 
@@
 %%%%%%%%%%%%%%%%%%
-\subsection*{Node statistics}
+\subsubsection*{Node statistics}
 
@@
 
-\begin{example}[Ancestry proportions] \label{ex:ancestry_props} \revpoint{2}{5}
+\begin{example}[Ancestry proportions] \label{ex:ancestry_props}
     If $\iw = \bone_S$ are the indicator weights of the set $S$ of $n$ samples,
@@
 %%%%%%%%%%%%%%%%%%
-\subsection*{Site statistics}
+\subsubsection*{Site statistics}
 
@@
     contributes $f(x_1) + \cdots + f(x_k) = (1 - x_1/n) + \cdots + (1 - x_k/n) = k - 1$
-    to the statistic. \revpoint{2}{6}
+    to the statistic.
     Therefore, $\site(f, \iw)_{[a,b)}$, is the minimum number of derived mutations per unit length,
@@
 This is more general than previous definitions, but agrees with them for biallelic sites.
-\revpoint{2}{7}
 
 %%%%%%%%%%%%%%%%%%
-\subsection*{Branch statistics}
+\subsubsection*{Branch statistics}
 
@@
 
-\begin{example}[Patterson's $f_4$] \label{ex:branch_f4} \revpoint{2}{7}
+\begin{example}[Patterson's $f_4$] \label{ex:branch_f4}
     Suppose that the four subsets each consist of only a single sample.
@@
 %%%%%%%%%%%%%%%%%%
-\section*{Duality of site and branch statistics}
+\subsection*{Duality of site and branch statistics}
 
@@
 which we indeed find
-(the genome-wide mean Branch $f_4$ is around -700 generations, \revpoint{2}{12}
+(the genome-wide mean Branch $f_4$ is around -700 generations,
 as is the Site $f_4$ divided by mutation rate;
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%
+\section*{Data and code availability}
+
+All methods described here are implemented in Python and C in the package \tskit{},
+available from \url{https://github.com/tskit-dev/tskit}
+under the terms of the MIT license.
+All code used to produce the figures in this paper
+is available from \url{https://github.com/petrelharp/treestats_ms}.
+
 %%%%%%%%%%%%%%%%%%
-\section*{Application to 1000 Genomes tree sequences}
+\section*{Results}
+
+\subsection*{Application to 1000 Genomes tree sequences}
 
@@
 of the 1000 Genomes dataset~\citep{1k2015global} using dated trees estimated
-by Relate~\citep{speidel2019method}. \revpoint{2}{9}
+by Relate~\citep{speidel2019method}.
 (Although Relate estimates a succession of essentially independent marginal trees rather
@@
 %%%%%%%%%%%%%%%%%%
-\section*{Algorithm and implementation}
+\subsection*{Algorithm and implementation}
 
@@
 \algstep{B3.}{Edge removal loop.}{If $k = |\Et|$ or $t_l \neq
-    \Et[\indexout_k]\prop{right}$ go to \algref{B6}.}
+    \Et[\indexout_k]\prop{right}$ go to \algorithmref{B6}.}
 
@@
     $F_v \leftarrow f(x_v)$, $s \leftarrow s + \beta_v F_v$
-    and $v \leftarrow \pi_v$. Afterwards, go to \algref{B3}.
+    and $v \leftarrow \pi_v$. Afterwards, go to \algorithmref{B3}.
 }
@@
 \algstep{B6.}{Edge insertion loop.}{If $j = |\Et|$ or $t_l \neq
-    \Et[\indexin_j]\prop{left}$ go to \algref{B9}.}
+    \Et[\indexin_j]\prop{left}$ go to \algorithmref{B9}.}
 
@@
     $F_v \leftarrow f(x_v)$, $s \leftarrow s + \beta_v F_v$
-    and $v \leftarrow \pi_v$. Afterwards, go to \algref{B6}.
+    and $v \leftarrow \pi_v$. Afterwards, go to \algorithmref{B6}.
 }
@@
     Finally, set $\sigma \leftarrow \sigma + (t_r - t_l)s$,  $t_l \leftarrow t_r$
-    and go to \algref{B2}.
+    and go to \algorithmref{B2}.
 }
@@
 
-Algorithm \algref{B} (named ``B'' for ``Branch'')
+Algorithm \algorithmref{B} (named ``B'' for ``Branch'')
 begins by setting the initial state for $\pi$, $\beta$,
@@
 removal indexes, respectively. After completing initialization in
-\algref{B1}, we enter the main tree loop in \algref{B2} which is
+\algorithmref{B1}, we enter the main tree loop in \algorithmref{B2} which is
 run once for each tree in the sequence. As we are processing each
@@
 current tree. These must be the edges in which the right coordinate
-is equal to $t_l$, and so \algref{B3} loops over these edges using
-the edge removal index $\indexout$. Step \algref{B4} then removes
+is equal to $t_l$, and so \algorithmref{B3} loops over these edges using
+the edge removal index $\indexout$. Step \algorithmref{B4} then removes
 the branch $u \mapsto v$ corresponding to a single edge, by
@@
 in the subtree rooted at $u$ becoming disconnected from the rest of the
-tree. Step \algref{B5} ensures that the state of the rest of the tree
+tree. Step \algorithmref{B5} ensures that the state of the rest of the tree
 is correctly maintained by propagating the loss of the state $x_u$
@@
 Once all of the edges with right coordinate equal to $t_l$ have been removed in
-steps \algref{B3--B5}, we then insert edges that start in the current tree
-(i.e., with left coordinate $t_l$) in steps \algref{B6--B8}. We update the
+steps \algorithmref{B3--B5}, we then insert edges that start in the current tree
+(i.e., with left coordinate $t_l$) in steps \algorithmref{B6--B8}. We update the
 state to account for the new branch $u \mapsto v$ by setting the parent of
 $u$ to $v$, computing the branch length of $u$ and adding the new contribution
-of $u$ to $s$ in \algref{B7}. Then, as adding a new edge can connect the
+of $u$ to $s$ in \algorithmref{B7}. Then, as adding a new edge can connect the
 subtree rooted at $u$ to the larger tree at $v$, we propagate the gain of $x_u$
-up the tree in step \algref{B8} (note the symmetry with step \algref{B5}).
+up the tree in step \algorithmref{B8} (note the symmetry with step \algorithmref{B5}).
 Finally, once we have removed and inserted all of the relevant edges, we are
 ready to add the contribution of the current tree to the overall statistic,
-$\sigma$ and move on to the next tree in step \algref{B9}. To do this, we first
+$\sigma$ and move on to the next tree in step \algorithmref{B9}. To do this, we first
 compute the right-hand coordinate of the tree $t_r$ (a process slightly
@@
 by no edges). Then, we add the running sum $s$ weighted by the span of the
-current tree $t_r - t_l$ to $\sigma$, and return to \algref{B2} to process the
+current tree $t_r - t_l$ to $\sigma$, and return to \algorithmref{B2} to process the
 next tree. If we have reached the end of the sequence, we then divide $\sigma$
@@
 
-To analyze Algorithm~\algref{B} we will assume that the tree sequence has
+To analyze Algorithm~\algorithmref{B} we will assume that the tree sequence has
 been simulated under the standard coalescent with a sample of size
@@
 \citep{kelleher2016efficient}.  Clearly, each edge is examined
-exactly once in steps \algref{B6} and \algref{B7} in order to create
-the trees, and at most once in steps \algref{B3} and \algref{B4} (we do
+exactly once in steps \algorithmref{B6} and \algorithmref{B7} in order to create
+the trees, and at most once in steps \algorithmref{B3} and \algorithmref{B4} (we do
 not remove the edges for the final tree). Additionally, each edge that
@@
 may incur the cost of traversing up the tree as
-far as the root in steps \algref{B5} and \algref{B8}.
+far as the root in steps \algorithmref{B5} and \algorithmref{B8}.
 As coalescent genealogies are asymptotically
@@
 Therefore, the overall running time
-of Algorithms~\algref{B} is of order $n + \rho \log(n) \log_2(n)$, which is
+of Algorithms~\algorithmref{B} is of order $n + \rho \log(n) \log_2(n)$, which is
 \begin{equation}\label{eqn:alg_complexity}
@@
 %%%%%%%%%%%%%%%%%%%%%
-\subsection*{Efficiency}
+\subsubsection*{Efficiency}
 
@@
 matrix, importing it into another package and computing from the matrix.
-
-%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Data and code accessibility}
-
-All methods described here are implemented in Python and C in the package \tskit{},
-available from \url{https://github.com/tskit-dev/tskit}
-under the terms of the MIT license.
-All code used to produce the figures in this paper
-is available from \url{https://github.com/petrelharp/treestats_ms}.
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%
-\includereviews
-
 \end{document}

### 2020-04-30 submit/cover_letter.tex
cover letter

@@
 
-    We are writing to submit our manuscript, entitled
-    \emph{Efficiently summarizing relationships in large samples: a general
-        duality between statistics of genealogies and genomes},
-    for consideration for publication in \textit{Genetics}.
+We are writing with great joy and the final version of our accepted manuscript,
+``Efficiently summarizing relationships in large samples: a general duality between statistics of genealogies and genomes''.
+We have addressed both of Reviewer 2's final suggestions,
+by adding an explanatory note to Figure 1's caption, and deleting the word "significant",
+which didn't add anything to the sentence it was in.
+No other changes have been made,
+other than moving the \LaTeX over to the Genetics template.
 
-    An important part of the paper is that it introduces fast methods for computation
-    of population genetic statistics from large numbers of genomes using tree sequences,
-    available through the tested and validated python package \texttt{tskit}.
-    However, that's not the main point ---
-    if it was, we'd be submitting to a more bioinformatics-focused journal. 
-    It is well-known that population genetics statistics say something about the underlying genealogies,
-    although it is not common for researchers to explicitly discuss this.
-    The bigger point of our manuscript is to describe a general framework that makes it clear
-    precisely what aspect of the underlying genealogies is described by each statistic,
-    under the infinite-sites model of mutation,
-    and to provide methods to compute these statistics of tree shape as well.
-    We think this makes population genetics statistics conceptually cleaner,
-    but has some other possible benefits, besides speed:
-    (1) we make it easy to invent and implement new statistics;
-    (2) the framework allows separation of mutational from demographic noise;
-    and (3) if unbiased tree sequences can be inferred from genomic data,
-    summarizing the trees themselves provides a descriptive method
-    insensitive to variation in substitution rate or perhaps even marker ascertainment.
-    The last is dependent on testing and improvement of tree sequence inference,
-    but would be tremendously useful, if it works out.
+We have not added section titles for ``Materials and Methods'' or ``Results'',
+since we don't feel those are appropriate for our theoretically-oriented paper.
+We hope this is OK with you,
+as we see other fine Genetics papers have not included these section headings
+(e.g., Jain \& Stephan 2017, \url{https://www.genetics.org/content/206/1/389}).
+But, if you'd prefer we'll gladly put them in --- just let us know.
 
-    We have put a lot of effort into making the paper accessible to the readership of \textit{Genetics}.
-    The paper begins with a careful treatment of the theoretical framework and many examples,
-    then illustrates points with simulation and a tree sequence inferred from 1000 Genomes data
-    (by Speidel et al, 2019),
-    and finally describes the fundamental algorithm and compares computational efficiency
-    to other methods.
-    The definitions and framework might seem lengthy,
-    but in our assessment they are necessarily so:
-    there are many false paths that intuition can lead even experienced theoreticians down,
-    and choices we have made in our framework that might seem strange at first
-    turn out to be natural once carefully considered.
-
-    We are enthusiastic about the potential impact of this paper
-    and its interest to the readers of \textit{Genetics}.
-    We hope that you agree.
-
-\closing{Sincerely,}
+\closing{Thanks very much,}
 
